\SourcePage{485}{488}
\section{Vibrational energy levels}

In a quantum-mechanical treatment, the vibrational energy of a molecule is determined by the eigenvalues of the Hamiltonian
\begin{equation}
 \widehat H^{(v)}=\frac12\sum_\alpha\sum_{i=1}^{f_\alpha}
 \left(\widehat P_{\alpha i}^{2}+\omega_\alpha^2Q_{\alpha i}^2\right),
 \tag{101.1}
\end{equation}
where $\widehat P_{\alpha i}=-i\hbar\,\partial/\partial Q_{\alpha i}$ are the momentum operators corresponding to the normal coordinates $Q_{\alpha i}$. Since this Hamiltonian separates into a sum of independent terms (the expressions in parentheses), the energy levels are given by the sums
\begin{equation}
 E^{(v)}=\hbar\sum_\alpha\omega_\alpha\sum_i\left(v_{\alpha i}+\frac12\right)
 =\sum_\alpha\hbar\omega_\alpha\left(v_\alpha+\frac{f_\alpha}{2}\right),
 \tag{101.2}
\end{equation}
where $v_\alpha=\sum_i v_{\alpha i}$, while $f_\alpha$ is the multiplicity of the frequency $\omega_\alpha$. The wave

\SourcePage{486}{489}
functions, in turn, are products of the corresponding wave functions of linear harmonic oscillators:
\begin{equation}
 \psi=\prod_\alpha\psi_\alpha,
 \tag{101.3}
\end{equation}
where
\begin{equation}
 \psi_\alpha=\mathrm{const}\cdot\exp\left(-\frac12c_\alpha^2\sum_iQ_{\alpha i}^2\right)
 \prod_iH_{v_{\alpha i}}(c_\alpha Q_{\alpha i});
 \tag{101.4}
\end{equation}
$H_v$ denotes the Hermite polynomial of degree $v$, and $c_\alpha=\sqrt{\omega_\alpha/\hbar}$.

If some of the frequencies $\omega_\alpha$ are multiple, the vibrational energy levels are, in general, degenerate. The energy (101.2) depends only on the sum $v_\alpha=\sum_i v_{\alpha i}$. Consequently, the degeneracy of a level equals the number of ways in which the given set of numbers $v_\alpha$ can be formed from the numbers $v_{\alpha i}$. For a single number $v_\alpha$, this number is\footnote{This is the number of ways of distributing $v_\alpha$ balls among $f_\alpha$ boxes.}
\[
 \frac{(v_\alpha+f_\alpha-1)!}{v_\alpha!(f_\alpha-1)!}.
\]
Thus the total degeneracy is
\begin{equation}
 \prod_\alpha\frac{(v_\alpha+f_\alpha-1)!}{v_\alpha!(f_\alpha-1)!}.
 \tag{101.5}
\end{equation}
For doubly occurring frequencies, the factors in this product are $v_\alpha+1$, and for triply occurring ones they are $(1/2)(v_\alpha+1)(v_\alpha+2)$.

It must be kept in mind that this degeneracy occurs only insofar as the vibrations are treated as purely harmonic. When terms of higher degree in the normal coordinates are included in the Hamiltonian (anharmonicity of the vibrations), the degeneracy is generally removed, though not entirely (see \S~104 for more detail).

The wave functions (101.3) belonging to one and the same degenerate vibrational term furnish a certain representation (in general, a reducible one) of the molecular symmetry group. Functions belonging to different frequencies, however, transform independently of one another. Hence the representation furnished by all the functions (101.3) is the product of the representations furnished by the functions (101.4), and it is therefore sufficient to consider only the latter.

The exponential factor in (101.4) is invariant under every symmetry transformation. In the Hermite polynomials, terms of each fixed degree transform only among themselves (a symmetry transformation plainly does not change the degree of any term). On the other hand, since each Hermite polynomial is determined completely by its highest-degree term, after writing

\SourcePage{487}{490}
\[
 \prod_{i=1}^{f_\alpha}H_{v_{\alpha i}}(c_\alpha Q_{\alpha i})
 =\mathrm{const}\cdot Q_{\alpha1}^{v_{\alpha1}}Q_{\alpha2}^{v_{\alpha2}}\ldots
 Q_{\alpha f_\alpha}^{v_{\alpha f_\alpha}}+\text{terms of lower degree},
\]
it is enough to consider only the highest-degree term.

The same term contains the functions for which the sum $v_\alpha=\sum_i v_{\alpha i}$ has the same value. Thus we have a representation furnished by products of $v_\alpha$ quantities $Q_{\alpha i}$; this is precisely the symmetric product (see \S~94), taken $v_\alpha$ times, of the irreducible representation furnished by the quantities $Q_{\alpha i}$ with itself (L.~Tisza, 1933).

For one-dimensional representations, finding the characters of their $v$-fold symmetric products with themselves is trivial\footnote{Here we use the notation $\chi_v(G)$ in place of the cumbersome $[\chi^v](G)$.}:
\[
 \chi_v(G)=[\chi(G)]^v.
\]
For two- and three-dimensional representations, the following mathematical device is convenient\footnote{Applied to this purpose by A.~S.~Kompaneets (1940).}. The sum of the squares of the basis functions of an irreducible representation is invariant under all symmetry transformations. We may therefore formally regard them as the components of a two- or three-dimensional vector, and the symmetry transformations as certain rotations (or reflections) acting on these vectors. We emphasize that, in general, these rotations and reflections have nothing in common with the actual symmetry transformations and that, for every given element of the group $G$, they also depend on the particular representation under consideration.

Let us examine two-dimensional representations in greater detail. Let $\chi(G)$ be the character of some group element in a given two-dimensional representation, with $\chi(G)\ne0$. Under a rotation through an angle $\varphi$ in the plane, the sum of the diagonal elements of the transformation matrix for the components $x$, $y$ of a two-dimensional vector equals $2\cos\varphi$. By setting
\begin{equation}
 2\cos\varphi=\chi(G),
 \tag{101.6}
\end{equation}
we find the angle of the rotation formally corresponding to the element $G$ in the given irreducible representation. The $v$-fold symmetric product of the representation with itself is a representation with a basis consisting of the $v+1$ quantities $x^v,x^{v-1}y,\ldots,y^v$. The characters

\SourcePage{488}{491}
of this representation are\footnote{For the calculation it is convenient to choose the basis functions in the form
\[
 (x+iy)^v,\quad (x+iy)^{v-1}(x-iy),\ldots,(x-iy)^v;
\]
the rotation matrix is then diagonal, and the sum of its diagonal elements has the form
\[
 e^{iv\varphi}+e^{i(v-2)\varphi}+\ldots+e^{-iv\varphi}.
\]}
\begin{equation}
 \chi_v(G)=\frac{\sin(v+1)\varphi}{\sin\varphi}.
 \tag{101.7}
\end{equation}
The case $\chi(G)=0$ requires separate consideration, since a zero character corresponds both to a rotation through $\pi/2$ and to a reflection. If $\chi(G^2)=-2$, the transformation is a rotation through $\pi/2$, and for $\chi_v(G)$ we obtain
\begin{equation}
 \chi_v(G)=(-1)^{v/2}\frac{1+(-1)^v}{2}.
 \tag{101.8}
\end{equation}
If, however, $\chi(G^2)=2$, then $\chi(G)$ must be regarded as the character of a reflection (that is, the transformation $x\to x$, $y\to-y$); in this case
\begin{equation}
 \chi_v(G)=\frac{1+(-1)^v}{2}.
 \tag{101.9}
\end{equation}

Formulas for symmetric products of three-dimensional representations can be obtained in the same way. The rotation (or reflection) that formally corresponds to a group element in a given representation is readily found with the aid of Table~7. It is the transformation corresponding to the given $\chi(G)$ in that one of the isomorphic groups in which the coordinates transform according to this representation. Thus, for the representation $F_1$ of the groups $O$ and $T_d$, the transformation must be taken from the group $O$, and for the representation $F_2$, from the group $T_d$. We shall not pursue the derivation of the corresponding formulas for the characters $\chi_v(G)$ here.
