\SourcePage{36}{41}
\section{Polarization of the photon}

The polarization vector~$\mathbf{e}$ plays for the photon the role of the ``spin part'' of the wave function (with the reservations expressed in \secref{6} regarding the notion of photon spin).

The various cases arising for the polarization of a photon are in no way distinct from the possible types of polarization of a classical electromagnetic wave (see~II, \S~48).

An arbitrary polarization~$\mathbf{e}$ can be represented as a superposition of two mutually orthogonal polarizations~$\mathbf{e}^{(1)}$ and~$\mathbf{e}^{(2)}$ ($\mathbf{e}^{(1)}\mathbf{e}^{(2)*} = 0$), chosen in some definite manner. In the decomposition
\begin{equation}
\mathbf{e} = e_1\mathbf{e}^{(1)} + e_2\mathbf{e}^{(2)}
\tag{8.1}
\end{equation}
the squared moduli of the coefficients~$e_1$ and~$e_2$ give the probability that the photon has polarization~$\mathbf{e}^{(1)}$ or~$\mathbf{e}^{(2)}$.

As these two basis polarizations one may choose two mutually perpendicular linear polarizations. Alternatively, any polarization can be decomposed into two circular ones with opposite senses of rotation. We shall denote the right and left circular polarization vectors by~$\mathbf{e}^{(+1)}$ and~$\mathbf{e}^{(-1)}$ respectively; in a coordinate system~$\xi\eta\zeta$ with the $\zeta$~axis along the photon's direction $\mathbf{n} = \mathbf{k}/\omega$,
\begin{equation}
\mathbf{e}^{(+1)} = -\frac{i}{\sqrt{2}}\bigl(\mathbf{e}^{(\xi)} + i\mathbf{e}^{(\eta)}\bigr), \qquad \mathbf{e}^{(-1)} = \frac{i}{\sqrt{2}}\bigl(\mathbf{e}^{(\xi)} - i\mathbf{e}^{(\eta)}\bigr).
\tag{8.2}
\end{equation}

The existence of two distinct polarizations of the photon (at a given momentum) means, in other words, that every eigenvalue of the momentum is doubly degenerate. This circumstance is intimately connected with the fact that the photon mass equals zero.

For a freely moving particle of nonvanishing mass a rest frame always exists. It is clear that the intrinsic symmetry properties of the particle as such manifest themselves precisely in this frame. The symmetry to be considered is with respect to all possible rotations about the center,
\SourcePage{37}{42}
that is, with respect to the full spherical symmetry group. The quantity characterizing the symmetry properties of the particle under this group is its spin~$s$, which determines the multiplicity of the degeneracy (the number $2s + 1$ of distinct wave functions that transform into one another). In particular, a particle whose wave function is a vector (three components) has spin~1.

For a particle of zero mass, however, no rest frame exists --- in every reference frame it travels at the speed of light. For such a particle there is always a preferred spatial direction --- the direction of the momentum vector~$\mathbf{k}$ (the $\zeta$~axis). In this situation the full three-dimensional rotation symmetry is evidently absent, and one may speak only of axial symmetry about the distinguished axis.

Under axial symmetry only the \emph{helicity} of the particle is conserved --- the projection of angular momentum onto the $\zeta$~axis; we denote it by~$\lambda$\,\footnote{In contrast to the projection~$m$ of the angular momentum onto a fixed direction (the $z$~axis) in space, discussed in the preceding section.}. If one further demands symmetry under reflections in planes passing through the $\zeta$~axis, then states differing in the sign of~$\lambda$ are mutually degenerate; for $\lambda \neq 0$ there is therefore a twofold degeneracy\,\footnote{We note that the electronic terms of a diatomic molecule are classified in exactly the same way (see~III, \S~78).}. The state of a photon with a definite momentum belongs to precisely this type of twofold-degenerate state. It is described by a ``spin'' wave function that is a vector~$\mathbf{e}$ in the $\xi\eta$~plane; the two components of this vector transform into each other under all rotations about the $\zeta$~axis and under reflections in planes containing that axis.

The various polarization cases of a photon are in a definite correspondence with the possible values of its helicity. This correspondence can be established using formulas~III, (57.9), which relate the components of a vector wave function to the components of the equivalent rank-2 spinor\,\footnote{Recall that the wave-function components, viewed as probability amplitudes for the various values of the angular-momentum projection of the particle (which is what is meant here), correspond to the contravariant components of the spinor.}. The projections $\lambda = +1$ and~$-1$ correspond to vectors~$\mathbf{e}$ that have only the component $e_\xi - ie_\eta$ or $e_\xi + ie_\eta$ nonzero, i.e.\ $\mathbf{e} = \mathbf{e}^{(+1)}$ or $\mathbf{e} = \mathbf{e}^{(-1)}$ respectively. In other words, $\lambda = +1$ and~$-1$ correspond to the right and left circular polarizations of the photon (in \secref{16} the same result will be derived by directly computing the eigenfunctions of the spin-projection operator). Thus the projection of the photon's angular momentum onto the direction of its motion can take only the two values~$(\pm 1)$; the value~0 is not possible.
\SourcePage{38}{43}

A photon state with definite momentum and polarization is a pure state (in the sense explained in~III, \S~14); it is described by a wave function and constitutes a complete quantum-mechanical specification of the particle (photon) state. There also exist ``mixed'' photon states corresponding to a less complete description, carried out not by a wave function but only by a density matrix.

Let us examine a photon state that is mixed in its polarization but possesses a definite momentum~$\mathbf{k}$. Such a state (called a state of \emph{partial polarization}) admits a ``coordinate'' wave function.

The polarization density matrix of a photon is a second-rank tensor~$\rho_{\alpha\beta}$ defined in the plane orthogonal to~$\mathbf{n}$ (the $\xi\eta$~plane; the indices~$\alpha$, $\beta$ assume only two values). This tensor is Hermitian:
\begin{equation}
\rho_{\alpha\beta} = \rho_{\beta\alpha}^*,
\tag{8.3}
\end{equation}
and normalized by the condition
\begin{equation}
\rho_{\alpha\alpha} = \rho_{11} + \rho_{22} = 1.
\tag{8.4}
\end{equation}
By virtue of~(8.3) the diagonal components~$\rho_{11}$ and~$\rho_{22}$ are real, and one determines the other through condition~(8.4). The component~$\rho_{12}$ is complex, while $\rho_{21} = \rho_{12}^*$. Altogether, then, the density matrix is characterized by three real parameters.

Once the polarization density matrix is known, one can find the probability that the photon has any given definite polarization~$\mathbf{e}$. This probability is obtained by ``projecting'' the tensor~$\rho_{\alpha\beta}$ onto the direction of the vector~$\mathbf{e}$, i.e.\ it equals the quantity
\begin{equation}
\rho_{\alpha\beta}e_\alpha^*e_\beta.
\tag{8.5}
\end{equation}
Thus the components~$\rho_{11}$ and~$\rho_{22}$ are the probabilities of linear polarizations along the $\xi$ and~$\eta$ axes. Projection onto the vectors~(8.2) yields the probabilities of the two circular polarizations:
\begin{equation}
\tfrac{1}{2}\bigl[1 \pm i(\rho_{12} - \rho_{21})\bigr].
\tag{8.6}
\end{equation}

In both form and substance the properties of the tensor~$\rho_{\alpha\beta}$ coincide with those of the tensor~$J_{\alpha\beta}$ that describes partially polarized light in the classical theory (see~II, \S~50). Let us recall some of these properties here.

For a pure state of definite polarization~$\mathbf{e}$, the tensor~$\rho_{\alpha\beta}$ reduces to products of the components of the vector~$\mathbf{e}$:
\begin{equation}
\rho_{\alpha\beta} = e_\alpha e_\beta^*.
\tag{8.7}
\end{equation}
In this case the determinant $|\rho_{\alpha\beta}| = 0$. In the opposite case of an unpolarized photon, all polarization directions are equally probable, i.e.\
\begin{equation}
\rho_{\alpha\beta} = \delta_{\alpha\beta}/2,
\tag{8.8}
\end{equation}
and then $|\rho_{\alpha\beta}| = 1/4$.
\SourcePage{39}{44}

For the general case of partial polarization it is convenient to introduce three real Stokes parameters~$\xi_1$, $\xi_2$, $\xi_3$\,\footnote{The notation for the parameters should not be confused with that for the $\xi$~axis!}; the density matrix is then expressible as
\begin{equation}
\rho_{\alpha\beta} = \frac{1}{2}\begin{pmatrix}1 + \xi_3 & \xi_1 - i\xi_2 \\ \xi_1 + i\xi_2 & 1 - \xi_3\end{pmatrix}.
\tag{8.9}
\end{equation}
All three parameters range between~$-1$ and~$+1$. In the unpolarized state $\xi_1 = \xi_2 = \xi_3 = 0$; for a completely polarized photon $\xi_1^2 + \xi_2^2 + \xi_3^2 = 1$.

The parameter~$\xi_3$ characterizes linear polarization along the $\xi$ or~$\eta$ axis; the probability of linear polarization along these axes is $(1 + \xi_3)/2$ and $(1 - \xi_3)/2$ respectively. The values $\xi_3 = +1$ or~$-1$ therefore correspond to complete polarization in these directions.

The parameter~$\xi_1$ characterizes linear polarization along directions making an angle $\varphi = \pi/4$ or $\varphi = -\pi/4$ with the $\xi$~axis. The probability of linear polarization in these directions is $(1 + \xi_1)/2$ or $(1 - \xi_1)/2$; this is easily verified by projecting the tensor~$\rho_{\alpha\beta}$ onto the directions $\mathbf{e} = (1, \pm 1)/\sqrt{2}$.

Lastly, the parameter~$\xi_2$ represents the degree of circular polarization; by~(8.6), the probability of right or left circular polarization is $(1 + \xi_2)/2$ or $(1 - \xi_2)/2$ respectively. Because these two polarizations answer to helicities $\lambda = \pm 1$, it is evident that in general~$\xi_2$ coincides with the mean helicity of the photon. Note further that for a pure state with polarization~$\mathbf{e}$,
\begin{equation}
\xi_2 = i[\mathbf{ee}^*]\mathbf{n}.
\tag{8.10}
\end{equation}

Recall (see~II, \S~50) that the quantities~$\xi_2$ and~$\sqrt{\xi_1^2 + \xi_3^2}$ are invariant under Lorentz transformations.

In what follows we shall also need to know how the Stokes parameters transform under time reversal. One readily verifies that they are invariant under this operation. Since the result clearly does not depend on whether the polarization state is pure or mixed, it is enough to check invariance for a pure state. In quantum mechanics, time reversal replaces the wave function with its complex conjugate (see~III, \S~18). For a plane-polarized wave the corresponding substitution is\,\footnote{The extra sign reversal of~$\mathbf{e}$ is due to the fact that time reversal changes the sign of the electromagnetic vector potential. The scalar potential is unaffected; consequently, for the 4-vector~$e$ time reversal takes the form
\begin{equation}
(e_0, \mathbf{e}) \to (e_0^*, -\mathbf{e}^*).
\tag{8.11a}
\end{equation}}
\begin{equation}
\mathbf{k} \to -\mathbf{k}, \qquad \mathbf{e} \to -\mathbf{e}^*.
\tag{8.11}
\end{equation}
\SourcePage{40}{45}
Under this transformation the symmetric part of the density matrix, $(1/2)(e_\alpha e_\beta^* + e_\beta e_\alpha^*)$, and hence the parameters~$\xi_1$ and~$\xi_3$ remain unchanged. The invariance of~$\xi_2$ under the same transformation is apparent from~(8.10); it is also obvious already from the physical meaning of~$\xi_2$ as the mean helicity. Indeed, helicity is the projection of the angular momentum~$\mathbf{j}$ onto the direction~$\mathbf{n}$, that is, the product~$\mathbf{jn}$; time reversal reverses the sign of both these vectors.

For the calculations that follow, the photon density matrix must be cast in four-dimensional form, that is, expressed as a certain 4-tensor~$\rho_{\mu\nu}$. When the photon is polarized and described by a 4-vector~$e_\mu$, a natural definition of this tensor is
\begin{equation}
\rho_{\mu\nu} = e_\mu e_\nu^*.
\tag{8.12}
\end{equation}
In the three-dimensionally transverse gauge, $e = (0, \mathbf{e})$, if one of the spatial coordinate axes is chosen along~$\mathbf{n}$, the nonvanishing components of this 4-tensor coincide with~(8.7).

For an unpolarized photon the three-dimensionally transverse gauge yields a tensor~$\rho_{\mu\nu}$ with components
\begin{equation}
\rho_{ik} = \tfrac{1}{2}(\delta_{ik} - n_in_k), \qquad \rho_{0i} = \rho_{i0} = \rho_{00} = 0
\tag{8.13}
\end{equation}
(if one of the axes coincides with the direction of~$\mathbf{n}$, we return to~(8.8)). Using the tensor~$\rho_{\mu\nu}$ directly in this three-dimensional form is, however, inconvenient. We may instead perform a gauge transformation; for the density matrix this is a transformation of the type
\begin{equation}
\rho_{\mu\nu} \to \rho_{\mu\nu} + \chi_\mu k_\nu + \chi_\nu k_\mu,
\tag{8.14}
\end{equation}
where~$\chi_\mu$ are arbitrary functions. Setting
\[
\chi_0 = -1/(4\omega), \qquad \chi_i = k_i/(4\omega^2)
\]
we obtain in place of~(8.13) the simple four-dimensional expression
\begin{equation}
\rho_{\mu\nu} = -g_{\mu\nu}/2.
\tag{8.15}
\end{equation}

The four-dimensional representation of the density matrix of a partially polarized photon is readily obtained by first rewriting the two-dimensional tensor~(8.9) in three-dimensional form.
\SourcePage{41}{46}
In the three-dimensionally transverse gauge: $e^{(1)} = (0, \mathbf{e}^{(1)})$, $e^{(2)} = (0, \mathbf{e}^{(2)})$. The four-dimensional photon density matrix is therefore
\begin{equation}
\begin{split}
\rho_{\mu\nu} &= (1/2)(e_\mu^{(1)}e_\nu^{(1)} + e_\mu^{(2)}e_\nu^{(2)}) + (\xi_1/2)(e_\mu^{(1)}e_\nu^{(2)} + e_\mu^{(2)}e_\nu^{(1)}) -\\
&\quad - (i\xi_2/2)(e_\mu^{(1)}e_\nu^{(2)} - e_\mu^{(2)}e_\nu^{(1)}) + (\xi_3/2)(e_\mu^{(1)}e_\nu^{(1)} - e_\mu^{(2)}e_\nu^{(2)}).
\end{split}
\tag{8.17}
\end{equation}
The convenience of any particular choice of the 4-vectors~$e^{(1)}$, $e^{(2)}$ depends on the specific conditions of the problem at hand.

It should be borne in mind that the conditions~(8.16) do not fix the choice of~$e^{(1)}$ and~$e^{(2)}$ uniquely. If any 4-vector~$e_\mu$ satisfies these conditions, so does any 4-vector of the form $e_\mu + \chi k_\mu$ (since $k^2 = 0$). This ambiguity is tied to the gauge freedom of the density matrix.

The first term in~(8.17) corresponds to the unpolarized state. It could therefore be replaced, by~(8.15), with~$-g_{\mu\nu}/2$. Such a replacement is again equivalent to a gauge transformation. When working with 4-tensors of the form~(8.17), decomposed in two independent 4-vectors, it is useful to adopt the following formal device. Writing~(8.17) as
\[
\rho_{\mu\nu} = \sum_{a,b=1}^{2}\rho^{(ab)}e_\mu^{(a)}e_\nu^{(b)},
\]
we represent the coefficients~$\rho^{(ab)}$ as a $2\times 2$ matrix
\[
\rho = \begin{pmatrix}\rho^{(11)} & \rho^{(12)} \\ \rho^{(21)} & \rho^{(22)}\end{pmatrix}.
\]
Any Hermitian $2\times 2$ matrix can be expanded over four linearly independent $2\times 2$ matrices --- the Pauli matrices~$\sigma_x$, $\sigma_y$, $\sigma_z$ and the identity matrix~1:
\begin{equation}
\rho = (1/2)(1 + \boldsymbol{\xi\sigma}), \qquad \boldsymbol{\xi} = (\xi_1, \xi_2, \xi_3),
\tag{8.18}
\end{equation}
as is easily verified by direct comparison with~(8.17) using the well-known explicit forms of the Pauli matrices~(18.5) (grouping the three quantities~$\xi_1$, $\xi_2$, $\xi_3$ into a ``vector''~$\boldsymbol{\xi}$ is of course purely formal and serves only to shorten the notation).

The requisite generalization is achieved by replacing these 3-vectors by spacelike real 4-vectors $e^{(1)}$, $e^{(2)}$, orthogonal to each other and to the photon 4-momentum~$k$:
\begin{equation}
e^{(1)2} = e^{(2)2} = -1, \qquad e^{(1)}e^{(2)} = 0, \qquad e^{(1)}k = e^{(2)}k = 0.
\tag{8.16}
\end{equation}

\medskip\noindent\textbf{Problem}

Write the photon density matrix in the representation whose ``coordinate axes'' are the circular basis vectors~(8.2).

\noindent\textbf{Solution.} The components of the tensor~$\rho'_{\alpha\beta}$ in the new axes ($\alpha$, $\beta = \pm 1$) are obtained by projecting the tensor~(8.9) onto the basis vectors~(8.2):
\[
\rho'_{11} = \rho_{\alpha\beta}e_\alpha^{(+1)*}e_\beta^{(+1)}, \qquad \rho'_{1,-1} = \rho_{\alpha\beta}e_\alpha^{(+1)*}e_\beta^{(-1)}, \quad \ldots,
\]
\[
\rho' = \frac{1}{2}\begin{pmatrix}1 + \xi_2 & -\xi_3 + i\xi_1 \\ -\xi_3 - i\xi_1 & 1 - \xi_2\end{pmatrix}.
\]

