\SourcePage{567}{570}
\section{Current density in a magnetic field}

\input{manuscript/chapters/ch15/fragments/sec115-opening-p567-fresh-20260718.translation.en.tex}
\begin{equation}
 \delta H=-\frac1c\int\mathbf j\,\delta\mathbf A\,dV,
 \tag{115.1}
\end{equation}
\input{manuscript/chapters/ch15/fragments/sec115-p567-p568-hamilton-variation-agent.translation.en.tex}
\begin{equation}
 \overline H=\int\Psi^*\left[
 \frac1{2m}\left(\hat{\mathbf p}-\frac ec\mathbf A\right)^2
 -\frac\mu s\mathbf H\hat{\mathbf s}\right]\Psi\,dV.
 \tag{115.2}
\end{equation}
Varying this expression and recalling that
$\delta\mathbf H=\Curl{\delta\mathbf A}$, we obtain
\begin{align}
 \delta\overline H={}&\int\Psi^*\left[-\frac e{2mc}
 (\hat{\mathbf p}\,\delta\mathbf A+\delta\mathbf A\hat{\mathbf p})
 +\frac{e^2}{mc^2}\mathbf A\,\delta\mathbf A\right]\Psi\,dV\notag\\
 &-\frac\mu s\int\Curl{\delta\mathbf A}\cdot
 \Psi^*\hat{\mathbf s}\Psi\,dV.
 \tag{115.3}
\end{align}
We transform the term containing $\hat{\mathbf p}\,\delta\mathbf A$ by
integration by parts:
\[
 \int\Psi^*\hat{\mathbf p}\,\delta\mathbf A\Psi\,dV
 =-i\hbar\int\Psi^*\nabla(\delta\mathbf A\Psi)\,dV
 =i\hbar\int\delta\mathbf A\Psi\nabla\Psi^*\,dV
\]
(as usual, the integral over the surface at infinity vanishes). We also
integrate the last term in (115.3) by parts, using the familiar vector-analysis
identity
\[
 \mathbf a\Curl{\mathbf b}
 =-\Div{(\mathbf a\times\mathbf b)}
 +\mathbf b\Curl{\mathbf a}.
\]

\SourcePage{568}{571}
The integral of the divergence term vanishes, leaving
\[
 \int\Psi^*\hat{\mathbf s}\Psi\Curl{\delta\mathbf A}\,dV
 =\int\delta\mathbf A\Curl{(\Psi^*\hat{\mathbf s}\Psi)}\,dV.
\]
The final result is therefore
\begin{align*}
 \delta\overline H={}&-\frac{ie\hbar}{2mc}\int\delta\mathbf A
 (\Psi\nabla\Psi^*-\Psi^*\nabla\Psi)\,dV\\
 &+\frac{e^2}{mc^2}\int\mathbf A\,\delta\mathbf A\Psi\Psi^*\,dV
 -\frac\mu s\int\delta\mathbf A\Curl{(\Psi^*\hat{\mathbf s}\Psi)}\,dV.
\end{align*}
Comparison with (115.1) gives the following expression for the current
density:
\begin{equation}
 \mathbf j=\frac{ie\hbar}{2m}
 [ (\nabla\Psi^*)\Psi-\Psi^*\nabla\Psi]
 -\frac{e^2}{mc}\mathbf A\Psi^*\Psi
 +\frac\mu s c\Curl{(\Psi^*\hat{\mathbf s}\Psi)}.
 \tag{115.4}
\end{equation}
We emphasize that although the vector potential occurs explicitly in this
expression, the result is nevertheless unambiguous, as it must be. This can be
verified directly by observing that the gauge transformation of the vector
potential must, in accordance with (111.8), be accompanied by the wave-function
transformation (111.9).

It is also readily checked that the current (115.4), together with the charge
density $\rho=e|\Psi|^2$, obeys the continuity equation, as required:
\[
 \frac{\partial\rho}{\partial t}+\Div{\mathbf j}=0.
\]
The final term of (115.4) is the contribution to the current density arising
from the particle's magnetic moment. It has the form $c\Curl{\mathbf m}$, where
\begin{equation}
 \mathbf m=\frac\mu s\Psi^*\hat{\mathbf s}\Psi
 =\Psi^*\hat{\boldsymbol\mu}\Psi
 \tag{115.5}
\end{equation}
is the spatial density of magnetic moment.

Expression (115.4) is the mean value of the current density. It may be regarded
as a diagonal matrix element of an operator, the current-density operator
$\hat{\mathbf j}$. This operator has its simplest form in the
second-quantization representation: one merely replaces $\Psi$ and $\Psi^*$ by
the operators $\hat\Psi$ and $\hat\Psi^+$ (and, by the general rule,
$\hat\Psi^+$ must stand to the left of $\hat\Psi$ in every term). The
off-diagonal matrix elements of this operator can likewise be defined:
\begin{equation}
 \mathbf j_{nm}=\frac{ie\hbar}{2m}
 [ (\nabla\Psi_n^*)\Psi_m-\Psi_n^*\nabla\Psi_m]
 -\frac{e^2}{mc}\mathbf A\Psi_n^*\Psi_m
 +\frac\mu s c\Curl{(\Psi_n^*\hat{\mathbf s}\Psi_m)}.
 \tag{115.6}
\end{equation}
