\SourcePage{533}{536}
\section{$6j$ symbols}

In \S~106 we defined the $3j$ symbols as the coefficients in the sum (106.4), which represents the wavefunction of a three-particle system with zero total angular momentum. From the standpoint of transformation properties under rotations, this sum is a scalar. It follows that the set of $3j$ symbols for specified values of $j_1,j_2,j_3$ (and all possible $m_1,m_2,m_3$) may be regarded as a collection of quantities which, under rotations, transform contragrediently to the products $\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3}$, thereby making the entire sum invariant.

\input{manuscript/chapters/ch14/fragments/sec108-lines006-009-fresh-d-20260718.translation.en.tex}
\begin{equation}
 \sum_m(-1)^{j-m}
 \begin{pmatrix}j&\cdots\\m&\cdots\end{pmatrix}
 \begin{pmatrix}j&\cdots\\-m&\cdots\end{pmatrix}
 \tag{108.1}
\end{equation}
(compare the construction of the scalar (106.2)).

Since every ``contraction'' involves a pair of numbers $m$, a complete scalar must be constructed from a product of an even number of $3j$ symbols. By their orthogonality property, contracting a product of two $3j$ symbols gives the trivial result
\[
\begin{aligned}
&\sum_{m_1m_2m_3}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\begin{pmatrix}j_1&j_2&j_3\\-m_1&-m_2&-m_3\end{pmatrix}
(-1)^{j_1+j_2+j_3-m_1-m_2-m_3}\\
&\qquad=\sum_{m_1m_2m_3}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}^{2}=1
\end{aligned}
\]
(here the equality $m_1+m_2+m_3=0$ and formulas (106.6) and (106.12) have been used). Hence the smallest number of factors required to obtain a nontrivial scalar is four. In each $3j$ symbol, the three numbers $j$ geometrically form a closed triangle. Since every number $j$ must occur in two $3j$ symbols during the contraction, it is clear that a scalar formed from a product of four $3j$ symbols contains six numbers $j$. Geometrically, they must be represented by the edge lengths of an irregular tetrahedron (Fig.~45), with one face corresponding to each $3j$ symbol. A definite convention for carrying out the contractions is adopted in defining the scalar in question; it is expressed by the formula
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
={}&\sum_{\text{all }m}(-1)^{\sum_i(j_i-m_i)}
\begin{pmatrix}j_1&j_2&j_3\\-m_1&-m_2&-m_3\end{pmatrix}
\begin{pmatrix}j_1&j_5&j_6\\m_1&-m_5&m_6\end{pmatrix}\\
&\times\begin{pmatrix}j_4&j_2&j_6\\m_4&m_2&-m_6\end{pmatrix}
\begin{pmatrix}j_4&j_5&j_3\\-m_4&m_5&m_3\end{pmatrix}.
\end{aligned}
\tag{108.2}
\end{equation}
The summation here is over all possible values of every number $m$; however, since the sum of the three $m$'s in each $3j$ symbol must vanish, only

\SourcePage{535}{538}
three of the six $m$'s are actually independent. The quantities defined by (108.2) are called $6j$ symbols or Racah coefficients\footnote{The notation
\[
W(j_1j_2j_4j_5;j_3j_6)=(-1)^{j_1+j_2+j_4+j_5}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
\]
is also used in the literature.}.

\input{manuscript/chapters/ch14/fragments/sec108-line050-fresh-c-20260718.translation.en.tex}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,line width=.45pt,line cap=round,font=\small]
 \coordinate (L) at (-1.55,0.45); \coordinate (R) at (1.55,0.45);
 \coordinate (T) at (0.55,1.65); \coordinate (B) at (-0.65,-1.85);
 \draw (L)--(R)--(T)--(L)--(B)--(R);
 \draw[dashed] (T)--(B);
 \node[above left] at ($(L)!0.56!(T)$) {$j_1$};
 \node[above right] at ($(T)!0.48!(R)$) {$j_2$};
 \node[below] at ($(L)!0.40!(R)$) {$j_3$};
 \node[right] at ($(R)!0.48!(B)$) {$j_4$};
 \node[left] at ($(L)!0.50!(B)$) {$j_5$};
 \node[right] at ($(T)!0.55!(B)$) {$j_6$};
\end{tikzpicture}
\caption{}
\label{fig:45}
\end{figure}

The $6j$ symbols also have another, less evident symmetry property, which equates symbols having different sets of numbers $j$:
\begin{equation}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}=
\left\{\begin{matrix}
j_1&\frac12(j_2+j_5+j_3-j_6)&\frac12(j_3+j_6+j_2-j_5)\\
j_4&\frac12(j_2+j_5+j_6-j_3)&\frac12(j_3+j_6+j_5-j_2)
\end{matrix}\right\}.
\tag{108.3}
\end{equation}
($T.\ Regge$, 1959)\footnote{See the note on p.~529.}.

We give a useful relation between $6j$ and $3j$ symbols that follows from definition (108.2):
\begin{equation}
\begin{aligned}
&\sum_{m_4m_5m_6}(-1)^{j_4+j_5+j_6-m_4-m_5-m_6}
\begin{pmatrix}j_1&j_5&j_6\\m_1&-m_5&m_6\end{pmatrix}
\begin{pmatrix}j_4&j_2&j_6\\m_4&m_2&-m_6\end{pmatrix}\\
&\qquad\times\begin{pmatrix}j_4&j_5&j_3\\-m_4&m_5&m_3\end{pmatrix}
=\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}.
\end{aligned}
\tag{108.4}
\end{equation}
\input{manuscript/chapters/ch14/fragments/sec108-lines093-100-fresh-d-20260718.translation.en.tex}

\input{manuscript/chapters/ch14/fragments/sec108-lines102-104-fresh-d-20260718.translation.en.tex}

\input{manuscript/chapters/ch14/fragments/sec108-line106-fresh-d-20260718.translation.en.tex}
\begin{equation}
 \psi_{j_{23}JM}=\sum_{j_{12}}\langle j_{12}|j_{23}\rangle\psi_{j_{12}JM}.
 \tag{108.5}
\end{equation}
It is clear on physical grounds that the coefficients of this transformation do not depend on $M$: they cannot depend on the orientation of the entire system in space. They therefore depend only on the six angular momenta $j_1,j_2,j_3,j_{12},j_{23},J$, not on their projections, and so are scalar quantities (in the sense specified above). These coefficients can be calculated directly as follows.

\SourcePage{537}{540}
Applying formula (106.9) twice, we find
\[
\begin{aligned}
\psi_{j_{23}JM}&=\sum_{(m)}\langle m_1m_{23}|JM\rangle\psi_{j_1m_1}\psi_{j_{23}m_{23}}\\
&=\sum_{(m)}\langle m_1m_{23}|JM\rangle\langle m_2m_3|j_{23}m_{23}\rangle
\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3},\\
\psi_{j_{12}JM}&=\sum_{(m)}\langle m_3m_{12}|JM\rangle\langle m_1m_2|j_{12}m_{12}\rangle
\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3}
\end{aligned}
\]
(the sign $(m)$ under the summation sign means that the summation is over all the numbers $m_1,m_2,\ldots$ occurring in the expression). Using the orthonormality of the functions $\psi_{jm}$, we now obtain
\[
\begin{aligned}
\langle j_{12}|j_{23}\rangle&\equiv\int\psi_{j_{12}JM}^{*}\psi_{j_{23}JM}\,dq\\
&=\sum_{(m)}\langle m_3m_{12}|JM\rangle\langle m_1m_{23}|JM\rangle
\langle m_1m_2|j_{12}m_{12}\rangle\langle m_2m_3|j_{23}m_{23}\rangle.
\end{aligned}
\]
The sum on the right is taken at a specified value of $M$, but the result of the summation is in fact independent of $M$, for the reason given above. The summation may therefore be extended over the values of $M$ as well, provided that the factor $1/(2J+1)$ is placed before the sum. Then, expressing the coefficients $\langle m_1m_2|jm\rangle$ in terms of $3j$ symbols according to (106.10), we obtain
\begin{equation}
\langle j_{12}|j_{23}\rangle=(-1)^{j_1+j_2+j_3+J}\sqrt{(2j_{12}+1)(2j_{23}+1)}
\left\{\begin{matrix}j_1&j_2&j_{12}\\j_3&J&j_{23}\end{matrix}\right\}.
\tag{108.6}
\end{equation}
The connection of the $6j$ symbols with the transformation coefficients (108.5) makes it easy to derive several useful formulas for sums of products of $6j$ symbols.

First, by the unitarity of transformation (108.5) (and the reality of its coefficients), we have
\begin{equation}
\sum_j(2j+1)(2j''+1)
\left\{\begin{matrix}j_1&j_2&j'\\j_3&j_4&j\end{matrix}\right\}
\left\{\begin{matrix}j_3&j_2&j\\j_1&j_4&j''\end{matrix}\right\}
=\delta_{j'j''}.
\tag{108.7}
\end{equation}
Next consider three schemes for coupling three angular momenta, whose intermediate sums are respectively $j_{12},j_{23}$, and $j_{31}$. By the rule for matrix multiplication, the corresponding transformation coefficients (108.6) are related by
\[
\sum_{j_{23}}\langle j_{12}|j_{23}\rangle\langle j_{23}|j_{31}\rangle=\langle j_{12}|j_{31}\rangle.
\]

\SourcePage{538}{541}
Substitution of (108.6), followed by a change of index notation, gives
\begin{equation}
\sum_j(-1)^{j+j_3+j_6}(2j+1)
\left\{\begin{matrix}j_2&j_4&j_6\\j_1&j_5&j\end{matrix}\right\}
\left\{\begin{matrix}j_4&j_1&j_6\\j_2&j_5&j_3\end{matrix}\right\}
=\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}.
\tag{108.8}
\end{equation}
Finally, consideration of different schemes for coupling four angular momenta yields\footnote{See the book by Edmonds cited above.} the following addition formula for products of three $6j$ symbols:
\begin{equation}
\begin{aligned}
&\sum_j(-1)^{j+\sum_{1}^{9}j_i}(2j+1)
\left\{\begin{matrix}j_4&j_2&j_6\\j_9&j_8&j\end{matrix}\right\}
\left\{\begin{matrix}j_2&j_1&j_3\\j_7&j&j_9\end{matrix}\right\}
\left\{\begin{matrix}j_4&j_3&j_5\\j_7&j_8&j\end{matrix}\right\}\\
&\qquad=\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
\left\{\begin{matrix}j_6&j_1&j_5\\j_7&j_8&j_9\end{matrix}\right\}.
\end{aligned}
\tag{108.9}
\end{equation}
($L.\ C.\ Biedenharn$, $J.\ P.\ Elliott$, 1953).

\input{manuscript/chapters/ch14/fragments/sec108-line176-fresh-c-20260718.translation.en.tex}
\begingroup\footnotesize
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
={}&\Delta(j_1j_2j_3)\Delta(j_1j_5j_6)\Delta(j_4j_2j_6)\Delta(j_4j_5j_3)\\
&\times\sum_z\frac{(-1)^z(z+1)!}{(z-j_1-j_2-j_3)!(z-j_1-j_5-j_6)!(z-j_4-j_2-j_6)!(z-j_4-j_5-j_3)!}\\
&\times(j_1+j_2+j_4+j_5-z)!(j_2+j_3+j_5+j_6-z)!(j_3+j_1+j_6+j_4-z)! ,
\end{aligned}
\tag{108.10}
\end{equation}
\endgroup
where
\[
\Delta(abc)=\left[\frac{(a+b-c)!(a-b+c)!(-a+b+c)!}{(a+b+c+1)!}\right]^{1/2},
\]
\input{manuscript/chapters/ch14/fragments/sec108-line192-fresh-e-20260718.translation.en.tex}

We conclude with several remarks about higher-order scalars formed from $3j$ symbols.

After the $6j$ symbol, the next in complexity is the scalar formed by contracting a product of six $3j$ symbols. These $3j$ symbols contain 18 pairwise equal numbers $j$,

\SourcePage{539}{542}
so that the resulting scalar depends on nine parameters $j$. It is conventionally called a $9j$ symbol and is defined as follows ($E.\ Wigner$, 1951)\footnote{According to the general contraction rule (108.1), the arguments $m$ in the last three $3j$ symbols should be written with minus signs, and the factor $(-1)^{\sum(j-m)}$ should be inserted under the summation sign. Using property (106.6) of the $3j$ symbols, however, and noting that in this case, as is readily seen, the sum $\sum m$ over all nine numbers $m$ is zero, we arrive at definition (108.11).}:
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_{11}&j_{12}&j_{13}\\j_{21}&j_{22}&j_{23}\\j_{31}&j_{32}&j_{33}\end{matrix}\right\}
={}&\sum_{\text{all }m}
\begin{pmatrix}j_{11}&j_{12}&j_{13}\\m_{11}&m_{12}&m_{13}\end{pmatrix}
\begin{pmatrix}j_{21}&j_{22}&j_{23}\\m_{21}&m_{22}&m_{23}\end{pmatrix}\\
&\times\begin{pmatrix}j_{31}&j_{32}&j_{33}\\m_{31}&m_{32}&m_{33}\end{pmatrix}
\begin{pmatrix}j_{11}&j_{21}&j_{31}\\m_{11}&m_{21}&m_{31}\end{pmatrix}\\
&\times\begin{pmatrix}j_{12}&j_{22}&j_{32}\\m_{12}&m_{22}&m_{32}\end{pmatrix}
\begin{pmatrix}j_{13}&j_{23}&j_{33}\\m_{13}&m_{23}&m_{33}\end{pmatrix}.
\end{aligned}
\tag{108.11}
\end{equation}

\begin{table}[p]
\centering\small
\caption{Formulas for $6j$ symbols}
\label{tab:10}
\renewcommand{\arraystretch}{1.55}
\begin{tabular}{@{}l@{}}
\hline
$\displaystyle \left\{\begin{matrix}a&b&c\\0&c&b\end{matrix}\right\}=\frac{(-1)^s}{\sqrt{(2b+1)(2c+1)}},\qquad s=a+b+c$\\ \hline
$\displaystyle \left\{\begin{matrix}a&b&c\\\frac12&c-\frac12&b+\frac12\end{matrix}\right\}=(-1)^s\left[\frac{(s-2b)(s-2c+1)}{(2b+1)(2b+2)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\\frac12&c-\frac12&b-\frac12\end{matrix}\right\}=(-1)^s\left[\frac{(s+1)(s-2a)}{2b(2b+1)2c(2c+1)}\right]^{1/2}$\\ \hline
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b-1\end{matrix}\right\}=(-1)^s\left[\frac{s(s+1)(s-2a-1)(s-2a)}{(2b-1)2b(2b+1)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b\end{matrix}\right\}=(-1)^s\left[\frac{2(s+1)(s-2a)(s-2b)(s-2c+1)}{2b(2b+1)(2b+2)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b+1\end{matrix}\right\}=(-1)^s\left[\frac{(s-2b-1)(s-2b)(s-2c+1)(s-2c+2)}{(2b+1)(2b+2)(2b+3)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c&b\end{matrix}\right\}=(-1)^{s+1}\frac{2[b(b+1)+c(c+1)-a(a+1)]}{[2b(2b+1)(2b+2)2c(2c+1)(2c+2)]^{1/2}}$\\ \hline
\end{tabular}
\end{table}

This quantity can also be represented as a sum of

\SourcePage{540}{543}
products of three $6j$ symbols:
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_{11}&j_{12}&j_{13}\\j_{21}&j_{22}&j_{23}\\j_{31}&j_{32}&j_{33}\end{matrix}\right\}
={}&\sum_j(-1)^{2j}(2j+1)
\left\{\begin{matrix}j_{11}&j_{21}&j_{31}\\j_{32}&j_{33}&j\end{matrix}\right\}\\
&\times\left\{\begin{matrix}j_{12}&j_{22}&j_{32}\\j_{21}&j&j_{23}\end{matrix}\right\}
\left\{\begin{matrix}j_{13}&j_{23}&j_{33}\\j&j_{11}&j_{12}\end{matrix}\right\}.
\end{aligned}
\tag{108.12}
\end{equation}
\input{manuscript/chapters/ch14/fragments/sec108-lines245-249-fresh-e-20260718.translation.en.tex}
