\SourcePage{520}{523}
\section{$3j$ Symbols}

The rule for adding angular momenta obtained in \S~31 determines the possible values of the total angular momentum of a system composed of two particles (or more complex parts) having angular momenta $j_1$ and $j_2$\footnote{Strictly, throughout this discussion we tacitly mean a system whose parts interact so weakly that their angular momenta may, in the first approximation, be regarded as conserved. The results below apply, of course, not only to the addition of the total angular momenta of two particles (or systems), but also to the addition of the orbital angular momentum and spin of one and the same system when the spin--orbit coupling is sufficiently weak.}. This rule is in fact closely connected with the behavior of wave functions under spatial rotations and follows directly from the properties of spinors.

The wave functions of particles with angular momenta $j_1$ and $j_2$ are symmetric spinors of ranks $2j_1$ and $2j_2$, while the system's wave function is their product
\begin{equation}
 \psi^{(1)}_{\lambda\mu\ldots}\psi^{(2)}_{\rho\sigma\ldots}.
 \tag{106.1}
\end{equation}
Symmetrizing this product over all indices gives a symmetric spinor of rank $2(j_1+j_2)$, corresponding to a state of total angular momentum $j_1+j_2$. Next contract the product (106.1) over one pair of indices, one belonging to $\psi^{(1)}$ and the other to $\psi^{(2)}$ (otherwise the result is zero). Because each spinor is symmetric, it is immaterial which indices are chosen from $\lambda,\mu,\ldots$ and $\rho,\sigma,\ldots$ After symmetrization we obtain a symmetric spinor of rank $2(j_1+j_2-1)$, corresponding to angular momentum $j_1+j_2-1$\footnote{To avoid confusion, the following observation is useful. The wave function of a two-particle system is always a spinor of rank $2(j_1+j_2)$, which in general differs from $2j$, where $j$ is the system's total angular momentum. Such a spinor may, however, be equivalent to one of lower rank. For example, the wave function of two particles with $j_1=j_2=1/2$ is a second-rank spinor; if the total angular momentum is $j=0$, this spinor is antisymmetric and therefore reduces to a scalar. In general, $j$ determines the symmetry of the system's spinor wave function: it is symmetric in $2j$ indices and antisymmetric in the remaining indices.}.

\SourcePage{521}{524}
Continuing this procedure, we find in accordance with the rule already known that $j$ assumes, once each, the values from $j_1+j_2$ down to $|j_1-j_2|$.

Mathematically, this is the decomposition of the direct product $D^{(j_1)}\times D^{(j_2)}$ of two irreducible representations of the rotation group (of dimensions $2j_1+1$ and $2j_2+1$) into irreducible parts. In these terms the angular-momentum addition rule is
\[
 D^{(j_1)}\times D^{(j_2)}=D^{(j_1+j_2)}+D^{(j_1+j_2-1)}+\cdots+D^{(|j_1-j_2|)}.
\]

To solve the angular-momentum addition problem completely, we must also determine how to construct the wave function of a system with a specified total angular momentum from the wave functions of its two constituent particles.

We begin with the simplest case: two angular momenta adding to zero. Evidently $j_1=j_2$ and $m_1=-m_2$. Let $\psi_{jm}$ be normalized wave functions for the states of one particle with angular momentum $j$ and projection $m$ (in a nonspinor representation). The required system wave function $\psi_0$ is a sum of products of the two particles' wave functions with opposite values of $m$:
\begin{equation}
 \psi_0=\frac{1}{\sqrt{2j+1}}\sum_{m=-j}^{j}(-1)^{j-m}\psi^{(1)}_{jm}\psi^{(2)}_{j,-m}.
 \tag{106.2}
\end{equation}
Here $j$ is the common value of $j_1$ and $j_2$. The factor before the sum follows from normalization. All coefficients in the sum must have the same absolute value, since all projections $m$ of the particle angular momenta are equally probable. Their alternating signs in (106.2) are readily found from the spinor representation. In spinor notation the sum in (106.2) is the scalar (the system's total angular momentum is zero)
\begin{equation}
 \psi^{(1)}_{\lambda\mu\ldots}\psi^{(2)\lambda\mu\ldots},
 \tag{106.3}
\end{equation}
formed from two spinors of rank $2j$. The signs in (106.2) then follow directly from (57.3).

Only the relative signs of the terms in (106.2) are in general unambiguous; the overall sign may depend on the ``order of addition'' of the angular momenta. Indeed, lowering every spinor index of $\psi^{(1)}$ (there are $j+m$ ones and $j-m$ twos) and raising those of $\psi^{(2)}$ multiplies the scalar (106.3) by $(-1)^{2j}$, so its sign changes for half-integral $j$.

\SourcePage{522}{525}
Now consider a system of three particles, with angular momenta $j_1,j_2,j_3$ and projections $m_1,m_2,m_3$, whose total angular momentum is zero. This condition implies $m_1+m_2+m_3=0$ and requires each $j_i$ to be obtainable by vector addition of the other two. Geometrically, $j_1,j_2,j_3$ must be the sides of a closed triangle; thus each is no smaller than the difference and no greater than the sum of the other two:
\[
 |j_1-j_2|\leq j_3\leq j_1+j_2\quad\text{and so on}.
\]
The algebraic sum $j_1+j_2+j_3$ is therefore an integer.

The wave function of this system has the form
\begin{equation}
 \psi_0=\sum_{m_1m_2m_3}
 \begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
 \psi^{(1)}_{j_1m_1}\psi^{(2)}_{j_2m_2}\psi^{(3)}_{j_3m_3},
 \tag{106.4}
\end{equation}
where each $m_i$ is summed from $-j_i$ to $j_i$. The coefficients in this expression are called Wigner $3j$ symbols. By definition, they are nonzero only when $m_1+m_2+m_3=0$.

Permuting the indices 1, 2, 3 can change (106.4) only by an inessential phase factor. The $3j$ symbols can in fact be defined as purely real (see below), so the only ambiguity in $\psi_0$ is its overall sign, as for (106.2). Hence a permutation of the columns of a $3j$ symbol either leaves it unchanged or reverses its sign.

The customary definition uses the following, maximally symmetric prescription. In spinor notation $\psi_0$ is a scalar formed from the product of three spinors $\psi^{(1)}_{\lambda\mu\ldots}$, $\psi^{(2)}_{\lambda\mu\ldots}$, $\psi^{(3)}_{\lambda\mu\ldots}$ by contracting all index pairs, each pair belonging to two different spinors. For every pair belonging to particles 1 and 2, write the index upstairs on $\psi^{(1)}$ and downstairs on $\psi^{(2)}$; for a pair belonging to particles 2 and 3, upstairs on $\psi^{(2)}$ and downstairs on $\psi^{(3)}$; and for a pair belonging to particles 3 and 1, upstairs on $\psi^{(3)}$ and downstairs on $\psi^{(1)}$. There are respectively $j_1+j_2-j_3$, $j_2+j_3-j_1$, and $j_1+j_3-j_2$ pairs of these three kinds. This prescription fixes the sign of $\psi_0$ uniquely.

\SourcePage{523}{526}
A cyclic permutation of 1, 2, 3 plainly leaves $\psi_0$ unchanged, and therefore does not change the $3j$ symbol. Interchanging any two indices requires raising the lower and lowering the upper indices in all $j_1+j_2+j_3$ pairs. Thus $\psi_0$ is multiplied by $(-1)^{j_1+j_2+j_3}$, and
\begin{equation}
 \begin{pmatrix}j_2&j_1&j_3\\m_2&m_1&m_3\end{pmatrix}
 =(-1)^{j_1+j_2+j_3}\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
 \quad\text{and so on}.
 \tag{106.5}
\end{equation}
Thus the sign changes when two columns are interchanged if $j_1+j_2+j_3$ is odd.

Finally,
\begin{equation}
 \begin{pmatrix}j_1&j_2&j_3\\-m_1&-m_2&-m_3\end{pmatrix}
 =(-1)^{j_1+j_2+j_3}\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}.
 \tag{106.6}
\end{equation}
Indeed, reversing the signs of all $z$ components may be regarded as a rotation through $\pi$ about the $y$ axis. This transformation is equivalent to raising every lower and lowering every upper spinor index; see (58.5).

Equation (106.4) leads to an important formula for the wave function $\psi_{jm}$ of a two-particle system with prescribed $j$ and $m$. Regard particles 1 and 2 together as a single system. Since its angular momentum $j$ combines with that of particle 3 to give zero, $j=j_3$ and $m=-m_3$. From (106.2),
\begin{equation}
 \psi_0=\frac1{\sqrt{2j+1}}\sum_m(-1)^{j-m}\psi_{jm}\psi^{(3)}_{j,-m}.
 \tag{106.7}
\end{equation}
\SourcePage{524}{527}
This must be compared with (106.4), with $j,-m$ written for $j_3,m_3$. First, however, the contraction convention in (106.7), as specified by (106.3), must be reconciled with that in (106.4). This requires interchanging upper and lower indices in the pairs belonging to particles 1 and 3, and introduces an additional factor $(-1)^{j_1-j_2+j_3}$. We obtain\footnote{Under time reversal the wave functions transform, according to (60.2), as $\psi_{jm}\to(-1)^{j-m}\psi_{j,-m}$. It is readily verified that applying this transformation to $\psi_{j_1m_1}$ and $\psi_{j_2m_2}$ on the right-hand side of (106.8) transforms $\psi_{jm}$ on the left in the same way.}
\begin{equation}
 \psi_{jm}=(-1)^{j_1-j_2+m}\sqrt{2j+1}\sum_{m_1m_2}
 \begin{pmatrix}j_1&j_2&j\\m_1&m_2&-m\end{pmatrix}
 \psi^{(1)}_{j_1m_1}\psi^{(2)}_{j_2m_2},
 \tag{106.8}
\end{equation}
where $m_1,m_2$ are summed subject to $m_1+m_2=m$.

Formula (106.8) is the desired expansion of the system wave function in wave functions of the two particles with specified angular momenta $j_1,j_2$. It may be written
\begin{equation}
 \psi_{jm}=\sum_{m_1m_2}(m_1m_2\mid jm)\psi^{(1)}_{j_1m_1}\psi^{(2)}_{j_2m_2},\qquad m_2=m-m_1.
 \tag{106.9}
\end{equation}
The coefficients
\begin{equation}
 (m_1m_2\mid jm)=(-1)^{j_1-j_2+m}\sqrt{2j+1}
 \begin{pmatrix}j_1&j_2&j\\m_1&m_2&-m\end{pmatrix}
 \tag{106.10}
\end{equation}
form the transformation matrix from the complete orthonormal set of $(2j_1+1)(2j_2+1)$ state wave functions $|m_1m_2\rangle$ to the corresponding set $|jm\rangle$, for fixed $j_1,j_2$. They are called vector-addition coefficients, or Clebsch--Gordan coefficients. The notation $(m_1m_2\mid jm)$ follows the general notation (11.18) for expansion coefficients between two systems of functions. For brevity the quantum numbers $j_1,j_2$, common to both systems, have been omitted; when needed they are included as $(j_1m_1j_2m_2\mid j_1j_2jm)$\footnote{The notations $C^{jm}_{j_1m_1j_2m_2}$ and $C^{j_1j_2j}_{m_1m_2m}$ are also used for Clebsch--Gordan coefficients.}.

The transformation (106.9) is unitary (see \S~12). Hence the coefficients of the inverse transformation
\begin{equation}
 \psi^{(1)}_{j_1m_1}\psi^{(2)}_{j_2m_2}
 =\sum_{j=|j_2-j_1|}^{j_1+j_2}(jm\mid m_1m_2)\psi_{jm}
 \tag{106.11}
\end{equation}
are the complex conjugates of those in (106.9). As shown below, these coefficients are real; therefore

\SourcePage{525}{528}
\[
 (m_1m_2\mid jm)=(jm\mid m_1m_2).
\]
By the general rules of quantum mechanics, the squares of the expansion coefficients in (106.11) give the probabilities for the system to have the various values $j,m$, when $j_1,m_1$ and $j_2,m_2$ are specified.

Unitarity of (106.9) gives the orthogonality relations
\begin{align}
 \sum_{m_1m_2}(m_1m_2\mid jm)(m_1m_2\mid j'm')&=\delta_{jj'}\delta_{mm'}, \tag{106.12}\\
 \sum_{jm}(m_1m_2\mid jm)(m'_1m'_2\mid jm)&=\delta_{m_1m'_1}\delta_{m_2m'_2}. \tag{106.13}
\end{align}

The explicit general expression for a $3j$ symbol is rather cumbersome. It can be written\footnote{The expansion coefficients (106.9) were first calculated by Wigner (1931). Their symmetry properties and the symmetric expression (106.14) were found by Racah (G. Racah, 1942). Probably the most direct calculation passes from the suitably normalized spinor representation of $\psi_0$ to the sum (106.4), using the correspondence formula (57.6); the reality of its coefficient automatically establishes the reality of the $3j$ symbols. Another derivation is given in Edmonds's book (see the note on p.~272). The table below is also taken from that book.}
\begingroup\small
\begin{align}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
={}&\left[\frac{(j_1+j_2-j_3)!(j_1-j_2+j_3)!(-j_1+j_2+j_3)!}{(j_1+j_2+j_3+1)!}\right]^{1/2}\notag\\
&\times[(j_1+m_1)!(j_1-m_1)!(j_2+m_2)!(j_2-m_2)!(j_3+m_3)!(j_3-m_3)!]^{1/2}\notag\\
&\times\sum_z\frac{(-1)^{z+j_1-j_2-m_3}}{z!(j_1+j_2-j_3-z)!(j_1-m_1-z)!(j_2+m_2-z)!}\notag\\[-2pt]
&\hspace{25mm}\times\frac1{(j_3-j_2+m_1+z)!(j_3-j_1-m_2+z)!}.
\tag{106.14}
\end{align}
\endgroup
The sum extends over all integral $z$; since the factorial of a negative number is infinite, only finitely many terms actually contribute. The factor preceding the sum is manifestly symmetric in 1, 2, 3; symmetry of the sum appears after an appropriate change of summation variable.

\SourcePage{526}{529}
Besides (106.5) and (106.6), which follow simply from the definition, $3j$ symbols possess further symmetry properties whose derivation is more involved and will not be given here. They are conveniently stated by introducing the square $(3\times3)$ array
\begin{equation}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\longleftrightarrow
\begin{bmatrix}
j_2+j_3-j_1&j_3+j_1-j_2&j_1+j_2-j_3\\
j_1-m_1&j_2-m_2&j_3-m_3\\
j_1+m_1&j_2+m_2&j_3+m_3
\end{bmatrix}.
\tag{106.15}
\end{equation}
The sum of the entries in every row and column is $j_1+j_2+j_3$. Then: 1) interchanging any two columns multiplies the $3j$ symbol by $(-1)^{j_1+j_2+j_3}$, reproducing (106.5); 2) the same holds for any two rows (for the last two rows this reproduces (106.6)); 3) the symbol is unchanged when rows and columns are interchanged\footnote{See T. Regge, Nuovo Cimento, 1958, vol.~10, p.~544; 1959, vol.~11, p.~116. For deeper mathematical aspects of (106.15), and of the $6j$-symbol property (108.3) below, see the review by Ya.~A. Smorodinsky and L.~A. Shelepin, Sov. Phys. Usp., 1972, vol.~15, p.~1.}.

Some simpler special cases are
\begin{equation}
 \begin{pmatrix}j&j&0\\m&-m&0\end{pmatrix}=\frac{(-1)^{j-m}}{\sqrt{2j+1}},
 \tag{106.16}
\end{equation}
which corresponds to (106.2), and
\begingroup\small
\begin{align}
\begin{pmatrix}j_1&j_2&j_1+j_2\\m_1&m_2&-m_1-m_2\end{pmatrix}
={}&(-1)^{j_1-j_2+m_1+m_2}\notag\\[-2pt]
&\times\left[\frac{(2j_1)!(2j_2)!(j_1+j_2+m_1+m_2)!(j_1+j_2-m_1-m_2)!}{(j_1+j_2+1)!(j_1+m_1)!(j_1-m_1)!(j_2+m_2)!(j_2-m_2)!}\right]^{1/2},
\tag{106.17}
\end{align}
\endgroup
which follows directly from (106.14).

\SourcePage{527}{530}
The derivation of
\begin{align}
\begin{pmatrix}j_1&j_2&j_3\\0&0&0\end{pmatrix}
={}&(-1)^p\left[\frac{(j_1+j_2-j_3)!(j_1-j_2+j_3)!(-j_1+j_2+j_3)!}{(2p+1)!}\right]^{1/2}\notag\\[-2pt]
&\times\frac{p!}{(p-j_1)!(p-j_2)!(p-j_3)!},
\tag{106.18}
\end{align}
where $2p=j_1+j_2+j_3$ is even, requires additional calculation\footnote{See the book by Edmonds cited above.}. For odd $2p$ this $3j$ symbol vanishes by (106.6).

For reference, Table~9 lists $3j$ symbols for $j_3=1/2,1,3/2,2$. For every $j_3$ it gives the smallest set from which all remaining symbols follow using (106.5) and (106.6).

\setcounter{table}{8}
\begin{table}[htbp]
\centering\small
\caption{Formulas for $3j$ symbols}
\renewcommand{\arraystretch}{1.55}
\begin{tabular}{@{}c c l@{}}
\toprule
$j_3$ & $j_1$ & \multicolumn{1}{c}{Value of $(-1)^{j-m}\begin{pmatrix}j_1&j&j_3\\m_1&-m-m_3&m_3\end{pmatrix}$}\\
\midrule
$\frac12$ & $j$ & $m_3=\frac12:\quad -\left[\dfrac{j-m}{(2j+1)(2j+2)}\right]^{1/2}$\\
 & $j+1$ & $\left[\dfrac{j+m+1}{(2j+1)(2j+2)}\right]^{1/2}$\\
\midrule
$1$ & $j$ & $m_3=0:\quad \dfrac{2m}{[2j(2j+1)(2j+2)]^{1/2}}$\\
 & $j+1$ & $\left[\dfrac{2(j+m+1)(j-m+1)}{(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
 & $j$ & $m_3=1:\quad \left[\dfrac{2(j-m)(j+m+1)}{2j(2j+1)(2j+2)}\right]^{1/2}$\\
 & $j+1$ & $\left[\dfrac{(j-m)(j-m+1)}{(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
\bottomrule
\end{tabular}
\end{table}

\SourcePage{528}{531}
\begin{table}[htbp]
\centering\scriptsize
\caption*{Table 9 (continued)}
\renewcommand{\arraystretch}{1.55}
\begin{tabular}{@{}c c l@{}}
\toprule
$j_3$ & $j_1$ & \multicolumn{1}{c}{Value}\\
\midrule
$\frac32$ & $j+\frac12$ & $m_3=\frac12:\quad -(j+3m+\frac32)\left[\dfrac{j-m+\frac12}{2j(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
&$j+\frac32$&$\left[\dfrac{3(j-m+\frac12)(j-m+\frac32)(j+m+\frac32)}{(2j+1)(2j+2)(2j+3)(2j+4)}\right]^{1/2}$\\
&$j+\frac12$&$m_3=\frac32:\quad-\left[\dfrac{3(j-m-\frac12)(j-m+\frac12)(j+m+\frac32)}{2j(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
&$j+\frac32$&$\left[\dfrac{(j-m-\frac12)(j-m+\frac12)(j-m+\frac32)}{(2j+1)(2j+2)(2j+3)(2j+4)}\right]^{1/2}$\\
\midrule
$2$&$j$&$m_3=0:\quad\dfrac{2[3m^2-j(j+1)]}{[(2j-1)2j(2j+1)(2j+2)(2j+3)]^{1/2}}$\\
&$j+1$&$-2m\left[\dfrac{6(j+m+1)(j-m+1)}{2j(2j+1)(2j+2)(2j+3)(2j+4)}\right]^{1/2}$\\
&$j+2$&$\left[\dfrac{6(j+m+2)(j+m+1)(j-m+2)(j-m+1)}{(2j+1)(2j+2)(2j+3)(2j+4)(2j+5)}\right]^{1/2}$\\
&$j$&$m_3=1:\quad(1+2m)\left[\dfrac{6(j+m+1)(j-m)}{(2j-1)2j(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
&$j+1$&$-2(j+2m+2)\left[\dfrac{(j-m+1)(j-m)}{2j(2j+1)(2j+2)(2j+3)(2j+4)}\right]^{1/2}$\\
&$j+2$&$2\left[\dfrac{(j+m+2)(j-m+2)(j-m+1)(j-m)}{(2j+1)(2j+2)(2j+3)(2j+4)(2j+5)}\right]^{1/2}$\\
&$j$&$m_3=2:\quad\left[\dfrac{6(j-m-1)(j-m)(j+m+1)(j+m+2)}{(2j-1)2j(2j+1)(2j+2)(2j+3)}\right]^{1/2}$\\
&$j+1$&$-2\left[\dfrac{(j-m-1)(j-m)(j-m+1)(j+m+2)}{2j(2j+1)(2j+2)(2j+3)(2j+4)}\right]^{1/2}$\\
&$j+2$&$\left[\dfrac{(j-m-1)(j-m)(j-m+1)(j-m+2)}{(2j+1)(2j+2)(2j+3)(2j+4)(2j+5)}\right]^{1/2}$\\
\bottomrule
\end{tabular}
\end{table}

\SourcePage{529}{532}
\ProblemHeading

Determine the angular dependence of the wave functions of a spin-$1/2$ particle in states with specified orbital angular momentum $l$, total angular momentum $j$, and projection $m$.

\SolutionHeading The general formula (106.8) solves the problem if $\psi^{(1)}$ denotes an orbital-angular-momentum eigenfunction (the spherical function $Y_{lm_l}$) and $\psi^{(2)}$ the spin wave function $\chi(\sigma)$, where $\sigma=\pm1/2$:
\[
 \psi_{jm}=(-1)^{l+m-1/2}\sqrt{2j+1}\sum_\sigma
 \begin{pmatrix}l&\frac12&j\\m-\sigma&\sigma&-m\end{pmatrix}Y_{l,m-\sigma}\chi(\sigma).
\]
Substitution of the $3j$ symbols gives
\begin{align*}
 \psi_{l+1/2,m}&=\sqrt{\frac{j+m}{2j}}\,\chi\!\left(\frac12\right)Y_{l,m-1/2}
 +\sqrt{\frac{j-m}{2j}}\,\chi\!\left(-\frac12\right)Y_{l,m+1/2},\\
 \psi_{l-1/2,m}&=-\sqrt{\frac{j-m+1}{2j+2}}\,\chi\!\left(\frac12\right)Y_{l,m-1/2}
 +\sqrt{\frac{j+m+1}{2j+2}}\,\chi\!\left(-\frac12\right)Y_{l,m+1/2}.
\end{align*}

% The assigned range ends immediately before the heading of \S 107.
