\SourcePage{533}{536}
\section{$6j$ symbols}

In \S~106 we defined the $3j$ symbols as the coefficients in the sum (106.4), which represents the wave function of a system of three particles having zero total angular momentum. From the standpoint of transformation properties under rotations, this sum is a scalar. It follows that the set of $3j$ symbols for specified values of $j_1,j_2,j_3$ (and all possible $m_1,m_2,m_3$) may be regarded as a collection of quantities which, under rotations, transform by a law contragredient to the transformation law of the products $\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3}$, thereby ensuring invariance of the whole sum.

This point of view suggests the question of constructing a scalar made solely from $3j$ symbols. Such a scalar must depend only on the numbers $j$, and not on the

\SourcePage{534}{537}
numbers $m$, which change under rotations. In other words, it must be expressible as sums over all the numbers $m$. Each such summation consists in contracting a product of two $3j$ symbols according to the rule
\begin{equation}
 \sum_m(-1)^{j-m}
 \begin{pmatrix}j&\cdots\\m&\cdots\end{pmatrix}
 \begin{pmatrix}j&\cdots\\-m&\cdots\end{pmatrix}
 \tag{108.1}
\end{equation}
(compare the construction of the scalar (106.2)).

Since each contraction involves a pair of numbers $m$, products of an even number of $3j$ symbols must be considered in order to form a complete scalar. By their orthogonality property, contraction of the product of two $3j$ symbols gives the trivial result
\[
\begin{aligned}
&\sum_{m_1m_2m_3}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\begin{pmatrix}j_1&j_2&j_3\\-m_1&-m_2&-m_3\end{pmatrix}
(-1)^{j_1+j_2+j_3-m_1-m_2-m_3}\\
&\qquad=\sum_{m_1m_2m_3}
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}^{2}=1
\end{aligned}
\]
(here the equality $m_1+m_2+m_3=0$ and formulas (106.6) and (106.12) have been used). Hence the smallest number of factors needed to obtain a nontrivial scalar is four. In every $3j$ symbol the three numbers $j$ form a geometrically closed triangle. Since, in a contraction, each number $j$ must occur in two $3j$ symbols, it is clear that a scalar constructed from a product of four $3j$ symbols contains six numbers $j$. Geometrically, they must be represented by the edge lengths of an irregular tetrahedron (Fig.~45), with each $3j$ symbol corresponding to one face. The definition of the required scalar adopts a definite convention for carrying out the contractions, expressed by the formula
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
={}&\sum_{\text{all }m}(-1)^{\sum_i(j_i-m_i)}
\begin{pmatrix}j_1&j_2&j_3\\-m_1&-m_2&-m_3\end{pmatrix}
\begin{pmatrix}j_1&j_5&j_6\\m_1&-m_5&m_6\end{pmatrix}\\
&\times\begin{pmatrix}j_4&j_2&j_6\\m_4&m_2&-m_6\end{pmatrix}
\begin{pmatrix}j_4&j_5&j_3\\-m_4&m_5&m_3\end{pmatrix}.
\end{aligned}
\tag{108.2}
\end{equation}
The summation here extends over all possible values of all the numbers $m$; since, however, the sum of the three $m$'s in each $3j$ symbol must vanish, only

\SourcePage{535}{538}
three of the six $m$'s are actually independent. The quantities defined by (108.2) are called $6j$ symbols, or Racah coefficients\footnote{The notation
\[
W(j_1j_2j_4j_5;j_3j_6)=(-1)^{j_1+j_2+j_4+j_5}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
\]
is also used in the literature.}.

From definition (108.2), with the symmetry properties of the $3j$ symbols taken into account, it is readily verified that a $6j$ symbol is unchanged by any permutation of its three columns, and that the upper and lower entries may be interchanged in any pair of columns. Owing to these symmetry properties, the sequence of numbers $j_1\ldots j_6$ in a $6j$ symbol can be displayed in 24 equivalent forms\footnote{If the tetrahedron in Fig.~45 is imagined to be regular, the 24 equivalent permutations of the numbers $j$ can be obtained by applying the 24 symmetry transformations (rotations and reflections) of the tetrahedron.}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[x=1cm,y=1cm,line width=.45pt,line cap=round,font=\small]
 \coordinate (L) at (-1.55,0.45); \coordinate (R) at (1.55,0.45);
 \coordinate (T) at (0.55,1.65); \coordinate (B) at (-0.65,-1.85);
 \draw (L)--(R)--(T)--(L)--(B)--(R);
 \draw[dashed] (T)--(B);
 \node[above left] at ($(L)!0.56!(T)$) {$j_1$};
 \node[above right] at ($(T)!0.48!(R)$) {$j_2$};
 \node[below] at ($(L)!0.50!(R)$) {$j_3$};
 \node[right] at ($(R)!0.48!(B)$) {$j_4$};
 \node[left] at ($(L)!0.50!(B)$) {$j_5$};
 \node[right] at ($(T)!0.55!(B)$) {$j_6$};
\end{tikzpicture}
\caption{}
\label{fig:45}
\end{figure}

The $6j$ symbols also possess another, less evident, symmetry property, which gives an equality between symbols with different sets of numbers $j$:
\begin{equation}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}=
\left\{\begin{matrix}
j_1&\frac12(j_2+j_5+j_3-j_6)&\frac12(j_3+j_6+j_2-j_5)\\
j_4&\frac12(j_2+j_5+j_6-j_3)&\frac12(j_3+j_6+j_5-j_2)
\end{matrix}\right\}.
\tag{108.3}
\end{equation}
($T.\ Regge$, 1959)\footnote{See the note on p.~529.}.

We give a useful relation between the $6j$ and $3j$ symbols which follows from definition (108.2):
\begin{equation}
\begin{aligned}
&\sum_{m_4m_5m_6}(-1)^{j_4+j_5+j_6-m_4-m_5-m_6}
\begin{pmatrix}j_1&j_5&j_6\\m_1&-m_5&m_6\end{pmatrix}
\begin{pmatrix}j_4&j_2&j_6\\m_4&m_2&-m_6\end{pmatrix}\\
&\qquad\times\begin{pmatrix}j_4&j_5&j_3\\-m_4&m_5&m_3\end{pmatrix}
=\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}.
\end{aligned}
\tag{108.4}
\end{equation}
The expression summed on the left-hand side differs from the sum in (108.2) by the absence of one factor (a $3j$ symbol). The sum in (108.4) may therefore be represented by a tetrahedron (Fig.~45) with one face removed; this is what distinguishes the sum from a scalar. In other words, in its transfor-

\SourcePageJoin{536}{539}
mation properties it corresponds to a single $3j$ symbol---the one on the right-hand side of (108.4), to which it must be proportional. The coefficient of proportionality (the $6j$ symbol on the right-hand side) is readily obtained by multiplying both sides by
\[
\begin{pmatrix}j_1&j_2&j_3\\m_1&m_2&m_3\end{pmatrix}
\]
and summing over the remaining numbers $m_1,m_2,m_3$.

The $6j$ symbols arise naturally in the following question connected with the addition of three angular momenta.

Let three angular momenta $j_1,j_2,j_3$ combine to give a resultant angular momentum $J$. Specification of $J$ (and of its projection $M$) does not, however, determine the state of the system uniquely; the state also depends on the manner in which the angular momenta are added (or, as it is called, on their coupling scheme).

Consider, for example, the following two coupling schemes: 1) first $j_1$ and $j_2$ are combined into the total angular momentum $j_{12}$, after which $j_{12}$ and $j_3$ are combined into the final angular momentum $J$; 2) $j_2$ and $j_3$ are combined into $j_{23}$, and then $j_{23}$ and $j_1$ into $J$. The first scheme corresponds to states in which (besides $j_1,j_2,j_3,J,M$) the quantity $j_{12}$ has a definite value; denote their wave functions by $\psi_{j_{12}JM}$, omitting for brevity the repeated indices $j_1j_2j_3$. Similarly, denote the wave functions of the second coupling scheme by $\psi_{j_{23}JM}$. In both cases the values of the ``intermediate'' angular momentum ($j_{12}$ or $j_{23}$) are in general not unique. Thus, for specified $J,M$, there are two different sets of states distinguished by the values of $j_{12}$ or $j_{23}$. By the general rules, the functions of these two sets are related by a certain unitary transformation
\begin{equation}
 \psi_{j_{23}JM}=\sum_{j_{12}}\langle j_{12}|j_{23}\rangle\psi_{j_{12}JM}.
 \tag{108.5}
\end{equation}
It is evident on physical grounds that the coefficients of this transformation do not depend on $M$: they cannot depend on the orientation of the entire system in space. Thus they depend only on the values of the six angular momenta $j_1,j_2,j_3,j_{12},j_{23},J$, not on their projections; that is, they are scalar quantities in the sense stated above. These coefficients can be evaluated directly as follows.

\SourcePage{537}{540}
Applying formula (106.9) twice, we find
\[
\begin{aligned}
\psi_{j_{23}JM}&=\sum_{(m)}\langle m_1m_{23}|JM\rangle\psi_{j_1m_1}\psi_{j_{23}m_{23}}\\
&=\sum_{(m)}\langle m_1m_{23}|JM\rangle\langle m_2m_3|j_{23}m_{23}\rangle
\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3},\\
\psi_{j_{12}JM}&=\sum_{(m)}\langle m_3m_{12}|JM\rangle\langle m_1m_2|j_{12}m_{12}\rangle
\psi_{j_1m_1}\psi_{j_2m_2}\psi_{j_3m_3}
\end{aligned}
\]
(the symbol $(m)$ beneath a summation sign means that the summation is over all the numbers $m_1,m_2,\ldots$ occurring in the expression). Using the orthonormality of the functions $\psi_{jm}$, we now obtain
\[
\begin{aligned}
\langle j_{12}|j_{23}\rangle&\equiv\int\psi_{j_{12}JM}^{*}\psi_{j_{23}JM}\,dq\\
&=\sum_{(m)}\langle m_3m_{12}|JM\rangle\langle m_1m_{23}|JM\rangle
\langle m_1m_2|j_{12}m_{12}\rangle\langle m_2m_3|j_{23}m_{23}\rangle.
\end{aligned}
\]
The sum on the right-hand side is taken at a specified value of $M$, but the result is in fact independent of $M$, for the reason given above. The summation may therefore be extended over the values of $M$ if a factor $1/(2J+1)$ is placed before the sum. Then, expressing the coefficients $\langle m_1m_2|jm\rangle$ in terms of $3j$ symbols according to (106.10), we obtain
\begin{equation}
\langle j_{12}|j_{23}\rangle=(-1)^{j_1+j_2+j_3+J}\sqrt{(2j_{12}+1)(2j_{23}+1)}
\left\{\begin{matrix}j_1&j_2&j_{12}\\j_3&J&j_{23}\end{matrix}\right\}.
\tag{108.6}
\end{equation}
The relation between the $6j$ symbols and the transformation coefficients (108.5) makes it easy to obtain several useful summation formulas for products of $6j$ symbols.

First, by the unitarity of transformation (108.5), and since its coefficients are real, we have
\begin{equation}
\sum_j(2j+1)(2j''+1)
\left\{\begin{matrix}j_1&j_2&j'\\j_3&j_4&j\end{matrix}\right\}
\left\{\begin{matrix}j_3&j_2&j\\j_1&j_4&j''\end{matrix}\right\}
=\delta_{j'j''}.
\tag{108.7}
\end{equation}
Next consider three coupling schemes for three angular momenta, with intermediate sums $j_{12},j_{23}$, and $j_{31}$, respectively. By the rule of matrix multiplication, the corresponding transformation coefficients (108.6) obey
\[
\sum_{j_{23}}\langle j_{12}|j_{23}\rangle\langle j_{23}|j_{31}\rangle=\langle j_{12}|j_{31}\rangle.
\]

\SourcePage{538}{541}
Substitution of (108.6), followed by a change of index notation, gives
\begin{equation}
\sum_j(-1)^{j+j_3+j_6}(2j+1)
\left\{\begin{matrix}j_2&j_4&j_6\\j_1&j_5&j\end{matrix}\right\}
\left\{\begin{matrix}j_4&j_1&j_6\\j_2&j_5&j_3\end{matrix}\right\}
=\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}.
\tag{108.8}
\end{equation}
Finally, consideration of different coupling schemes for four angular momenta yields\footnote{See the book by Edmonds cited above.} the following addition formula for products of three $6j$ symbols:
\begin{equation}
\begin{aligned}
&\sum_j(-1)^{j+\sum_{1}^{9}j_i}(2j+1)
\left\{\begin{matrix}j_4&j_2&j_6\\j_9&j_8&j\end{matrix}\right\}
\left\{\begin{matrix}j_2&j_1&j_3\\j_7&j&j_9\end{matrix}\right\}
\left\{\begin{matrix}j_4&j_3&j_5\\j_7&j_8&j\end{matrix}\right\}\\
&\qquad=\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
\left\{\begin{matrix}j_6&j_1&j_5\\j_7&j_8&j_9\end{matrix}\right\}.
\end{aligned}
\tag{108.9}
\end{equation}
($L.\ C.\ Biedenharn$, $J.\ P.\ Elliott$, 1953).

For reference, we give some explicit expressions for $6j$ symbols. In the general case a $6j$ symbol can be represented by the following sum:
\begingroup\footnotesize
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_1&j_2&j_3\\j_4&j_5&j_6\end{matrix}\right\}
={}&\Delta(j_1j_2j_3)\Delta(j_1j_5j_6)\Delta(j_4j_2j_6)\Delta(j_4j_5j_3)\\
&\times\sum_z\frac{(-1)^z(z+1)!}{(z-j_1-j_2-j_3)!(z-j_1-j_5-j_6)!(z-j_4-j_2-j_6)!(z-j_4-j_5-j_3)!}\\
&\times(j_1+j_2+j_4+j_5-z)!(j_2+j_3+j_5+j_6-z)!(j_3+j_1+j_6+j_4-z)! ,
\end{aligned}
\tag{108.10}
\end{equation}
\endgroup
where
\[
\Delta(abc)=\left[\frac{(a+b-c)!(a-b+c)!(-a+b+c)!}{(a+b+c+1)!}\right]^{1/2},
\]
and the sum is taken over all positive integral values of $z$ for which none of the factorials in the denominator has a negative argument. Table~10 gives formulas for $6j$ symbols in the cases where one of the parameters is $0$, $1/2$, or $1$.

We conclude with several remarks on scalars of higher order constructed from $3j$ symbols.

The scalar next in complexity after the $3j$ symbol is formed by contracting a product of six $3j$ symbols. These $3j$ symbols contain 18 numbers $j$ equal in pairs,

\SourcePage{539}{542}
so that the resulting scalar depends on nine parameters $j$. It is customarily called a $9j$ symbol and is defined as follows ($E.\ Wigner$, 1951)\footnote{According to the general contraction rule (108.1), the arguments $m$ in the last three $3j$ symbols should be written with minus signs and a factor $(-1)^{\sum(j-m)}$ should be inserted under the summation sign. Using property (106.6) of the $3j$ symbols, however, and noting that in the present case the sum $\sum m$ of all nine numbers $m$ is zero, as is readily seen, we arrive at definition (108.11).}:
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_{11}&j_{12}&j_{13}\\j_{21}&j_{22}&j_{23}\\j_{31}&j_{32}&j_{33}\end{matrix}\right\}
={}&\sum_{\text{all }m}
\begin{pmatrix}j_{11}&j_{12}&j_{13}\\m_{11}&m_{12}&m_{13}\end{pmatrix}
\begin{pmatrix}j_{21}&j_{22}&j_{23}\\m_{21}&m_{22}&m_{23}\end{pmatrix}\\
&\times\begin{pmatrix}j_{31}&j_{32}&j_{33}\\m_{31}&m_{32}&m_{33}\end{pmatrix}
\begin{pmatrix}j_{11}&j_{21}&j_{31}\\m_{11}&m_{21}&m_{31}\end{pmatrix}\\
&\times\begin{pmatrix}j_{12}&j_{22}&j_{32}\\m_{12}&m_{22}&m_{32}\end{pmatrix}
\begin{pmatrix}j_{13}&j_{23}&j_{33}\\m_{13}&m_{23}&m_{33}\end{pmatrix}.
\end{aligned}
\tag{108.11}
\end{equation}

\begin{table}[p]
\centering\small
\caption{Formulas for $6j$ symbols}
\label{tab:10}
\renewcommand{\arraystretch}{1.55}
\begin{tabular}{@{}l@{}}
\hline
$\displaystyle \left\{\begin{matrix}a&b&c\\0&c&b\end{matrix}\right\}=\frac{(-1)^s}{\sqrt{(2b+1)(2c+1)}},\qquad s=a+b+c$\\ \hline
$\displaystyle \left\{\begin{matrix}a&b&c\\\frac12&c-\frac12&b+\frac12\end{matrix}\right\}=(-1)^s\left[\frac{(s-2b)(s-2c+1)}{(2b+1)(2b+2)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\\frac12&c-\frac12&b-\frac12\end{matrix}\right\}=(-1)^s\left[\frac{(s+1)(s-2a)}{2b(2b+1)2c(2c+1)}\right]^{1/2}$\\ \hline
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b-1\end{matrix}\right\}=(-1)^s\left[\frac{s(s+1)(s-2a-1)(s-2a)}{(2b-1)2b(2b+1)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b\end{matrix}\right\}=(-1)^s\left[\frac{2(s+1)(s-2a)(s-2b)(s-2c+1)}{2b(2b+1)(2b+2)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c-1&b+1\end{matrix}\right\}=(-1)^s\left[\frac{(s-2b-1)(s-2b)(s-2c+1)(s-2c+2)}{(2b+1)(2b+2)(2b+3)(2c-1)2c(2c+1)}\right]^{1/2}$\\
$\displaystyle \left\{\begin{matrix}a&b&c\\1&c&b\end{matrix}\right\}=(-1)^{s+1}\frac{2[b(b+1)+c(c+1)-a(a+1)]}{[2b(2b+1)(2b+2)2c(2c+1)(2c+2)]^{1/2}}$\\ \hline
\end{tabular}
\end{table}

This quantity can also be represented as a sum

\SourcePage{540}{543}
of products of three $6j$ symbols:
\begin{equation}
\begin{aligned}
\left\{\begin{matrix}j_{11}&j_{12}&j_{13}\\j_{21}&j_{22}&j_{23}\\j_{31}&j_{32}&j_{33}\end{matrix}\right\}
={}&\sum_j(-1)^{2j}(2j+1)
\left\{\begin{matrix}j_{11}&j_{21}&j_{31}\\j_{32}&j_{33}&j\end{matrix}\right\}\\
&\times\left\{\begin{matrix}j_{12}&j_{22}&j_{32}\\j_{21}&j&j_{23}\end{matrix}\right\}
\left\{\begin{matrix}j_{13}&j_{23}&j_{33}\\j&j_{11}&j_{12}\end{matrix}\right\}.
\end{aligned}
\tag{108.12}
\end{equation}
The equivalence of (108.11) and (108.12) can be verified by substituting definition (108.2) into (108.12) and using the orthogonality properties of the $3j$ symbols.

The $9j$ symbol has a high degree of symmetry, following directly from definition (108.11) and the symmetry properties of the $3j$ symbols. It is readily verified that interchange of any two of its rows or any two of its columns multiplies the $9j$ symbol by $(-1)^{\sum j}$. Moreover, the $9j$ symbol is unchanged by transposition, that is, by interchange of rows and columns.

Scalars of still higher orders depend on still more parameters $j$. Clearly, their number must always be a multiple of three (the $3nj$ symbols). We shall not discuss the properties of these quantities here. We mention only that for every $n>3$ there is more than one distinct, mutually irreducible type of $3nj$ symbol. Thus there are two different types of $12j$ symbols\footnote{For a fuller account of the theory of $9j$ symbols and the properties of $3nj$ symbols, see the book by Edmonds cited on p.~272 and the books: A.~P.~Yutsis, I.~B.~Levinson, V.~V.~Vanagas, \emph{The Mathematical Apparatus of the Theory of Angular Momentum}, Vilnius, 1960; D.~A.~Varshalovich, A.~N.~Moskalev, V.~K.~Khersonskii, \emph{Quantum Theory of Angular Momentum}, Moscow: Nauka, 1975.}.
