\SourcePage{255}{258}
\section{Spinors}

For zero spin the wave function has only one component, $\psi(0)$. Application
of the spin operator gives zero: $\hat{\mathbf s}\psi=0$. Since $\mathbf s$
is connected with the operator of infinitesimal rotations, this means that
the wave function of a spin-zero particle is unchanged by rotations of the
coordinate system; it is therefore a scalar.

The wave function of a spin-$1/2$ particle has two components, $\psi(1/2)$
and $\psi(-1/2)$. For convenience in later generalizations, distinguish them
by the upper indices 1 and 2, respectively. The two-component quantity
\begin{equation}
 \psi=\binom{\psi^1}{\psi^2}=
 \binom{\psi(1/2)}{\psi(-1/2)}
 \tag{56.1}
\end{equation}
is called a \emph{spinor}.

Under an arbitrary rotation of the coordinate system, the spinor components
undergo the linear transformation
\begin{equation}
 \psi^{1'}=a\psi^1+b\psi^2,\qquad
 \psi^{2'}=c\psi^1+d\psi^2.
 \tag{56.2}
\end{equation}
This can be written
\begin{equation}
 \psi^{\lambda'}=(\hat U\psi)^\lambda,\qquad
 \hat U=\begin{pmatrix}a&b\\c&d\end{pmatrix},
 \tag{56.3}
\end{equation}
where $\hat U$ is the transformation matrix.\footnote{The notation
$\hat U\psi$ means multiplication of the rows of $\hat U$ by the column
$\psi$.} In general, the elements of this matrix are complex functions of the
angles through which the coordinate axes are rotated. Relations among them
follow directly from the physical requirements imposed on a spinor as a
particle's wave function.

Consider the bilinear form
\begin{equation}
 \psi^1\varphi^2-\psi^2\varphi^1,
 \tag{56.4}
\end{equation}
where $\psi$ and $\varphi$ are two spinors. A direct calculation gives
\[
 \psi^{1'}\varphi^{2'}-\psi^{2'}\varphi^{1'}
 =(ad-bc)(\psi^1\varphi^2-\psi^2\varphi^1).
\]
Thus (56.4) transforms into itself under a rotation of the coordinate system.
If there is only one function transforming into itself, it can be regarded as
corresponding to zero spin and must consequently be a scalar, remaining
unchanged under every rotation. We therefore obtain
\SourcePage{256}{259}
\begin{equation}
 ad-bc=1;
 \tag{56.5}
\end{equation}
the determinant of the transformation matrix is unity.\footnote{Such a
transformation of two quantities is called \emph{binary}.}

Further relations arise from the requirement that
\begin{equation}
 \psi^1\psi^{1*}+\psi^2\psi^{2*}
 \tag{56.6}
\end{equation}
be a scalar. This expression determines the probability of finding the
particle at a given point in space. A transformation that leaves invariant
the sum of the squared moduli of the transformed quantities is unitary, so
$\hat U^+=\hat U^{-1}$ (see \secref{12}). Subject to (56.5), the inverse matrix is
\[
 \hat U^{-1}=\begin{pmatrix}d&-b\\-c&a\end{pmatrix}.
\]
Equating it to the adjoint matrix
\[
 \hat U^+=\begin{pmatrix}a^*&c^*\\b^*&d^*\end{pmatrix},
\]
we find
\begin{equation}
 a=d^*,\qquad b=-c^*.
 \tag{56.7}
\end{equation}

By (56.5) and (56.7), the four complex quantities $a,b,c,d$ actually contain
only three independent real parameters, corresponding to the three angles
that specify a rotation of a three-dimensional coordinate system.

Comparison of the scalar expressions (56.4) and (56.6) shows that
$\psi^{1*},\psi^{2*}$ must transform like $\psi^2,-\psi^1$. It is readily
verified from (56.5) and (56.7) that they do.\footnote{This property is closely
connected with time-reversal symmetry. Time reversal corresponds (see
\secref{18}) to replacing the wave function by its complex conjugate, but it also
reverses the signs of angular-momentum projections. Functions complex
conjugate to the components $\psi^1=\psi(1/2)$ and
$\psi^2=\psi(-1/2)$ must therefore be equivalent in their properties to the
components corresponding to spin projections $-1/2$ and $1/2$, respectively.}

Spinor algebra can be cast in a form analogous to tensor algebra. Alongside
the contravariant components $\psi^1,\psi^2$, bearing upper indices, introduce
\SourcePage{257}{260}the \emph{covariant} components, bearing lower indices,
by the definition
\begin{equation}
 \psi_1=\psi^2,\qquad \psi_2=-\psi^1.
 \tag{56.8}
\end{equation}
The invariant combination (56.4) of two spinors then takes the form of the
scalar product
\begin{equation}
 \psi_\lambda\varphi^\lambda
 =\psi_1\varphi^1+\psi_2\varphi^2.
 \tag{56.9}
\end{equation}
Here and below, as in tensor algebra, summation is understood over any twice
repeated (dummy) index. The following rule must be kept in mind in spinor
algebra. We have
$\psi^\lambda\varphi_\lambda=\psi^1\varphi_1+\psi^2\varphi_2
=-\psi^2\varphi^1-\psi^1\varphi^2$, and therefore
\begin{equation}
 \psi^\lambda\varphi_\lambda=-\psi_\lambda\varphi^\lambda.
 \tag{56.10}
\end{equation}
It follows at once that the scalar product of any spinor with itself is zero:
\begin{equation}
 \psi_\lambda\psi^\lambda=0.
 \tag{56.11}
\end{equation}

As stated above, $\psi_1,\psi_2$ transform like
$\psi^{1*},\psi^{2*}$; hence
\begin{equation}
 \psi'_{\lambda}=(\hat U^*\psi)_\lambda.
 \tag{56.12}
\end{equation}
The product $\hat U^*\psi$ may also be written $\psi\hat U^*$ with the
transposed matrix $\hat U^*$. Since $\hat U$ is unitary,
$\hat U^*=\widetilde{\hat U}^{-1}$, so
$\widetilde\psi'=(\widetilde\psi\hat U^{-1})$, or\footnote{Notation of the
form $\psi\hat U$, with $\psi$ to the left of $\hat U$, means multiplication
of the row of components $(\psi_1,\psi_2)$ by the columns of $\hat U$.}
\begin{equation}
 \psi'_\lambda=\psi_\mu(U^{-1})^\mu{}_{\lambda}.
 \tag{56.13}
\end{equation}

Just as tensors are introduced from vectors in ordinary tensor algebra, one
can introduce \emph{spinors of higher rank}. A second-rank spinor, for
example, is a four-component quantity $\psi^{\lambda\mu}$ whose components
transform like the products $\psi^\lambda\varphi^\mu$ of components of two
first-rank spinors. Along with the contravariant components
$\psi^{\lambda\mu}$, one may consider the covariant components
$\psi_{\lambda\mu}$ and mixed components $\psi_\lambda{}^\mu$, transforming
respectively like $\psi_\lambda\varphi_\mu$ and
$\psi_\lambda\varphi^\mu$. Spinors of any rank are defined similarly.
\SourcePage{258}{261}
Passage from contravariant to covariant spinor components and back can be
written
\begin{equation}
 \psi_\lambda=g_{\lambda\mu}\psi^\mu,\qquad
 \psi^\lambda=g^{\mu\lambda}\psi_\mu,
 \tag{56.14}
\end{equation}
where
\begin{equation}
 g_{11}=g_{22}=0,\qquad g_{12}=-g_{21}=1
 \tag{56.15}
\end{equation}
is the \emph{metric spinor} in a two-dimensional vector space.\footnote{The
matrix (56.15) coincides with $i\hat\sigma_y$.} Similarly, for example,
\[
 \psi_\lambda{}^\mu=g_{\lambda\nu}\psi^{\nu\mu},\qquad
 \psi_{\lambda\mu}=g_{\lambda\nu}g_{\mu\rho}\psi^{\nu\rho},
\]
so that $\psi_1{}^2=-\psi^{11}$, $\psi_2{}^1=-\psi^{21}$,
$\psi_{11}=\psi^{22}$, and so forth.

The $g_{\lambda\mu}$ themselves form the antisymmetric unit spinor of second
rank. It is readily verified that their components remain unchanged under
coordinate transformations and that
\begin{equation}
 g_{\lambda\nu}g^{\mu\nu}=\delta_\lambda^\mu,
 \tag{56.16}
\end{equation}
where $\delta_1^1=\delta_2^2=1$ and
$\delta_1^2=\delta_2^1=0$.

As in ordinary tensor algebra, spinor algebra has two fundamental operations:
multiplication and simplification (or contraction) over a pair of indices.
The product of two spinors is a spinor of higher rank. Thus the second- and
third-rank spinors $\psi_{\lambda\mu}$ and
$\varphi^{\nu\rho\sigma}$ form the fifth-rank spinor
$\psi_{\lambda\mu}\varphi^{\nu\rho\sigma}$. Simplification over a pair of
indices, that is, summation of components over equal values of one covariant
and one contravariant index, lowers the rank by two. For example, simplifying
$\psi_{\lambda\mu}{}^{\nu\rho\sigma}$ over $\mu$ and $\nu$ gives the
third-rank spinor $\psi_{\lambda\mu}{}^{\mu\rho\sigma}$; simplifying
$\varphi_\lambda{}^\mu$ gives the scalar
$\varphi_\lambda{}^\lambda$. A rule analogous to (56.10) applies: interchanging
the upper and lower positions of the indices over which simplification is
performed changes the sign, so
$\varphi_\lambda{}^\lambda=-\varphi^\lambda{}_{\lambda}$. In particular, if
a spinor is symmetric in any two of its indices, simplification over those
indices gives zero. Thus a symmetric second-rank spinor
$\varphi_\lambda{}^\mu$ has $\varphi_\lambda{}^\lambda=0$.

A \emph{symmetric spinor} of rank $n$ is one symmetric in all its indices. A
symmetric spinor can be constructed from an asymmetric one by symmetrization,
that is, by summing the components obtained from all possible permutations of
the indices. It follows from the foregoing that the components of a
\SourcePage{259}{262}symmetric spinor cannot be contracted to form a spinor
of lower rank.

A spinor antisymmetric in all its indices, on the other hand, can only have
rank two. Indeed, because every index takes only two values, with three or more
indices at least two must have equal values, and the spinor components then
vanish identically. Every antisymmetric second-rank spinor reduces to a scalar
multiplied by the unit spinor $g_{\lambda\mu}$. We record the following
consequence:
\begin{equation}
 g_{\lambda\mu}\psi_\nu+g_{\mu\nu}\psi_\lambda
 +g_{\nu\lambda}\psi_\mu=0,
 \tag{56.17}
\end{equation}
where $\psi_\lambda$ is an arbitrary spinor. This rule follows simply because
the expression on the left is, as can readily be verified, an antisymmetric
third-rank spinor.

The spinor formed by multiplying $\psi_{\lambda\mu}$ by itself and
contracting one pair of indices is antisymmetric in the other pair; indeed,
\[
 \psi_{\lambda\nu}\psi_\mu{}^\nu
 =-\psi_\lambda{}^\nu\psi_{\mu\nu}.
\]
It must therefore reduce to $g_{\lambda\mu}$ multiplied by a scalar. Fixing
the latter so that contraction over the second pair gives the correct result,
we find
\begin{equation}
 \psi_{\lambda\nu}\psi_\mu{}^\nu
 =-\frac12\psi_{\rho\sigma}\psi^{\rho\sigma}g_{\lambda\mu}.
 \tag{56.18}
\end{equation}

The components of the spinor $\varphi^{\lambda\mu\ldots}$ complex conjugate
to $\psi_{\lambda\mu\ldots}$ transform as components of a contravariant
spinor, and conversely. The sum of the squared moduli of the components of any
spinor is therefore invariant.
