\SourcePage{272}{275}
\section{Partial polarization of particles}

For any given spinor $\psi^\lambda$, the wave function of a spin-$1/2$
particle, the direction of the $z$ axis can always be chosen so that one of
its components, say $\psi^2$, vanishes. This is already clear because a
direction in space is specified by two quantities (angles): the number of
parameters at our disposal is exactly equal to the number of quantities, the
real and imaginary parts of the complex $\psi^2$, that are to be made zero.

Physically, this means that if a spin-$1/2$ particle---for definiteness, an
electron---is in a state described by a certain spin wave function, there is
a direction in space along which its spin projection has the definite value
$\sigma=1/2$. The electron in such a state may be said to be \emph{fully
polarized}.

There are, however, electron states that may be called \emph{partially
polarized}. They are described not by wave functions but only by density
matrices; that is, they are mixed states with respect to spin (see \secref{14}).

The spin (or \emph{polarization}) density matrix of the electron is a
second-rank spinor $\rho^\lambda{}_mu$, normalized by
\begin{equation}
 \rho^\lambda{}_lambda=\rho^1{}_1+\rho^2{}_2=1
 \tag{59.1}
\end{equation}
and satisfying the ``Hermiticity'' condition
\begin{equation}
 (\rho^\lambda{}_mu)^*=\rho^\mu{}_lambda.
 \tag{59.2}
\end{equation}
For a pure, that is fully polarized, electron spin state, the spinor
$\rho^\lambda{}_mu$ reduces to a product of wave-function components:
\begin{equation}
 \rho^\lambda{}_mu=\psi^\lambda\psi_\mu^*.
 \tag{59.3}
\end{equation}
\SourcePage{273}{276}
The diagonal components of the density matrix give the probabilities of the
values $+1/2$ and $-1/2$ of the electron's spin projection on the $z$ axis.
The mean value of this projection is therefore
\[
 \bar s_z=\frac12(\rho^1{}_1-\rho^2{}_2),
\]
or, on using (59.1),
\begin{equation}
 \rho^1{}_1=\frac12+\bar s_z,\qquad
 \rho^2{}_2=\frac12-\bar s_z.
 \tag{59.4}
\end{equation}
In a pure state, the mean values of $s_\pm=s_x\pm is_y$ are calculated as
\[
 \bar s_\pm=\psi_\lambda^*\hat s_\pm\psi^\lambda.
\]
Since, by (55.6) and (55.7), the operators $\hat s_\pm$ have the matrices
\[
 \hat s_+=\begin{pmatrix}0&1\\0&0\end{pmatrix},\qquad
 \hat s_-=\begin{pmatrix}0&0\\1&0\end{pmatrix},
\]
we find
\[
 \bar s_+=\psi_1^*\psi^2,\qquad \bar s_-=\psi_2^*\psi^1.
\]
Accordingly, in a mixed state,
\begin{equation}
 \rho^1{}_2=\bar s_-,\qquad \rho^2{}_1=\bar s_+.
 \tag{59.5}
\end{equation}
With the Pauli matrices, (59.4) and (59.5) can be combined as
\begin{equation}
 \rho^\lambda{}_mu=\frac12
 (\delta^\lambda{}_mu+2\hat{\boldsymbol\sigma}^\lambda{}_mu\bar{\mathbf s}).
 \tag{59.6}
\end{equation}

Thus every component of the electron's polarization density matrix is
expressed in terms of the mean components of its spin vector. In other words,
the real vector $\bar{\mathbf s}$ completely determines the polarization
properties of a spin-$1/2$ particle. In the limiting case of full
polarization, one component of this vector, with a suitable choice of axes,
equals $1/2$, while the other two vanish. In the opposite, unpolarized case,
all three components vanish. For arbitrary partial polarization and an
arbitrary choice of coordinate system, in general $0\leqslant p\leqslant1$,
where
\[
 p=2(\bar s_x^2+\bar s_y^2+\bar s_z^2)^{1/2}
\]
may be called the electron's \emph{degree of polarization}.
\SourcePage{274}{277}
For a particle of arbitrary spin $s$, the density matrix is the rank-$4s$
spinor $\rho^{\lambda\mu\ldots}{}_{\rho\sigma\ldots}$, symmetric in its
first $2s$ and its last $2s$ indices, and satisfying
\begin{equation}
 \rho^{\lambda\mu\ldots}{}_{\lambda\mu\ldots}=1,
 \tag{59.7}
\end{equation}
\begin{equation}
 (\rho^{\lambda\mu\ldots}{}_{\rho\sigma\ldots})^*
 =\rho^{\rho\sigma\ldots}{}_{\lambda\mu\ldots}.
 \tag{59.8}
\end{equation}
To count the independent density-matrix components, note that among the
possible sets of values of $\lambda,\mu,\ldots$ (or of
$\rho,\sigma,\ldots$) only $2s+1$ are essentially different. Since the
components of $\rho^{\lambda\mu\ldots}{}_{\rho\sigma\ldots}$ are also related
by the single condition (59.7), their number is
$(2s+1)^2-1=4s(s+1)$. Although these components are complex, (59.8) ensures
that this does not increase the total number of independent quantities
characterizing a partially polarized state; that number is therefore
$4s(s+1)$.\footnote{Specifying these quantities is equivalent to specifying
the mean values of the components of $\mathbf s$ and all their powers and
products of degrees 2, 3, \ldots, $2s$ which cannot be reduced to still lower
degrees (see Problem 3 of \secref{55}).} For comparison, a fully polarized state is
described by only $4s$ quantities: the $2s+1$ complex components of
$\psi^{\lambda\mu\ldots}$, subject to one normalization condition and
containing one common phase irrelevant to the description of the state.

Like any spinor of rank $4s$, the spinor
$\rho^{\lambda\mu\ldots}{}_{\rho\sigma\ldots}$ is equivalent to a set of
irreducible tensors of ranks $4s$, $4s-2$, \ldots, 0. Here there is exactly
one tensor of each rank. Indeed, by the symmetry properties of
$\rho^{\lambda\mu\ldots}{}_{\rho\sigma\ldots}$, each contraction can be
performed in only one way: over any one of the indices
$\lambda,\mu,\ldots$ and any one of $\rho,\sigma,\ldots$. Moreover, the
scalar (rank-zero tensor) is absent altogether, since condition (59.7)
reduces it to unity.
