\SourcePage{246}{249}
\section{Spin}

In both classical and quantum mechanics, conservation of angular momentum
follows from the isotropy of space with respect to a closed system. This
already reveals the connection between angular momentum and rotational
symmetry. In quantum mechanics, however, that connection is especially deep:
it becomes essentially the principal content of the concept of angular
momentum. Moreover, the classical definition of a particle's angular momentum
as the product $\mathbf r\times\mathbf p$ loses its direct meaning here,
because the radius vector and momentum cannot be measured simultaneously.

We saw in \secref{28} that specifying $l$ and $m$ determines the angular dependence
of a particle's wave function, and hence all its symmetry properties under
rotations. In its most general form, the statement of these properties amounts
to giving the transformation law of wave functions when the coordinate system
is rotated.

The wave function $\psi_{LM}$ of a system of particles, with specified angular
momentum $L$ and projection $M$, remains unchanged\footnote{Apart from an
inessential phase factor.} only when the coordinate system is rotated about
the $z$ axis. Any rotation that changes the direction of the $z$ axis makes
the projection of the angular momentum on that axis cease to have a definite
value. Thus, in the new coordinate axes, the wave function generally becomes
a superposition (linear combination) of the $2L+1$ functions corresponding to
the possible values of $M$ for the given $L$. One may say that rotations of
the coordinate system transform these $2L+1$ functions $\psi_{LM}$ into one
another.\footnote{In mathematical terminology, these functions realize the
so-called \emph{irreducible representations} of the rotation group. The number
of functions transformed into one another is called the \emph{dimension of
the representation}; it is understood that no choice of other linear
combinations of the functions can reduce this number.} The transformation
law---the superposition coefficients as functions of the rotation angles of
the coordi\SourcePageJoin{247}{250}nate axes---is completely fixed by the
value of $L$. Angular momentum thereby acquires the meaning of a quantum
number that classifies the states of a system by their transformation
properties under rotations of the coordinate system. This aspect of angular
momentum in quantum mechanics is particularly important because it is not
directly tied to an explicit angular dependence of the wave functions: their
law of transformation into one another can be stated independently, without
reference to such a dependence.

Consider a composite particle (an atomic nucleus, for example) that is at rest
as a whole and is in a definite internal state. Besides a definite internal
energy, it has angular momentum of definite magnitude $L$, associated with the
motion of the particles within it; this angular momentum may have $2L+1$
different spatial orientations. In other words, when treating the motion of a
composite particle as a whole, we must assign to it, in addition to its
coordinates, one discrete variable: the projection of its internal angular
momentum on some selected direction in space.

With the meaning of angular momentum understood as above, however, its origin
becomes immaterial. We are led naturally to the idea of an ``intrinsic''
angular momentum that must be assigned to a particle irrespective of whether
it is ``composite'' or ``elementary.''

Thus quantum mechanics requires an elementary particle to possess an
``intrinsic'' angular momentum unrelated to its motion through space. This is
a specifically quantum property of elementary particles (it disappears in
the limit $\hbar\to0$) and therefore admits no classical interpretation in
principle.\footnote{In particular, it would be entirely meaningless to picture
the ``intrinsic'' angular momentum of an elementary particle as arising from
rotation ``about its own axis.''}

The intrinsic angular momentum of a particle is called its \emph{spin}, in
distinction to the angular momentum associated with its motion in space,
which is called \emph{orbital angular momentum}.\footnote{The physical idea
that the electron possesses intrinsic angular momentum was proposed by
Uhlenbeck and Goudsmit (G.~Uhlenbeck, S.~Goudsmit, 1925). Spin was introduced
into quantum mechanics by Pauli (W.~Pauli, 1927).} This may refer either to an
elementary particle or to a particle which, though composite, behaves as an
elementary one in the range of phenomena under consideration (an atomic
nucleus, for example). We denote a particle's spin, measured like orbital
angular momentum in units of $\hbar$, by $s$.
\SourcePage{248}{251}
For a particle with spin, a wave-function description of the state must
determine not only the probabilities of its various positions in space but
also those of the possible orientations of its spin. Thus the wave function
must depend not only on three continuous variables, the particle's
coordinates, but also on a discrete \emph{spin variable}. This variable gives
the spin projection on a selected direction in space (the $z$ axis) and takes
a finite set of discrete values, denoted below by $\sigma$.

Let $\psi(x,y,z;\sigma)$ be such a wave function. It is, in essence, a set of
several coordinate functions, one for each value of $\sigma$; these functions
will be called the \emph{spin components} of the wave function. The integral
\[
 \int |\psi(x,y,z;\sigma)|^2\,dV
\]
gives the probability that the particle has a specified value of $\sigma$.
The probability that it is in the volume element $dV$, with arbitrary
$\sigma$, is
\[
 dV\sum_\sigma |\psi(x,y,z;\sigma)|^2.
\]

When applied to the wave function, the quantum-mechanical spin operator acts
on the spin variable $\sigma$. In other words, it transforms the components
of the wave function into one another in a certain way. The form of this
operator will be established below. From quite general considerations,
however, it is already easy to see that $\hat s_x$, $\hat s_y$, and
$\hat s_z$ obey the same commutation relations as the orbital-angular-momentum
operators.

The angular-momentum operator essentially coincides with the operator of an
infinitesimal rotation. In deriving the orbital-angular-momentum operator in
\secref{26}, we considered the result of applying a rotation to a function of the
coordinates. Such a derivation is meaningless for spin angular momentum,
since its operator acts on the spin variable rather than on the coordinates.
To obtain the required commutation relations, we must therefore consider an
infinitesimal rotation in general, as a rotation of the coordinate system.
Carry out successive infinitesimal rotations about the $x$ and $y$ axes, and
then about the same axes in the reverse order. Direct calculation readily
shows that the difference between the two resulting operations is equivalent
to an infinitesimal \SourcePage{249}{252}rotation about the $z$ axis, through
an angle equal to the product of the angles about $x$ and $y$. We shall not
give this simple calculation here. It again yields the usual commutation
relations between the component operators of angular momentum, which must
therefore hold also for the spin operators:
\begin{equation}
 \{\hat s_x,\hat s_y\}=i\hat s_z
 \tag{54.1}
\end{equation}
with all their physical consequences.

The commutation relations (54.1) determine the possible magnitudes and
components of spin. The entire derivation in \secref{27}, formulas (27.7)--(27.9),
used only the commutation relations and applies here without change, with $s$
substituted for $L$. Formulas (27.7) show that the eigenvalues of the spin
projection form a sequence differing by unity. We can no longer assert,
however, that these values themselves must be integers, as they were for the
orbital projection $L_z$. The derivation at the beginning of \secref{27} does not
apply, since it uses (26.14), the expression for $\hat l_z$ specific to
orbital angular momentum.

The sequence of eigenvalues $s_z$ is bounded above and below by values equal
in magnitude and opposite in sign; denote them by $\pm s$. The difference
$2s$ between the largest and smallest values of $s_z$ must be an integer or
zero. Hence $s$ may have the values $0$, $1/2$, $1$, $3/2$,~\dots

Thus the eigenvalues of the square of the spin are
\begin{equation}
 s^2=s(s+1),
 \tag{54.2}
\end{equation}
where $s$ may be either an integer, including zero, or a half-integer. For a
given $s$, the spin component $s_z$ may take the $2s+1$ values
$s,s-1,\ldots,-s$. Correspondingly, the wave function of a particle of spin
$s$ has $2s+1$ components.\footnote{Since $s$ is a fixed number for each kind
of particle, the spin angular momentum $\hbar s$ tends to zero in the
classical limit $\hbar\to0$. This reasoning does not apply to orbital angular
momentum, because $l$ may take arbitrary values. The classical limit requires
$\hbar$ to tend to zero and $l$ simultaneously to infinity, while $\hbar l$
remains finite.}
\SourcePage{250}{253}
Experiment shows that most elementary particles---electrons, positrons,
protons, neutrons, $\mu$ mesons, and all hyperons ($\Lambda$, $\Sigma$,
$\Xi$)---have spin $1/2$. There are also elementary particles, the $\pi$
mesons and $K$ mesons, whose spin is zero.

The total angular momentum of a particle is the sum of its orbital angular
momentum $\mathbf l$ and spin $\mathbf s$. Their operators act on functions
of entirely different variables and therefore commute with each other.

The eigenvalues of the total angular momentum
\begin{equation}
 \mathbf j=\mathbf l+\mathbf s
 \tag{54.3}
\end{equation}
are determined by the same ``vector-model'' rule as the sum of the orbital
angular momenta of two different particles (\secref{31}). For specified $l$ and
$s$, the total angular momentum may have the values
$l+s,l+s-1,\ldots,|l-s|$. Thus an electron of spin $1/2$ with nonzero orbital
angular momentum $l$ may have total angular momentum $j=l\pm1/2$; for $l=0$,
of course, $j$ has only the single value $j=1/2$.

The operator $\mathbf J$ of the total angular momentum of a system of
particles is the sum of their individual $\mathbf j$ operators, so its values
are again determined by the vector-model rules. It may be written
\begin{equation}
 \mathbf J=\sum_a\mathbf j_a=\mathbf L+\mathbf S,
 \qquad \mathbf L=\sum_a\mathbf l_a,\quad
 \mathbf S=\sum_a\mathbf s_a,
 \tag{54.4}
\end{equation}
where $\mathbf S$ may be called the total spin and $\mathbf L$ the total
orbital angular momentum of the system.

If the total spin of a system is half-integral (or integral), the same is true
of its total angular momentum, since orbital angular momentum is always
integral. In particular, a system of an even number of identical particles
necessarily has integral total spin, and therefore integral total angular
momentum as well.

The total-angular-momentum operators $\mathbf j$ of a particle (or
$\mathbf J$ of a system) obey the same commutation rules as the orbital- or
spin-angular-momentum operators, since these are the general commutation rules
for any angular momentum. Formulas (27.13) for the matrix elements of angular
momentum, which follow from these rules, likewise apply to any angular
momentum, provided the matrix elements are defined with respect to the
eigenfunctions of that same angular momentum. With the appropriate changes of
notation, formulas (29.7)--(29.10) for matrix elements of arbitrary vector
quantities also remain valid.
\SourcePage{251}{254}
\ProblemHeading

A particle of spin $1/2$ is in a state with the definite value $s_z=1/2$.
Find the probabilities of the possible values of the spin projection on an
axis $z'$ inclined at angle $\theta$ to the $z$ axis.

\SolutionHeading

The mean spin vector $\bar{\mathbf s}$ is plainly directed along the $z$ axis
and has magnitude $1/2$. Its projection on the $z'$ axis gives the mean spin
in that direction as $\bar s_{z'}=(1/2)\cos\theta$. On the other hand,
$\bar s_{z'}=(1/2)(w_+-w_-)$, where $w_\pm$ are the probabilities of the
values $s_{z'}=\pm1/2$. Together with $w_++w_-=1$, this gives
\[
 w_+=\cos^2(\theta/2),\qquad w_-=\sin^2(\theta/2).
\]
