\SourcePage{251}{254}
\section{The spin operator}

In the remainder of this chapter we shall not be concerned with the coordinate
dependence of wave functions. When discussing, for example, the behavior of
$\psi(x,y,z;\sigma)$ under a rotation of the coordinate system, one may
suppose the particle to be at the origin. Its coordinates then remain
unchanged by the rotation, and the results characterize specifically the
dependence of $\psi$ on the spin variable $\sigma$.

The variable $\sigma$ differs from ordinary variables (coordinates) in being
discrete. The most general linear operator acting on functions of a discrete
variable $\sigma$ may be written
\begin{equation}
 (\hat f\psi)(\sigma)=\sum_{\sigma'}f_{\sigma\sigma'}\psi(\sigma'),
 \tag{55.1}
\end{equation}
where $f_{\sigma\sigma'}$ are constants. The parentheses around $\hat f\psi$
emphasize that the following spin argument belongs not to the original
function $\psi$, but to the function produced by $\hat f$. It is readily seen
that the $f_{\sigma\sigma'}$ coincide with the matrix elements of the
operator, defined by the usual rule (11.5).\footnote{Notice that the indices
of the matrix elements on the right-hand side of (55.1) occur in an order
which is, in a certain sense, the reverse of the usual order in (11.11).}
Integration over the coordinates in (11.5) is now replaced by summation over
the discrete variable, so the matrix element is defined by
\begin{equation}
 f_{\sigma_2\sigma_1}=\sum_\sigma
 \psi_{\sigma_2}^*(\sigma)[\hat f\psi_{\sigma_1}(\sigma)].
 \tag{55.2}
\end{equation}
\SourcePage{252}{255}
Here $\psi_{\sigma_1}(\sigma)$ and $\psi_{\sigma_2}(\sigma)$ are
eigenfunctions of $\hat s_z$ with eigenvalues $s_z=\sigma_1$ and
$s_z=\sigma_2$. Each function describes a state in which the particle has a
definite $s_z$; only one component of the wave function is nonzero.\footnote{More
precisely, one should write
$\psi_{\sigma_1}(\sigma)=\psi(x,y,z)\delta_{\sigma_1\sigma}$; coordinate
factors irrelevant here have been omitted in (55.3).

We emphasize once again the need to distinguish the specified eigenvalue
$s_z$ ($\sigma_1$ or $\sigma_2$) from the independent variable $\sigma$.
This is the reason for the difference between the notation in (11.11) and
(55.1).}
\begin{equation}
 \psi_{\sigma_1}(\sigma)=\delta_{\sigma\sigma_1},\qquad
 \psi_{\sigma_2}(\sigma)=\delta_{\sigma\sigma_2}.
 \tag{55.3}
\end{equation}
According to (55.1),
\[
 (\hat f\psi_{\sigma_1})(\sigma)=
 \sum_{\sigma'}f_{\sigma\sigma'}\psi_{\sigma_1}(\sigma')=
 \sum_{\sigma'}f_{\sigma\sigma'}\delta_{\sigma'\sigma_1}
 =f_{\sigma\sigma_1}.
\]
Substitution of this result and $\psi_{\sigma_2}(\sigma)$ in (55.2) makes
that equation an identity, proving the assertion.

Operators acting on functions of $\sigma$ can thus be represented by matrices
of order $2s+1$. This applies in particular to the spin operator itself,
whose action on a wave function is, by (55.1),
\begin{equation}
 (\hat{\mathbf s}\psi)(\sigma)=
 \sum_{\sigma'}\mathbf s_{\sigma\sigma'}\psi(\sigma').
 \tag{55.4}
\end{equation}
As stated above (at the end of \secref{54}), the matrices $\hat s_x$, $\hat s_y$,
$\hat s_z$ coincide with the matrices $L_x$, $L_y$, $L_z$ obtained in
\secref{27}, with $L,M$ replaced by $s,\sigma$:
\begin{equation}
 \begin{aligned}
 (s_x)_{\sigma,\sigma-1}&=(s_x)_{\sigma-1,\sigma}
   =\frac12\sqrt{(s+\sigma)(s-\sigma+1)},\\
 (s_y)_{\sigma,\sigma-1}&=-(s_y)_{\sigma-1,\sigma}
   =-\frac i2\sqrt{(s+\sigma)(s-\sigma+1)},\\
 (s_z)_{\sigma\sigma}&=\sigma.
 \end{aligned}
 \tag{55.5}
\end{equation}
This defines the spin operator.

In the most important case of spin $1/2$ ($s=1/2$, $\sigma=\pm1/2$), these
are matrices of order two. They are written
\begin{equation}
 \hat{\mathbf s}=\frac12\hat{\boldsymbol\sigma},
 \tag{55.6}
\end{equation}
\SourcePage{253}{256}
where\footnote{When the matrices are written as in (55.7), their rows and
columns are numbered by the values of $\sigma$, the row number corresponding
to the first matrix-element index and the column number to the second. Here
these numbers are $+1/2,-1/2$. According to (55.4), the operator acts by
multiplying row $\sigma$ of the matrix by the wave-function components
arranged as the column
$\displaystyle\psi=\binom{\psi(1/2)}{\psi(-1/2)}$.}
\begin{equation}
 \hat\sigma_x=\begin{pmatrix}0&1\\1&0\end{pmatrix},\qquad
 \hat\sigma_y=\begin{pmatrix}0&-i\\i&0\end{pmatrix},\qquad
 \hat\sigma_z=\begin{pmatrix}1&0\\0&-1\end{pmatrix}.
 \tag{55.7}
\end{equation}
The matrices (55.7) are called the \emph{Pauli matrices}. The matrix
$\hat s_z=\frac12\hat\sigma_z$ is diagonal, as it must be when defined with
respect to the eigenfunctions of $s_z$ itself.\footnote{Using the same letter
for the spin projection and the Pauli matrices causes no ambiguity: the Pauli
matrices carry a hat.}

We note several special properties of the Pauli matrices. Direct
multiplication of (55.7) gives
\begin{equation}
 \hat\sigma_y\hat\sigma_z=i\hat\sigma_x,\qquad
 \hat\sigma_z\hat\sigma_x=i\hat\sigma_y,\qquad
 \hat\sigma_x\hat\sigma_y=i\hat\sigma_z.
 \tag{55.8}
\end{equation}
Combining these equalities with the general commutation rules (54.1), we find
\begin{equation}
 \hat\sigma_i\hat\sigma_k+\hat\sigma_k\hat\sigma_i=0\quad(i\ne k),
 \tag{55.9}
\end{equation}
so the Pauli matrices anticommute. These relations readily give the useful
formulas
\begin{equation}
 \hat{\boldsymbol\sigma}^{,2}=3,\qquad
 (\hat{\boldsymbol\sigma}\mathbf a)
 (\hat{\boldsymbol\sigma}\mathbf b)=
 \mathbf a\mathbf b+i\hat{\boldsymbol\sigma}(\mathbf a\times\mathbf b),
 \tag{55.10}
\end{equation}
where $\mathbf a$ and $\mathbf b$ are arbitrary vectors.\footnote{Terms on the
right-hand sides of (55.8)--(55.10) that do not depend on
$\hat{\boldsymbol\sigma}$ are, of course, to be understood as constants
multiplied by the unit matrix of order two.} It follows from these relations
that any scalar polynomial expression in the matrices $\hat\sigma_i$ reduces
to terms independent of $\hat{\boldsymbol\sigma}$ and terms of first degree
in $\hat{\boldsymbol\sigma}$. Hence any scalar function whatsoever of the
operator $\hat{\boldsymbol\sigma}$ reduces to a linear function (see Problem
1). Finally, the traces (sums of diagonal elements) of the Pauli matrices and
their products are
\begin{equation}
 \Tr\hat\sigma_i=0,\qquad \Tr(\hat\sigma_i\hat\sigma_k)=2\delta_{ik}.
 \tag{55.11}
\end{equation}
\SourcePage{254}{257}
The following sections of this chapter give a detailed study of the spin
properties of wave functions, including their behavior under arbitrary
rotations of the coordinate system. Here we immediately note one important
property: their behavior under rotations about the $z$ axis.

Perform an infinitesimal rotation through $\delta\varphi$ about the $z$ axis.
The operator of this rotation is expressed through the angular-momentum
operator, here the spin, as $1+i\delta\varphi\hat s_z$. The functions
$\psi(\sigma)$ therefore become $\psi(\sigma)+\delta\psi(\sigma)$, where
\[
 \delta\psi(\sigma)=i\delta\varphi\hat s_z\psi(\sigma)
 =i\sigma\psi(\sigma)\delta\varphi.
\]
Writing this as $d\psi/d\varphi=i\sigma\psi(\sigma)$ and integrating, we find
that a finite rotation through $\varphi$ transforms $\psi(\sigma)$ into
\begin{equation}
 \psi'(\sigma)=e^{i\sigma\varphi}\psi(\sigma).
 \tag{55.12}
\end{equation}
In particular, a rotation through $2\pi$ multiplies the functions by
$e^{2\pi i\sigma}$, the same factor for every $\sigma$, equal to
$(-1)^{2s}$ (the integer $2\sigma$ always has the same parity as $2s$).
Thus a complete rotation of the coordinate system about the $z$ axis restores
the original wave functions of a particle with integral spin, whereas those
of a particle with half-integral spin change sign.

\ProblemHeading

1. Reduce an arbitrary function of the scalar
$a+\mathbf b\hat{\boldsymbol\sigma}$, which is linear in the Pauli matrices,
to a linear function.

\SolutionHeading

To determine the coefficients in
\[
 f(a+\mathbf b\hat{\boldsymbol\sigma})=A+B\mathbf b\hat{\boldsymbol\sigma},
\]
choose the $z$ axis along $\mathbf b$. The eigenvalues of
$a+\mathbf b\hat{\boldsymbol\sigma}$ are then $a\pm b$, and the corresponding
eigenvalues of $f(a+\mathbf b\hat{\boldsymbol\sigma})$ are $f(a\pm b)$.
Hence
\[
 A=\frac12[f(a+b)+f(a-b)],\qquad
 B=\frac1{2b}[f(a+b)-f(a-b)].
\]

2. Determine the values of the scalar product $\mathbf s_1\mathbf s_2$ of
the spins ($1/2$) of two particles in states where their resultant spin
$\mathbf S=\mathbf s_1+\mathbf s_2$ has a definite value, 0 or 1.

\SolutionHeading

The general formula (31.3), valid for the addition of any two angular
momenta, gives
\[
 \mathbf s_1\mathbf s_2=1/4\quad\text{for }S=1,\qquad
 \mathbf s_1\mathbf s_2=-3/4\quad\text{for }S=0.
\]

3. Which powers of the operator $s_z$ for arbitrary spin $s$ are independent?

\SolutionHeading

The operator
\[
 (\hat s_z-s)(\hat s_z-s+1)\cdots(\hat s_z+s),
\]
formed from the differences between $\hat s_z$ and all its possible
eigenvalues, gives zero when applied to any wave function and is therefore
itself \SourcePageJoin{255}{258}zero. It follows that $(\hat s_z)^{2s+1}$
can be expressed in terms of lower powers of $\hat s_z$, so only its powers
from 1 through $2s$ are independent.
