\SourcePage{259}{262}
\section{Wave Functions of Particles with Arbitrary Spin}

Having developed the formal algebra of spinors of arbitrary rank, we can turn
to our immediate task: studying the properties of wave functions for
particles of arbitrary spin.

A convenient approach is to consider a collection of $n$ particles of spin
$1/2$. The greatest possible value of the $z$ component of the system's total
spin is $n/2$; it is attained when every particle has $s_z=1/2$, so that all
spins point in the same direction, along the $z$ axis. In this case the total
spin $S$ of the system must itself equal $n/2$.

\SourcePage{260}{263}
All components of the system's wave function
$\psi(\sigma_1,\sigma_2,\ldots,\sigma_n)$ then vanish except the single
component $\psi(1/2,1/2,\ldots,1/2)$. Write the wave function as a product of
$n$ spinors $\psi^\lambda\varphi^\mu\ldots$, one for each particle. In each
spinor only the component with $\lambda,\mu,\ldots=1$ is nonzero, so only the
product $\psi^1\varphi^1\ldots$ survives. The collection of all these products
forms an $n$th-rank spinor symmetric in all its indices. After a transformation
of the coordinate system, so that the spins no longer point along $z$, this
becomes a general $n$th-rank spinor, but it remains symmetric.

The spin properties of wave functions are, in essence, their behavior under
rotations of the coordinate system. They are identical for a particle of spin
$s$ and for a system of $n=2s$ spin-$1/2$ particles oriented so that the
system's total spin is $s$. It follows that the wave function of a particle of
spin $s$ is a symmetric spinor of rank $n=2s$.

As required, a symmetric spinor of rank $2s$ has $2s+1$ independent
components. Indeed, the distinct components are only those whose indices
contain $2s$ ones and no twos, $2s-1$ ones and one two, and so on, down to no
ones and $2s$ twos.

Mathematically, symmetric spinors classify the possible transformation laws
of quantities under rotations of the coordinate system. If $2s+1$ distinct
quantities transform linearly among themselves, and no choice of linear
combinations can reduce their number, their transformation law is equivalent
to that of the components of a symmetric rank-$2s$ spinor. Any collection of
any number of functions which transform linearly among themselves under
coordinate rotations can, by a suitable linear transformation, be reduced to
one or more symmetric spinors.\footnote{In other words, symmetric spinors
realize the irreducible representations of the rotation group (see \S~98).}

Thus an arbitrary $n$th-rank spinor $\psi^{\lambda\mu\nu\ldots}$ can be
reduced to symmetric spinors of ranks $n,n-2,n-4,\ldots$. The reduction can
be carried out as follows. Symmetrizing $\psi^{\lambda\mu\nu\ldots}$ over all
its indices gives a symmetric spinor of the same rank $n$. Next, contracting

\SourcePage{261}{264}
the original spinor $\psi^{\lambda\mu\nu\ldots}$ over different pairs of
indices gives spinors of rank $n-2$, such as
$\psi^\lambda{}_{\lambda}{}^{\nu\ldots}$; symmetrizing these in turn produces
symmetric spinors of rank $n-2$. Symmetrizing the spinors obtained after
contracting $\psi^{\lambda\mu\ldots}{}_{\lambda\mu}$ over two pairs of
indices gives symmetric spinors of rank $n-4$, and so forth.

It remains to connect the components of a symmetric rank-$2s$ spinor with the
$2s+1$ functions $\psi(\sigma)$, where
$\sigma=s,s-1,\ldots,-s$. The component
\[
 \psi^{\overbrace{11\ldots1}^{s+\sigma}
              \overbrace{22\ldots2}^{s-\sigma}},
\]
in which 1 occurs $s+\sigma$ times and 2 occurs $s-\sigma$ times,
corresponds to spin projection $\sigma$ on the $z$ axis. To see this, replace
the one spin-$s$ particle once more by a system of $n=2s$ spin-$1/2$
particles. The component displayed above corresponds to the product
\[
 \underbrace{\psi^1\varphi^1\ldots}_{s+\sigma}
 \underbrace{\chi^2\rho^2\ldots}_{s-\sigma}.
\]
This product describes a state in which $s+\sigma$ particles have spin
projection $+1/2$, and $s-\sigma$ have projection $-1/2$; the total
projection is consequently
$\frac12(s+\sigma)-\frac12(s-\sigma)=\sigma$. Finally, choose the factor of
proportionality between this spinor component and $\psi(\sigma)$ so that
\begin{equation}
 \sum_{\sigma=-s}^{+s}|\psi(\sigma)|^2
 =\sum_{\lambda,\mu,\ldots=1}^{2}|\psi^{\lambda\mu\ldots}|^2
 \tag{57.1}
\end{equation}
holds. This sum is a scalar, as it must be, since it determines the probability
of finding the particle at the specified point in space. In the sum on the
right, components having $s+\sigma$ indices equal to 1 occur
\[
 \frac{(2s)!}{(s+\sigma)!(s-\sigma)!}
\]
times. The correspondence between the functions $\psi(\sigma)$ and the
spinor components is therefore
\begin{equation}
 \psi(\sigma)=\left[\frac{(2s)!}{(s+\sigma)!(s-\sigma)!}\right]^{1/2}
 \psi^{\overbrace{11\ldots1}^{s+\sigma}
       \overbrace{22\ldots2}^{s-\sigma}}.
 \tag{57.2}
\end{equation}

\SourcePage{262}{265}
Relation (57.2) ensures not only (57.1), but, as is readily verified, also the
more general relation
\begin{equation}
 \psi^{\lambda\mu\ldots}\varphi_{\lambda\mu\ldots}
 =\sum_\sigma(-1)^{s-\sigma}\psi(\sigma)\varphi(-\sigma),
 \tag{57.3}
\end{equation}
where $\psi^{\lambda\mu\ldots}$ and $\varphi_{\lambda\mu\ldots}$ are two
different spinors of equal rank, and $\psi(\sigma)$, $\varphi(\sigma)$ are
the functions associated with them by (57.2). The factor
$(-1)^{s-\sigma}$ occurs because raising all the indices of a spinor component
changes its sign once for every index equal to 2.

Equations (55.5) define the action of the spin operator on the wave functions
$\psi(\sigma)$. It is straightforward to determine how these operators act
when the wave function is written as a rank-$2s$ spinor. For spin $1/2$, the
functions $\psi(+1/2)$ and $\psi(-1/2)$ coincide with the spinor components
$\psi^1$ and $\psi^2$. According to (55.6) and (55.7), the action of the spin
operators on them is
\begin{equation}
 \begin{aligned}
 (\hat s_x\psi)^1&=\tfrac12\psi^2,&
 (\hat s_y\psi)^1&=-\tfrac i2\psi^2,&
 (\hat s_z\psi)^1&=\tfrac12\psi^1,\\
 (\hat s_x\psi)^2&=\tfrac12\psi^1,&
 (\hat s_y\psi)^2&=\tfrac i2\psi^1,&
 (\hat s_z\psi)^2&=-\tfrac12\psi^2.
 \end{aligned}
 \tag{57.4}
\end{equation}
To pass to arbitrary spin, consider again a system of $2s$ spin-$1/2$
particles and write its wave function as a product of $2s$ spinors. The spin
operator of the system is the sum of the spin operators of the particles;
each acts only on its corresponding spinor, with its action given by (57.4).
Returning then to arbitrary symmetric spinors, or equivalently to wave
functions of a spin-$s$ particle, gives
\begin{equation}
 \begin{aligned}
 (\hat s_x\psi)^{\overbrace{11\ldots}^{s+\sigma}
                         \overbrace{22\ldots}^{s-\sigma}}
 &=\frac{s+\sigma}{2}\psi^{\overbrace{11\ldots}^{s+\sigma-1}
                                  \overbrace{22\ldots}^{s-\sigma+1}}
 +\frac{s-\sigma}{2}\psi^{\overbrace{11\ldots}^{s+\sigma+1}
                                  \overbrace{22\ldots}^{s-\sigma-1}},\\
 (\hat s_y\psi)^{\overbrace{11\ldots}^{s+\sigma}
                         \overbrace{22\ldots}^{s-\sigma}}
 &=-i\frac{s+\sigma}{2}\psi^{\overbrace{11\ldots}^{s+\sigma-1}
                                   \overbrace{22\ldots}^{s-\sigma+1}}
 +i\frac{s-\sigma}{2}\psi^{\overbrace{11\ldots}^{s+\sigma+1}
                                  \overbrace{22\ldots}^{s-\sigma-1}},\\
 (\hat s_z\psi)^{\overbrace{11\ldots}^{s+\sigma}
                         \overbrace{22\ldots}^{s-\sigma}}
 &=\sigma\psi^{\overbrace{11\ldots}^{s+\sigma}
                       \overbrace{22\ldots}^{s-\sigma}}.
 \end{aligned}
 \tag{57.5}
\end{equation}
So far, spinors have been discussed as wave functions of the intrinsic angular
momentum of elementary particles. Formally, however, there is no distinction
between the spin of one particle and the total angular momentum of any system
viewed

\SourcePage{263}{266}
as a whole, with its internal structure disregarded. The transformation
properties of spinors therefore apply equally to the behavior under spatial
rotations of the wave functions $\psi_{jm}$ of any particle, or system of
particles, having total angular momentum $j$, regardless of whether its origin
is orbital or spin. There must accordingly be a definite correspondence
between the transformation laws of the eigenfunctions $\psi_{jm}$ under
rotations of the coordinate system and those of the components of a symmetric
spinor of rank $2j$.

In establishing this correspondence, however, two aspects of a wave
function's dependence on the angular-momentum projection $m$, for fixed $j$,
must be distinguished clearly. A wave function may be viewed as a probability
amplitude for the various values of $m$, or as an eigenfunction belonging to
a specified value of $m$.

These two aspects appeared already at the beginning of \S~55, where the
eigenfunction $\delta_{\sigma\sigma_0}$ of $s_z$, belonging to
$s_z=\sigma_0$, was considered. Their mathematical distinction is especially
transparent for a particle with $s=1/2$. As a function of the variable
$\sigma$, the spin function is a contravariant spinor of first rank; in spinor
notation it must therefore be written $\delta^\lambda{}_{\sigma_0}$. As a
function of $\sigma_0$, it is consequently a covariant spinor.

This fact is plainly general: the eigenfunctions $\psi_{jm}$ can be associated
with the components of a covariant symmetric rank-$2j$ spinor by formulas
analogous to (57.2).\footnote{The same result may be approached in another
way. Expand the wave function $\psi$ of a particle in a state of angular
momentum $j$ in the eigenfunctions $\psi_{jm}$:
$\psi=\sum_m a_m\psi_{jm}$. The coefficients $a_m$ are probability amplitudes
for the different values of $m$. In this sense they correspond to the
``components'' $\psi(m)$ of the spin wave function, which fixes their
transformation law. On the other hand, the value of $\psi$ at a given point
in space cannot depend on the choice of coordinate system; hence the sum
$\sum a_m\psi_{jm}$ must be a scalar. Comparison with the scalar (57.3) shows
that $a_m$ must transform as $(-1)^{j-m}\psi_{j,-m}$.}
\begin{equation}
 \psi_{jm}=\left[\frac{(2j)!}{(j+m)!(j-m)!}\right]^{1/2}
 \psi_{\overbrace{11\ldots}^{j+m}\overbrace{22\ldots}^{j-m}}.
 \tag{57.6}
\end{equation}
For integral angular momentum $j$, the eigenfunctions are the spherical
functions $Y_{jm}$. The case $j=1$ is especially important.

\SourcePage{264}{267}
The three spherical functions $Y_{1m}$ are
\[
 Y_{10}=i\sqrt{\frac3{4\pi}}\cos\theta=i\sqrt{\frac3{4\pi}}n_z,
 \qquad
 Y_{1,\pm1}=\mp i\sqrt{\frac3{8\pi}}\sin\theta e^{\pm i\varphi}
 =\mp i\sqrt{\frac3{8\pi}}(n_x\pm in_y),
\]
where $\mathbf n$ is a unit vector in the direction of the radius vector. It
is evident that, by their transformation properties, these three functions
are equivalent to the components of a vector $\mathbf a$. We write the
correspondence as
\begin{equation}
 \psi_{10}=ia_z,\qquad
 \psi_{11}=-\frac i{\sqrt2}(a_x+ia_y),\qquad
 \psi_{1,-1}=\frac i{\sqrt2}(a_x-ia_y).
 \tag{57.7}
\end{equation}
Comparison with (57.6) shows that the components of a symmetric second-rank
spinor may be associated with those of a vector by
\begin{equation}
 \psi_{11}=-\frac i{\sqrt2}(a_x+ia_y),\qquad
 \psi_{12}=\psi_{21}=\frac i{\sqrt2}a_z,\qquad
 \psi_{22}=\frac i{\sqrt2}(a_x-ia_y),
 \tag{57.8}
\end{equation}
or
\begin{equation}
 \psi^{11}=\frac i{\sqrt2}(a_x-ia_y),\qquad
 \psi^{12}=\psi^{21}=-\frac i{\sqrt2}a_z,\qquad
 \psi^{22}=-\frac i{\sqrt2}(a_x+ia_y).
 \tag{57.9}
\end{equation}
Conversely,
\begin{equation}
 a_z=i\sqrt2\psi^{12},\qquad
 a_x=\frac i{\sqrt2}(\psi^{22}-\psi^{11}),\qquad
 a_y=\frac1{\sqrt2}(\psi^{11}+\psi^{22}).
 \tag{57.10}
\end{equation}
With this definition one readily verifies that
\begin{equation}
 \psi^{\lambda\mu}\varphi_{\lambda\mu}=\mathbf a\mathbf b,
 \tag{57.11}
\end{equation}
where $\mathbf a$ and $\mathbf b$ are the vectors corresponding to the
symmetric spinors $\psi^{\lambda\mu}$ and $\varphi_{\lambda\mu}$. The
following spinor and vector are likewise found to correspond:\footnote{The
mixed components of a symmetric spinor may be written
$\psi^\lambda{}_{\mu}$, without distinguishing between
$\psi^{\lambda\mu}$ and $\psi^{\mu\lambda}$.}
\begin{equation}
 \psi^\lambda{}_{\rho}\varphi^{\rho\mu}
 -\varphi^\lambda{}_{\rho}\psi^{\rho\mu}
 \longleftrightarrow i\sqrt2(\mathbf a\times\mathbf b).
 \tag{57.12}
\end{equation}
Equations (57.10) can be written compactly with the Pauli matrices:
\begin{equation}
 a_i=\frac i{\sqrt2}(\hat\sigma_i)^\lambda{}_{\mu}
 \psi^\mu{}_{\lambda}.
 \tag{57.13}
\end{equation}

\SourcePage{265}{268}
The matrix indices on $\sigma$ are placed above and below to match the
positions of the spinor indices on $\psi^\mu{}_{\lambda}$. The origin of this
formula is easily understood from the special case in which the second-rank
spinor $\psi^\mu{}_{\lambda}$ reduces to the product of a first-rank spinor
$\psi^\mu$ and its complex conjugate $\psi_{\lambda}^*$. Then
\[
 \frac12\psi_\lambda^*\hat{\boldsymbol\sigma}^\lambda{}_{\mu}\psi^\mu
\]
is the mean spin (for a particle whose wave function is $\psi^\mu$), so its
vector character is evident.

The correspondence (57.8) or (57.9) is a special case of a general rule: any
symmetric spinor of even rank $2j$, where $j$ is an integer, can be associated
with a symmetric tensor of half that rank, $j$, which vanishes upon
contraction over any pair of indices. We shall call such a tensor
irreducible. This follows already because spinor and tensor have the same
number of independent components, namely $2j+1$, as a direct count readily
shows.\footnote{In other words, the $2j+1$ components of an irreducible
rank-$j$ tensor ($j$ integral), the set of $2j+1$ spherical functions
$Y_{jm}$, and the $2j+1$ components of a symmetric rank-$2j$ spinor all
realize the same irreducible representation of the rotation group.} The
correspondence between the spinor and tensor components can be found from
(57.8)--(57.10) by treating the spinor of the given rank as a product of
several second-rank spinors, and the tensor as a product of vectors.

\ProblemHeading

1. Rewrite definition (57.4) of the spin-$1/2$ operator using the spinor
components of the vector $\mathbf s$.

\SolutionHeading

Using (57.9), which relates the vector $\mathbf s$ to the spinor
$s^{\lambda\mu}$, definition (57.4) becomes
\[
 s^{\lambda\mu}\psi^\nu=\frac i{2\sqrt2}
 (\psi^\lambda g^{\mu\nu}+\psi^\mu g^{\lambda\nu}).
\]

2. Write formulas defining the action of the spin operator on the vector wave
function of a spin-1 particle.

\SolutionHeading

Equations (57.9) relate the components of the vector function
$\boldsymbol\psi$ to those of the spinor $\psi^{\lambda\mu}$, while the last
of equations (57.5) gives
\[
 \hat s_z\psi_+=-\psi_+,
 \qquad \hat s_z\psi_-=\psi_-,
 \qquad \hat s_z\psi_z=0,
\]
where $\psi_\pm=\psi_x\pm i\psi_y$, or
\[
 \hat s_z\psi_x=-i\psi_y,\qquad
 \hat s_z\psi_y=i\psi_x,\qquad
 \hat s_z\psi_z=0.
\]
The remaining formulas follow by cyclic permutation of $x,y,z$. Together
they can be written
\[
 \hat s_i\psi_k=-ie_{ik\ell}\psi_\ell.
\]

The complex vector $\boldsymbol\psi$ can be represented as
$\boldsymbol\psi=e^{i\alpha}(\mathbf u+i\mathbf v)$, where $\mathbf u$ and
$\mathbf v$ are real vectors. By a suitable choice of the common phase
$\alpha$, they can be made mutually perpendicular. The two vectors
$\mathbf u$ and $\mathbf v$ determine a plane with the following property:
the spin projection on the direction perpendicular to it can take only the
values $\pm1$.
