\SourcePage{580}{583}
\section{The shell model}

Many nuclear properties admit a good description by the \emph{shell model},
whose basic ideas resemble those used for the electron shell of an atom. In
this description every nucleon is regarded as moving in a self-consistent
field produced by all the other nucleons (because nuclear forces have a short
range, this field falls off rapidly outside the volume bounded by the nuclear
``surface''). The state of the nucleus as a whole is accordingly specified by
listing the states of its individual nucleons.

\SourcePage{581}{584}
The self-consistent field is spherically symmetric, with its centre naturally
at the centre of mass of the nucleus. This, however, raises a difficulty. In
the self-consistent-field method the wavefunction of the system is constructed
as a product (or an appropriately symmetrized sum of products) of the
one-particle wavefunctions. Such a function does not ensure that the centre of
mass is at rest: although it gives zero for the mean velocity of the centre of
mass, it also assigns finite probabilities to nonzero velocities\footnote{For
electrons in an atom this difficulty did not arise, since the centre of mass
was automatically fixed by its coincidence with the position of the fixed
heavy nucleus.}.

The difficulty can be avoided by first eliminating centre-of-mass motion when
any physical quantity is calculated with the self-consistent-field functions
$\psi(\mathbf r_1,\ldots,\mathbf r_A)$. Let $f(\mathbf r_i,\mathbf p_i)$ be
some physical quantity, a function of the coordinates and momenta of the
nucleons. In calculating its matrix elements with the functions $\psi$, one
must leave $\psi(\mathbf r_i)$ unchanged and replace the arguments of $f$ by
\begin{equation}
 \mathbf r_i\to\mathbf r_i-\mathbf R,\qquad
 \mathbf p_i\to\mathbf p_i-\frac{\mathbf P}{A},
 \tag{118.1}
\end{equation}
where $\mathbf R$ is the radius vector of the nuclear centre of mass, $A$ the
number of particles in the nucleus, and $\mathbf P$ its momentum as a whole.
The second replacement in (118.1) corresponds to subtracting the
centre-of-mass velocity $\mathbf V$ from the nucleon velocities,
$\mathbf v_i\to\mathbf v_i-\mathbf V$, with
$\mathbf P=Am_p\mathbf V$ (S.~Gartenhaus, C.~Schwartz, 1957).

For example, the nuclear dipole-moment operator is
$\hat{\mathbf d}=e\sum_p\mathbf r_p$, where the sum extends over all protons
in the nucleus. For matrix elements in the self-consistent-field method it
must instead be replaced by $e\sum_p(\mathbf r_p-\mathbf R)$. The coordinate
of the nuclear centre is
\[
 \mathbf R=\frac1A\left(\sum_p\mathbf r_p+\sum_n\mathbf r_n\right),
\]
where the sums run over all protons and neutrons. Since the number of protons
is $Z$, the final replacement of the dipole-moment operator is
\begin{equation}
 e\sum_p\mathbf r_p\to
 e\left(1-\frac ZA\right)\sum_p\mathbf r_p
 -e\frac ZA\sum_n\mathbf r_n.
 \tag{118.2}
\end{equation}

\SourcePage{582}{585}
Thus the protons enter with the ``effective charge'' $e(1-Z/A)$ and the
neutrons with the ``charge'' $-eZ/A$. As (118.2) shows, the correction terms
arising in a dipole-moment calculation have relative order unity. The
corrections for magnetic and higher electric multipole moments, on the other
hand, are readily seen to have relative order $\sim1/A$.

In the nonrelativistic approximation the interaction of a nucleon with the
self-consistent field is independent of its spin. Such a dependence could
appear only through a term proportional to $\hat{\mathbf s}\mathbf n$, where
$\mathbf n$ is the unit vector along the nucleon's radius vector $\mathbf r$;
this product is a pseudoscalar rather than a true scalar.

A spin dependence of the nucleon energy does arise when relativistic terms
depending on the particle velocity are included. The largest of them is
linear in the velocity. The three vectors $\mathbf s$, $\mathbf n$, and
$\mathbf v$ form the true scalar $(\mathbf n\times\mathbf v)\mathbf s$.
Consequently the operator of the nucleon's \emph{spin--orbit coupling} in the
nucleus has the form
\begin{equation}
 \hat V_{sl}=-\varphi(r)(\mathbf n\times\hat{\mathbf v})\hat{\mathbf s},
 \tag{118.3}
\end{equation}
where $\varphi(r)$ is a function of $r$ (see also the note on p.~579). Since
$m_p(\mathbf r\times\mathbf v)$ is the orbital angular momentum
$\hbar\hat{\mathbf l}$ of the particle, (118.3) may also be written
\begin{equation}
 \hat V_{sl}=-f(r)\hat{\mathbf l}\hat{\mathbf s},
 \tag{118.4}
\end{equation}
where $f=\hbar\varphi/(rm_p)$. Notice that this interaction is first order in
$v/c$, whereas the spin--orbit coupling of an atomic electron is a
second-order effect (\S~72). The distinction is connected with the fact that
nuclear forces already depend on spin in the nonrelativistic approximation,
while the nonrelativistic interaction of electrons (the Coulomb forces) does
not.

The spin--orbit interaction energy is concentrated mainly near the nuclear
surface; that is, $f(r)$ decreases towards the interior. Indeed, an
interaction of this kind could not exist at all in unbounded nuclear matter:
by homogeneity such a system has no distinguished direction along which
$\mathbf n$ could point.

The interaction (118.4) splits a nucleon level of orbital angular momentum
$l$ into two levels with $j=$

\SourcePage{583}{586}
$l\pm1/2$. Since
\begin{equation}
 \begin{aligned}
  \mathbf l\mathbf s&=l/2 &&\text{for }j=l+1/2,\\
  \mathbf l\mathbf s&=-(l+1)/2 &&\text{for }j=l-1/2,
 \end{aligned}
 \tag{118.5}
\end{equation}
by (31.3), the splitting is
\begin{equation}
 \Delta E=E_{l-1/2}-E_{l+1/2}=\overline{f(r)}(l+1/2).
 \tag{118.6}
\end{equation}
Experiment shows that the $j=l+1/2$ level (parallel $\mathbf l$ and
$\mathbf s$) is deeper than the $j=l-1/2$ level; hence $f(r)>0$.

The spin--orbit coupling of a nucleon in a nucleus is relatively weak compared
with its interaction with the self-consistent field. In general, however, it
is large compared with the energy of the direct interaction between two
nucleons in the nucleus, because the latter falls more rapidly as the atomic
weight increases.

This relation between the interaction energies makes the $jj$-coupling scheme
appropriate for classifying nuclear levels. The spin and orbital angular
momentum of each nucleon combine into a total angular momentum
$\mathbf j=\mathbf l+\mathbf s$, which remains definite because the direct
interaction between the particles does not destroy the coupling of
$\mathbf l$ and $\mathbf s$ (M.~G\"oppert-Mayer, 1949; O.~Haxel,
J.~H.~Jensen, H.~E.~Suess, 1949)\footnote{Only for the very lightest nuclei is
the coupling closer to the $LS$ type.}. The individual nucleon vectors
$\mathbf j$ then combine into the total nuclear angular momentum $\mathbf J$
(usually called simply the \emph{nuclear spin}, as though the nucleus were an
elementary particle). In this respect nuclear-level classification differs
essentially from that of atomic levels: in an atomic electron shell the
relativistic spin--orbit coupling is generally small compared with the direct
electric and exchange interactions, and the levels are therefore usually
classified by $LS$ coupling.

The state of each nucleon in a nucleus is characterized by its angular
momentum $j$ and parity. Although neither $\mathbf l$ nor $\mathbf s$ is
separately conserved, the magnitude of the nucleon's orbital angular momentum
is nevertheless definite. A given $j$ can arise either from $l=j-1/2$ or from
$l=j+1/2$. For fixed half-integral $j$ these states have different parities
$(-1)^l$; specification of $j$ and parity therefore also determines $l$.

\SourcePage{584}{587}
Nucleon states with the same $l$ and $j$ are conventionally numbered, in order
of increasing energy, by a ``principal quantum number'' $n$ taking integral
values beginning with 1\footnote{This differs from the convention for atomic
electron levels, in which $n$ begins with $l+1$.}. The states are denoted
$1s_{1/2}$, $1p_{1/2}$, $1p_{3/2}$, and so forth. The digit preceding the
letter is the principal quantum number, the letters $s$, $p$, $d$, \ldots
indicate $l$ in the usual way, and the suffix gives $j$. A state with fixed
$n,l,j$ can contain at most $2j+1$ neutrons and the same number of protons.

The characteristics of a nuclear state as a whole (for a specified
configuration) are conventionally written as the value of $J$ with a $+$ or
$-$ suffix indicating its parity. In the shell model the latter is fixed by
whether the algebraic sum of the $l$ values of all nucleons is even or odd.

Analysis of experimental data on nuclear properties reveals several regular
features in the arrangement of nuclear levels.

First, nucleon-level energy increases with orbital angular momentum $l$. As
$l$ increases, the centrifugal energy of the particle increases and its
binding energy consequently decreases.

Next, for fixed $l$ the level $j=l+1/2$, corresponding to parallel
$\mathbf l$ and $\mathbf s$, lies deeper than the level $j=l-1/2$. This rule
was already noted above in discussing the nuclear spin--orbit coupling.

Another rule concerns nuclear isospin. Recall that the projection $T_\zeta$
is already determined by the mass and charge numbers of the nucleus (see
(116.1)). At fixed $T_\zeta$, the magnitude of the isospin may take any value
satisfying $T\geqslant|T_\zeta|$. The nuclear ground state usually has the
smallest allowed isospin:
\begin{equation}
 T_{\mathrm{ground}}=|T_\zeta|=(1/2)(N-Z).
 \tag{118.7}
\end{equation}
This rule follows from the character of the neutron--proton interaction: in an
$np$ system the state with $T=0$ (the deuteron state) has greater binding
energy than the state with $T=1$ (see the note on p.~580).

Some rules can also be stated for the spins of nuclear ground states. They
specify how the angular momenta $\mathbf j$ of the separate nucleons combine
into the total nuclear spin. They express the tendency of protons and neutrons
occupying identical sta\SourcePageJoin{585}{588}tes to form $pp$ and $nn$ pairs with opposite angular
momenta (the binding energy of such pairs is of order 1--2 MeV). This
pairing tendency means, for example, that when both the proton and neutron
numbers are even (an \emph{even--even nucleus}), all nucleon angular momenta
cancel in pairs and the total nuclear angular momentum is zero.

If the nucleus has an odd number of either protons or neutrons and all
nucleons outside closed shells occupy identical states, its total angular
momentum usually equals that of one nucleon: after all possible proton and
neutron pairs have formed, only one nucleon appears to retain an uncompensated
angular momentum. The total angular momenta of closed shells vanish
automatically.

For \emph{odd--odd nuclei}, where both $Z$ and $N$ are odd, there is no
sufficiently general rule determining the ground-state spin.

A detailed treatment of the actual order in which nuclear shells fill would
require an extensive analysis of experimental data and lies outside the scope
of this book. We give only a few further general remarks.

In studying atoms we found that their electron states can be divided into
groups such that, on completing one group and proceeding to the next, the
electron binding energy falls. An analogous situation occurs in nuclei, the
nucleon states being distributed among the following groups:
\begin{equation}
 \begin{array}{@{}ll@{}}
  1s_{1/2} & 2\text{ nucleons},\\
  1p_{3/2},\ 1p_{1/2} & 6\text{ nucleons},\\
  1d_{5/2},\ 1d_{3/2},\ 2s_{1/2} & 12\text{ nucleons},\\
  1f_{7/2},\ 2p_{3/2},\ 1f_{5/2},\ 2p_{1/2},\ 1g_{9/2}
    & 30\text{ nucleons},\\
  2d_{5/2},\ 1g_{7/2},\ 1h_{11/2},\ 2d_{3/2},\ 3s_{1/2}
    & 32\text{ nucleons},\\
  2f_{7/2},\ 1h_{9/2},\ 1i_{13/2},\ 2f_{5/2},\ 3p_{3/2},\ 3p_{1/2}
    & 44\text{ nucleons}.
 \end{array}
 \tag{118.8}
\end{equation}
For each group the total number of proton or neutron vacancies is indicated.
Thus a group is filled when the total proton number $Z$ or neutron number $N$
in the nucleus reaches one of the values
\[
 2,\ 8,\ 20,\ 50,\ 82,\ 126.
\]
These are conventionally called \emph{magic numbers}\footnote{The
$1f_{7/2}$ states, with their 8 vacancies, are sometimes separated into a
distinct group, in accordance with the fact that 28 also has some of the
properties of a magic number.}.

\SourcePage{586}{589}
Especially stable are the so-called \emph{doubly magic nuclei}, in which both
$Z$ and $N$ are magic numbers. Compared with neighbouring nuclei, they have
an anomalously small affinity for one more nucleon, and their first excited
levels lie anomalously high\footnote{Examples are
${}^{4}_{2}\mathrm{He}_{2}$, ${}^{16}_{8}\mathrm{O}_{8}$,
${}^{40}_{20}\mathrm{Ca}_{20}$, and ${}^{208}_{82}\mathrm{Pb}_{126}$;
${}^{4}\mathrm{He}$ cannot bind one more nucleon at all. The lower-right
suffix gives the value of $N$.}.

The states in each group in (118.8) are listed approximately in their order of
filling along the sequence of nuclei. The actual filling displays substantial
irregularities. Moreover, in heavy nuclei far from magic numbers, separations
between levels can become comparable with the ``pairing energy''; under these
conditions the very notion of individual states of the two members of a pair
largely loses its meaning.

We now make several remarks on calculating the nuclear magnetic moment in the
shell model. By the magnetic moment of a nucleus we naturally mean the moment
averaged over the motion of its constituent particles. This mean moment
$\overline{\boldsymbol\mu}$ is directed along the nuclear spin $\mathbf J$,
the only distinguished direction in the nucleus. Its operator is therefore
\begin{equation}
 \hat{\overline{\boldsymbol\mu}}=\mu_0g\hat{\mathbf J},
 \tag{118.9}
\end{equation}
where $\mu_0$ is the nuclear magneton and $g$ the gyromagnetic factor. The
eigenvalue of its projection is $\overline\mu_z=\mu_0gM_J$. Usually (compare
(111.1)) the nuclear magnetic moment $\mu$ means simply the maximum projection,
$\mu=\mu_0gJ$; with this notation
\begin{equation}
 \hat{\overline{\boldsymbol\mu}}=\mu(\hat{\mathbf J}/J).
 \tag{118.10}
\end{equation}

The nuclear magnetic moment is the sum of the moments of nucleons outside
closed shells, since the nucleon moments in closed shells cancel. Each nucleon
produces a moment consisting of two parts, spin and (for a proton) orbital,
and hence expressible as $g_s\hat{\mathbf s}+g_l\hat{\mathbf l}$. Here and
below we omit the factor $\mu_0$, adopting the customary convention that
magnetic moments are measured in nuclear magnetons. The orbital and spin
gyromagnetic factors are $g_l=1$, $g_s=5{.}585$ for the proton and $g_l=0$,
$g_s=-3{.}826$ for the neutron.

\SourcePage{587}{590}
After averaging over the nucleon's motion in the nucleus, its magnetic moment
is proportional to $\mathbf j$. Writing it as $g_j\hat{\mathbf j}$, we have
\[
 g_j\hat{\mathbf j}=g_s\hat{\mathbf s}+g_l\hat{\mathbf l}
 =\frac12(g_l+g_s)\hat{\mathbf j}
 +\frac12(g_l-g_s)(\hat{\mathbf l}-\hat{\mathbf s}).
\]
Multiplication by $\hat{\mathbf j}=\hat{\mathbf l}+\hat{\mathbf s}$ and
replacement by eigenvalues gives
\[
 g_jj(j+1)=\frac12(g_l+g_s)j(j+1)
 +\frac12(g_l-g_s)[l(l+1)-s(s+1)].
\]
Putting $s=1/2$ and $j=l\pm1/2$, we find
\begin{equation}
 g_j=g_l\pm\frac{g_s-g_l}{2l+1}\quad\text{for }j=l\pm1/2.
 \tag{118.11}
\end{equation}
With the gyromagnetic factors given above this yields, for the proton magnetic
moment $\mu_p=g_jj$,
\begin{equation}
 \begin{aligned}
  \mu_p&=\left(1-\frac{2{.}29}{j+1}\right)j
     &&\text{for }j=l-1/2,\\
  \mu_p&=j+2{.}29 &&\text{for }j=l+1/2,
 \end{aligned}
 \tag{118.12}
\end{equation}
and, for the neutron magnetic moment,
\begin{equation}
 \begin{aligned}
  \mu_n&=\frac{1{.}91}{j+1}j &&\text{for }j=l-1/2,\\
  \mu_n&=-1{.}91 &&\text{for }j=l+1/2.
 \end{aligned}
 \tag{118.13}
\end{equation}
(T.~Schmidt, 1937).

If there is only one nucleon outside closed shells, (118.12) or (118.13)
directly gives the nuclear magnetic moment. For two nucleons their magnetic
moments are also combined straightforwardly (see Problem 1). With more
nucleons the magnetic moment must be averaged using the system wavefunction,
properly constructed from the individual nucleon functions. Specification of
the nucleon configuration and of the state of the nucleus as a whole makes
this construction unique when only one system state with the given $J$ and
$T$ can correspond to the configuration (see, for example, Problem 3).
Otherwise the nuclear state is a mixture of several independent states with
the same $J,T$, and the coefficients in the linear combination forming the
nuclear wavefunction are in general unknown\footnote{The accuracy of this
``one-particle'' scheme for nuclear magnetic moments is in fact low. The two
values in each of (118.12) and (118.13) act more like upper and lower bounds
than exact moment values.}.

\SourcePage{588}{591}
Finally, spin--orbit coupling of the nucleons gives protons in the nucleus an
additional magnetic moment relative to (118.9) (M.~G\"oppert-Mayer,
J.~H.~Jensen, 1952). When the interaction operator depends explicitly on
particle velocity, an external field is introduced by the replacement
$\hat{\mathbf p}\to\hat{\mathbf p}-(e/c)\mathbf A$. Applying it in (118.3)
and using (111.7) for the vector potential, we find an additional term in the
proton Hamiltonian:
\[
 \varphi(r)\frac{e}{cm_p}(\mathbf n\times\mathbf A)\hat{\mathbf s}
 =f(r)\frac{e}{2c\hbar}\{\mathbf r\times(\mathbf H\times\mathbf r)\}
   \hat{\mathbf s}
 =f(r)\frac{e}{2c\hbar}\{\mathbf r\times(\hat{\mathbf s}\times\mathbf r)\}
   \mathbf H.
\]
This term is equivalent to an additional magnetic moment with operator
\begin{equation}
 \hat{\boldsymbol\mu}_{\mathrm{add}}
 =-\frac{e}{2c\hbar}f(r)\{\mathbf r\times(\hat{\mathbf s}\times\mathbf r)\}
 =-\frac{e}{2c\hbar}r^2f(r)
 \{\hat{\mathbf s}-(\hat{\mathbf s}\mathbf n)\mathbf n\}.
 \tag{118.14}
\end{equation}

\ProblemHeading 1. Determine the magnetic moment of a two-nucleon system, with total
angular momentum $\mathbf J=\mathbf j_1+\mathbf j_2$, in terms of the
individual moments $\mu_1$ and $\mu_2$.

\SolutionHeading As in the derivation of (118.11), one obtains
\[
 \frac\mu J=\frac12\left(\frac{\mu_1}{j_1}+\frac{\mu_2}{j_2}\right)
 +\frac12\left(\frac{\mu_1}{j_1}-\frac{\mu_2}{j_2}\right)
 \frac{(j_1-j_2)(j_1+j_2+1)}{J(J+1)}.
\]

\ProblemHeading 2. Find the possible states of a system of three nucleons with
$j=3/2$ and equal principal quantum numbers.

\SolutionHeading We proceed as in \S~67 when finding the possible states of
equivalent electrons. Each nucleon can occupy one of eight states with the
following pairs $(m_j,\tau_\zeta)$:
\[
 \begin{gathered}
 (3/2,1/2),\quad(1/2,1/2),\quad(-1/2,1/2),\quad(-3/2,1/2),\\
 (3/2,-1/2),\quad(1/2,-1/2),\quad(-1/2,-1/2),\quad(-3/2,-1/2).
 \end{gathered}
\]
Combining three distinct states gives the following pairs $(M_J,T_\zeta)$:
\[
 (7/2,1/2),\quad2(5/2,1/2),\quad(3/2,3/2),\quad4(3/2,1/2),
 \quad(1/2,3/2),\quad5(1/2,1/2),
\]
where a prefactor gives the number of corresponding states; states with
negative $M_J,T_\zeta$ need not be listed. They correspond to system states
with
\[
 (J,T)=(7/2,1/2),\quad(5/2,1/2),\quad(3/2,3/2),\quad
 (3/2,1/2),\quad(1/2,1/2).
\]

\ProblemHeading 3. Determine the magnetic moment of the ground state of a
configuration of two neutrons and one proton in $p_{3/2}$ states (with equal
$n$), taking isotopic invariance into account\footnote{The nucleus
${}^{7}\mathrm{Li}$ has this configuration outside the closed
$(1s_{1/2})^4$ shell.}.

\SourcePage{589}{592}
\SolutionHeading The ground state of this configuration has $J=3/2$, and by
the rule stated in the text its isospin has the least possible value,
$T=|T_\zeta|=1/2$.

We determine the system wavefunction corresponding to the largest possible
$M_J=3/2$. Subject to the Pauli principle for the two identical nucleons, this
value is realized by the following triples of $m_j$ values for $p,n,n$:
\[
 (3/2,3/2,-3/2),\quad(3/2,1/2,-1/2),\quad
 (1/2,3/2,-1/2),\quad(1/2,3/2,1/2).
\]
Hence the required wavefunction $\Psi^{JM_J}_{TT_\zeta}$ is a linear
combination
\begin{equation}
 \begin{split}
 \Psi^{3/2\;3/2}_{1/2\;-1/2}
 &=a[\psi^{3/2}_{1/2}\psi^{3/2}_{-1/2}\psi^{-3/2}_{-1/2}]
 +b[\psi^{3/2}_{1/2}\psi^{1/2}_{-1/2}\psi^{-1/2}_{-1/2}]\\
 &\quad+c[\psi^{-3/2}_{-1/2}\psi^{1/2}_{1/2}\psi^{-1/2}_{-1/2}]
 +d[\psi^{3/2}_{-1/2}\psi^{1/2}_{-1/2}\psi^{-1/2}_{1/2}],
 \end{split}
 \tag{1}
\end{equation}
where $[\ldots]$ denotes the normalized antisymmetrized product (a determinant
of the form (61.5)) of individual nucleon functions
$\psi^{m_j}_{\tau_\zeta}$.

The function (1) must be annihilated by
\[
 \hat T_- =\sum_{i=1}^3\hat\tau_-^{(i)}
 \quad\text{and}\quad
 \hat J_+=\sum_{i=1}^3\hat j_+^{(i)}
\]
(compare the problem in \S~67). The operators $\hat\tau_-^{(i)}$ turn the
proton function of nucleon $i$ into a neutron function, and annihilate a
neutron function. Thus $\hat T_-$ turns the first term of (1) into a
determinant with two identical rows, hence zero, while the determinants from
the other three terms become identical. Therefore $b+c+d=0$. For a nucleon
with $j=3/2$, (27.12) gives
\[
 \hat j_+\psi^{3/2}=0,\quad
 \hat j_+\psi^{1/2}=\sqrt3\psi^{3/2},\quad
 \hat j_+\psi^{-1/2}=2\psi^{1/2},\quad
 \hat j_+\psi^{-3/2}=\sqrt3\psi^{-1/2}.
\]
It follows that
\[
 \hat J_+\Psi^{3/2\;3/2}_{1/2\;-1/2}
 =\sqrt3(a+b-c)[\psi^{3/2}_{1/2}\psi^{3/2}_{-1/2}\psi^{-1/2}_{-1/2}]
 +2(c-d)[\psi^{3/2}_{-1/2}\psi^{1/2}_{1/2}\psi^{1/2}_{-1/2}],
\]
where some signs change on permuting determinant rows. Setting this expression
to zero gives
\[
 a+b-c=0,\qquad c-d=0.
\]
Together with normalization of (1), these relations give
\[
 a=\frac3{\sqrt{15}},\qquad b=-\frac2{\sqrt{15}},\qquad
 c=d=\frac1{\sqrt{15}}.
\]
The mean projection of a proton or neutron moment in a state of given $m_j$ is
$\mu_pm_j/j$ or $\mu_nm_j/j$. Averaging with (1), therefore, gives
\[
 \overline\mu=\overline\mu_z
 =\frac9{15}\mu_p+\frac4{15}\mu_p
 +\frac1{15}\left(\frac13\mu_p+\frac23\mu_n\right)
 +\frac1{15}\left(-\frac13\mu_p+\frac34\mu_n\right)
 =\frac1{15}(13\mu_p+\mu_n).
\]
For a nucleon in the $p_{3/2}$ state, (118.12) and (118.13) give
$\mu_n=-1{.}91$ and $\mu_p=3{.}79$. Hence $\mu=3{.}03$.
