\SourcePage{575}{578}
\section{Nuclear forces}

The distinctive feature of the \emph{nuclear forces} acting between nucleons
is, first of all, their short range: they decrease exponentially over
distances of order $10^{-13}$ cm.

In the nonrelativistic limit the nuclear forces may be regarded as independent
of the nucleon velocities and as having a potential (the velocities of
nucleons in a nucleus are approximately one quarter of the speed of light;
see below). The potential energy $U$ of interaction between two nucleons
depends not only on their separation $r$, but also on their spins, and the spin
dependence is by no means weak\footnote{In this respect the interaction of
nucleons differs essentially from that of electrons. For electrons the
spin--spin interaction has a purely relativistic origin and is small (in
atoms).}. Its precise dependence on $r$ could, of course, be established only
by a consistent theory of nuclear forces. The form of its spin dependence,
however, follows already from simple considerations involving the properties
of spin operators.

There are only three vectors on which the interaction energy $U$ can depend:
the unit vector $\mathbf n$ directed along the radius vector between the two
nucleons, and their spins $\mathbf s_1$ and $\mathbf s_2$. From the general
properties of a spin-$1/2$ operator, any function of it reduces to a linear
function (see §~55). It must also be taken into account that the product
$\mathbf n\mathbf s$ is a pseudoscalar rather than a true scalar (because
$\mathbf n$ is a polar vector and $\mathbf s$ an axial vector). Consequently,
only two independent scalar quantities, each linear in either spin, can be
formed from $\mathbf n$, $\mathbf s_1$, and $\mathbf s_2$:
$\mathbf s_1\mathbf s_2$ and
$(\mathbf n\mathbf s_1)(\mathbf n\mathbf s_2)$\footnote{Here the nuclear
forces are understood to be invariant under spatial inversion, so that they
cannot contain pseudoscalar terms. At present there is no experimental
evidence to the contrary.}.

Thus, as regards its dependence on the spins, the interaction operator for two
nucleons can be represented as a sum of three independent terms,
\begin{equation}
 \hat U_{\text{ordinary}}=U_1(r)+U_2(r)(\hat{\mathbf s}_1\hat{\mathbf s}_2)
 +U_3(r)\bigl[3(\hat{\mathbf s}_1\mathbf n)
 (\hat{\mathbf s}_2\mathbf n)-\hat{\mathbf s}_1\hat{\mathbf s}_2\bigr],
 \tag{117.1}
\end{equation}
one of which is spin-independent and two spin-dependent. The third term has
been written in this form so that it vanishes upon averaging over the
directions of $\mathbf n$; the forces described by this term are usually
called \emph{tensor forces}.

The subscript ``ordinary'' was attached to the interaction (117.1) to
emphasize that this operator does not change the

\SourcePage{576}{579}
charge state of the nucleons. An interaction in which a proton turns into a
neutron, and conversely, is also possible. The operator of this ``exchange''
interaction differs in form from (117.1) by the presence of the particle
permutation operator (116.4):
\begin{equation}
 \hat U_{\text{exch}}=\left\{U_4(r)+U_5(r)
 (\hat{\mathbf s}_1\hat{\mathbf s}_2)+U_6(r)
 \bigl[3(\hat{\mathbf s}_1\mathbf n)(\hat{\mathbf s}_2\mathbf n)
 -\hat{\mathbf s}_1\hat{\mathbf s}_2\bigr]\right\}\hat P.
 \tag{117.2}
\end{equation}
The complete interaction operator is the sum
\begin{equation}
 \hat U=\hat U_{\text{ordinary}}+\hat U_{\text{exch}}.
 \tag{117.3}
\end{equation}
The interaction between two nucleons is therefore characterized by six
different functions of their separation. In general, all these terms have the
same order of magnitude\footnote{We also note that an interaction depending
on the nucleon velocities is described, to first order in those velocities,
by an operator of the form
\[
 [\varphi_1(r)+\varphi_2(r)\hat P]\hat{\mathbf L}\hat{\mathbf S},
\]
where $\mathbf L=\mathbf r\times\mathbf p$ is the orbital angular momentum of
the relative motion of the nucleons ($\mathbf p$ is its momentum), and
$\mathbf S=\mathbf s_1+\mathbf s_2$; this operator contains two functions of
$r$. Terms of the forms $\mathbf p\mathbf n$ and $\mathbf S\mathbf n$ are
excluded by the requirements of invariance under inversion and time
reversal.}.

The spin operators occurring in (117.1) and (117.2) can be expressed in terms
of the total-spin operator $\mathbf S$. Indeed, on squaring the equalities
$\hat{\mathbf S}=\hat{\mathbf s}_1+\hat{\mathbf s}_2$ and
$\hat{\mathbf S}\mathbf n=\hat{\mathbf s}_1\mathbf n+
\hat{\mathbf s}_2\mathbf n$, and using
$\hat{\mathbf s}_1^2=\hat{\mathbf s}_2^2=3/4$ and
$(\hat{\mathbf s}_1\mathbf n)^2=(\hat{\mathbf s}_2\mathbf n)^2=1/4$
(see (55.10)), we obtain
\begin{equation}
 \hat{\mathbf s}_1\hat{\mathbf s}_2
 =\frac12\left(\hat{\mathbf S}^2-\frac32\right),\qquad
 (\hat{\mathbf s}_1\mathbf n)(\hat{\mathbf s}_2\mathbf n)
 =\frac12\left[(\hat{\mathbf S}\mathbf n)^2-\frac12\right].
 \tag{117.4}
\end{equation}

The operator $\hat{\mathbf S}^2$ commutes with $\hat{\mathbf S}$, and hence
the interactions described by the first two terms of (117.1) and (117.2)
preserve the total-spin vector of the system. The tensor interaction, by
contrast, contains the operator $(\hat{\mathbf S}\mathbf n)^2$, which
commutes with the square $\hat{\mathbf S}^2$ but not with the vector
$\hat{\mathbf S}$ itself. As a result, only the magnitude of the total spin,
and not its direction, is conserved.

The total spin $S$ of a two-nucleon system can take the values 0 and 1. Its
total isotopic spin $T$ has the same two possible values. All possible states
of this system consequently fall into four groups distinguished by the pair
of numbers $S,T$. For the states in each such group

\SourcePage{577}{580}
there is a corresponding interaction operator of the form $A(r)$ (when
$S=0$), or
$A(r)+B(r)[(\hat{\mathbf S}\mathbf n)^2-2/3]$ (when $S=1$), to which the
general operator (117.3) reduces in these cases (see Problem 1)\footnote{The
experimental properties of the deuteron show that for $T=0$, $S=1$ the
nucleon interaction includes a strong attraction with a deep potential well
(the presence of tensor forces makes it difficult to state this fact as a
property of the functions $A(r)$ and $B(r)$). It can also be concluded, from
the sign of the observed quadrupole moment of the deuteron, that the
coefficient $B(r)$ of the tensor forces is negative in this state. Nucleon
scattering data show that attraction is also present for $T=1$, $S=0$, but it
is weaker and, in particular, does not produce a stable two-particle system.}.

For fixed values of $S$ and $T$, the states of the system are classified by
the total angular momentum $J$ and the parity. As we know, $T=0$ corresponds
to states with symmetric wave functions $\psi$, whereas $T=1$ corresponds to
antisymmetric ones. On the other hand, $S$ determines the symmetry of the wave
function with respect to the spin variables (symmetric for $S=1$ and
antisymmetric for $S=0$). It is therefore clear that specifying the pair
$S,T$ also determines the symmetry of the wave function with respect to the
spatial variables, that is, the parity of the state. States of a system with
isotopic spin $T=0$ can evidently be only even triplets ($S=1$) or odd
singlets ($S=0$); states with isotopic spin $T=1$ are odd triplets or even
singlets.

Since spin is not conserved as a vector, orbital angular momentum is not, in
general, conserved either (only the sum $\mathbf J=\mathbf L+\mathbf S$ is
conserved). Nevertheless, the magnitude $L$ may be conserved simply because
specified values of $J$, $S$, and parity (or of $J$, $S$, and $T$) can be
compatible with only one definite value of $L$ (recall that the parity of a
two-particle system is $(-1)^L$). Thus, an odd state with $S=1$, $J=1$ can
have only $L=1$, and is therefore a ${}^3P_1$ state. In other cases two
different values of $L$ may correspond to the prescribed $J$, $S$, and
parity, so that $L$ is not conserved. For example, in an odd state with
$S=1$, $J=2$, either $L=1$ or $L=3$ is possible; such a state is a
superposition ${}^3P_2+{}^3F_2$.

We thereby obtain the following possible states of a two-nucleon system (the
subscript $\pm$ denotes the parity):

\SourcePage{578}{581}
\begin{align*}
 \text{for }T=1:&\quad {}^3P_0^-,\ {}^3P_1^-,\
 ({}^3P_2+{}^3F_2)^-,\ {}^3F_3^-,\ldots;\\
 &\quad {}^1S_0^+,\ {}^1D_2^+,\ {}^1G_4^+,\ldots,\\
 \text{for }T=0:&\quad ({}^3S_1+{}^3D_1)^+,\ {}^3D_2^+,\\
 &\quad ({}^3D_3+{}^3G_3)^+,\ldots;\
 {}^1P_1^-,\ {}^1F_3^-,\ldots
\end{align*}

Nuclear forces are not additive in general. This means that the interaction
in a system of more than two nucleons does not reduce to a sum of the
interactions of every pair of particles. Three-particle and higher
interactions apparently play a relatively small role in comparison with pair
interactions, however, so the properties of complex nuclei can to a large
extent be treated on the basis of the properties of pair interactions.

Experimental data on nuclei show that, as the number of particles $A$
increases, the nucleon system begins to behave like macroscopic ``nuclear
matter,'' whose volume and energy grow in proportion to $A$ (apart from
effects associated with the Coulomb interaction of protons and with the free
surface of the nucleus). The property of nuclear forces responsible for this
phenomenon is called their \emph{saturation}.

The existence of this property imposes certain restrictions on the functions
$U_1,\ldots,U_6$ defining the pair interactions between nucleons. Imagine all
particles concentrated in a volume whose dimensions are of the order of the
range of the nuclear forces; every pair then interacts. If there is a
configuration of some set of nucleons (and an orientation of their spins) in
which attractive forces act between every pair, the potential energy of the
system will be a negative quantity proportional to $A^2$. Its kinetic energy,
on the other hand, is positive and proportional to $A^{5/3}$, hence to a lower
power of $A$\footnote{The density $n$ with which the particles are
concentrated in the given volume is proportional to their number $A$, while
the kinetic energy of each is then proportional to $n^{2/3}$ (cf. (70.1)).
The total kinetic energy is consequently proportional to $AA^{2/3}$.}. Under
these conditions, a sufficiently large collection of nucleons would indeed
be concentrated in a small volume independent of $A$, and would therefore not
form nuclear matter. The saturation condition for the nuclear forces must
thus amount to conditions excluding configurations that yield a negative
interaction energy proportional to $A^2$ (see Problem 2).

The proportionality between the volume of nuclear matter and the number of
particles is expressed by a relation of the form
\begin{equation}
 R=r_0A^{1/3},
 \tag{117.5}
\end{equation}

\SourcePage{579}{582}
which relates the nuclear radius $R$ to its particle number $A$. Experimental
data (from electron scattering by nuclei) give
$r_0\approx1.1\cdot10^{-13}$ cm.

Let us determine the limiting momentum of nucleons in nuclear matter (cf.
§~70). The phase-space volume corresponding to particles that occupy unit
volume in physical space and have momenta $p\leqslant p_0$ is
$4\pi p_0^3/3$. Dividing this by $(2\pi\hbar)^3$ gives the number of
``cells,'' each of which can contain two protons and two neutrons. Taking the
number of protons equal to the number of neutrons, we obtain
\[
 4\frac{4\pi}{3}\left(\frac{p_0}{2\pi\hbar}\right)^3=\frac AV,
\]
where $V$ is the nuclear volume. Substitution of (117.5) gives
\[
 p_0=\left(\frac{3\pi^2A}{2V}\right)^{1/3}\hbar
 =\frac{(9\pi)^{1/3}\hbar}{2r_0}
 =1.4\cdot10^{-14}\ \text{g}\cdot\text{cm}\cdot\text{s}^{-1}.
\]
The corresponding energy $p_0^2/2m_p$ ($m_p$ is the nucleon mass) is about
$30$ MeV, and the velocity is $p_0/m_p\approx c/4$.

\subsection*{Problems}

\ProblemHeading{1.} Find the interaction operators for two nucleons in states
with specified values of $S$ and $T$.

\SolutionHeading The required operators $\hat U_{ST}$ are obtained from the
general expression (117.1)--(117.3), with relations (116.3) and (117.4) taken
into account:
\begin{align*}
 \hat U_{00}&=U_1-\frac34U_2+U_4-\frac34U_5,\\
 \hat U_{01}&=U_1-\frac34U_2-U_4+\frac34U_5,\\
 \hat U_{10}&=U_1+\frac14U_2+U_4+\frac14U_5
 +\frac12(U_3+U_6)\bigl[3(\hat{\mathbf S}\mathbf n)^2-2\bigr],\\
 \hat U_{11}&=U_1+\frac14U_2-U_4-\frac14U_5
 +\frac12(U_3-U_6)\bigl[3(\hat{\mathbf S}\mathbf n)^2-2\bigr].
\end{align*}

\ProblemHeading{2.} Find the conditions for saturation of the nuclear forces,
assuming that tensor forces are absent and that the ranges of all the other
types of force are equal.

\SolutionHeading Consider certain limiting cases (between which all other
possible cases lie) for the state of a system of $A$ nucleons, and write the
conditions under which the interaction energy of an ``average'' nucleon pair
in this system is positive.

Let the total spin and isotopic spin of the nucleus have their greatest
possible values, $S_{\text{nuc}}=T_{\text{nuc}}=A/2$ (all particles in the
system are protons with parallel spins). For every nucleon pair we then have
$S=T=1$, giving the condition
\begin{equation}
 U_{11}>0.
 \tag{1}
\end{equation}

\SourcePage{580}{583}
Now let $T_{\text{nuc}}=A/2$, $S_{\text{nuc}}=0$. Every pair then has $T=1$,
whereas the mean value of $s_z$ for an individual nucleon is zero. The latter
means that the nucleon can have $s_z=1/2$ or $s_z=-1/2$ with equal
probability; under these conditions the probabilities that a nucleon pair is
in a state with $S=0$ or $S=1$ are respectively $1/2$ and $3/4$ (they are
proportional to the number $2S+1$ of possible values of $S_z$). Hence the
condition that the mean pair energy be positive is
\begin{equation}
 \frac14U_{01}+\frac34U_{11}>0.
 \tag{2}
\end{equation}

Similarly, consideration of the state with $T_{\text{nuc}}=0$,
$S_{\text{nuc}}=A/2$ leads to
\begin{equation}
 \frac14U_{10}+\frac34U_{11}>0.
 \tag{3}
\end{equation}
In the state with $T_{\text{nuc}}=S_{\text{nuc}}=0$, the probability that a
pair has $S=T=1$ is $3/4\cdot3/4$, the probability that it has $T=1$, $S=0$
is $3/4\cdot1/4$, and so forth. We thereby find the condition
\begin{equation}
 \frac9{16}U_{11}+\frac3{16}(U_{10}+U_{01})+\frac1{16}U_{00}>0.
 \tag{4}
\end{equation}

Finally, let the system consist of $A/2$ protons and $A/2$ neutrons, with the
spins of all the protons parallel to one another and antiparallel to the spins
of all the neutrons. An individual nucleon can with equal probability be a
$p$ or an $n$, that is, it can have $T_\zeta=1/2$ or $T_\zeta=-1/2$; the
probability that a nucleon pair has $T=0$ is $1/4$. One nucleon in the pair is
then a $p$ and the other an $n$, and therefore $S_z=0$. This value of $S_z$
can arise with equal probability from states with $S=0$ or $S=1$. Thus the
probabilities that the pair is in a state with $T=0$, $S=0$ or with $T=0$,
$S=1$ are each $1/4\cdot1/2=1/8$. The state with $T=1$, $S=0$ has the same
probability, and the remaining $5/8$ belongs to the state with $T=S=1$.
Taking all this into account gives the condition
\begin{equation}
 \frac18(U_{00}+U_{01}+U_{10})+\frac58U_{11}>0.
 \tag{5}
\end{equation}
Inequalities (1)--(5) are the required system of saturation conditions for
the nuclear forces.
