\SourcePage{597}{600}
\section{Isotope shift}

Properties specific to the nucleus---its finite mass, dimensions, and spin---
distinguish it from a fixed point centre of a Coulomb field and have a definite
effect on the electronic energy levels of an atom.

One such effect is the so-called \emph{isotope shift} of the levels: the change
in a level's energy when one isotope of an element is replaced by another. In
practice, of course, it is not the change in one level that is of interest, but
the change in the separation between two levels, observed as a spectral line.
For this reason, one must consider not the energy of the entire electron shell
as a whole, but only the part associated with the electron taking part in the
spectral transition in question.

In light atoms the principal source of the isotope shift is the finite mass of
the nucleus. Allowance for nuclear motion introduces into the atomic
Hamiltonian the term
\[
 \frac1{2M}\left(\sum_i\hat{\mathbf p}_i\right)^2,
\]
where $M$ is the nuclear mass and $\mathbf p_i$ are the electron
momenta\footnote{In the centre-of-mass frame of the atom, the sum of the
nuclear and electron momenta vanishes:
$\mathbf p_{\text{nuc}}+\sum_i\mathbf p_i=0$. Their total kinetic energy is
therefore
\[
 \frac{\mathbf p_{\text{nuc}}^2}{2M}+\frac1{2m}\sum_i\mathbf p_i^2
 =\frac1{2M}\left(\sum_i\mathbf p_i\right)^2
 +\frac1{2m}\sum_i\mathbf p_i^2.
\]
}. The isotope shift due to this effect is consequently found as the mean
value
\begin{equation}
 \frac12\left(\frac1{M_1}-\frac1{M_2}\right)
 \overline{\left(\sum_i\mathbf p_i\right)^2},
 \tag{120.1}
\end{equation}
calculated with the wavefunction of the given atomic state ($M_1$ and $M_2$
are the nuclear masses of the isotopes).

\SourcePage{598}{601}
In heavy atoms the main contribution to the isotope shift is associated with
the finite extent of the nucleus. In practice this effect is appreciable only
for levels of an outer electron in an $s$ state, since an $s$-state
wavefunction, unlike those of states with $l\ne0$, does not vanish as
$r\to0$. The probability of finding the electron within the ``nuclear
volume'' is therefore comparatively large. We shall calculate the isotope
shift for this case\footnote{The calculation below neglects relativistic
effects in the motion of the electron near the nucleus and is valid when
$Ze^2/(\hbar c)\ll1$.}.

Let $\varphi(r)$ denote the actual electrostatic potential of the nucleus, as
distinct from the potential $Ze/r$ of the Coulomb field of a point charge
$Ze$. The change in the electron energy relative to its value in the pure
Coulomb field $Ze/r$ is then given by
\begin{equation}
 \Delta E=-e\int\left(\varphi-\frac{Ze}{r}\right)\psi^2(r)\,dV,
 \tag{120.2}
\end{equation}
where $\psi(r)$ is the electron wavefunction (for an $s$ state it is
spherically symmetric and real). Although the integration is formally over
all space, the difference $\varphi-Ze/r$ in the integrand is nonzero only
inside the nucleus.

On the other hand, as $r\to0$ the $s$-state wavefunction tends to a constant
limit (see \S~32), and this constant value is already attained, to sufficient
accuracy, outside the nucleus. Thus $\psi^2$ may be taken outside the integral
after replacing $\psi(r)$ by its value at $r=0$, calculated for the Coulomb
field of a point charge.

To transform the integral further, we use the identity $\Delta r^2=6$ and
rewrite (120.2) as
\[
 \Delta E=-\frac16e\psi^2(0)\int
 \left(\varphi-\frac{Ze}{r}\right)\Delta r^2\,dV
 =-\frac16e\psi^2(0)\int r^2\Delta
 \left(\varphi-\frac{Ze}{r}\right)dV
\]
(in transforming the volume integral, the resulting integral over a surface
at infinity has been taken to vanish). But
$\Delta(1/r)=-4\pi\delta(\mathbf r)$, while
$r^2\delta(\mathbf r)=0$ for every $r$. The electrostatic Poisson equation is
$\Delta\varphi=-4\pi\rho$, where here $\rho$ is the density of electric charge
in the nucleus. We therefore obtain
\begin{equation}
 \Delta E=\frac{2\pi}{3}\psi^2(0)Ze^2\overline{r^2},
 \tag{120.3}
\end{equation}

\SourcePage{599}{602}
where
\[
 \overline{r^2}=\frac1{Ze}\int\rho r^2\,dV
\]
is the proton mean-square radius of the nucleus (for a uniform distribution
of protons in the nucleus one would have
$\overline{r^2}=3R^2/5$, where $R$ is the geometrical radius of the nucleus).
The isotope shift of a level is determined by the difference between the
expressions (120.3) for the two isotopes.

In \S~71 the magnitude of $\psi(0)$ was estimated and was found to depend on
the atomic number (assumed large) as $\sqrt Z$. Hence the magnitude of the
splitting (120.3) is proportional to $R^2Z^2$.
