\SourcePage{367}{370}
\section{Crossing of electronic terms}

The electronic terms of a diatomic molecule may be represented graphically as functions of the distance $r$ between the nuclei, with the energy plotted against $r$. The question whether the curves representing different terms can cross is of considerable interest.

Let $U_1(r)$ and $U_2(r)$ be two different electronic terms. If they cross at some point, then near that point the functions $U_1$ and $U_2$ must have close values. It is convenient to formulate the question of whether such a crossing is possible in the following way.

Consider a point $r_0$ at which the values of $U_1(r)$ and $U_2(r)$ are very close but not equal; denote them by $E_1$ and $E_2$. We ask whether $U_1$ and $U_2$ can be made equal by shifting the point through a small amount $\delta r$. The energies $E_1$ and $E_2$ are eigenvalues of the Hamiltonian $\widehat H_0$ for the electrons in the field of nuclei separated by the distance $r_0$. If $r$ is increased by $\delta r$, the Hamiltonian becomes $\widehat H_0+\widehat V$, where $\widehat V=\delta r\,\partial\widehat H_0/\partial r$ is a small correction; the values of $U_1$ and $U_2$ at $r_0+\delta r$ may be regarded as eigenvalues of the new Hamiltonian. This formulation makes it possible to find the values of the terms $U_1(r)$ and $U_2(r)$ at $r_0+\delta r$ by perturbation theory, with $\widehat V$ treated as a perturbation of $\widehat H_0$.

The ordinary method of perturbation theory is not applicable here, however, because the energy eigenvalues $E_1$ and $E_2$ of the unperturbed problem lie very close to each other, and their difference need not be large compared with the perturbation (condition (38.9) is not fulfilled). As the limit $E_2-E_1=0$ gives the case of degenerate eigenvalues, it is natural to treat close eigenvalues by a method analogous to that developed in \S~39.

Let $\psi_1$ and $\psi_2$ be eigenfunctions of the unperturbed operator $\widehat H_0$, belonging to the energies $E_1$ and $E_2$. As the initial zeroth approximation, instead of $\psi_1$ and $\psi_2$ themselves, take their linear combinations
\begin{equation}
 \psi=c_1\psi_1+c_2\psi_2.
 \tag{79.1}
\end{equation}

\SourcePage{368}{371}
Substitution of this expression into the perturbed equation
\begin{equation}
 (\widehat H_0+\widehat V)\psi=E\psi,
 \tag{79.2}
\end{equation}
gives
\[
 c_1(E_1+\widehat V-E)\psi_1+c_2(E_2+\widehat V-E)\psi_2=0.
\]
Multiplying this equation on the left in turn by $\psi_1^*$ and $\psi_2^*$ and integrating, we obtain the two algebraic equations
\begin{equation}
 \left.\begin{aligned}
 c_1(E_1+V_{11}-E)+c_2V_{12}&=0,\\
 c_1V_{21}+c_2(E_2+V_{22}-E)&=0.
 \end{aligned}\right\}
 \tag{79.3}
\end{equation}
Since $\widehat V$ is Hermitian, the matrix elements $V_{11}$ and $V_{22}$ are real, while $V_{12}=V_{21}^*$. The compatibility condition for these equations is
\[
 \begin{vmatrix}E_1+V_{11}-E&V_{12}\\V_{21}&E_2+V_{22}-E\end{vmatrix}=0,
\]
and hence
\begin{equation}
 E=\frac12(E_1+E_2+V_{11}+V_{22})\pm
 \sqrt{\frac14(E_1-E_2+V_{11}-V_{22})^2+|V_{12}|^2}.
 \tag{79.4}
\end{equation}
This formula determines the required energy eigenvalues in the first approximation.

If the energies of the two terms become equal at $r_0+\delta r$, so that the terms cross, the two values of $E$ given by (79.4) must coincide. The expression under the square root in (79.4) must therefore vanish. Since it is a sum of two squares, the conditions for a crossing point are
\begin{equation}
 E_1-E_2+V_{11}-V_{22}=0,\qquad V_{12}=0.
 \tag{79.5}
\end{equation}
But only one arbitrary parameter is available to specify the perturbation $\widehat V$, namely the displacement $\delta r$. Thus, in general, the two equations (79.5) cannot be fulfilled simultaneously. Here we assume that $\psi_1$ and $\psi_2$ have been chosen real, so that $V_{12}$ is real as well.

It may happen, however, that the matrix element $V_{12}$ vanishes identically. Only one

\SourcePage{369}{372}
equation (79.5) then remains, and it can be fulfilled by a suitable choice of $\delta r$. This occurs whenever the two terms under consideration have different symmetries. Here symmetry includes every possible kind: symmetry with respect to rotations about the axis, reflections in planes, inversion, and permutations of electrons. For a diatomic molecule, therefore, the terms may differ in $\Lambda$, parity, or multiplicity; for $\Sigma$ terms they may also be $\Sigma^+$ and $\Sigma^-$.

This statement follows from the fact that the perturbation operator, like the Hamiltonian itself, commutes with all the symmetry operators of the molecule: the operator of angular momentum about the axis, the reflection and inversion operators, and the electron-permutation operators. It was shown in \S\S~29 and 30 that, for a scalar quantity whose operator commutes with the angular-momentum and inversion operators, only matrix elements for transitions between states of equal angular momentum and parity can be nonzero. Essentially the same proof applies to an arbitrary symmetry operator. We shall not repeat it here, especially since another general proof, based on group theory, will be given in \S~97.

We therefore reach the result that in a diatomic molecule only terms of different symmetry can cross; terms of the same symmetry cannot cross (E.~Wigner, J.~von Neumann, 1929). If some approximate calculation were to give two crossing terms of the same symmetry, the next approximation would separate them, as the solid curves in Fig.~27 show.

It should be stressed that this result is not confined to a diatomic molecule. It is in fact a general theorem of quantum mechanics, valid whenever the Hamiltonian contains a parameter and its eigenvalues are consequently functions of that parameter.

In the language of group theory (see \S~96), the general requirement for terms to be capable of crossing is that they belong to different irreducible representations of the symmetry group of the system's Hamiltonian\footnote{The electronic terms of the $\mathrm H_2^+$ ion appear to be an exception to this rule. These terms are characterized by the angular-momentum projection $\Lambda$ and by two elliptic quantum numbers $n_\xi$, $n_\eta$ (see the problem in \S~78). Since all these numbers are associated with functions of different variables, there is in general no reason to prevent terms $E(R)$ with different pairs $n_\xi$, $n_\eta$ but the same $\Lambda$ from crossing, although such terms have the same symmetry with respect to rotations and reflections. In reality, however, separability of the variables in the Schrödinger equation for this system means that its Hamiltonian has a higher symmetry than its geometrical properties alone suggest; relative to this full symmetry group, states with different values of $n_\xi$, $n_\eta$ belong to different types.}.

\SourcePage{370}{373}
In a polyatomic molecule the electronic terms depend not on one parameter but on several: the distances between the various nuclei.

Let $s$ be the number of independent internuclear distances. In an $N$-atom molecule ($N>2$), with an arbitrary arrangement of the nuclei, this number is $s=3N-6$. Geometrically, each term $U_n(r_1,\ldots,r_s)$ is a surface in a space of $s+1$ dimensions. One may then speak of intersections of these surfaces along manifolds of various dimensions, from 0 (intersection at a point) to $s-1$.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm]
 \draw[->] (0,0)--(3.6,0) node[right] {$r$};
 \draw[->] (0,0)--(0,2.8) node[right] {$U(r)$};
 \draw[dashed] (.25,.45)--(3.2,2.35);
 \draw[dashed] (.25,1.35)--(3.25,1.05);
 \draw[smooth] plot coordinates {(.25,1.34)(.8,1.27)(1.35,1.29)(1.8,1.45)(2.25,1.82)(2.8,2.35)};
 \draw[smooth] plot coordinates {(.25,.45)(.85,.84)(1.35,1.16)(1.75,1.28)(2.25,1.30)(2.8,1.19)(3.25,1.06)};
\end{tikzpicture}
\caption*{Fig. 27}
\end{figure}

The foregoing derivation remains fully valid, except that the perturbation $\widehat V$ is now determined not by one parameter but by $s$ parameters, the displacements $\delta r_1,\ldots,\delta r_s$. With as few as two parameters, however, the two equations (79.5) can in general be fulfilled. We thus arrive at the result that any two terms in a polyatomic molecule can cross. For terms of the same symmetry, the crossing is specified by the two conditions (79.5), so the manifold along which the crossing occurs has dimension $s-2$. For terms of different symmetry only one condition remains, and the crossing occurs along a manifold of dimension $s-1$.

Thus, for $s=2$, the terms are represented by surfaces in a three-dimensional coordinate system. If the terms have different symmetries, these surfaces intersect along lines ($s-1=1$); if they have the same symmetry, they intersect at points ($s-2=0$). The shape of the surfaces near a crossing point in the latter case is readily found. The energy values near such points are given by (79.4). In this expression the matrix elements $V_{11}$, $V_{22}$, and $V_{12}$ are linear functions of the displacements $\delta r_1$, $\delta r_2$, and hence linear

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=.9cm,y=.9cm]
 \draw[->] (0,0)--(0,3.1) node[above] {$U$};
 \draw[->] (0,0)--(2.8,0) node[right] {$r_2$};
 \draw[->] (0,0)--(-1.1,-1.0) node[below] {$r_1$};
 \foreach \h/\rx/\ry in {.35/.25/.10,.7/.5/.2,1.05/.75/.3,1.4/1/.4}
   {\draw (-\rx,1.55+\h) arc[start angle=180,end angle=360,x radius=\rx,y radius=\ry];
    \draw (-\rx,1.55-\h) arc[start angle=180,end angle=360,x radius=\rx,y radius=\ry];}
 \draw (-1,2.95)--(0,1.55)--(1,2.95);
 \draw (-1,.15)--(0,1.55)--(1,.15);
\end{tikzpicture}
\caption*{Fig. 28}
\end{figure}

\SourcePage{371}{374}
functions of the distances $r_1$, $r_2$ themselves. As is known from analytic geometry, an equation of this form describes an elliptic cone. Thus, near their crossing points the terms are represented by a two-napped elliptic cone of arbitrary orientation (Fig.~28).
