\SourcePage{401}{404}
\section{Symmetry of molecular terms}

In \S~78 we have already considered some symmetry properties of the terms of a diatomic molecule. Those properties described the behavior of the wave functions under transformations that leave the nuclear coordinates unchanged. Thus, symmetry of the molecule under reflection in a plane through its axis gives the distinction between $\Sigma^+$ and $\Sigma^-$ terms; symmetry under reversal of the signs of all electronic coordinates (for a molecule composed of identical atoms)\footnote{The origin of coordinates is assumed to lie on the molecular axis, midway between the two nuclei.} divides the terms into even and odd ones.

These symmetry properties characterize an electronic term and are the same for all rotational levels belonging to that term.

The states of a molecule (like those of any system of particles; see \S~30) are also characterized by their behavior under inversion, that is, simultaneous reversal of the signs of the coordinates of all electrons and nuclei.

Accordingly, all molecular terms are divided into \emph{positive} terms, whose wave functions are unchanged when the signs of the electron and nuclear coordinates are reversed, and \emph{negative} terms, whose wave functions change sign under inversion\footnote{We retain the accepted terminology. It is unfortunate, since for an atom the behavior of terms under inversion is described by their parity, rather than by their sign.

For $\Sigma$-terms, this sign must not be confused with the $+$ and $-$ signs written as superscripts!}.

For $\Lambda\ne0$, every term is doubly degenerate, corresponding to the two possible directions of the angular momentum relative to the axis

\SourcePage{402}{405}
of the molecule. Under inversion the angular momentum itself does not change sign, but the molecular axis reverses direction (the atoms interchange places), and hence the direction of $\Lambda$ relative to the molecule also reverses. The two wave functions belonging to a given energy level are therefore transformed into one another, and from them one can always form one linear combination invariant under inversion and another that changes sign under it. Each term thus gives two states, one positive and the other negative. In fact, every term with $\Lambda\ne0$ does undergo a splitting (see \S~88), so these two states correspond to different energies.

The signs of $\Sigma$-terms require separate consideration. First, it is clear that spin has no bearing on the sign of a term: inversion acts only on particle coordinates and leaves the spin part of the wave function unchanged. Consequently, all components of the multiplet structure of any one term have the same sign. In other words, the sign of the term depends only on $K$, not on $J$\footnote{Recall that $\Sigma$-terms ordinarily belong to case $b$, and hence the quantum numbers $K$ and $J$ must be used.}.

The molecular wave function is a product of electronic and nuclear wave functions. It was shown in \S~82 that, in a $\Sigma$ state, the motion of the nuclei is equivalent to the motion of one particle with orbital angular momentum $K$ in a central field $U(r)$. The nuclear wave function is therefore multiplied by $(-1)^K$ when the signs of the nuclear coordinates are reversed (see (30.7)).

The electronic wave function characterizes the electronic term. To determine its behavior under inversion, consider it in a coordinate system rigidly attached to the nuclei and rotating with them. Let $xyz$ be a space-fixed coordinate system and $\xi\eta\zeta$ a rotating system in which the molecule as a whole is at rest. Choose the directions of the $\xi\eta\zeta$ axes so that the $\zeta$ axis coincides with the molecular axis and points, say, from nucleus 1 to nucleus 2, while the relative orientation of the positive $\xi\eta\zeta$ directions is the same as in the $xyz$ system (that is, if $xyz$ is right-handed, $\xi\eta\zeta$ must also be right-handed). Inversion reverses the directions of the $xyz$ axes, turning a right-handed system into a left-handed one. The $\xi\eta\zeta$ system must likewise become left-handed. The $\zeta$ axis, however, is rigidly attached to the nuclei and retains its former direction; hence the direction

\SourcePage{403}{406}
of either the $\xi$ or the $\eta$ axis must be reversed. Thus inversion in the fixed coordinate system is equivalent, in the moving system, to reflection in a plane through the molecular axis. Under such a reflection the electronic wave function of a $\Sigma^+$ term is unchanged, while that of a $\Sigma^-$ term changes sign.

It follows that the sign of the rotational components of a $\Sigma^+$ term is determined by the factor $(-1)^K$: all levels with even $K$ are positive and those with odd $K$ are negative. For a $\Sigma^-$ term the sign of the rotational levels is determined by $(-1)^{K+1}$, and all levels with even $K$ are negative while those with odd $K$ are positive.

If the molecule consists of identical atoms\footnote{The two atoms must belong not only to the same element but also to the same isotope.}, its Hamiltonian is also invariant under interchange of the coordinates of the two nuclei. A term is called symmetric with respect to the nuclei if its wave function is unchanged when the nuclei are interchanged, and antisymmetric if the wave function changes sign. Symmetry with respect to the nuclei is closely related to the parity and sign of a term. Interchanging the nuclear coordinates is equivalent to reversing the signs of the coordinates of all particles (electrons and nuclei), followed by reversing only the electronic coordinates. Hence, if a term is even (odd) and at the same time positive (negative), it is symmetric with respect to the nuclei. If a term is even (odd) and at the same time negative (positive), it is antisymmetric with respect to the nuclei.

At the end of \S~62 a general theorem was established: the coordinate wave function of a system of two identical particles is symmetric for even total spin and antisymmetric for odd total spin. Applying this result to the two nuclei of a molecule composed of identical atoms, we find that the symmetry of the term is related to the parity of the resultant spin $I$ obtained by combining the spins $i$ of the two nuclei. The term is symmetric for even $I$ and antisymmetric for odd $I$\footnote{In view of the relation among parity, sign, and symmetry of terms, it follows that for even resultant nuclear spin $I$, positive levels are even and negative levels are odd; for odd $I$ the reverse holds.}. In particular, if the nuclei have no spin ($i=0$), then $I$ is also zero; the molecule will therefore have no antisymmetric terms at all. Thus nuclear spin has an important indirect effect on molecular terms, although its

\SourcePage{404}{407}
direct effect (the hyperfine structure of the terms) is altogether negligible.

Allowance for nuclear spin gives an additional degeneracy of the levels. In the same \S~62 the numbers of states with even and odd values of $I$ obtained by combining two spins $i$ were counted. Thus, for half-integral $i$, the number of states with even $I$ is $i(2i+1)$ and that with odd $I$ is $(i+1)(2i+1)$. It follows from the preceding discussion that the ratio of the degeneracies $g_s,g_a$\footnote{In this connection, the degeneracy of a level is often called its \emph{statistical weight}. Formulas (86.1) and (86.2) determine the ratios of the nuclear statistical weights of symmetric and antisymmetric levels.} of the symmetric and antisymmetric terms for half-integral $i$ is
\begin{equation}
 \frac{g_s}{g_a}=\frac{i}{i+1}.
 \tag{86.1}
\end{equation}
For integral $i$, the corresponding ratio is
\begin{equation}
 \frac{g_s}{g_a}=\frac{i+1}{i}.
 \tag{86.2}
\end{equation}

We have seen that the sign of the rotational components of a $\Sigma^+$ term is determined by $(-1)^K$. Thus, for example, the rotational components of a ${}^1\Sigma_g^+$ term are positive and hence symmetric for even $K$, while for odd $K$ they are negative and hence antisymmetric. In view of the results just obtained, the nuclear statistical weights of the rotational components of a ${}^1\Sigma_g^+$ level with successive values of $K$ alternate in the ratio (86.1) or (86.2). The same applies to the levels ${}^1\Sigma_u^-$, and also to ${}^1\Sigma_g^-$ and ${}^1\Sigma_u^+$. In particular, for $i=0$ the statistical weights vanish for levels with even $K$ in the terms ${}^1\Sigma_u^-$, ${}^1\Sigma_g^-$, and for levels with odd $K$ in the terms ${}^1\Sigma_g^+$, ${}^1\Sigma_u^+$. In other words, in the electronic states ${}^1\Sigma_u^-$, ${}^1\Sigma_g^-$ there are no rotational states with even $K$, while in the states ${}^1\Sigma_g^+$, ${}^1\Sigma_u^+$ there are no rotational states with odd $K$.

Because the interaction of the nuclear spins with the electrons is extremely weak, the probability of a change in $I$ is very small even in molecular collisions. Molecules differing in the parity of $I$, and consequently possessing only symmetric or only antisymmetric terms, therefore behave in practice as distinct modifications of the substance. Examples are the so-called orthohydrogen and parahydrogen: in a molecule of the former, the spins $i=1/2$ of the two nuclei are parallel ($I=1$), while in the latter they are antiparallel ($I=0$).
