\SourcePage{405}{408}
\section{Matrix elements for a diatomic molecule}

This section gives several general formulas for matrix elements of physical quantities of a diatomic molecule. We begin with matrix elements for transitions between states of zero spin.

Let $\mathbf A$ be some vector physical quantity characterizing the molecule with its nuclei fixed (for example, its electric or magnetic dipole moment). First consider this quantity in the coordinate system $\xi\eta\zeta$ rotating with the molecule, with the $\zeta$ axis along the molecular axis. The angular momentum of the molecule relative to this system (that is, the electronic angular momentum $\mathbf L$) is not conserved in full, but its $\zeta$ component is conserved. The selection rules for the quantum number $L_\zeta=\Lambda$ therefore remain valid (and coincide with the selection rules for $M$ in \S~29). Thus the nonzero vector matrix elements are
\begin{equation}
\langle n'\Lambda|A_\zeta|n\Lambda\rangle,\quad
\langle n'\Lambda|A_\xi+iA_\eta|n,\Lambda-1\rangle,\quad
\langle n',\Lambda-1|A_\xi-iA_\eta|n\Lambda\rangle
\tag{87.1}
\end{equation}
($n$ enumerates the electronic terms for a given $\Lambda$).

If both terms are $\Sigma$ terms, one must also take account of the selection rule associated with reflection symmetry in a plane containing the molecular axis. Under such a reflection the $\zeta$ component of an ordinary (polar) vector does not change, whereas that of an axial vector changes sign. It follows that for a polar vector $A_\zeta$ has nonzero matrix elements only for the transitions $\Sigma^+\to\Sigma^+$ and $\Sigma^-\to\Sigma^-$, and for an axial vector only for $\Sigma^+\to\Sigma^-$. We do not discuss the components $A_\xi,A_\eta$, since transitions without a change in $\Lambda$ are impossible for them in any case.

If the molecule consists of identical atoms, there is also a parity selection rule. The components of a polar vector change sign under inversion. Their matrix elements are therefore nonzero only for transitions between states of opposite parity (and, for an axial vector, conversely). In particular, all diagonal matrix elements of the components of a polar vector vanish identically.

The connection between the matrix elements (87.1) and those of the same vector in the fixed coordinate system $xyz$ is furnished by the general formulas derived below, in \S~110, for any axially symmetric physical system.

After separating out the dependence on the quantum number $M_K$ common to every vector (the $z$ projection of the molecule's total angular momentum

\SourcePage{406}{409}
$\mathbf K$), there remain the reduced matrix elements $\langle n'K'\Lambda'\|A\|nK\Lambda\rangle$. Their relation to (87.1) is specified by (110.7), with $k=k'=1$ (corresponding to a vector) and the appropriate change of notation for the quantum numbers (recall that, by (82.4), $\Lambda$ equals the $\zeta$ component of the total angular momentum $\mathbf K$). Using the relation (107.1) between the components of a spherical tensor of rank one and the Cartesian vector components, and taking the $3j$ symbols from Table~9 (p.~530), we obtain for the elements diagonal in $\Lambda$:
\begin{equation}
\begin{split}
\langle n'K\Lambda\|A\|nK\Lambda\rangle&=\Lambda\sqrt{\frac{2K+1}{K(K+1)}}\langle n'\Lambda|A_\zeta|n\Lambda\rangle,\\
\langle n',K-1,\Lambda\|A\|nK\Lambda\rangle&=i\sqrt{\frac{K^2-\Lambda^2}{K}}\langle n'\Lambda|A_\zeta|n\Lambda\rangle
\end{split}
\tag{87.2}
\end{equation}
and for the elements nondiagonal in $\Lambda$:
\begin{equation}
\begin{split}
\langle n'K\Lambda\|A\|nK,\Lambda-1\rangle
&=\left[\frac{(2K+1)(K+\Lambda)(K-\Lambda+1)}{4K(K+1)}\right]^{1/2}\\[-2pt]
&\qquad\times\langle n'\Lambda|A_\xi+iA_\eta|n,\Lambda-1\rangle,\\
\langle n'K\Lambda\|A\|n,K-1,\Lambda-1\rangle
&=i\left[\frac{(K+\Lambda)(K+\Lambda-1)}{4K}\right]^{1/2}\\[-2pt]
&\qquad\times\langle n'\Lambda|A_\xi+iA_\eta|n,\Lambda-1\rangle,\\
\langle n',K-1,\Lambda\|A\|nK,\Lambda-1\rangle
&=i\left[\frac{(K-\Lambda)(K-\Lambda+1)}{4K}\right]^{1/2}\\[-2pt]
&\qquad\times\langle n'\Lambda|A_\xi+iA_\eta|n,\Lambda-1\rangle.
\end{split}
\tag{87.3}
\end{equation}
The remaining nonzero elements follow from these by using the Hermiticity relations for the reduced matrix elements,
\[
\langle nK\Lambda\|A\|n'K'\Lambda'\rangle=\langle n'K'\Lambda'\|A\|nK\Lambda\rangle^*,
\]
and for the matrix elements in the $\xi\eta\zeta$ system,
\[
\langle n\Lambda|A_\xi-iA_\eta|n'\Lambda'\rangle=\langle n'\Lambda'|A_\xi+iA_\eta|n\Lambda\rangle^*,\qquad
\langle n\Lambda|A_\zeta|n'\Lambda'\rangle=\langle n'\Lambda'|A_\zeta|n\Lambda\rangle^*.
\]

\SourcePage{407}{410}
We record separately the formulas for the matrix elements of the vector $\mathbf A=\mathbf n$, the unit vector along the molecular axis. Here simply $A_\xi=A_\eta=0$, $A_\zeta=1$, so in the $\xi\eta\zeta$ system only the diagonal elements are nonzero: $\langle n\Lambda|A_\zeta|n\Lambda\rangle=1$. The reduced matrix elements are diagonal in every index except $K$; displaying only that index, we have
\begin{equation}
\langle K\|n\|K\rangle=\Lambda\sqrt{\frac{2K+1}{K(K+1)}},\qquad
\langle K-1\|n\|K\rangle=i\sqrt{\frac{K^2-\Lambda^2}{K}}.
\tag{87.4}
\end{equation}
(H.~Hönl, F.~London, 1925). For $\Lambda=0$ these formulas give
\[
\langle K\|n\|K\rangle=0,\qquad \langle K-1\|n\|K\rangle=i\sqrt K,
\]
which, as expected, are precisely the matrix elements of a unit vector for motion in a centrally symmetric field (see (29.14)).

We now indicate how the formulas must be modified for transitions between states of nonzero spin. It is important here whether the states belong to case $a$ or case $b$.

If both states belong to case $a$, the formulas change essentially only in notation. The quantum numbers $K,M_K$ do not exist; in their place are the total angular momentum $J$ and its $z$ projection $M_J$. In addition there are the numbers $S$ and $\Omega=\Lambda+\Sigma$, so the reduced matrix elements are written
\[
\langle n'J'S'\Omega'\Lambda'\|A\|nJS\Omega\Lambda\rangle.
\]
Let $\mathbf A$ be an orbital vector, that is, one independent of spin. Its operator commutes with the spin operator $\widehat{\mathbf S}$, so its matrix is diagonal in $S$ and $S_\zeta=\Sigma$; consequently $\Omega=\Lambda+\Sigma$ changes together with $\Lambda$ (that is, $\Omega'-\Omega=\Lambda'-\Lambda$). Formulas (87.2)--(87.4) change only by the addition of indices to the matrix elements and the replacement of $K,\Lambda$ by $J,\Omega$ in the remaining factors. For example, the first formula in (87.2) is replaced by
\[
\langle n'JS\Omega\Lambda\|A\|nJS\Omega\Lambda\rangle
=\Omega\sqrt{\frac{2J+1}{J(J+1)}}\langle n'\Omega\Lambda|A_\zeta|n\Omega\Lambda\rangle
\]
(the diagonal index $S$ is omitted).

Now let $\mathbf A=\mathbf S$. Since the spin operator commutes with the orbital angular momentum and with the Hamiltonian, its matrix is diagonal in $n,\Lambda$. It is not, however, diagonal in $S$ and $\Sigma$ (or $\Omega$). The matrix elements of $A_\xi,A_\eta,A_\zeta$ for

\SourcePage{408}{411}
transitions $S,\Sigma\to S',\Sigma'$ are given by (27.13), with $S,\Sigma$ in place of $L,M$. The change to the $xyz$ coordinate system is then made by (87.2), (87.3), replacing $K,\Lambda$ by $J,\Omega$. In this way, for example, we obtain
\begin{equation}
\begin{split}
\langle J\Omega\|S\|J,\Omega-1\rangle
&=\left[\frac{(2J+1)(J+\Omega)(J-\Omega+1)}{4J(J+1)}\right]^{1/2}
\langle\Omega|S_\xi+iS_\eta|\Omega-1\rangle\\
&=\left[\frac{(2J+1)(J+\Omega)(J-\Omega+1)(S+\Sigma)(S-\Sigma+1)}{4J(J+1)}\right]^{1/2}.
\end{split}
\tag{87.5}
\end{equation}
(the diagonal indices $n,S,\Lambda$ are omitted).

Next let both states belong to case $b$, and let $\mathbf A$ be an orbital vector. The matrix elements are calculated in two stages. First consider the rotating molecule before coupling $\mathbf S$ and $\mathbf K$; the matrix elements are diagonal in $S$ and are given by the same formulas (87.2), (87.3). At the second stage $\mathbf K$ is coupled with $\mathbf S$ to form the total angular momentum $\mathbf J$, and the passage to the new matrix elements is made by the general formulas (109.3) (with $K,S,J$ playing the roles of $j_1,j_2,J$ in those formulas). Thus, for elements diagonal in $J,K,\Lambda$, we first obtain
\[
\langle n'JK\Lambda\|A\|nJK\Lambda\rangle
=(-1)^{K+J+S+1}(2J+1)
\begin{Bmatrix}K&J&S\\J&K&1\end{Bmatrix}
\langle n'K\Lambda\|A\|nK\Lambda\rangle,
\]
and then, taking the value of the $6j$ symbol from Table~10 (p.~542) and the reduced matrix element from (87.2), finally
\[
\langle n'JK\Lambda\|A\|nJK\Lambda\rangle
=\Lambda\left[\frac{2J+1}{J(J+1)}\right]^{1/2}
\frac{J(J+1)+K(K+1)-S(S+1)}{2K(K+1)}
\langle n'\Lambda|A_\zeta|n\Lambda\rangle.
\]

Matrix elements for transitions between two states, one belonging to case $a$ and the other to case $b$, are calculated similarly; we shall not pursue that calculation here.

\ProblemHeading[1] Determine the Stark splitting of the terms of a diatomic molecule possessing a permanent dipole moment; the term belongs to case $a$.

\SourcePage{409}{412}
\SolutionHeading The energy of a dipole $\mathbf d$ in an electric field $\boldsymbol{\mathcal E}$ is $-\mathbf d\boldsymbol{\mathcal E}$. Symmetry makes it clear that the dipole moment of a diatomic molecule is directed along its axis: $\mathbf d=d\mathbf n$ ($d$ is constant). Choosing the field direction as the $z$ axis, the perturbation operator is $-dn_z\mathcal E$.

Using the formulas derived above to find the diagonal matrix elements of $n_z$, we obtain for case $a$ the following level splitting:\footnote{This may appear to contradict the general statement (see \S~76) that the linear Stark effect is absent. There is, of course, no contradiction: here the linear effect is associated with the twofold degeneracy of levels having $\Omega\ne0$. The formula therefore applies provided that the Stark-splitting energy is large compared with the energy of the so-called $\Lambda$-doubling (see \S~88).}
\[
\Delta E=-\mathcal E dM_J\frac{\Omega}{J(J+1)}.
\]

\ProblemHeading[2] The same, for a term belonging to case $b$ (with $\Lambda\ne0$).

\SolutionHeading In the same way we find
\[
\Delta E=-\mathcal E dM_J\Lambda\frac{J(J+1)-S(S+1)+K(K+1)}{2K(K+1)J(J+1)}.
\]

\ProblemHeading[3] The same for a ${}^1\Sigma$ term.

\SolutionHeading For $\Lambda=0$ the linear effect is absent, and second-order perturbation theory is required. In the sum in the general formula (38.10), it is sufficient to retain only terms corresponding to transitions between rotational components of the same electronic term (the energy differences in the denominators of the other terms are large). Thus
\[
\Delta E=d^2\mathcal E^2\left\{\frac{|\langle KM_K|n_z|K-1,M_K\rangle|^2}{E_K-E_{K-1}}+\frac{|\langle KM_K|n_z|K+1,M_K\rangle|^2}{E_K-E_{K+1}}\right\},
\]
where $E_K=BK(K+1)$. A direct calculation gives
\[
\Delta E=\frac{d^2\mathcal E^2}{B}\frac{K(K+1)-3M_K^2}{2K(K+1)(2K-1)(2K+3)}.
\]
