\section{Multipole moments}

In classical theory the electrical properties of a system are characterized by its multipole moments of various orders, expressed in terms of the charges and coordinates of the particles. In quantum theory these quantities retain the same definitions, but must be regarded as operators.

The first multipole moment is the \emph{dipole moment}, the vector
\[\mathbf d=\sum e\mathbf r\]
(the sum is over all particles in the system; the particle index is omitted for brevity). The matrix of

\SourcePage{343}{346}
this operator, like that of any polar vector (see \S~30), has nonzero elements only for transitions between states of different parity. All its diagonal elements therefore vanish. In other words, the mean dipole moment of any system of particles, for example an atom, is zero in every stationary state.\footnote{To avoid misunderstanding, we emphasize that this statement concerns a closed system of particles, or particles in a centrally symmetric external electric field. Thus, if the nuclei are regarded as ``fixed,'' the general assertion applies to the electron system of an atom, but not of a molecule.

It is also assumed that the energy level has no additional, or ``accidental,'' degeneracy besides degeneracy with respect to the direction of the total angular momentum. Otherwise stationary-state wave functions without definite parity can be formed, and the corresponding diagonal elements of the dipole moment need not vanish.}

The same clearly applies to all $2^l$-pole moments with odd $l$. Their components are odd polynomials of degree $l$ in the coordinates and, like a polar-vector component, change sign under coordinate inversion; the same parity selection rule therefore applies.

The \emph{quadrupole moment} of a system is defined as the symmetric tensor
\begin{equation}\widehat Q_{ik}=\sum e(3x_ix_k-\delta_{ik}\mathbf r^2),\tag{75.1}\end{equation}
whose diagonal sum is zero. To determine these quantities in a particular state of a system, say an atom, the operator (75.1) must be averaged with the corresponding wave function. It is useful to perform this averaging in two stages (cf. \S~72).

Let $\widehat Q_{ik}$ denote the quadrupole-moment operator averaged over electron states having a specified total angular momentum $J$, but not a specified projection $M_J$. The resulting operator can depend only on operators describing the state of the atom as a whole. The only such vector is $\widehat{\mathbf J}$. Hence
\begin{equation}
 \widehat Q_{ik}=\frac{3Q}{2J(2J-1)}\left(\widehat J_i\widehat J_k+\widehat J_k\widehat J_i-\frac23\widehat{\mathbf J}^{2}\delta_{ik}\right),
 \tag{75.2}
\end{equation}
where the expression in parentheses is symmetric in $i,k$ and contracts to zero over those indices (the meaning of $Q$ is given below). The opera-

\SourcePage{344}{347}
tors $\widehat J_i$ are the familiar matrices (\S~27, 54) between states of different $M_J$; $\widehat{\mathbf J}^{2}$ may of course be replaced by its eigenvalue $J(J+1)$.

Since the three components of $\mathbf J$ cannot have definite values simultaneously, neither can the components of $Q_{ik}$. For $Q_{zz}$,
\[\widehat Q_{zz}=\frac{3Q}{J(2J-1)}\left(\widehat J_z^2-\frac13\widehat{\mathbf J}^{2}\right).\]
In a state with $\mathbf J^2=J(J+1)$ and $J_z=M_J$, $Q_{zz}$ also has a definite value:
\begin{equation}Q_{zz}=\frac{3Q}{J(2J-1)}\left[M_J^2-\frac13J(J+1)\right].\tag{75.3}\end{equation}
For $M_J=J$, when the angular momentum is directed ``entirely'' along $z$, $Q_{zz}=Q$; this quantity is ordinarily called simply the quadrupole moment.

For $J=0$ all angular-momentum matrix elements vanish, and so do the operators (75.2). They also vanish identically for $J=1/2$, as follows directly by multiplying the Pauli matrices (55.7), which represent the components of any angular momentum equal to $1/2$.

This is a special case of the general rule that a $2^l$-pole tensor with even $l$ is nonzero only for states whose total angular momentum satisfies
\begin{equation}J\geqslant l/2.\tag{75.4}\end{equation}
Such a multipole tensor is an irreducible tensor of rank $l$ (see II, \S~41), and (75.4) follows from the general angular-momentum selection rules for its matrix elements: it is the condition under which diagonal elements may be nonzero (\S~107). As noted above, parity also requires $l$ to be even.

Electric multipole moments are purely ``orbital'' quantities: their operators contain no spin operators. Thus, if spin--orbit interaction can be neglected and $L,S$ are separately conserved, their matrix elements obey selection rules in $L$ as well as $J$.

\SourcePage{345}{348}
\ProblemHeading{1.} Find the relation between the atomic quadrupole-moment operators in states belonging to different fine-structure components of a level, that is, different $J$ for fixed $L,S$.

\SolutionHeading For fixed $L,S$, the quadrupole operator, being purely orbital, depends only on $\widehat{\mathbf L}$ and has the form (75.2) with $\widehat{\mathbf J}$ replaced by $\widehat{\mathbf L}$ and with another constant $Q$. Equation (75.2) results by further averaging over a state of specified $J$:
\begin{equation}
\begin{split}
 \widehat Q_{ik}&=\frac{3Q_J}{2J(J-1)}[\widehat J_i\widehat J_k+\widehat J_k\widehat J_i-\tfrac23J(J+1)\delta_{ik}]\\
 &=\frac{3Q_L}{2L(L-1)}\overline{[\widehat L_i\widehat L_k+\widehat L_k\widehat L_i-\tfrac23L(L+1)\delta_{ik}]}.
\end{split}\tag{1}
\end{equation}
The displayed denominators are retained exactly as printed in the source. To relate $Q_J,Q_L$, multiply (1) on the left by $\widehat J_i$ and on the right by $\widehat J_k$, summing over $i,k$, and pass to eigenvalues of the diagonal operators. Then
\[\widehat J_i\widehat L_i\widehat L_k\widehat J_k=(\mathbf{JL})^2,\qquad 2\mathbf{JL}=J(J+1)+L(L+1)-S(S+1).\]
Using
\[\{\widehat L_i,\widehat L_k\}=ie_{ikl}\widehat L_l,\qquad \{\widehat J_i,\widehat L_l\}=ie_{ilm}\widehat L_m,\]
as in the problem to \S~29 gives
\[\widehat J_i\widehat L_k\widehat L_i\widehat J_k=(\mathbf{JL})^2-(\mathbf{JL}),\]
and similarly
\[\widehat J_i\widehat J_i\widehat J_k\widehat J_k=(\mathbf J^2)^2,\qquad \widehat J_i\widehat J_k\widehat J_i\widehat J_k=\mathbf J^2(\mathbf J^2-1).\]
Thus
\begin{equation}Q_J=Q_L\frac{3(\mathbf{JL})(2\mathbf{JL}-1)-2J(J+1)L(L+1)}{(J+1)(2J+3)L(2L-1)}.\tag{2}\end{equation}
For $S=1/2$,
\begin{equation}\begin{aligned}Q_J&=Q_L,&&J=L+\tfrac12,\\Q_J&=Q_L\frac{(L-1)(2L+3)}{L(2L+1)},&&J=L-\tfrac12.\end{aligned}\tag{3}\end{equation}

\ProblemHeading{2.} Express the quadrupole moment of an electron of charge $-|e|$ and orbital angular momentum $l$ in terms of its mean-square distance from the centre.

\SolutionHeading Average
\[Q_{zz}=-|e|\,\overline{r^2}(3\cos^2\theta-1)=-|e|\,\overline{r^2}(3n_z^2-1)\]

\SourcePage{346}{349}
in the state with angular momentum $l$ and projection $m=l$. The angular factor follows directly from the formula obtained in the problem to \S~29, with $\widehat l_z$ replaced by $l$:
\begin{equation}Q_l=|e|\,\overline{r^2}\frac{2l}{2l+3}.\tag{1}\end{equation}
Its sign is opposite to the electron charge, as it should be: a particle whose angular momentum points along $z$ is found mainly near the plane $z=0$, and hence $\cos^2\theta<1/3$. For specified $j=l\pm1/2$, (3) gives
\begin{equation}Q_j=|e|\,\overline{r^2}\frac{2j-1}{2j+2}.\tag{2}\end{equation}

\ProblemHeading{3.} Determine the quadrupole moment, in its ground state, of an atom in which all $\nu$ electrons outside closed shells occupy equivalent states of orbital angular momentum $l$.

\SolutionHeading Since closed shells have zero total quadrupole moment,
\[\widehat Q_{ik}=\frac{3|e|\,\overline{r^2}}{(2l-1)(2l+3)}\sum[\widehat l_i\widehat l_k+\widehat l_k\widehat l_i-\tfrac23l(l+1)\delta_{ik}],\]
summed over the $\nu$ outer electrons (formula (4) was used here).

First suppose $\nu\leqslant2l+1$, so at most half the shell is occupied. By Hund's rule (\S~67), all $\nu$ spins are parallel, $S=\nu/2$. The spin wave function is symmetric, and the coordinate wave function is therefore antisymmetric in these electrons. Their $m$ values must all differ, giving
\[L=(M_L)_{\max}=\sum_{m=l-\nu+1}^{l}m=\tfrac12\nu(2l-\nu+1).\]
The required $Q_L$ is the $Q_{zz}$ eigenvalue at $M_L=L$:
\[Q_L=\frac{6|e|\,\overline{r^2}}{(2l-1)(2l+3)}\sum_{m=l-\nu+1}^{l}[m^2-l(l+1)/3],\]
whence
\begin{equation}Q_L=\frac{2l(2l-2\nu+1)}{(2l-1)(2l+3)}|e|\,\overline{r^2}.\tag{1}\end{equation}
The final passage from $Q_L$ to $Q_J$ uses (2).

For a shell more than half full, use holes instead of electrons. The answer is the same formula (6) with the overall sign reversed, since a hole has charge $+|e|$, and $\nu$ now denotes the number of vacancies rather than electrons.
