\SourcePage{331}{334}
\section{Mendeleev's periodic system of the elements}

To explain the origin of the periodic variation of properties established by D.~I.~Mendeleev (1869) in the sequence of elements arranged by increasing atomic number, we must consider the features of the successive filling of atomic electron shells (N.~Bohr, 1922).

\SourcePage{332}{335}
In passing from one atom to the next, the nuclear charge increases by one and one electron is added to the shell. At first sight, one might expect the binding energies of the successively added electrons to vary monotonically as the atomic number increases. In fact they do not.

The normal state of the hydrogen atom contains a single electron in the $1s$ state. In the atom of the next element, helium, another electron is added in the same $1s$ state. The binding energy of each $1s$ electron in helium is, however, considerably greater than that of the electron in hydrogen. This follows naturally from the difference between the field experienced by the electron in H and that entered by an electron added to He$^+$: at large distances the fields are approximately the same, but near the nucleus the field of the He$^+$ ion with $Z=2$ is stronger than that of the hydrogen nucleus with $Z=1$.

In lithium ($Z=3$), the third electron enters the $2s$ state, since no more than two electrons can occupy $1s$ states simultaneously. At a given $Z$ the $2s$ level lies above $1s$; as the nuclear charge increases, both levels descend. In passing from $Z=2$ to $Z=3$, however, the first effect greatly outweighs the second, and the binding energy of the third electron in Li is much smaller than that of the electrons in helium. From Be ($Z=4$) through Ne ($Z=10$), another $2s$ electron and then six $2p$ electrons are added in succession. Owing to the increasing nuclear charge, the binding energies of the electrons added along this sequence generally increase. The next electron, added in passing to Na ($Z=11$), enters a $3s$ state; the effect of passing to a higher shell outweighs the increased nuclear charge, and the binding energy again falls sharply.

This pattern of shell filling is characteristic of the whole sequence of elements. All electron states may be divided into groups filled successively: while a group is filled along the sequence, the binding energy generally increases, but it drops sharply when filling of the next group begins.

Figure~24 shows the ionization potentials of the elements known from spectroscopic data; they determine the binding energies of the electrons added in passing from each element to the next.

\SourcePage{333}{336}
\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=.105cm,y=.23cm,line width=.35pt]
 \draw[->] (0,0)--(94,0) node[right] {$Z$};
 \draw[->] (0,0)--(0,26) node[above] {$E,\ \text{eV}$};
 \foreach \y in {5,10,15,20,25} \draw (-1,\y)--(0,\y) node[left] {\small\y};
 \foreach \x in {5,10,...,90} \draw (\x,0)--(\x,-.5) node[below] {\small\x};
 \draw plot[smooth] coordinates {(1,13.6)(2,24.6)(3,5.4)(4,9.3)(5,8.3)(6,11.3)(7,14.5)(8,13.6)(9,17.4)(10,21.6)(11,5.1)(12,7.6)(13,6.0)(14,8.2)(15,10.5)(16,10.4)(17,13.0)(18,15.8)(19,4.3)(20,6.1)(21,6.6)(22,6.8)(23,6.7)(24,6.8)(25,7.4)(26,7.9)(27,7.9)(28,7.6)(29,7.7)(30,9.4)(31,6.0)(32,7.9)(33,9.8)(34,9.8)(35,11.8)(36,14.0)(37,4.2)(38,5.7)(39,6.2)(40,6.6)(41,6.8)(42,7.1)(43,7.3)(44,7.4)(45,7.5)(46,8.3)(47,7.6)(48,9.0)(49,5.8)(50,7.3)(51,8.6)(52,9.0)(53,10.5)(54,12.1)(55,3.9)(56,5.2)(57,5.6)(58,5.5)(59,5.5)(60,5.5)(61,5.6)(62,5.6)(63,5.7)(64,6.1)(65,5.9)(66,5.9)(67,6.0)(68,6.1)(69,6.2)(70,6.3)(71,5.4)(72,7.0)(73,7.9)(74,7.9)(75,7.8)(76,8.4)(77,9.1)(78,9.0)(79,9.2)(80,10.4)(81,6.1)(82,7.4)(83,7.3)(84,8.4)(85,9.3)(86,10.7)(87,4.0)(88,5.3)(89,5.2)(90,6.3)};
 \foreach \x/\y/\t in {1/13.6/H,2/24.6/He,3/5.4/Li,4/9.3/Be,5/8.3/B,6/11.3/C,7/14.5/N,8/13.6/O,9/17.4/F,10/21.6/Ne,11/5.1/Na,12/7.6/Mg,14/8.2/Si,15/10.5/P,16/10.4/S,18/15.8/Ar,19/4.3/K,20/6.1/Ca,21/6.6/Sc,22/6.8/Ti,23/6.7/V,24/6.8/Cr,25/7.4/Mn,26/7.9/Fe,30/9.4/Zn,31/6.0/Ga,32/7.9/Ge,33/9.8/As,34/9.8/Se,35/11.8/Br,36/14/Kr,37/4.2/Rb,38/5.7/Sr,41/6.8/Nb,42/7.1/Mo,46/8.3/Pd,48/9/Cd,49/5.8/In,50/7.3/Sn,51/8.6/Sb,52/9/Te,54/12.1/Xe,55/3.9/Cs,56/5.2/Ba,57/5.6/La,71/5.4/Lu,72/7/Hf,73/7.9/Ta,74/7.9/W,75/7.8/Re,78/9/Pt,80/10.4/Hg,81/6.1/Tl,82/7.4/Pb,83/7.3/Bi,84/8.4/Po,86/10.7/Rn,87/4/Fr,88/5.3/Ra,89/5.2/Ac} \node[font=\tiny,anchor=south] at (\x,\y+.15) {\t};
\end{tikzpicture}
\caption{}
\label{fig:ionization-potentials}
\end{figure}

\SourcePage{334}{337}
The states are distributed among successively filled groups as follows:
\begin{equation}
\begin{array}{r@{\ }l}
1s&2\text{ electrons},\\
2s,,2p&8\text{ electrons},\\
3s,,3p&8\text{ electrons},\\
4s,,3d,,4p&18\text{ electrons},\\
5s,,4d,,5p&18\text{ electrons},\\
6s,,4f,,5d,,6p&32\text{ electrons},\\
7s,,6d,,5f&\ldots
\end{array}
\tag{73.1}
\end{equation}
The first group is filled in H and He; filling of the second and third corresponds to the first two short periods of the periodic system, each containing eight elements. Next come two long periods of 18 elements and a long period that includes the rare-earth elements and contains 32 elements in all. The last group is not filled completely in naturally occurring elements or in artificial transuranic elements.

To understand the variation of properties as the states of each group are filled, the following feature of $d$ and $f$ states, which distinguishes them from $s$ and $p$ states, is important. After a rapid, almost vertical fall near the origin, the effective-potential-energy curves of the central field (the electrostatic plus centrifugal field) for an electron in a heavy atom have a deep minimum and then rise toward zero asymptotically. For $s$ and $p$ states these curves lie very close together on their rising portions. Thus the electron lies at approximately the same distances from the nucleus in these states. The curves for $d$ and especially $f$ states lie much farther to the left; the classically accessible region bounded by them ends much nearer the nucleus than for $s$ and $p$ states at the same total electron energy. In other words, in $d$ and $f$ states the electron is mainly much closer to the nucleus than in $s$ and $p$ states.

A number of atomic properties, including the chemical properties of the elements (see \S~81), depend mainly on the outer regions of the electron shells. The feature of $d$ and $f$ states just described is therefore very important. When $4f$ states are filled in the rare-earth elements (see below), the added electrons lie much closer to the nucleus than electrons in the states already filled. Consequently,

\SourcePage{335}{338}
these electrons have almost no effect on chemical properties, and all rare-earth elements are chemically very similar.

Elements with filled $d$ and $f$ shells, or with no such shells, are called \emph{main-group elements}; those in which these states are being filled are called \emph{transition-group elements}. It is convenient to consider the two groups separately.

We begin with the main-group elements. Hydrogen and helium have the normal states
\[
 {}_1\mathrm H\ 1s\,{}^2S_{1/2};\qquad {}_2\mathrm{He}\ 1s^2\,{}^1S_0
\]
(the subscript to the left of a chemical symbol always denotes the atomic number). The electron configurations of the remaining main-group elements are shown in Table~3.

\begin{table}[ht]
\centering\small
\caption{Electron configurations of the main-group elements}
\begin{tabular}{c|cccccccc|l}
&$s$&$s^2$&$s^2p$&$s^2p^2$&$s^2p^3$&$s^2p^4$&$s^2p^5$&$s^2p^6$&\\\hline
$n=2$&$_3$Li&$_4$Be&$_5$B&$_6$C&$_7$N&$_8$O&$_9$F&$_{10}$Ne&$1s^2$\\
3&$_{11}$Na&$_{12}$Mg&$_{13}$Al&$_{14}$Si&$_{15}$P&$_{16}$S&$_{17}$Cl&$_{18}$Ar&$2s^22p^6$\\
4&$_{19}$K&$_{20}$Ca&&&&&&&$3s^23p^6$\\
4&$_{29}$Cu&$_{30}$Zn&$_{31}$Ga&$_{32}$Ge&$_{33}$As&$_{34}$Se&$_{35}$Br&$_{36}$Kr&$3d^{10}$\\
5&$_{37}$Rb&$_{38}$Sr&&&&&&&$4s^24p^6$\\
5&$_{47}$Ag&$_{48}$Cd&$_{49}$In&$_{50}$Sn&$_{51}$Sb&$_{52}$Te&$_{53}$I&$_{54}$Xe&$4d^{10}$\\
6&$_{55}$Cs&$_{56}$Ba&&&&&&&$5s^25p^6$\\
6&$_{79}$Au&$_{80}$Hg&$_{81}$Tl&$_{82}$Pb&$_{83}$Bi&$_{84}$Po&$_{85}$At&$_{86}$Rn&$4f^{14}5d^{10}$\\
7&$_{87}$Fr&$_{88}$Ra&&&&&&&$6s^26p^6$\\\hline
&$^2S_{1/2}$&$^1S_0$&$^2P_{1/2}$&$^3P_0$&$^4S_{3/2}$&$^3P_2$&$^2P_{3/2}$&$^1S_0$&
\end{tabular}
\end{table}

In each atom the shells shown in the right-hand column on the same line, and those on every line above it, are completely filled. The configuration in the shells being filled is shown at the top; the principal quantum number for these states is the number in the left-hand column on the same line. The normal states of the whole atom are shown at the bottom. Thus Al has the configuration $1s^22s^22p^63s^23p\,{}^2P_{1/2}$.

The values of $L$ and $S$ in the normal atomic state can be found from the electron configuration by

\SourcePage{336}{339}
Hund's rule (\S~67), and $J$ by the rule given in \S~72.

The noble-gas atoms (He, Ne, Ar, Kr, Xe, Rn) occupy a special position: each completes the filling of one of the groups in (73.1). Their electron configurations are particularly stable (their ionization potentials are the largest in the corresponding sequences), which accounts for their chemical inertness.

The filling of states among the main-group elements is very regular: first the $s$ and then the $p$ states for each principal quantum number $n$ are filled. The electron configurations of their ions are equally regular, so long as ionization does not affect electrons in $d$ and $f$ shells: each ion has the configuration of the preceding atom. Thus Mg$^+$ has the Na configuration, and Mg$^{++}$ that of Ne.

We next consider transition-group elements. The $3d$, $4d$, and $5d$ shells are filled in the groups called, respectively, the \emph{iron}, \emph{palladium}, and \emph{platinum} groups.

Table~4 gives the experimentally determined electron configurations and terms of the atoms in these groups. Filling of the $d$ shells is plainly much less regular than filling of $s$ and $p$ shells in main-group atoms. A characteristic feature is the competition between $s$ and $d$ states. Instead of the regular sequence $d^ps^2$ with increasing $p$, configurations $d^{p+1}s$ or $d^{p+2}$ are often more favorable. In the iron group, for example, Cr is $3d^54s$, not $3d^44s^2$; Ni, with eight $d$ electrons, is followed directly by Cu with a completely filled $d$ shell (and thus assigned above to the main groups). The terms of the ions show a similar irregularity: ionic configurations usually do not coincide with those of the preceding atoms. Thus V$^+$ is $3d^4$, not $3d^24s^2$ as in Ti, and Fe$^+$ is $3d^64s$, rather than the $3d^54s^2$ configuration of Mn. All ions occurring naturally in crystals and solutions have only $d$ electrons, and no $s$ or $p$ electrons, in their unfilled shells. Iron, for example, occurs in crystals and solutions only as Fe$^{++}$ and Fe$^{+++}$, with configurations $3d^6$ and $3d^5$ respectively.

\SourcePage{337}{340}
\begin{table}[ht]
\centering\scriptsize
\caption{Electron configurations of atoms in the iron, palladium, and platinum groups}
\begin{tabular}{c|cccccccc}
\multicolumn{9}{c}{Iron group}\\
&$_{21}$Sc&$_{22}$Ti&$_{23}$V&$_{24}$Cr&$_{25}$Mn&$_{26}$Fe&$_{27}$Co&$_{28}$Ni\\
&$3d4s^2$&$3d^24s^2$&$3d^34s^2$&$3d^54s$&$3d^54s^2$&$3d^64s^2$&$3d^74s^2$&$3d^84s^2$\\
&$^2D_{3/2}$&$^3F_2$&$^4F_{3/2}$&$^7S_3$&$^6S_{5/2}$&$^5D_4$&$^4F_{9/2}$&$^3F_4$\\\hline
\multicolumn{9}{c}{Palladium group}\\
&$_{39}$Y&$_{40}$Zr&$_{41}$Nb&$_{42}$Mo&$_{43}$Tc&$_{44}$Ru&$_{45}$Rh&$_{46}$Pd\\
&$4d5s^2$&$4d^25s^2$&$4d^45s$&$4d^55s$&$4d^55s^2$&$4d^75s$&$4d^85s$&$4d^{10}$\\
&$^2D_{3/2}$&$^3F_2$&$^6D_{1/2}$&$^7S_3$&$^6S_{5/2}$&$^5F_5$&$^4F_{9/2}$&$^1S_0$\\\hline
\multicolumn{9}{c}{Platinum group}\\
&$_{57}$La&&&&&&&\\
&$5d6s^2$&&&&&&&\\
&$^2D_{3/2}$&&&&&&&\\
&$_{71}$Lu&$_{72}$Hf&$_{73}$Ta&$_{74}$W&$_{75}$Re&$_{76}$Os&$_{77}$Ir&$_{78}$Pt\\
&$5d6s^2$&$5d^26s^2$&$5d^36s^2$&$5d^46s^2$&$5d^56s^2$&$5d^66s^2$&$5d^76s^2$&$5d^96s$\\
&$^2D_{3/2}$&$^3F_2$&$^4F_{3/2}$&$^5D_0$&$^6S_{5/2}$&$^5D_4$&$^4F_{9/2}$&$^3D_3$
\end{tabular}
\end{table}

The same situation occurs while the $4f$ shell is filled along the series called the \emph{rare-earth elements} (Table~5)\footnote{Chemistry courses usually include Lu among the rare-earth elements as well. This is incorrect, however, since its $4f$ shell is already filled; Lu belongs to the platinum group, as in Table~4.}. Filling of the $4f$ shell is likewise not entirely regular, being characterized by competition among the $4f$, $5d$, and $6s$ states.

The last group of transition elements begins with actinium. Its $6d$ and $5f$ shells are filled in a manner analogous to the rare-earth series (Table~6).

\SourcePage{338}{341}
\begin{table}[ht]
\centering\scriptsize
\caption{Electron configurations of rare-earth atoms}
\begin{tabular}{ccccccc}
$_{58}$Ce&$_{59}$Pr&$_{60}$Nd&$_{61}$Pm&$_{62}$Sm&$_{63}$Eu\\
$4f5d6s^2$&$4f^36s^2$&$4f^46s^2$&$4f^56s^2$&$4f^66s^2$&$4f^76s^2$\\
$^1G_4$&$^4I_{9/2}$&$^5I_4$&$^6H_{5/2}$&$^7F_0$&$^8S_{7/2}$\\\hline
$_{64}$Gd&$_{65}$Tb&$_{66}$Dy&$_{67}$Ho&$_{68}$Er&$_{69}$Tm&$_{70}$Yb\\
$4f^75d6s^2$&$4f^96s^2$&$4f^{10}6s^2$&$4f^{11}6s^2$&$4f^{12}6s^2$&$4f^{13}6s^2$&$4f^{14}6s^2$\\
$^9D_2$&$^6H_{15/2}$&$^5I_8$&$^4I_{15/2}$&$^3H_6$&$^2F_{7/2}$&$^1S_0$
\end{tabular}
\end{table}

\begin{table}[ht]
\centering\scriptsize
\caption{Electron configurations of atoms in the actinide group}
\begin{tabular}{cccccccc}
$_{89}$Ac&$_{90}$Th&$_{91}$Pa&$_{92}$U&$_{93}$Np&$_{94}$Pu&$_{95}$Am&$_{96}$Cm\\
$6d7s^2$&$6d^27s^2$&$5f^26d7s^2$&$5f^36d7s^2$&$5f^46d7s^2$&$5f^67s^2$&$5f^77s^2$&$5f^76d7s^2$\\
$^2D_{3/2}$&$^3F_2$&$^4K_{11/2}$&$^5L_6$&$^6L_{11/2}$&$^7F_0$&$^8S_{7/2}$&$^9D_2$
\end{tabular}
\end{table}

We conclude this section with an interesting application of the Thomas--Fermi method. We have seen that $p$-shell electrons first occur in the fifth element (B), $d$ electrons at $Z=21$ (Sc), and $f$ electrons at $Z=58$ (Ce). These values of $Z$ can be predicted by the Thomas--Fermi method as follows.

An electron of orbital angular momentum $l$ in a complex atom moves with an ``effective potential energy''\footnote{Atomic units are used, as in \S~70.}
\[
 U_l(r)=-\varphi(r)+\frac{(l+1/2)^2}{2r^2}.
\]
The first term is the potential energy in the electric field described by the Thomas--Fermi potential $\varphi(r)$. The second is the centrifugal energy, in which $(l+1/2)^2$ is used instead of

\SourcePage{339}{342}
$l(l+1)$ because the motion is quasiclassical. Since the electron's total energy in the atom is negative, if $U_l(r)>0$ for all $r$ at given $Z$ and $l$, the atom cannot contain electrons with this $l$. If $l$ is fixed and $Z$ varied, then for sufficiently small $Z$ one indeed has $U_l(r)>0$ everywhere. As $Z$ increases, a point is reached at which the curve $U_l=U_l(r)$ touches the abscissa axis; at larger $Z$ there is a region in which $U_l(r)<0$. Thus electrons of a given $l$ first appear when the curve $U_l(r)$ is tangent to the axis, i.e. when
\[
 U_l(r)=-\varphi(r)+\frac{(l+1/2)^2}{2r^2}=0,
 \qquad U'_l(r)=-\varphi'(r)-\frac{(l+1/2)^2}{r^3}=0.
\]
Substitution of (70.6) for the potential gives
\begin{equation}
\begin{split}
 Z^{2/3}\frac{\chi(x)}x&=\left(\frac2{3\pi}\right)^{2/3}\frac{(l+1/2)^2}{2x^2},\\
 Z^{2/3}\frac{x\chi'(x)-\chi(x)}{x^2}&=-2\left(\frac2{3\pi}\right)^{2/3}\frac{(l+1/2)^2}{2x^3}.
\end{split}
\tag{73.2}
\end{equation}
Dividing the second equation term by term by the first gives
\[
 \frac{\chi'(x)}{\chi(x)}=-\frac1x,
\]
after which the first of (73.2) determines $Z$. Numerical calculation gives
\[
 Z=0.155(2l+1)^3.
\]
This formula determines the values of $Z$ at which electrons with a given $l$ first appear, with an error of about 10\%. The exact integer values result if 0.155 is replaced by 0.17:
\begin{equation}
 Z=0.17(2l+1)^3.
 \tag{73.3}
\end{equation}
For $l=1,2,3$, rounding to the nearest integers gives precisely 5, 21, and 58. For $l=4$, (73.3) gives $Z=124$; thus $g$ electrons would first appear only in element 124.
