\SourcePage{318}{321}
\section{The Thomas--Fermi equation}

Numerical calculations of the charge and field distributions in an atom by the self-consistent-field method are extremely laborious, especially for complex atoms. For complex atoms in particular, however, there is another approximate method whose merit lies in its simplicity, although it gives results considerably less accurate than those of the self-consistent-field method.

This method (E.~Fermi, L.~Thomas, 1927) rests on the fact that, in complex atoms containing many electrons, most of the electrons have comparatively large principal quantum numbers. The semiclassical approximation is applicable under these conditions. We may therefore use the notion of ``cells in phase space'' (\S~48) for the states of the individual atomic electrons.

The phase-space volume corresponding to electrons that have momenta less than $p$ and lie in a physical-space volume element $dV$ is $\frac43\pi p^3dV$.

\SourcePage{319}{322}
This volume contains $4\pi p^3dV/[3(2\pi)^3]$ cells\footnote{Atomic units are used in this section.}, that is, possible states which can contain at most
\[
 2\frac{4\pi p^3}{3(2\pi)^3}dV=\frac{p^3}{3\pi^2}dV
\]
electrons simultaneously (two electrons with mutually opposite spins in each cell). In the normal state of the atom, the electrons present in every volume element $dV$ must occupy, in phase space, the cells corresponding to momenta from zero up to some maximum value $p_0$. The kinetic energy of the electrons at every point then has the least possible value. If the number of electrons in $dV$ is written as $n\,dV$, where $n$ is the electron number density, the maximum momentum $p_0$ at each point is related to $n$ by
\[
 \frac{p_0^3}{3\pi^2}=n.
\]
Consequently, the maximum kinetic energy of an electron at a point where the electron density is $n$ is
\begin{equation}\frac{p_0^2}{2}=\frac12(3\pi^2n)^{2/3}.\tag{70.1}\end{equation}

Let $\varphi(r)$ be the electrostatic potential, taken to vanish at infinity. The total energy of an electron is $p^2/2-\varphi$. Clearly, every electron must have negative total energy; otherwise it would escape to infinity. Denote the maximum total electron energy at every point by $-\varphi_0$, where $\varphi_0$ is a positive constant (if it were not constant, electrons would pass from points having smaller $\varphi_0$ to points having larger $\varphi_0$).

We may thus write
\begin{equation}\frac{p_0^2}{2}=\varphi-\varphi_0.\tag{70.2}\end{equation}
Equating (70.1) and (70.2), we find
\begin{equation}n=\frac1{3\pi^2}[2(\varphi-\varphi_0)]^{3/2},\tag{70.3}\end{equation}
which relates the electron density to the potential at every point of the atom.

\SourcePage{320}{323}
When $\varphi=\varphi_0$, the density $n$ vanishes. Evidently, $n$ must also be set equal to zero throughout the region in which $\varphi<\varphi_0$, where (70.2) would give a negative maximum kinetic energy.

The equation $\varphi=\varphi_0$ therefore determines the boundary of the atom. Outside a centrally symmetric charge distribution of zero total charge, however, there is no field. Hence $\varphi=0$ at the boundary of a neutral atom, and for such an atom the constant $\varphi_0$ must be set equal to zero. For an ion, in contrast, $\varphi_0$ is nonzero.

We consider a neutral atom below and accordingly put $\varphi_0=0$. The electrostatic Poisson equation gives $\Delta\varphi=4\pi n$; substitution of (70.3) yields the fundamental \emph{Thomas--Fermi equation}
\begin{equation}\Delta\varphi=\frac{8\sqrt2}{3\pi}\varphi^{3/2}.\tag{70.4}\end{equation}
The field distribution in the normal state is determined by the centrally symmetric solution of this equation satisfying the following boundary conditions: as $r\to0$, the field must approach the Coulomb field of the nucleus, so that $\varphi r\to Z$; as $r\to\infty$, $\varphi r\to0$. Introduce a new variable $x$ in place of $r$ by
\begin{equation}r=xbZ^{-1/3},\qquad b=\frac12\left(\frac{3\pi}{4}\right)^{2/3}=0.885,\tag{70.5}\end{equation}
and a new unknown function $\chi$ in place of $\varphi$:
\begin{equation}\varphi(r)=\frac Zr\chi\left(\frac{rZ^{1/3}}b\right)=\frac{Z^{4/3}}b\frac{\chi(x)}x\footnote{In ordinary units: $\displaystyle\varphi(r)=\frac{Ze}{r}\chi\left(\frac{rZ^{1/3}me^2}{0.885\hbar^2}\right)$.}.\tag{70.6}\end{equation}
We then obtain
\begin{equation}x^{1/2}\frac{d^2\chi}{dx^2}=\chi^{3/2}\tag{70.7}\end{equation}
with boundary conditions $\chi=1$ at $x=0$ and $\chi=0$ at $x=\infty$. This equation contains no parameters, and hence defines a universal function $\chi(x)$. Table~2 gives this function as obtained by numerical integration of (70.7).

\SourcePage{321}{324}
The function $\chi(x)$ decreases monotonically and vanishes only at infinity\footnote{Equation (70.7) has the exact solution $\chi(x)=144x^{-3}$, which vanishes at infinity but does not satisfy the boundary condition at $x=0$. It could be used as an asymptotic expression for $\chi(x)$ at large $x$. It gives reasonably accurate values, however, only at very large $x$, whereas the Thomas--Fermi equation itself ceases to be applicable at large distances (see below).}. In other words, the Thomas--Fermi atom has no boundary and formally extends to infinity. The derivative at $x=0$ has the value $\chi'(0)=-1.59$. Thus, as $x\to0$, $\chi(x)\approx1-1.59x$, and the corresponding potential is
\begin{equation}\varphi(r)\approx Z/r-1.80\,Z^{4/3}.\tag{70.8}\end{equation}

\begin{table}[ht]
\centering
\caption{Values of the function $\chi(x)$}
\begin{tabular}{cc|cc|cc}
$x$&$\chi(x)$&$x$&$\chi(x)$&$x$&$\chi(x)$\\\hline
0.00&1.000&1.4&0.333&6&0.0594\\0.02&0.972&1.6&0.298&7&0.0461\\0.04&0.947&1.8&0.268&8&0.0366\\0.06&0.924&2.0&0.243&9&0.0296\\0.08&0.902&2.2&0.221&10&0.0243\\0.10&0.882&2.4&0.202&11&0.0202\\0.2&0.793&2.6&0.185&12&0.0171\\0.3&0.721&2.8&0.170&13&0.0145\\0.4&0.660&3.0&0.157&14&0.0125\\0.5&0.607&3.2&0.145&15&0.0108\\0.6&0.561&3.4&0.134&20&0.0058\\0.7&0.521&3.6&0.125&25&0.0035\\0.8&0.485&3.8&0.116&30&0.0023\\0.9&0.453&4.0&0.108&40&0.0011\\1.0&0.424&4.5&0.0919&50&0.00063\\1.2&0.374&5.0&0.0788&60&0.00039
\end{tabular}
\end{table}
The first term is the potential of the nuclear field, and the second is the potential produced by the electrons at the origin. Substitution of (70.6) into (70.3) gives the electron density in the form
\begin{equation}n=Z^2f\left(\frac{rZ^{1/3}}b\right),\qquad f(x)=\frac{32}{9\pi^3}\left(\frac\chi x\right)^{3/2}.\tag{70.9}\end{equation}
We see that in the Thomas--Fermi model the charge-density distributions in different atoms are similar, with

\SourcePageJoin{322}{325}
$Z^{-1/3}$ serving as the characteristic length parameter (in ordinary units, $\hbar^2/(me^2Z^{1/3})$, the Bohr radius divided by $Z^{1/3}$). If distances are measured in atomic units, then, in particular, the distances at which the electron density is greatest are the same for all $Z$. We may therefore say that most of the electrons in an atom of atomic number $Z$ lie at distances of order $Z^{-1/3}$ from the nucleus. Calculation shows that half the total electron charge is contained within a sphere of radius $1.33Z^{-1/3}$.

Similar reasoning shows that the mean electron speed in the atom (regarded in order of magnitude as the square root of the energy) is of order $Z^{2/3}$.

The Thomas--Fermi equation is inapplicable both at excessively small and at excessively large distances from the nucleus. At small $r$, its range of applicability is bounded by inequality (49.12); at still smaller distances in the Coulomb field of the nucleus, the semiclassical approximation fails. Putting $\alpha=Z$ in (49.12), we find $1/Z$ as the lower limiting distance. The semiclassical approximation also fails in a complex atom at large $r$. Indeed, it is readily seen that at $r\sim1$ the de Broglie wavelength of the electron becomes comparable with this distance itself, so the semiclassical condition is completely violated. This can be verified by estimating the terms in (70.2) and (70.4); the result is also evident beforehand, without calculation, since (70.4) contains no $Z$.

Thus the Thomas--Fermi equation is applicable only at distances large compared with $1/Z$ and small compared with 1. In a complex atom, however, most of the electrons are situated in this region.

The last circumstance means that the ``outer boundary'' of the atom in the Thomas--Fermi model is at $r\sim1$, so atomic dimensions are independent of $Z$. The energy of the outer electrons, and hence the ionization potential of the atom, is likewise independent of $Z$\footnote{This model does not, of course, represent the periodic dependence on $Z$ of atomic dimensions and ionization potentials which appears in the periodic system of the elements. Empirical data also show a slight systematic increase of atomic dimensions and decrease of ionization potentials as $Z$ increases.}.

The Thomas--Fermi method can be used to calculate the total ionization energy $E$, that is, the energy required to remove every electron from a neutral atom. For this purpose one must

\SourcePage{323}{326}
calculate the electrostatic energy of the Thomas--Fermi charge distribution in the atom. The required total energy equals half this electrostatic energy, since in a system of particles interacting by Coulomb's law, the mean kinetic energy is, by the virial theorem (see I, \S~10), minus one half the mean potential energy.

The dependence of $E$ on $Z$ can be found in advance from a simple argument. The electrostatic energy of $Z$ electrons in the field of a nucleus of charge $Z$, at a mean distance $Z^{-1/3}$ from it, is proportional to $Z\cdot Z/Z^{-1/3}=Z^{7/3}$. Numerical calculation gives $E=20.8Z^{7/3}$ eV. The dependence on $Z$ agrees well with experimental data; the empirical coefficient, however, is nearer 16.

We have already noted that nonzero positive values of $\varphi_0$ correspond to ionized atoms. If $\chi$ is defined by $\varphi-\varphi_0=Z\chi/r$, it obeys the same equation (70.7). We must now consider, however, solutions which vanish not at infinity, as for a neutral atom, but at finite values $x=x_0$; such solutions exist for every $x_0$. At $x=x_0$ the charge density vanishes together with $\chi$, while the potential remains finite.

The value of $x_0$ is related to the degree of ionization as follows. By Gauss's theorem, the total charge inside a sphere of radius $r$ is
\[
 -r^2\frac{\partial\varphi}{\partial r}=Z[\chi(x)-x\chi'(x)].
\]
The total ionic charge $z$ follows on setting $x=x_0$; since $\chi(x_0)=0$,
\begin{equation}z=-Zx_0\chi'(x_0).\tag{70.10}\end{equation}
In Fig.~23 the heavy curve shows $\chi=\chi(x)$ for a neutral atom, and the two curves below it correspond to ions with different degrees of ionization. Graphically, $z/Z$ is the length intercepted on the ordinate axis by the tangent to a curve at $x=x_0$.

Equation (70.7) also has solutions which nowhere vanish and which diverge at infinity. They may be regarded as corresponding to negative values of $\varphi_0$. Two such curves $\chi=\chi(x)$ are also shown in Fig.~23; they lie above the neutral-atom curve. At the point $x=x_1$ where
\begin{equation}\chi(x_1)-x_1\chi'(x_1)=0,\tag{70.11}\end{equation}

\SourcePage{324}{327}
the total charge within the sphere $x<x_1$ vanishes (graphically, this is evidently the point at which the tangent to the curve passes through the origin). If the curve is terminated at this point, it may be said to specify $\chi(x)$ for a neutral atom whose charge density remains nonzero at its boundary. Physically, this corresponds to a ``compressed'' atom confined within a prescribed finite volume\footnote{This treatment can be useful in studying the equation of state of matter at high compressions.}.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=.62cm,y=4.4cm]
 \draw[step=1cm,gray!55,very thin,xstep=1cm,ystep=.2] (0,0) grid (12,1);
 \draw[->] (0,0)--(12.25,0) node[below] {$x$}; \draw[->] (0,0)--(0,1.05) node[left] {$\chi$};
 \foreach \x in {1,...,12} \draw (\x,.012)--(\x,-.012) node[below] {\small\x};
 \foreach \y/\t in {.2/{0.2},.4/{0.4},.6/{0.6},.8/{0.8},1/{1.0}} \draw (.08,\y)--(-.08,\y) node[left] {\small\t};
 \draw[very thick,smooth] plot coordinates {(0,1)(.5,.62)(1,.43)(2,.24)(3,.16)(4,.108)(6,.059)(8,.037)(10,.024)(12,.017)};
 \draw[smooth] plot coordinates {(0,1)(.5,.58)(1,.37)(1.6,.16)(2.35,0)};
 \draw[smooth] plot coordinates {(0,1)(.5,.55)(1,.31)(1.5,.13)(2,0)};
 \draw[smooth] plot coordinates {(0,1)(.5,.67)(1,.48)(1.6,.40)(2.2,.39)(2.7,.43)};
 \draw[smooth] plot coordinates {(0,1)(.5,.64)(1,.43)(1.8,.27)(2.8,.23)(4.3,.29)};
 \draw[dashed] (0,.52)--(2.8,0); \draw[dashed] (0,.22)--(4.3,.12); \draw[dashed] (0,0)--(2.7,.43);
 \node at (2.35,.48) {$x_1$}; \node at (2.65,.06) {$x_0$};
\end{tikzpicture}
\caption*{Fig. 23}
\end{figure}

The Thomas--Fermi equation does not include exchange interaction between the electrons. The effects associated with it are of the next order in $Z^{-2/3}$. Consequently, including exchange interaction in the Thomas--Fermi method requires all effects of this order to be included at the same time\footnote{This was done by A.~S.~Kompaneets and E.~S.~Pavlovskii (ZhETF. 1956. Vol.~31. P.~427), and by D.~A.~Kirzhnits (ZhETF. 1957. Vol.~32. P.~115).}.

\ProblemHeading Find the relation between the electrostatic interaction energy of the electrons with one another and the energy of their interaction with the nucleus in a neutral atom in the Thomas--Fermi model.

\SolutionHeading The potential $\varphi_e$ of the field produced by the electrons is obtained by subtracting the nuclear-field potential $Z/r$ from the total potential $\varphi$. Hence the electron--electron interaction energy is
\[
 U_{ee}=-\frac12\int\varphi_en\,dV=\frac Z2\int\frac nr\,dV-\frac12\int\varphi n\,dV
 =\frac Z2\int\frac nr\,dV-\frac{(3\pi^2)^{2/3}}4\int n^{5/3}dV
\]

\SourcePage{325}{328}
(where $\varphi$ has been expressed in terms of $n$ by (70.3)). On the other hand, the interaction energy $U_{en}$ of the electrons with the nucleus and their kinetic energy $T$ are
\[
 U_{en}=-Z\int\frac nr\,dV,\qquad T=2\int\!\int_0^{p_0}\frac{p^2}{2}\frac{4\pi p^2dp\,dV}{(2\pi)^3}=\frac{3(3\pi^2)^{2/3}}{10}\int n^{5/3}dV.
\]
Comparison of these expressions with the preceding equality gives
\[
 U_{ee}=-\frac12U_{en}-\frac56T.
\]
At the same time, by the virial theorem (see I, \S~10), a system of particles interacting by Coulomb's law obeys $2T=-U=-U_{en}-U_{ee}$. It follows that
\[
 U_{ee}=-\frac17U_{en}.
\]
