\SourcePage{326}{329}
\section{Fine structure of atomic energy levels}

A systematic derivation of the formulas for relativistic effects in electron interactions belongs to another volume of this course (see IV, \S\S~33, 83). Here we give only a general account of these effects as applied to atomic terms.

The relativistic terms in the atomic Hamiltonian fall into two classes: some are linear in the electron spin operators, while others are quadratic in them. The former correspond, as it were, to an interaction between the electrons' orbital motion and their spins; this is called the \emph{spin--orbit interaction}. The latter describe the interaction between the electron spins (the \emph{spin--spin interaction}). Both interactions have the same, second, order in $v/c$, the ratio of the electron velocity to the velocity of light. In practice, however, in heavy atoms the spin--orbit interaction is considerably greater than the spin--spin interaction. The reason is that the spin--orbit interaction grows rapidly with atomic number, whereas the spin--spin interaction is, in the main, independent of $Z$ (see below).

\SourcePage{327}{330}
The spin--orbit interaction operator has the form
\begin{equation}
 \hat V_{sl}=\sum_a\hat{\mathbf s}_a\hat{\mathbf A}_a
 \tag{72.1}
\end{equation}
(the sum extends over all electrons in the atom), where $\hat{\mathbf s}_a$ are the electron spin operators and $\hat{\mathbf A}_a$ are certain ``orbital'' operators, i.e. operators acting on coordinate functions. In the self-consistent-field approximation the operators $\hat{\mathbf A}_a$ are proportional to the electron orbital-angular-momentum operators $\hat{\mathbf l}_a$, and $\hat V_{sl}$ can then be written
\begin{equation}
 \hat V_{sl}=\sum_a\alpha_a\hat{\mathbf s}_a\hat{\mathbf l}_a.
 \tag{72.2}
\end{equation}
The coefficients in this sum are expressed in terms of the potential energy $U(r)$ of an electron in the self-consistent field:
\begin{equation}
 \alpha_a=\frac{\hbar^2}{2m^2c^2r_a}\frac{dU(r_a)}{dr_a}.
 \tag{72.3}
\end{equation}
Since $|U(r)|$ decreases with distance from the nucleus, all the $\alpha_a$ are positive.

Treating (72.2) as a perturbation, we must average it over the unperturbed state to calculate the energy. The principal contribution to this energy comes from the region near the nucleus, at distances of the order of the Bohr radius ($\sim\hbar^2/Zme^2$) for a nucleus of charge $Ze$. In this region the nuclear field is practically unscreened, and
\[
 U\sim-\frac{Ze^2}{r}\sim-\frac{Z^2me^4}{\hbar^2},
\]
so that
\[
 \alpha\sim\frac{\hbar^2}{m^2c^2}\frac U{r^2}\sim Z^4\left(\frac{e^2}{\hbar c}\right)^2\frac{me^4}{\hbar^2}.
\]
The mean value of $\alpha$ follows by multiplying this expression by the probability $w$ of finding the electron near the nucleus. According to (71.3), $w\sim Z^{-2}$, and hence the spin--orbit interaction energy of an electron is ultimately
\[
 \alpha\sim\left(\frac{Ze^2}{\hbar c}\right)^2\frac{me^4}{\hbar^2},
\]

\SourcePage{328}{331}
i.e. it differs from the basic energy of an outer electron in the atom ($\sim me^4/\hbar^2$) only by the factor $(Ze^2/\hbar c)^2$. This factor grows rapidly with atomic number and becomes of order unity in heavy atoms.

The actual averaging of the perturbation operator (72.2) over the unperturbed states of an electron shell is carried out in two stages. First we average over the electronic state of an atom having specified magnitudes $L$ and $S$ of its total orbital angular momentum and spin, but not over their directions. After this averaging, $\hat V_{sl}$ is still an operator, but one that must now be expressible solely in terms of operators for quantities characterizing the atom as a whole, rather than its individual electrons. These are $\hat{\mathbf L}$ and $\hat{\mathbf S}$.

Denote the spin--orbit interaction operator averaged in this way by $\hat V_{SL}$. Since it is linear in $\hat{\mathbf S}$, it has the form
\begin{equation}
 \hat V_{SL}=A\hat{\mathbf S}\hat{\mathbf L},
 \tag{72.4}
\end{equation}
where $A$ is a constant characteristic of the given unsplit term: it depends on $S$ and $L$, but not on the atom's total angular momentum $J$\footnote{To clarify the meaning of this operation, recall that averaging in quantum mechanics generally means taking the corresponding diagonal matrix element. Partial averaging consists in forming the set of matrix elements diagonal in only some of the quantum numbers specifying the system's state. Thus, here averaging (72.2) means forming the matrix with elements $(nM'_LM'_S|\hat V_{sl}|nM_LM_S)$ for all possible $M_L$, $M'_L$, $M_S$, $M'_S$, diagonal in all the remaining quantum numbers (collectively denoted by $n$). Correspondingly, $\hat{\mathbf S}$ and $\hat{\mathbf L}$ are to be understood as the matrices $(M'_S|\mathbf S|M_S)$ and $(M'_L|\mathbf L|M_L)$, whose elements are given by (27.13). We shall repeatedly use this procedure of averaging in stages below.}.

To calculate the splitting of the degenerate level we must now solve the secular equation formed from the matrix elements of (72.4). In this case, however, we already know the proper zeroth-order functions in which the matrix of $\hat V_{SL}$ is diagonal. They are the wave functions of states with specified total angular momentum $J$. Averaging over such a state means replacing $\hat{\mathbf L}\hat{\mathbf S}$ by its eigenvalue, which according to (31.3) is
\[
 LS=\frac12[J(J+1)-L(L+1)-S(S+1)].
\]

\SourcePage{329}{332}
Since all components of a multiplet have the same $L$ and $S$, and only their relative positions are of interest, the splitting energy may be written
\begin{equation}
 \frac12AJ(J+1).
 \tag{72.5}
\end{equation}
The intervals between adjacent components, characterized by $J$ and $J-1$, are consequently
\begin{equation}
 \Delta E_{J,J-1}=AJ.
 \tag{72.6}
\end{equation}
This is the so-called \emph{Land\'e interval rule} (A.~Land\'e, 1923).

The constant $A$ may be either positive or negative. For $A>0$, the lowest component of the multiplet level has the smallest possible $J$, namely $J=|L-S|$; such multiplets are called \emph{normal}. If $A<0$, the lowest level has $J=L+S$ (an \emph{inverted multiplet}).

The sign of $A$ for normal atomic states is readily found when the electron configuration contains only one incompletely filled shell. If this shell is no more than half filled, Hund's rule (\S~67) says that all its $n$ electrons have parallel spins, so that the total spin has its greatest possible value, $S=n/2$. Substituting $\mathbf s_a=\mathbf S/n$ in (72.2) and taking $\alpha_a$, the same for all electrons in one shell, outside the sum gives
\[
 \hat V_{SL}=\frac{\alpha}{2S}\hat{\mathbf S}\hat{\mathbf L},
\]
and hence $A=\alpha/2S>0$. If the shell is more than half filled, first add and subtract in (72.2) the analogous sum over the vacant places---the holes in the incomplete shell. Since a completely filled shell would have $\hat V_{sl}=0$, the operator becomes $\hat V_{sl}=-\sum_a\alpha_a\hat{\mathbf l}_a\hat{\mathbf s}_a$, summed only over the holes, while the atom's total spin and orbital angular momentum are $\mathbf S=-\sum_a\mathbf s_a$, $\mathbf L=-\sum_a\mathbf l_a$. The same argument then gives $A=-\alpha/2S<0$.

It follows that a simple rule determines $J$ in the normal state of an atom with one incompletely filled shell. If that shell contains no more than half the maximum possible number of electrons, then $J=|L-S|$. If it is more than half filled, then $J=L+S$.

\SourcePage{330}{333}
As already mentioned, unlike the spin--orbit interaction, the spin--spin interaction is in the main independent of $Z$. This is clear from its nature as a direct interaction of electrons with one another, unrelated to the nuclear field.

By analogy with (72.4), the averaged spin--spin interaction operator must be quadratic in $\hat{\mathbf S}$. The quadratic expressions are $S^2$ and $(\hat{\mathbf L}\hat{\mathbf S})^2$. The eigenvalues of the first are independent of $J$, so it does not split the term. It may therefore be omitted, and we write
\begin{equation}
 \hat V_{ss}=B(\hat{\mathbf S}\hat{\mathbf L})^2,
 \tag{72.7}
\end{equation}
where $B$ is a constant. The eigenvalues of this operator contain terms independent of $J$, terms proportional to $J(J+1)$, and finally a term proportional to $J^2(J+1)^2$. The first do not cause splitting and are immaterial; the second may be absorbed into (72.5), which merely changes the constant $A$. The last give the following contribution to the term energy:
\begin{equation}
 \frac14BJ^2(J+1)^2.
 \tag{72.8}
\end{equation}

The construction of atomic levels described in \S\S~66, 67 rests on the assumption that the electron orbital angular momenta combine to form the atom's total orbital angular momentum $L$, while their spins combine to form its total spin $S$. As noted above, this description is possible only when relativistic effects are small; more precisely, the fine-structure intervals must be small compared with the separations between levels having different $L$ and $S$. This approximation is called the \emph{Russell--Saunders case} (H.~Russell, F.~Saunders, 1925), or $LS$ coupling.

Its actual range of applicability is limited. The levels of light atoms follow $LS$ coupling; as atomic number increases, relativistic interactions in the atom become stronger and the Russell--Saunders approximation ceases to apply\footnote{Although the quantitative formulas describing this coupling cease to apply, classification of the levels by this scheme may still be meaningful for heavier atoms, especially for the lowest states, including the normal state.}. In particular, the approximation also fails for highly excited levels in which one electron has a large $n$ and is therefore mainly at large distances from the nucleus (\S~68). The electrostatic interaction of this electron with the motion of the others is relatively weak, whereas the relativistic interaction in the ionic core does not decrease.

\SourcePage{331}{334}
At the opposite limiting case the relativistic interaction is large compared with the electrostatic interaction---more precisely, with the part responsible for the dependence of the energy on $L$ and $S$. Orbital angular momentum and spin can then no longer be considered separately, since neither is conserved. Individual electrons are characterized by their total angular momenta $j$, which combine to form the atom's total angular momentum $J$. This scheme is called $jj$ coupling. In fact, pure $jj$ coupling is not encountered; the levels of very heavy atoms exhibit various forms intermediate between $LS$ and $jj$ coupling\footnote{For further details on coupling types and the quantitative theory, see E.~Condon and G.~Shortley, \emph{The Theory of Atomic Spectra}.}.

A distinctive coupling occurs in some highly excited states. The ionic core may then be in a Russell--Saunders state, characterized by $L$ and $S$, while its coupling to the highly excited electron is of the $jj$ type (again because the electrostatic interaction is weak for this electron).

The fine structure of hydrogen-atom energy levels has certain special features; it is calculated in another volume of this course (see IV, \S~34). Here we note only that, for a given principal quantum number $n$, the energy depends solely on the electron's total angular momentum $j$. Thus the degeneracy is not completely removed: a level with given $n$ and $j$ corresponds to two states with orbital angular momenta $l=j\pm1/2$ (unless $j$ has its greatest possible value for the given $n$, $j=n-1/2$). Thus the $n=3$ level splits into three levels: one corresponds to the states $s_{1/2}$ and $p_{1/2}$, another to $p_{3/2}$ and $d_{3/2}$, and the third to $d_{5/2}$.
