\SourcePage{694}{697}
\section{Elastic collisions of fast electrons with atoms}

Elastic collisions of fast electrons with atoms can be treated in the Born
approximation if the speed of the incident electron is large compared with
the speeds of the atomic electrons.

Because the mass of an atom is much greater than that of an electron, the
atom may be regarded as stationary during the collision. The coordinate
system in which the center of mass is at rest then coincides with that in
which the atom is at rest. Thus $\mathbf p$ and $\mathbf p'$ in (126.7) are
the electron momenta before and after the collision, $m$ is the electron mass,
and $\theta$ coincides with the electron deflection angle $\vartheta$. The
potential energy $U(r)$ in (126.7), however, requires a proper definition.

In \S~126 we calculated the matrix element $U_{\mathbf p'\mathbf p}$ of the
interaction energy with respect to the free-particle wave functions before
and after the collision. For a collision with an atom, the wave functions
describing the atom's internal state must also be included. In elastic
scattering the atomic state does not change. Hence
$U_{\mathbf p'\mathbf p}$ must be defined as a matrix element with respect to
the electron wave functions $\psi_{\mathbf p}$ and $\psi_{\mathbf p'}$ which
is diagonal with respect to the atomic wave function. In other words, $U$ in
(126.7) is the potential energy of interaction between the electron and atom,
averaged over the latter's wave function. It equals $e\varphi(r)$, where
$\varphi(r)$ is the potential of the field produced at $r$ by the mean charge
distribution in the atom.

Denoting the atomic charge density by $\rho(r)$, the potential $\varphi$
satisfies Poisson's equation
\[
 \Delta\varphi=-4\pi\rho(r).
\]

\SourcePage{695}{698}
The required matrix element $U_{\mathbf p'\mathbf p}$ is essentially the
Fourier component of $U$, and hence of $\varphi$, corresponding to the wave
vector $\mathbf q=\mathbf k'-\mathbf k$. Applying Poisson's equation to each
Fourier component separately gives
\[
 \Delta(\varphi_{\mathbf q}e^{i\mathbf{qr}})=-q^2\varphi_{\mathbf q}e^{i\mathbf{qr}}
 =-4\pi\rho_{\mathbf q}e^{i\mathbf{qr}},
\]
so that $\varphi_{\mathbf q}=4\pi\rho_{\mathbf q}/q^2$, or
\begin{equation}
 \int\varphi e^{-i\mathbf{qr}}dV=\frac{4\pi}{q^2}\int\rho e^{-i\mathbf{qr}}dV.
 \tag{139.1}
\end{equation}
The charge density $\rho(r)$ consists of the electronic charge and the nuclear
charge:
\[
 \rho=-en(r)+Ze\delta(r),
\]
where $en(r)$ is the electronic charge density in the atom. Multiplication by
$e^{-i\mathbf{qr}}$ and integration give
\[
 \int\rho e^{-i\mathbf{qr}}dV=-e\int ne^{-i\mathbf{qr}}dV+Ze.
\]
The integral of interest is therefore
\begin{equation}
 \int Ue^{-i\mathbf{qr}}dV=\frac{4\pi e^2}{q^2}[Z-F(q)],
 \tag{139.2}
\end{equation}
where
\begin{equation}
 F(q)=\int ne^{-i\mathbf{qr}}dV
 \tag{139.3}
\end{equation}
is called the \emph{atomic form factor}. It is a function of the scattering
angle and of the speed of the incident electron.

Finally, substituting (139.2) in (126.7), we obtain the following expression
for the elastic scattering cross-section of fast electrons by an
atom\footnote{We neglect exchange effects between the scattered fast electron
and the atomic electrons; that is, we do not symmetrize the wave function of
the system. The validity of this neglect is evident beforehand: interference
between the rapidly oscillating wave function of the free particle and the
wave function of the atomic electrons in the ``exchange integral'' makes the
associated contribution to the scattering amplitude small.}:
\begin{equation}
 d\sigma=\frac{4m^2e^4}{\hbar^4q^4}[Z-F(q)]^2d\Omega,\qquad
 q=\frac{2mv}{\hbar}\sin\frac\vartheta2.
 \tag{139.4}
\end{equation}
Consider the limiting case $qa_0\ll1$, where $a_0$ is of the order of the
atomic size. Small $q$ correspond to small

\SourcePage{696}{699}
scattering angles: $\vartheta\ll v_0/v$, where
$v_0\sim\hbar/ma_0$ is of the order of the atomic-electron speeds.

Expand $F(q)$ in powers of $q$. The zeroth-order term is $\int n\,dV$, the
total number $Z$ of electrons in the atom. The first-order term is
proportional to $\int\mathbf r n(r)dV$, the mean dipole moment of the atom,
which vanishes identically (see \S~75). The expansion must therefore be
carried through the second-order term, giving
\[
 Z-F(q)=\frac{q^2}{6}\int nr^2dV.
\]
Substitution in (139.4) gives
\begin{equation}
 d\sigma=\left|\frac{me^2}{3\hbar^2}\int nr^2dV\right|^2d\Omega.
 \tag{139.5}
\end{equation}
Thus, in the small-angle region the cross-section is independent of the
scattering angle and is determined by the mean square distance of the atomic
electrons from the nucleus.

In the opposite limiting case of large $q$ ($qa_0\gg1$,
$\vartheta\gg v_0/v$), the factor $e^{-i\mathbf{qr}}$ in the integrand of
(139.3) oscillates rapidly, and the integral is consequently close to zero.
Thus $F(q)$ may be neglected in comparison with $Z$, leaving
\begin{equation}
 d\sigma=\left(\frac{Ze^2}{2mv^2}\right)^2
 \frac{d\Omega}{\sin^4(\vartheta/2)},
 \tag{139.6}
\end{equation}
the Rutherford cross-section for scattering by the atomic nucleus.

Let us also calculate the transport cross-section
\begin{equation}
 \sigma_{\mathrm{tr}}=\int(1-\cos\vartheta)d\sigma.
 \tag{139.7}
\end{equation}
In the region $\vartheta\ll v_0/v$, (139.5) gives
\[
 d\sigma=\mathrm{const}\cdot\sin\vartheta\,d\vartheta
 \approx\mathrm{const}\cdot\vartheta\,d\vartheta,
\]
where the constant is independent of $\vartheta$. The integrand in this region
is therefore proportional to $\vartheta^3d\vartheta$, and the integral
converges rapidly at its lower limit. In the region
$1\gg\vartheta\gg v_0/v$, however,
\[
 d\sigma\approx\mathrm{const}\cdot(d\vartheta/\vartheta^3),
\]
so the integrand is proportional to $d\vartheta/\vartheta$ and (139.7)
diverges logarithmically.

\SourcePage{697}{700}
It follows that this very angular region makes the principal contribution to
the integral, and integration may be restricted to it. The lower limit should
be of order $v_0/v$; write it as $e^2/(\gamma\hbar v)$, where $\gamma$ is a
dimensionless constant. The result is
\begin{equation}
 \sigma_{\mathrm{tr}}=4\pi\left(\frac{Ze^2}{mv^2}\right)^2
 \ln\frac{\gamma\hbar v}{e^2}.
 \tag{139.8}
\end{equation}
An exact calculation of $\gamma$ requires consideration of scattering through
angles $\vartheta>v_0/v$ and cannot be performed in general. The transport
cross-section depends only weakly on this constant, since it occurs under a
logarithm multiplied by the large quantity $\hbar v/e^2$.

For numerical calculation of the form factor of heavy atoms, the
Thomas--Fermi distribution of the density $n(r)$ may be used. We saw that in
the Thomas--Fermi model
\[
 n(r)=Z^2f(rZ^{1/3}/b)
\]
(all quantities in this and the following formulas are measured in atomic
units). It is readily seen that (139.3), calculated with this $n(r)$, contains
$q$ only in a specified combination with $Z$:
\begin{equation}
 F(q)=Z\varphi(bqZ^{-1/3}).
 \tag{139.9}
\end{equation}
Table~11 gives the values of the function $\varphi(x)$, which is universal for
all atoms\footnote{It must be kept in mind that this formula is inapplicable
at small $q$, consistently with the fact that the integral of $nr^2$ cannot
actually be calculated by the Thomas--Fermi method (see the note on p.~563).
It should also be remembered that the Thomas--Fermi model does not reproduce
the individual properties of atoms that disturb their systematic variation
with atomic number.}.

\setcounter{table}{10}
\begin{table}[htbp]
\centering
\caption{Thomas--Fermi atomic factor}
\begin{tabular}{@{}rr|rr|rr@{}}
\toprule
$x$ & $\varphi(x)$ & $x$ & $\varphi(x)$ & $x$ & $\varphi(x)$\\
\midrule
0    & 1.000 & 1.08 & 0.422 & 2.17 & 0.224\\
0.15 & 0.922 & 1.24 & 0.378 & 2.32 & 0.205\\
0.31 & 0.796 & 1.39 & 0.342 & 2.48 & 0.189\\
0.46 & 0.684 & 1.55 & 0.309 & 2.64 & 0.175\\
0.62 & 0.589 & 1.70 & 0.284 & 2.79 & 0.167\\
0.77 & 0.522 & 1.86 & 0.264 & 2.94 & 0.156\\
0.93 & 0.469 & 2.02 & 0.240 &      &      \\
\bottomrule
\end{tabular}
\end{table}

With the atomic form factor (139.9), (139.4) becomes
\begin{equation}
 d\sigma=\frac{4Z^2}{q^4}[1-\varphi(bqZ^{-1/3})]^2d\Omega
 =Z^{2/3}\Phi\left(Z^{-1/3}v\sin\frac\vartheta2\right)d\Omega,
 \tag{139.10}
\end{equation}

\SourcePage{698}{701}
where $\Phi(x)$ is another universal function. The total cross-section can be
obtained by integration. Small $\vartheta$ make the main contribution to the
integral. We may therefore write
\[
 d\sigma\approx Z^{2/3}\Phi(Z^{-1/3}v\vartheta/2)\cdot2\pi\vartheta\,d\vartheta
\]
and extend the integration over $d\vartheta$ to infinity:
\[
 \sigma=2\pi Z^{2/3}\int_0^\infty\Phi\left(Z^{-1/3}\frac{v\vartheta}{2}\right)
 \vartheta\,d\vartheta
 =\frac{8\pi}{v^2}Z^{4/3}\int_0^\infty x\Phi(x)dx.
\]
Thus
\begin{equation}
 \sigma=\mathrm{const}\cdot\frac{Z^{4/3}}{v^2}.
 \tag{139.11}
\end{equation}
It is similarly easy to verify that the constant $\gamma$ in (139.8) is
proportional to $Z^{-1/3}$.

\subsection*{Problem}

\ProblemHeading Calculate the elastic scattering cross-section of fast
electrons by a hydrogen atom in its ground state.

\SolutionHeading The ground-state wave function of the hydrogen atom is
$\psi=e^{-r}/\sqrt\pi$ in atomic units, so $n=e^{-2r}/\pi$. The angular
integration in (139.3), performed as in the derivation of (126.12), gives
\[
 F=\frac{4\pi}{q}\int_0^\infty n(r)\sin qr\,dr
 =\left(1+\frac{q^2}{4}\right)^{-2}.
\]
Substitution in (139.4) gives
\[
 d\sigma=\frac{4(8+q^2)^2}{(4+q^2)^4}d\Omega,
\]
where $q=2v\sin(\vartheta/2)$. To calculate the total cross-section, write
\[
 d\Omega=2\pi\sin\vartheta\,d\vartheta=\frac{2\pi}{v^2}q\,dq
\]
and integrate with respect to $q$ from zero to $2v$. Since $v$ is assumed
large and the integral converges, however, the upper limit may be replaced by
infinity. The result is
\[
 \sigma=\frac{7\pi}{3v^2}.
\]
The transport cross-section is calculated as
\[
 \sigma_{\mathrm{tr}}=\frac1{2v^2}\int q^2d\sigma.
\]
Changing the integration variable by $4+q^2=u$ and replacing the upper limit
by infinity everywhere except in the term $du/u$, we obtain
\[
 \sigma_{\mathrm{tr}}=\frac{4\pi}{v^2}\left(\ln v+\frac1{12}\right),
\]
in agreement with (139.8).
