\SourcePage{677}{680}
\section{The Rutherford formula}

Scattering in a Coulomb field is important in physical applications. It is
also of interest because in this case the quantum-mechanical collision problem
can be solved exactly and completely.

When a preferred direction is present (here the direction of the incident
particle), it is convenient to solve the Schrödinger equation in a Coulomb
field in parabolic coordinates $\xi$, $\eta$, $\varphi$ (see \S~37). The
problem of particle scattering in a central field

\SourcePage{678}{681}
has axial symmetry. The wave function $\psi$ is therefore independent of
$\varphi$. Write a particular solution of the Schrödinger equation (37.6) as
\begin{equation}
 \psi=f_1(\xi)f_2(\eta)
 \tag{135.1}
\end{equation}
((37.7) with $m=0$). Separation of variables then gives equations (37.8) with
$m=0$\footnote{Coulomb units are used in this section (see p.~154).}:
\begin{align}
 \frac{d}{d\xi}\left(\xi\frac{df_1}{d\xi}\right)
 +\left[\frac{k^2}{4}\xi-\beta_1\right]f_1&=0, \notag\\
 \frac{d}{d\eta}\left(\eta\frac{df_2}{d\eta}\right)
 +\left[\frac{k^2}{4}\eta-\beta_2\right]f_2&=0, \notag\\
 \beta_1+\beta_2&=1.
 \tag{135.2}
\end{align}
The energy of the scattered particle is, of course, positive; we have put
$E=k^2/2$. The signs in (135.2) correspond to a repulsive field. Exactly the
same final result is obtained for the scattering cross-section in an
attractive field.

We must find a solution of the Schrödinger equation which, for negative $z$
and large $r$, has the form of a plane wave:
\[
 \psi\sim e^{ikz}\quad\text{for}\quad-\infty<z<0,\quad r\to\infty,
\]
corresponding to a particle incident in the positive $z$ direction. We shall
see below that this condition can be satisfied by a single particular
integral (135.1), rather than a sum of integrals with different values of
$\beta_1$, $\beta_2$.

In parabolic coordinates the condition is
\[
 \psi\sim\exp\left[\frac{ik}{2}(\xi-\eta)\right]
 \quad\text{for}\quad\eta\to\infty\quad\text{and every }\xi.
\]
It can be satisfied only if
\begin{equation}
 f_1(\xi)=e^{ik\xi/2},
 \tag{135.3}
\end{equation}
while $f_2(\eta)$ obeys
\begin{equation}
 f_2(\eta)\sim e^{-ik\eta/2}\quad\text{as}\quad\eta\to\infty.
 \tag{135.4}
\end{equation}
Substitution of (135.3) in the first of (135.2) confirms that this function
does satisfy the equation if $\beta_1=ik/2$. With
$\beta_2=1-\beta_1$, the second of (135.2)

\SourcePage{679}{682}
then becomes
\[
 \frac{d}{d\eta}\left(\eta\frac{df_2}{d\eta}\right)
 +\left(\frac{k^2}{4}\eta-1+\frac{ik}{2}\right)f_2=0.
\]
Seek its solution in the form
\begin{equation}
 f_2(\eta)=e^{-ik\eta/2}w(\eta),
 \tag{135.5}
\end{equation}
where $w(\eta)$ tends to a constant limit as $\eta\to\infty$. The equation
for $w(\eta)$ is
\begin{equation}
 \eta w''+(1-ik\eta)w'-w=0,
 \tag{135.6}
\end{equation}
which is reduced, by introducing $\eta_1=ik\eta$, to the confluent
hypergeometric equation with parameters $\alpha=-i/k$, $\gamma=1$. We must
choose the solution of (135.6) which, when multiplied by $f_1(\xi)$, contains
only an outgoing (scattered), and no incoming, spherical wave. This solution
is
\[
 w=\mathrm{const}\cdot F\left(-\frac{i}{k},1,ik\eta\right).
\]

Collecting these expressions, we find the following exact solution of the
Schrödinger equation describing the scattering:
\begin{equation}
 \psi=\exp\left(-\frac{\pi}{2k}\right)\Gamma\left(1+\frac{i}{k}\right)
 \exp\left[\frac{ik}{2}(\xi-\eta)\right]
 F\left(-\frac{i}{k},1,ik\eta\right).
 \tag{135.7}
\end{equation}
The normalization constant in $\psi$ has been chosen so that the incident
plane wave has unit amplitude (see below).

To separate the incident and scattered waves in this function, consider its
form at large distances from the center. Using the first two terms of the
asymptotic expansion (d.14) of the confluent hypergeometric function, we find
at large $\eta$
\begin{align*}
 F\left(-\frac{i}{k},1,ik\eta\right)
 &\approx\frac{(-ik\eta)^{i/k}}{\Gamma(1+i/k)}
 \left(1+\frac{1}{ik^3\eta}\right)\\
 &\quad+\frac{(ik\eta)^{-i/k}e^{ik\eta}}{\Gamma(-i/k)ik\eta}\\
 &=\frac{e^{\pi/2k}}{\Gamma(1+i/k)}
 \left(1+\frac{1}{ik^3\eta}\right)
 \exp\left(\frac{i}{k}\ln k\eta\right)\\
 &\quad-\frac{(i/k)e^{\pi/2k}}{\Gamma(1-i/k)}
 \frac{e^{ik\eta}}{ik\eta}
 \exp\left(-\frac{i}{k}\ln k\eta\right).
\end{align*}
Substitution in (135.7), followed by passage to spherical coordinates
($\xi-\eta=2z$, $\eta=r-z=r(1-\cos\theta)$), gives the following

\SourcePage{680}{683}
final asymptotic expression for the wave function:
\begin{align}
 \psi={}&\left[1+\frac{1}{ik^3r(1-\cos\theta)}\right]
 \exp\left\{ikz+\frac{i}{k}\ln[kr(1-\cos\theta)]\right\}\notag\\
 &+\frac{f(\theta)}{r}\exp\left\{ikr-\frac{i}{k}\ln(2kr)\right\},
 \tag{135.8}
\end{align}
where
\begin{equation}
 f(\theta)=-\frac{1}{2k^2\sin^2(\theta/2)}
 \frac{\Gamma(1+i/k)}{\Gamma(1-i/k)}
 \exp\left(-\frac{2i}{k}\ln\sin\frac{\theta}{2}\right).
 \tag{135.9}
\end{equation}

The first term in (135.8) is the incident wave. Because the Coulomb field
decreases slowly, the incident plane wave is distorted even at large
distances from the center, as shown by the logarithmic term in the phase and
the term of order $1/r$ in the wave amplitude\footnote{The origin of this
distortion is already clear from the classical picture. Consider a family of
classical Coulomb hyperbolic trajectories with the same incident direction,
parallel to the $z$ axis. At large distances from the scattering center
($z\to-\infty$), the equation of a surface normal to them tends, as is readily
shown, not to $z=\mathrm{const}$ but to
$z+k^{-2}\ln k(r-z)=\mathrm{const}$. This surface is precisely a surface of
constant phase of the incident wave in (135.8).}. The distorting logarithmic
term in the phase is also present in the scattered spherical wave represented
by the second term in (135.8). These departures from the usual asymptotic
form (123.3) are nevertheless immaterial, since they give corrections to the
current density that tend to zero as $r\to\infty$.

Thus, from $d\sigma=|f(\theta)|^2d\Omega$, the scattering cross-section is
\[
 d\sigma=\frac{d\Omega}{4k^2\sin^4(\theta/2)},
\]
or, in ordinary units,
\begin{equation}
 d\sigma=\left(\frac{\alpha}{2mv^2}\right)^2
 \frac{d\Omega}{\sin^4(\theta/2)}
 \tag{135.10}
\end{equation}
($v=k\hbar/m$ is the particle speed). This is the familiar
\emph{Rutherford formula} obtained in classical mechanics. Thus quantum and
classical mechanics give the same result for scattering in a Coulomb field
(N.~Mott, W.~Gordon, 1928). Naturally, the Born formula (126.12) also gives
the same expression (135.10).

\SourcePage{681}{684}
For reference, we give the scattering amplitude (135.9) as an expansion in
spherical functions. It is obtained by substituting the phases (36.28) in
(124.5)\footnote{The quantity $\delta_l^{\mathrm C}$ in that formula differs
from the true (outgoing-wave) Coulomb phase by an amount independent of $l$.}:
\begin{equation}
 \exp(2i\delta_l^{\mathrm C})=\frac{\Gamma(l+1+i/k)}{\Gamma(l+1-i/k)}.
 \tag{135.11}
\end{equation}
Thus
\begin{equation}
 f(\theta)=\frac{1}{2ik}\sum_l(2l+1)
 \frac{\Gamma(l+1+i/k)}{\Gamma(l+1-i/k)}P_l(\cos\theta).
 \tag{135.12}
\end{equation}

The signs in (135.9) correspond to a repulsive Coulomb field. For an
attractive field, (135.9) must be replaced by its complex conjugate. In the
latter case $f(\theta)$ becomes infinite at the poles of
$\Gamma(1-i/k)$, that is, where the argument of the $\Gamma$-function is a
negative integer or zero (then $\operatorname{Im}k>0$ and $r\psi$ decays at
infinity). The corresponding energy values,
\[
 \frac{k^2}{2}=-\frac{1}{2n^2},\qquad n=1,2,3,\ldots,
\]
coincide with the discrete energy levels in the attractive Coulomb field
(compare \S~128).
