\SourcePage{681}{684}
\section{A system of continuous-spectrum wave functions}

In studying motion in a centrally symmetric field in Chapter V, we considered
stationary states in which the particle has specified values not only of the
energy but also of the orbital angular momentum $l$ and its projection $m$.
The wave functions of these states in the discrete spectrum ($\psi_{nlm}$)
and continuous spectrum ($\psi_{klm}$, with energy $\hbar^2k^2/2m$) together
form a complete system in which the wave function of an arbitrary state may
be expanded. Such a system is not, however, suited to the formulation of
problems in scattering theory. A different system is convenient here, one in
which the continuous-spectrum wave functions are characterized by specified
asymptotic behavior: at infinity there is a plane wave
$\exp(i\mathbf{kr})$ together with an outgoing spherical wave. In these
states the particle has a specified energy, but neither its angular momentum
nor its projection is specified.

\SourcePage{682}{685}
According to (123.6), (123.7), these wave functions, denoted here by
$\psi_{\mathbf k}^{(+)}$, are
\begin{equation}
 \psi_{\mathbf k}^{(+)}=\frac1{2k}\sum_{l=0}^{\infty}i^l(2l+1)e^{i\delta_l}R_{kl}(r)
 P_l\left(\frac{\mathbf{kr}}{kr}\right).
 \tag{136.1}
\end{equation}
The argument of the Legendre polynomials has been written as
$\cos\theta=\mathbf{kr}/kr$, so this expression is no longer tied to any
particular choice of coordinate axes (as it was in (123.6), where the $z$
axis coincided with the propagation direction of the plane wave). Giving
$\mathbf k$ all possible values produces a set of wave functions which, as
we shall now show, are mutually orthogonal and normalized by the usual
continuous-spectrum rule
\begin{equation}
 \int\psi_{\mathbf k'}^{(+)*}\psi_{\mathbf k}^{(+)}dV=(2\pi)^3\delta(\mathbf k'-\mathbf k).
 \tag{136.2}
\end{equation}

For the proof\footnote{Only the mutual orthogonality of the functions
$\psi_{\mathbf k}^{(+)}$ essentially requires a separate proof. Their
normalization could be established directly from their asymptotic form
(compare \S~21). In this sense, (136.2) is already evident from the fact that,
as $r\to\infty$, the only nondecreasing term in these functions is
$\psi_{\mathbf k}^{(+)}\approx e^{i\mathbf{kr}}$.}, note that the product
$\psi_{\mathbf k'}^{(+)*}\psi_{\mathbf k}^{(+)}$ is a double sum over $l$ and
$l'$ of terms containing the products
\[
 P_l\left(\frac{\mathbf{kr}}{kr}\right)P_{l'}\left(\frac{\mathbf{k'r}}{k'r}\right).
\]
Integration over the directions of $\mathbf r$ is carried out by
\begin{equation}
 \int P_l\left(\frac{\mathbf{kr}}{kr}\right)P_{l'}\left(\frac{\mathbf{k'r}}{k'r}\right)d\Omega
 =\delta_{ll'}\frac{4\pi}{2l+1}P_l\left(\frac{\mathbf{kk'}}{kk'}\right)
 \tag{136.3}
\end{equation}
(compare formula (c.12) in the Mathematical Supplements), after which there
remains
\begin{align*}
 \int\psi_{\mathbf k'}^{(+)*}\psi_{\mathbf k}^{(+)}dV
 ={}&\frac{\pi}{kk'}\sum_{l=0}^{\infty}(2l+1)
 \exp[i\delta_l(k)-i\delta_l(k')]P_l(\cos\gamma)\\
 &\times\int_0^\infty R_{k'l}R_{kl}r^2dr,
\end{align*}

\SourcePage{683}{686}
where $\gamma$ is the angle between $\mathbf k$ and $\mathbf k'$. The radial
functions $R_{kl}$ are orthogonal and normalized according to
\[
 \int_0^\infty R_{k'l}R_{kl}r^2dr=2\pi\delta(k'-k).
\]
We may therefore put $k=k'$ in the coefficients before the integrals. Using
also (124.3), we obtain
\begin{align*}
 \int\psi_{\mathbf k'}^{(+)*}\psi_{\mathbf k}^{(+)}dV
 &=\frac{2\pi^2}{k^2}\delta(k'-k)
 \sum_{l=0}^{\infty}(2l+1)P_l(\cos\gamma)\\
 &=\frac{8\pi^2}{k^2}\delta(k'-k)\delta(1-\cos\gamma).
\end{align*}
The expression on the right is zero for every
$\mathbf k\ne\mathbf k'$, while multiplication by
$2\pi k^2\sin\gamma\,d\gamma\,dk/(2\pi)^3$ and integration over all
$\mathbf k$ space gives unity, proving (136.2).

Besides the system $\psi_{\mathbf k}^{(+)}$, one may introduce a system
corresponding to states in which a plane wave and an incoming spherical wave
are present at infinity. Denote these functions by $\psi_{\mathbf k}^{(-)}$;
they are obtained from $\psi_{\mathbf k}^{(+)}$ by
\begin{equation}
 \psi_{\mathbf k}^{(-)}=\psi_{-\mathbf k}^{(+)*}.
 \tag{136.4}
\end{equation}
Indeed, complex conjugation changes the outgoing wave ($e^{ikr}/r$) into an
incoming wave ($e^{-ikr}/r$), while the plane wave becomes
$e^{-i\mathbf{kr}}$. To retain the former definition of $\mathbf k$, with
plane wave $e^{i\mathbf{kr}}$, one must also replace $\mathbf k$ by
$-\mathbf k$, as in (136.4). Noting that
$P_l(-\cos\theta)=(-1)^lP_l(\cos\theta)$, (136.1) gives
\begin{equation}
 \psi_{\mathbf k}^{(-)}=\frac1{2k}\sum_{l=0}^{\infty}i^l(2l+1)e^{-i\delta_l}R_{kl}(r)
 P_l\left(\frac{\mathbf{kr}}{kr}\right).
 \tag{136.5}
\end{equation}

The Coulomb field is a particularly important case. Here
$\psi_{\mathbf k}^{(+)}$ and $\psi_{\mathbf k}^{(-)}$ can be written in
closed form directly from (135.7). Express the parabolic coordinates by
\[
 \frac{k}{2}(\xi-\eta)=kz=\mathbf{kr},\qquad
 k\eta=k(r-z)=kr-\mathbf{kr}.
\]

\SourcePage{684}{687}
Thus, for a repulsive Coulomb field\footnote{Coulomb units are used.},
\begin{align}
 \psi_{\mathbf k}^{(+)}&=e^{-\pi/2k}\Gamma\left(1+\frac{i}{k}\right)e^{i\mathbf{kr}}
 F\left(-\frac{i}{k},1,i(kr-\mathbf{kr})\right), \tag{136.6}\\
 \psi_{\mathbf k}^{(-)}&=e^{-\pi/2k}\Gamma\left(1-\frac{i}{k}\right)e^{i\mathbf{kr}}
 F\left(\frac{i}{k},1,-i(kr+\mathbf{kr})\right). \tag{136.7}
\end{align}
The wave functions for an attractive Coulomb field follow by simultaneously
reversing the signs of $k$ and $r$:
\begin{align}
 \psi_{\mathbf k}^{(+)}&=e^{\pi/2k}\Gamma\left(1-\frac{i}{k}\right)e^{i\mathbf{kr}}
 F\left(\frac{i}{k},1,i(kr-\mathbf{kr})\right), \tag{136.8}\\
 \psi_{\mathbf k}^{(-)}&=e^{\pi/2k}\Gamma\left(1+\frac{i}{k}\right)e^{i\mathbf{kr}}
 F\left(-\frac{i}{k},1,-i(kr+\mathbf{kr})\right). \tag{136.9}
\end{align}

The effect of the Coulomb field on the particle's motion near the origin may
be characterized by the ratio of $|\psi_{\mathbf k}^{(+)}|^2$ or
$|\psi_{\mathbf k}^{(-)}|^2$ at $r=0$ to the squared modulus of the
free-motion wave function $\psi_{\mathbf k}=e^{i\mathbf{kr}}$. Using
\[
 \Gamma\left(1+\frac{i}{k}\right)\Gamma\left(1-\frac{i}{k}\right)
 =\frac{i}{k}\Gamma\left(\frac{i}{k}\right)\Gamma\left(1-\frac{i}{k}\right)
 =\frac{\pi}{k\sinh(\pi/k)},
\]
we readily find, for a repulsive field,
\begin{equation}
 \frac{|\psi_{\mathbf k}^{(+)}(0)|^2}{|\psi_{\mathbf k}|^2}
 =\frac{|\psi_{\mathbf k}^{(-)}(0)|^2}{|\psi_{\mathbf k}|^2}
 =\frac{2\pi}{k(e^{2\pi/k}-1)}
 \tag{136.10}
\end{equation}
and, for an attractive field,
\begin{equation}
 \frac{|\psi_{\mathbf k}^{(+)}(0)|^2}{|\psi_{\mathbf k}|^2}
 =\frac{|\psi_{\mathbf k}^{(-)}(0)|^2}{|\psi_{\mathbf k}|^2}
 =\frac{2\pi}{k(1-e^{-2\pi/k})}.
 \tag{136.11}
\end{equation}

The functions $\psi_{\mathbf k}^{(+)}$ and $\psi_{\mathbf k}^{(-)}$ play an
essential role in applications of perturbation theory to the continuous
spectrum. Suppose that a perturbation $\hat V$ causes a transition of a
particle between continuous-spectrum states. The transition probability is
determined by the matrix element
\begin{equation}
 \int\psi_f^*\hat V\psi_i\,dV.
 \tag{136.12}
\end{equation}
The question arises: which particular solutions of the wave equation should
be chosen as the initial ($\psi_i$) and final ($\psi_f$)

\SourcePage{685}{688}
wave functions in order to obtain the amplitude for a particle to pass from a
state of momentum $\hbar\mathbf k$ to a state of momentum
$\hbar\mathbf k'$ at infinity\footnote{One example of such a process is an
electron which, on colliding with a stationary heavy nucleus, emits a photon
and thereby changes its energy and direction of motion. The perturbation
$\hat V$ is the interaction of the electron with the radiation field, while
the Coulomb field of the nucleus is the field $U$ for which
$\psi_{\mathbf k}^{(+)}$ and $\psi_{\mathbf k}^{(-)}$ are defined (see IV,
\S~92, 96). Another example is the collision of an electron with an atom,
accompanied by ionization of the latter (see Problem 4 in \S~148).}? We shall
show that one must choose
\begin{equation}
 \psi_i=\psi_{\mathbf k}^{(+)},\qquad\psi_f=\psi_{\mathbf k'}^{(-)}
 \tag{136.13}
\end{equation}
(A.~Sommerfeld, 1931).

This becomes clear by considering how the question would be treated by
perturbation theory applied not only to $\hat V$ but also to the field $U(r)$
in which the particle moves. To zeroth order in $U$, the matrix element
(136.12) is
\[
 V_{\mathbf k'\mathbf k}=\int e^{-i\mathbf{k'r}}\hat V e^{i\mathbf{kr}}dV.
\]
At higher orders in $U$, this integral is replaced by a series whose terms are
integrals of the form
\[
 \int\frac{V_{\mathbf k'\mathbf k_1}U_{\mathbf k_1\mathbf k_2}\ldots U_{\mathbf k_n\mathbf k}}
 {(E_{\mathbf k}-E_{\mathbf k_1}+i0)\ldots(E_{\mathbf k}-E_{\mathbf k_n}+i0)}
 \,d^3k_1\ldots d^3k_n
\]
(compare \S~43, 130). The numerators contain matrix elements with respect to
unperturbed plane waves, arranged in various sequences, and every pole is
bypassed in the integrations according to the same specified rule. On the
other hand, this series may be obtained as the matrix element (136.12), with
$\psi_i$ and $\psi_f$ represented by perturbation series in the field $U$.
The fact that the result must be a sum of integrals in which all poles are
bypassed according to the same rule means that the poles in the series for
$\psi_i$ and $\psi_f^*$ are bypassed by that rule as well. Solving the wave
equation by perturbation theory with this bypass rule automatically produces
a solution whose asymptotic form contains an outgoing wave together with the
plane wave. In other words, the wave functions which at zeroth order in $U$
were
\[
 \psi_i=e^{i\mathbf{kr}},\qquad\psi_f^*=e^{-i\mathbf{k'r}},
\]

\SourcePage{686}{689}
must be replaced by the exact solutions $\psi_{\mathbf k}^{(+)}$ and
$\psi_{-\mathbf k'}^{(+)}=(\psi_{\mathbf k'}^{(-)})^*$, respectively. This
proves the rule ((136.13)).

The choice of $\psi_{\mathbf k'}^{(-)}$ as the final wave function applies
also to transitions from a discrete-spectrum state to a continuous-spectrum
state (the question of how to choose $\psi_i$ naturally does not arise in
that case).
