\SourcePage{686}{689}
\section{Collisions of identical particles}

The collision of two identical particles requires special consideration. In
quantum mechanics, the identity of the particles gives rise to a distinctive
exchange interaction between them. This interaction also has an essential
effect on scattering (N.~Mott, 1930)\footnote{The direct spin--orbit
interaction is still not considered here.}.

The orbital wave function of a two-particle system must be symmetric or
antisymmetric with respect to the particles according as their total spin is
even or odd (see \S~62). The wave function describing the scattering, obtained
by solving the ordinary Schrödinger equation, must therefore be symmetrized or
antisymmetrized in the particles. Interchanging the particles is equivalent
to reversing the direction of the radius vector joining them. In the
coordinate system in which the center of mass is at rest, this means that $r$
is unchanged while $\theta$ is replaced by $\pi-\theta$ (and consequently
$z=r\cos\theta$ becomes $-z$). Thus, instead of the asymptotic expression
(123.3), we must write
\begin{equation}
 \psi=e^{ikz}\pm e^{-ikz}+\frac1r e^{ikr}[f(\theta)\pm(\pi-\theta)].
 \tag{137.1}
\end{equation}
Because the particles are identical, one cannot, of course, say which is the
scattered particle and which is the scattering particle. In the center-of-mass
system there are two identical incident waves propagating toward each other,
$e^{ikz}$ and $e^{-ikz}$. The outgoing spherical wave in (137.1) accounts for
the scattering of both particles, and the current calculated from it gives
the probability that either particle is scattered into a specified element
$d\Omega$ of solid angle. The scattering cross-section is the ratio of this
current to the current density in each incident

\SourcePage{687}{690}
plane wave; hence, as before, it is determined by the squared modulus of the
coefficient of $e^{ikr}/r$ in (137.1).

Thus, if the total spin of the colliding particles is even, the scattering
cross-section is
\begin{equation}
 d\sigma_s=|f(\theta)+f(\pi-\theta)|^2d\Omega,
 \tag{137.2}
\end{equation}
while if it is odd,
\begin{equation}
 d\sigma_a=|f(\theta)-f(\pi-\theta)|^2d\Omega.
 \tag{137.3}
\end{equation}
The appearance of the interference term
$f(\theta)f^*(\pi-\theta)+f^*(\theta)f(\pi-\theta)$ is characteristic of the
exchange interaction. If the particles were distinguishable, as in ordinary
classical mechanics, the probability that either is scattered into the given
solid-angle element $d\Omega$ would simply be the sum of the probabilities
that one is deflected through $\theta$ and the oppositely moving one through
$\pi-\theta$. In other words, the cross-section would be
\[
 \{|f(\theta)|^2+|f(\pi-\theta)|^2\}d\Omega.
\]

In the limiting case of low velocities, the scattering amplitude, for an
interaction decreasing sufficiently rapidly with distance, tends to a
constant limit independent of angle (\S~132). Equation (137.3) shows that
$d\sigma_a$ then vanishes; only particles with even total spin scatter from
one another.

Equations (137.2), (137.3) assume that the total spin of the colliding
particles has a specified value. If the particles are not in specified spin
states, the cross-section must be found by averaging with all spin states
taken as equally probable.

It was shown in \S~62 that, among the $(2s+1)^2$ different spin states of a
system of two particles of spin $s$, $s(2s+1)$ states correspond to even total
spin and $(s+1)(2s+1)$ to odd total spin if $s$ is half-integral, while the
reverse holds if $s$ is integral. First suppose that the particle spin $s$ is
half-integral. The probability that the two-particle system has even $S$ is
then $s(2s+1)/(2s+1)^2=s/(2s+1)$, and the probability of odd $S$ is
$(s+1)/(2s+1)$. The scattering cross-section is therefore
\begin{equation}
 d\sigma=\frac{s}{2s+1}d\sigma_s+\frac{s+1}{2s+1}d\sigma_a.
 \tag{137.4}
\end{equation}

\SourcePage{688}{691}
Substitution of (137.2), (137.3) gives
\begin{equation}
 d\sigma\left\{|f(\theta)|^2+|f(\pi-\theta)|^2
 -\frac1{2s+1}[f(\theta)f(\pi-\theta)^*+f(\theta)^*f(\pi-\theta)]\right\}d\Omega.
 \tag{137.5}
\end{equation}
Similarly, for integral $s$,
\begin{equation}
 d\sigma=\left\{|f(\theta)|^2+|f(\pi-\theta)|^2
 +\frac1{2s+1}[f(\theta)f(\pi-\theta)^*+f(\theta)^*f(\pi-\theta)]\right\}d\Omega.
 \tag{137.6}
\end{equation}

As an example, write the formulas for the collision of two electrons
interacting by the Coulomb law, $U=e^2/r$. Substitution of (135.9) in (137.5)
with $s=1/2$ gives, after a simple calculation and in ordinary units,
\begin{align}
 d\sigma={}&\left(\frac{e^2}{m_0v^2}\right)^2
 \left[\frac1{\sin^4(\theta/2)}+\frac1{\cos^4(\theta/2)}\right.\notag\\
 &\left.-\frac1{\sin^2(\theta/2)\cos^2(\theta/2)}
 \cos\left(\frac{e^2}{\hbar v}\ln\tan^2\frac\theta2\right)\right]d\Omega
 \tag{137.7}
\end{align}
(the electron mass $m_0$ has been introduced in place of the reduced mass
$m=m_0/2$). This formula is considerably simplified when the speed is so high
that $e^2\ll v\hbar$; note that this is precisely the condition under which
perturbation theory applies to the Coulomb field. The cosine in the third term
may then be replaced by unity, giving
\begin{equation}
 d\sigma=\left(\frac{2e^2}{m_0v^2}\right)^2
 \frac{4-3\sin^2\theta}{\sin^4\theta}\,d\Omega.
 \tag{137.8}
\end{equation}
The opposite limiting case, $e^2\gg v\hbar$, corresponds to the transition to
classical mechanics (see the end of \S~127). This transition occurs in a
rather peculiar way in (137.7). When $e^2\gg v\hbar$, the cosine in the third
term in square brackets is a rapidly oscillating function. At each specified
$\theta$, (137.7) gives a scattering cross-section which in general differs
appreciably from the Rutherford value. But averaging over even a small
interval of $\theta$ makes the oscillating term in (137.7) disappear, and the
classical formula is recovered.

All the formulas above refer to the coordinate system in which the center of
mass is at rest. Passage to the system in which one particle was at rest
before the collision is made, according to (123.2), simply by replacing
$\theta$ with $2\vartheta$. Thus, for electron collisions, (137.7) gives

\SourcePage{689}{692}
\begin{align}
 d\sigma={}&\left(\frac{2e^2}{m_0v^2}\right)^2
 \left[\frac1{\sin^4\vartheta}+\frac1{\cos^4\vartheta}
 -\frac1{\sin^2\vartheta\cos^2\vartheta}\right.\notag\\
 &\left.\times\cos\left(\frac{e^2}{\hbar v}\ln\tan^2\vartheta\right)\right]
 \cos\vartheta\,d\Omega,
 \tag{137.9}
\end{align}
where $d\Omega$ is the solid-angle element in the new coordinate system. When
$\theta$ is replaced by $2\vartheta$, the solid-angle element $d\Omega$ must
be replaced by $4\cos\vartheta\,d\Omega$, since
$\sin\theta\,d\theta\,d\varphi=4\cos\vartheta\sin\vartheta\,d\vartheta\,d\varphi$.

\subsection*{Problem}

\ProblemHeading Determine the scattering cross-section of two identical
spin-$1/2$ particles having specified mean spin values
$\bar{\mathbf s}_1$ and $\bar{\mathbf s}_2$.

\SolutionHeading The dependence of the cross-section on the particle
polarizations must be expressed by a term proportional to the scalar
$\bar{\mathbf s}_1\bar{\mathbf s}_2$. Seek $d\sigma$ in the form
$a+b\bar{\mathbf s}_1\bar{\mathbf s}_2$. For unpolarized particles
($\bar{\mathbf s}_1=\bar{\mathbf s}_2=0$), the second term is absent and,
according to (137.4),
$d\sigma=a=(d\sigma_s+3d\sigma_a)/4$. If both particles are fully polarized
in the same direction ($\bar{\mathbf s}_1\bar{\mathbf s}_2=1/4$), the system
is certainly in a state with $S=1$; hence
$d\sigma=a+b/4=d\sigma_a$. Determining $a$ and $b$ from these two equalities,
we find
\[
 d\sigma=\frac14(d\sigma_s+3d\sigma_a)
 +(d\sigma_a-d\sigma_s)\bar{\mathbf s}_1\bar{\mathbf s}_2.
\]
