\SourcePage{705}{708}
\section{Regge poles}

In \S~128 we considered the analytic properties of the scattering amplitude
as a function of the complex variable $E$, the particle energy; the orbital
angular momentum $l$ was a parameter taking real integral values. Further
properties of considerable methodological importance emerge if $l$ is now
regarded as a continuous complex variable while the energy $E$ is kept
real\footnote{These properties were first studied by Regge (T.~Regge, 1958).}.

\SourcePage{706}{709}
As in \S~128, consider radial wave functions whose asymptotic form as
$r\to\infty$ is
\begin{equation}
 \chi_l=rR_l=A(l,E)\exp\left(-\frac{\sqrt{-2mE}}{\hbar}r\right)
 +B(l,E)\exp\left(\frac{\sqrt{-2mE}}{\hbar}r\right).
 \tag{141.1}
\end{equation}
These functions solve the Schrödinger equation (32.8), in which $l$ is now a
complex parameter. One of the two independent solutions is selected by
\begin{equation}
 R_l\approx\mathrm{const}\cdot r^l\quad\text{as }r\to0.
 \tag{141.2}
\end{equation}

This condition immediately imposes a restriction on the permissible values
of $l$. The general small-$r$ solution of (32.8) is
\[
 R_l\approx c_1r^l+c_2r^{-l-1}
\]
(see the end of \S~32). To distinguish the second solution uniquely against
the background of the first and exclude it, the term $r^{-l-1}$ must exceed
$r^l$ as $r\to0$. For complex $l$ this gives
$\operatorname{Re}l>\operatorname{Re}(-l-1)$, or
\begin{equation}
 \operatorname{Re}\left(l+\frac12\right)>0.
 \tag{141.3}
\end{equation}
Only this half-plane of complex $l$, to the right of the vertical line
$l=-1/2$, is considered below.

As a solution of a differential equation whose coefficients are analytic in
$l$, the wave function $R(r;l,E)$ is analytic in this parameter and has no
singularities in the half-plane (141.3). The same applies in particular to
the asymptotic expression (141.1), so $A(l,E)$ and $B(l,E)$ have no
singularities in $l$. It is understood, however, that retaining both terms of
(141.1) as $r\to\infty$ is legitimate. For $E>0$ this is always so; for
$E<0$ it is so if $U(r)$ satisfies (128.6) or (128.13). What matters here is
that the character of the asymptotic behavior in $r$ depends only on $E$, not
on $l$; making $l$ complex therefore does not alter the conditions for
passing to the asymptotic form.

Comparison of (141.1) with (128.15) gives the $S$-matrix element
\begin{equation}
 S(l,E)=\exp[2i\delta(l,E)]=e^{i\pi l}\frac{A(l,E)}{B(l,E)},
 \tag{141.4}
\end{equation}

\SourcePage{707}{710}
which remains valid for complex $l$, although the ``phase shift'' $\delta$ is
then no longer real.

For real $l$ and $E>0$, $A$ and $B$ are related by (128.4):
$A(l,E)=B^*(l,E)$. Hence, for complex $l$,
\begin{equation}
 A(l^*,E)=B^*(l,E)\quad\text{for }E>0,
 \tag{141.5}
\end{equation}
and $S(l,E)$ satisfies the condition of \emph{complex unitarity}
\begin{equation}
 S^*(l,E)S(l^*,E)=1.
 \tag{141.6}
\end{equation}

Since $A(l,E)$ and $B(l,E)$ have no singularities as functions of $l$, the
singularities, or poles, of $S(l,E)$ and of the partial scattering amplitude
$f(l,E)$ occur only at the zeros of $B(l,E)$. The poles of the scattering
amplitude in the complex $l$ plane are called \emph{Regge poles}. Their
positions depend, of course, on the real parameter $E$. The functions
$l=\alpha_i(E)$ specifying the pole positions are called \emph{Regge
trajectories}; as $E$ varies, the poles move along definite lines in the
$l$ plane. The index $i$ numbering the poles will be omitted below.

To begin the study of Regge trajectories, let us first show that all
$\alpha(E)$ are real for $E<0$. Consider the equation
\begin{equation}
 \chi''+\left[\frac{2m}{\hbar^2}(E-U)-\frac{\alpha(\alpha+1)}{r^2}\right]\chi=0,
 \tag{141.7}
\end{equation}
satisfied by the wave function with $l=\alpha$. Multiply by $\chi^*$ and
integrate with respect to $r$, integrating the first term by parts. This gives
\[
 -\int_0^\infty|\chi'|^2dr+\frac{2m}{\hbar^2}\int_0^\infty(E-U)|\chi|^2dr
 -\alpha(\alpha+1)\int_0^\infty\frac{|\chi|^2}{r^2}dr=0.
\]
Here it has been used that, when $B=0$, the condition defining a Regge pole,
the wave function decays exponentially as $r\to\infty$, so every integral
converges. The first two terms are real, and the integral in the last is real.
Therefore
\[
 \operatorname{Im}[\alpha(\alpha+1)]
 =\operatorname{Im}\left(\alpha+\frac12\right)^2
 =2\operatorname{Re}\left(\alpha+\frac12\right)\operatorname{Im}\alpha=0.
\]

\SourcePage{708}{711}
But only poles in the half-plane (141.3) are considered, so
$\operatorname{Re}(\alpha+1/2)>0$ and hence
\begin{equation}
 \operatorname{Im}\alpha(E)=0\quad\text{for }E<0.
 \tag{141.8}
\end{equation}

Next perform on (141.7) the operations analogous to the derivation of
(128.10): differentiate it with respect to $E$, multiply the resulting
equation by $\chi$, and multiply (141.7) by $\partial\chi/\partial E$.
Subtracting the equations gives the identity
\[
 \left[\chi'\frac{\partial\chi}{\partial E}
 -\chi\left(\frac{\partial\chi}{\partial E}\right)'\right]'
 -\frac{2m}{\hbar^2}\chi^2
 +\frac{\chi^2}{r^2}\frac{d\alpha(\alpha+1)}{dE}=0.
\]
Integrate from zero to infinity, again using the vanishing of $\chi$ as
$r\to\infty$. The integral of the first term vanishes, leaving
\begin{equation}
 \frac{d\alpha(\alpha+1)}{dE}\int_0^\infty\frac{\chi^2}{r^2}dr
 =\frac{2m}{\hbar^2}\int_0^\infty\chi^2dr.
 \tag{141.9}
\end{equation}
Since $\alpha$ is real, the wave function is real as well, and both integrals
in (141.9) are positive. Consequently,
\[
 \frac{d\alpha(\alpha+1)}{dE}
 =2\left(\alpha+\frac12\right)\frac{d\alpha}{dE}>0,
\]
and, because $\alpha+1/2>0$,
\[
 \frac{d\alpha}{dE}>0\quad\text{for }E<0.
\]
Thus, for $E<0$, the functions $\alpha(E)$ increase monotonically with $E$.

The negative values of $E$ at which $\alpha(E)$ takes ``physical'' values,
the integers $l=0,1,2,\ldots$, correspond to the discrete energy levels of
the system. This yields a new classification principle for bound states: the
Regge trajectories on which they lie.

As an example, consider the Regge trajectories for motion in an attractive
Coulomb field. The scattering-matrix elements are then\footnote{Compare
(135.11), where the signs before $k$ must be reversed to pass from repulsion
to attraction.}
\begin{equation}
 S_l=\frac{\Gamma(l+1-i/k)}{\Gamma(l+1+i/k)}.
 \tag{141.10}
\end{equation}

\SourcePage{709}{712}
Here $k$ is in Coulomb units. The poles occur where the argument of
$\Gamma(l+1-i/k)$ is a negative integer or zero. For $E<0$,
$k=i\sqrt{-2E}$, so
\begin{equation}
 \alpha(E)=-n_r-1+\frac1{\sqrt{-2E}},\qquad E<0,
 \tag{141.11}
\end{equation}
where $n_r=0,1,2,\ldots$ numbers the Regge trajectories. Equating
$\alpha(E)$ to an integer $l=0,1,2,\ldots$ gives the familiar Bohr formula for
the discrete energy levels in an attractive Coulomb field,
\[
 E=-\frac1{2(n_r+l+1)^2}.
\]
The number $n_r$ is the radial quantum number that specifies the number of
nodes of the radial wave function. Each Regge trajectory, each specified
$n_r$, corresponds to an infinite set of levels differing in orbital angular
momentum.

Turn now to the properties of $\alpha(E)$ for $E>0$. Recall from \S~128 that
$A(l,E)$ and $B(l,E)$ in (141.1), as functions of complex $E$, are defined on
a plane cut along the positive real semiaxis. The functions
$l=\alpha(E)$, the roots of $B(l,E)=0$, have the same cut. On its upper and
lower edges they take complex-conjugate values; on the upper edge
$\operatorname{Im}\alpha>0$. Without giving a formal proof, we present more
physical considerations that explain this result.

When $l$ is complex, the centrifugal energy, and hence the effective
potential energy $U_l=U+l(l+1)/(2mr^2)$, is also complex. Repeating the
derivation in \S~19 gives, in place of (19.6),
\[
 \frac{\partial}{\partial t}|\psi|^2+\Div{\mathbf j}
 =2|\psi|^2\operatorname{Im}U_l.
\]
For $l=\alpha$ with $\operatorname{Im}\alpha>0$, one also has
$\operatorname{Im}U_l>0$. The right-hand side is then positive, as though new
particles were being emitted within the field. Correspondingly, the
asymptotic wave function, which for $B=0$ contains only the first term of
(141.1), must represent an outgoing wave. This is precisely what occurs on
the upper edge of the cut; compare the passage from (128.1) to (128.3).

Since $\alpha(E)$ is complex for $E>0$, it cannot take the ``physical'' values
$l=0,1,$

\SourcePage{710}{713}
$2,\ldots$ there. It may, however, come close to such values in the complex
$l$ plane. We shall show that a resonance then appears in the partial
scattering amplitude for the corresponding integer $l$.

Let $l_0$ be an integer approached by $\alpha(E)$, and let $E_0$ be the real
positive energy for which $\operatorname{Re}\alpha(E_0)=l_0$. Near this point,
\begin{equation}
 \alpha(E)\approx l_0+i\eta+\beta(E-E_0),
 \tag{141.12}
\end{equation}
where $\eta=\operatorname{Im}\alpha(E_0)$ is real. Consider $\alpha(E)$ on the
upper edge of the cut. Then $\eta>0$, as stated above, and $\eta\ll1$ by the
assumed closeness to $l_0$. The constant $\beta=d\alpha/dE$ at $E=E_0$ may
also be taken as real and positive. Indeed, since $\alpha(E)$ is nearly real,
the wave function $\chi(r;\alpha,E)$ is nearly real. Neglecting quantities of
higher order in $\eta$, its imaginary part may be discarded, and the
positivity of $\beta$ follows from the positivity of the integrals in
(141.9)\footnote{To clarify the structure of these integrals, note that the
asymptotic region $r\gg a$, where $a$ is the range of the field and (141.1)
holds, contributes only a small amount when $\eta$ is small. If
$l=\alpha(E)$ is a zero of $B(l,E)$, then by (141.5)
$l=\alpha^*$ is a zero of $A(l,E)$. Hence $A(\alpha,E)$, and therefore
$\chi(r;\alpha,E)$ for $r\gg a$, is of order $\eta^{1/2}$ (see (134.11)). In
estimating the integrals, it is also important that on the upper edge of the
cut in $E$ the wave function contains $e^{ikr}$:
$\chi(r;\alpha,E)=A(\alpha,E)e^{ikr}$. On this edge $E$ may be understood as
$E+i\delta$, where $\delta\to+0$, so $k$ acquires a small positive imaginary
part that ensures convergence of the integrals in (141.9). Physically, the
small contribution from $r\gg a$ is associated with $E_0$ describing a
quasistationary state (see below): the particle reaches this region only
through the improbable decay of the state. The main contribution comes from
$r\sim a$, where the wave function is nearly real.}.

Since $l=\alpha(E)$ is a zero of $B(l,E)$, near $\alpha,E_0$ this function is
proportional to $\alpha-l$. Using (141.12),
\begin{equation}
 B(l_0,E)\approx\mathrm{const}\cdot[\beta(E-E_0)+i\eta].
 \tag{141.13}
\end{equation}
This has precisely the form of (134.6), with $E_0$ the energy and
$\Gamma=2\eta/\beta>0$ the width of a quasidiscrete level. Thus the closeness
of a Regge trajectory

\SourcePage{711}{714}
to integral $l$ at $E>0$ corresponds to quasistationary states of the system.
These states therefore admit the same classification principle as strictly
stationary states: each Regge trajectory may correspond to a whole family of
discrete and quasidiscrete levels.

Treating $l$ as a complex variable gives a useful integral representation of
the total scattering amplitude for $E>0$, which is given by (123.11):
\begin{equation}
 f(\mu)=\frac1{2ik}\sum_{l=0}^{\infty}(2l+1)[S(l,E)-1]P_l(\mu),\qquad\mu=\cos\theta.
 \tag{141.14}
\end{equation}

First, $P_l(\mu)$ must be defined not only for integers $l\ge0$ but also for
complex $l$. This is done by taking $P_l(\mu)$ to be the solution of (c.2),
\begin{equation}
 (1-\mu)^2P_l''(\mu)-2\mu P_l'(\mu)+l(l+1)P_l(\mu)=0,
 \tag{141.15}
\end{equation}
with boundary condition $P_l(1)=1$. So defined, $P_l(\mu)$ has no
singularities as a function of $l$ at finite values of that
variable\footnote{Comparison of (141.15) with (e.2) expresses $P_l(\mu)$ in
terms of a hypergeometric function:
\[
 P_l(\mu)=F\left(-l,l+1,1;\frac{1-\mu}{2}\right).
\]}.

It is readily seen that (141.14) equals the integral
\begin{equation}
 f(\mu)=\frac1{4k}\int_C\frac{2l+1}{\sin\pi l}[S(l,E)-1]P_l(-\mu)\,dl,
 \tag{141.16}
\end{equation}
taken over a path $C$ that encircles clockwise all the points
$l=0,1,2,\ldots$ on the real axis and closes at infinity:

\begin{center}
\begin{tikzpicture}[x=1.1cm,y=0.8cm]
  \draw[->] (-1,0) -- (4.8,0);
  \foreach \x/\lab in {1/1,2/2,3/3,4/4}
    \draw (\x,0.09)--(\x,-0.09) node[below] {\lab};
  \draw (-0.5,0.09)--(-0.5,-0.09)
    node[below left,xshift=-1pt] {$-\tfrac12$};
  \draw (0,0.09)--(0,-0.09)
    node[below right,xshift=1pt] {$0$};
  \draw[->,rounded corners=12pt] (4.5,0.48)--(-0.2,0.48)--(-0.42,0)--(-0.2,-0.48)--(4.5,-0.48);
  \node[right] at (4.5,-0.48) {$C$};
\end{tikzpicture}
\end{center}

All poles $l=\alpha_1,\alpha_2,\ldots$ of $S(l,E)$, which for $E>0$ do not
lie on the real axis, must remain outside $C$.

Indeed, (141.16) reduces to $-2\pi i$ times the sum of the residues of the
integrand at $l=0,1,2,\ldots$, the poles of $1/\sin\pi l$, whose residues are
$(-1)^l/\pi$. Using also $P_l(-\mu)=(-1)^lP_l(\mu)$ for integral $l$ reduces
(141.16) to (141.14)\footnote{A more detailed account of the questions
considered in this section, within nonrelativistic theory, may be found in the
book by de Alfaro and Regge cited in the note on p.~612.}.

\SourcePage{712}{715}
\subsection*{Problem}

\ProblemHeading Show that the phase shifts corresponding to successive
integral values of $l$ satisfy
\[
 \delta_{l+1}(E)-\delta_l(E)<\pi/2.
\]

\SolutionHeading Regard $l$ as a continuous real variable and differentiate
(32.10) with respect to $l$:
\[
 \left[\frac{\partial\chi}{\partial l}\right]''
 +\left[\frac{2m}{\hbar^2}(E-U)-\frac{l(l+1)}{r^2}\right]
 \frac{\partial\chi}{\partial l}=\frac{2l+1}{r^2}\chi.
\]
Multiply this equation by $\chi$, multiply the original equation by
$\partial\chi/\partial l$, and subtract. This gives
\[
 \left[\chi\frac{\partial\chi'}{\partial l}
 -\chi'\frac{\partial\chi}{\partial l}\right]'
 =(2l+1)\frac{\chi^2}{r^2}.
\]
Integrate from zero to infinity. At $r=0$ the expression in square brackets
vanishes, while as $r\to\infty$ the asymptotic expression (33.20) for $\chi$
may be used. The result is
\[
 4k\left(\frac\pi2-\frac{\partial\delta_l}{\partial l}\right)
 =(2l+1)\int_0^\infty\frac{\chi^2}{r^2}dr>0,
\]
so $\partial\delta_l/\partial l<\pi/2$. Integrating from $l$ to $l+1$ gives
the required inequality. Combining it with (133.17), one can prove that the
number $n_l$ of discrete levels does not increase with $l$. Indeed, as
$E\to\infty$, where the Born approximation applies, the scattering phases
tend to zero, so $\delta_l(\infty)=0$. Hence
\[
 n_{l+1}-n_l=-\frac1\pi[\delta_{l+1}(0)-\delta_l(0)]<\frac12,
 \qquad n_{l+1}-n_l\le0.
\]
