\SourcePage{627}{630}
\section{The semiclassical case}

Let us trace how the limiting transition from the quantum-mechanical theory
of scattering to the classical theory takes place.

Excluding the zero scattering angle $\theta$, we may write the scattering
amplitude given by the exact theory in the form (124.4)
\begin{equation}
 f(\theta)=\frac1{2ik}\sum_{l=0}^{\infty}(2l+1)P_l(\cos\theta)e^{2i\delta_l}.
 \tag{127.1}
\end{equation}
We know that semiclassical wave functions are characterized by a large value
of their phase. It is therefore natural to suppose beforehand that the
limiting transition in scattering theory corresponds to large phases
$\delta_l$. The value of the sum (127.1) is determined mainly by terms with
large $l$. We may consequently replace $P_l(\cos\theta)$ by the asymptotic
expression (49.7), written as
\[
 P_l(\cos\theta)\approx-\frac{i}{\sqrt{2\pi l\sin\theta}}
 \left\{\exp\left[i\left(l+\frac12\right)\theta+\frac{i\pi}4\right]
 -\exp\left[-i\left(l+\frac12\right)\theta-\frac{i\pi}4\right]\right\}.
\]
Substitution in (127.1) gives
\begingroup\small
\begin{equation}
 f(\theta)=\frac1k\sum_l\sqrt{\frac{l}{2\pi\sin\theta}}
 \left(\exp\left\{i\left[2\delta_l-\left(l+\frac12\right)\theta-\frac\pi4\right]\right\}
 -\exp\left\{i\left[2\delta_l+\left(l+\frac12\right)\theta+\frac\pi4\right]\right\}\right).
\tag{127.2}
\end{equation}
\endgroup

Regarded as functions of $l$, the exponential factors oscillate rapidly
(since their phases are large). Most terms in the sum (127.2) therefore
cancel one another. The sum is determined mainly by values of $l$ near the
one at which the exponent of either exponential has an extremum, that is,
near a root of
\begin{equation}
 2\frac{d\delta_l}{dl}\pm\theta=0.
 \tag{127.3}
\end{equation}
In this region there are many terms for which the exponential factors remain
nearly constant (the exponents vary slowly near their stationary point), and
these terms consequently do not cancel.

\SourcePage{628}{631}
In the semiclassical case the phases $\delta_l$ may be written (see \S~124)
as the limit, for $r\to\infty$, of the difference between the phase
\[
 \frac\pi4+\frac1\hbar\int_{r_0}^{r}
 \sqrt{2m[E-U(r)]-\frac{\hbar^2(l+1/2)^2}{r^2}}\,dr
\]
of the semiclassical wave function in the field $U(r)$ and the phase
$kr-\pi l/2$ of the free-motion wave function (see \S~33). Thus
\begin{equation}
 \delta_l=\int_{r_0}^{\infty}\left[\sqrt{\frac{2m}{\hbar^2}(E-U)-\frac{(l+1/2)^2}{r^2}}-k\right]dr
 +\frac\pi2\left(l+\frac12\right)-kr_0.
 \tag{127.4}
\end{equation}
This expression must be substituted in (127.3). In differentiating the
integral, one must remember that its lower limit $r_0$ also depends on $l$;
the resulting term $k\,dr_0/dl$, however, cancels the derivative of
$-kr_0$ in $\delta_l$. The quantity $\hbar(l+1/2)$ is the particle's angular
momentum. In classical mechanics it may be written as $m\rho v$, where
$\rho$ is the \emph{impact parameter} and $v$ is the particle's speed at
infinity.

Making this substitution, equation (127.3) finally becomes
\begin{equation}
 2\int_{r_0}^{\infty}\frac{\rho\,dr}{r^2\sqrt{1-U/E-\rho^2/r^2}}=\pi\pm\theta.
 \tag{127.5}
\end{equation}
In a repulsive field this equation has a root (in $\rho$) only with the minus
sign before $\theta$ on the right-hand side, and in an attractive field only
with the plus sign.

Equation (127.5) is precisely the classical equation that determines the
scattering angle from the impact parameter (see I, \S~18). It is also readily
verified that the classical expression is obtained for the cross-section.

For this purpose, expand the exponent in (127.2) in powers of
$l'=l-l_0(\theta)$, where $l_0(\theta)$ is defined by (127.3)--(127.5). For
definiteness consider the first term in (127.2), and accordingly take the
lower sign in (127.3), the repulsive case. Noting from (127.3) that
\[
 \frac{d\delta_l}{dl}=\frac12\frac{d\theta}{dl_0},
\]

\SourcePage{629}{632}
we have
\[
 i\left[2\delta_l-\left(l+\frac12\right)\theta-\frac\pi4\right]
 \approx i\left[2\delta_{l_0}-\left(l_0+\frac12\right)\theta-\frac\pi4\right]
 +i\frac12\frac{d\theta}{dl_0}l'^2.
\]
We now replace summation over $l$ in (127.2) by integration over $dl'$ near
$l'=0$. Regarding $l'$ as a complex variable, take the integration path near
this point along the direction of steepest descent of the exponential, at an
angle $\pi/4$ or $-\pi/4$ to the real axis according to the sign of
$d\theta/dl_0$. In other words, put $l'=\xi\exp(\pm i\pi/4)$ and integrate
over real $\xi$; because the integral converges rapidly, its limits may be
extended from $-\infty$ to $\infty$:
\[
 \int_{-\infty}^{\infty}\exp\left(-\frac12\left|\frac{d\theta}{dl_0}\right|\xi^2\right)d\xi
 =\sqrt{2\pi\left|\frac{dl_0}{d\theta}\right|}.
\]
The result is
\begin{equation}
 f(\theta)=\frac1k\left(\frac{l_0}{\sin\theta}\left|\frac{dl_0}{d\theta}\right|\right)^{1/2}
 \exp\left\{i\left[2\delta_{l_0}-\left(l_0+\frac12\right)\theta-\frac\pi4\right]\right\}.
 \tag{127.6}
\end{equation}
Hence
\begin{equation}
 d\sigma=|f|^2\,2\pi\sin\theta\,d\theta
 =2\pi\frac{l_0}{k^2}\left|\frac{dl_0}{d\theta}\right|d\theta,
 \tag{127.7}
\end{equation}
and, on introducing the impact parameter by $\rho=l_0/k$, we arrive at the
classical formula $d\sigma=2\pi\rho\,d\rho$.

Thus the condition for scattering through a specified angle $\theta$ to be
classical is that the value of $l$ satisfying (127.3) be large, and that
$\delta_l$ at this $l$ also be large\footnote{The relation between $\theta$
and $\rho$ given by (127.5) may be multivalued; more than one value of $\rho$
then corresponds to the same $\theta$. In that event the amplitude $f(\theta)$
is the sum of expressions (127.6) with the corresponding values of $l_0$. At
extrema of $\theta(\rho)$, the derivative $d\rho/d\theta$, and with it the
classical differential cross-section $d\sigma/d\Omega$, becomes infinite;
near this angle the classical approximation is, of course, insufficient (see
Problem 2).}. This condition has a simple meaning. To speak of classical
scattering through an angle $\theta$ for a particle passing with impact
parameter $\rho$, the quantum-mechanical uncertainties in both quantities
must be relatively small: $\Delta\rho\ll\rho$, $\Delta\theta\ll\theta$. The
uncertainty of the scattering angle is of order $\Delta\theta\sim\Delta p/p$,
where $p$ is the particle momentum and $\Delta p$ is the uncertainty of its
transverse component. Since $\Delta p\sim\hbar/\Delta\rho\gg\hbar/\rho$,

\SourcePage{630}{633}
we have $\Delta\theta\gg\hbar/p\rho$, and consequently, in any event,
\begin{equation}
 \theta\gg\frac{\hbar}{\rho mv}.
 \tag{127.8}
\end{equation}
Replacing the angular momentum $m\rho v$ by $\hbar l$ gives
$\theta l\gg1$, which agrees with the condition $\delta_l\gg1$ (since
$\delta_l\sim l\theta$, as follows from (127.3)).

The classical deflection angle of the particle may be estimated as the ratio
of the transverse momentum increment $\Delta p$ during the ``collision time''
$\tau\sim\rho/v$ to the initial momentum $mv$. The force acting on the
particle at distance $\rho$ in the field $U(r)$ is
$F=-dU(\rho)/d\rho$, and therefore $\Delta p\sim F\rho/v$, so that
$\theta\sim\rho F/mv^2$. Strictly, this estimate is valid only for
$\theta\ll1$, but as an order-of-magnitude estimate it can be extended to
$\theta\sim1$. Substitution in (127.8) gives the semiclassical condition for
scattering in the form
\begin{equation}
 |F|\rho^2\gg\hbar v.
 \tag{127.9}
\end{equation}
This inequality must hold for every value of $\rho$ for which
$|U(\rho)|\lesssim E$ still holds.

If the field $U(r)$ decreases faster than $1/r$, condition (127.9) must fail
at sufficiently large $\rho$. But large $\rho$ correspond to small $\theta$;
hence scattering through sufficiently small angles cannot be classical. If
the field falls off more slowly than $1/r$, scattering through small angles
is classical; whether scattering through large angles is classical then
depends on the behavior of the field at short distances.

For the Coulomb field $U=\alpha/r$, condition (127.9) holds if
$\alpha\gg\hbar v$. This is the reverse of the condition under which the
Coulomb field may be treated as a perturbation. We shall see, however, that
for accidental reasons the quantum theory of Coulomb scattering gives a
result identical with the classical one in all cases.

\subsection*{Problems}

\ProblemHeading 1. Find the total semiclassical scattering cross-section in
a field which at sufficiently large distances has the form $U=\alpha/r^n$,
where $n>2$.

\SolutionHeading Having in mind that the main role is played by phases
$\delta_l$ with large $l$, calculate them from (124.1):
\begin{equation}
 \delta_l=-\frac{m\alpha}{\hbar^2}\int_{l/k}^{\infty}\frac{dr}{r^n\sqrt{k^2-l^2/r^2}}
 =-\frac{m\alpha k^{n-2}}{2\hbar^2l^{n-1}}
 \frac{\Gamma(1/2)\Gamma((n-1)/2)}{\Gamma(n/2)}.
 \tag{1}
\end{equation}

\SourcePage{631}{634}
(For the evaluation of the integral, compare Problem 5 in \S~126.) Replacing
the summation in (123.11) by integration, write
\[
 \sigma=\frac{4\pi}{k^2}\int_0^\infty2l\sin^2\delta_l\,dl.
\]
After the substitution $\delta_l=u$ and an integration by parts with respect
to $du$, the integral reduces to a $\Gamma$-function. The result is
\begin{equation}
 \sigma=\frac{2\pi}{n-1}\sin\left[\frac\pi2\left(\frac{n-3}{n-1}\right)\right]
 \Gamma\left(\frac{n-3}{n-1}\right)
 \left[\frac{\sqrt\pi\,\Gamma((n-1)/2)}{\Gamma(n/2)}\right]^{2/(n-1)}
 \left(\frac{m\alpha}{\hbar^2k}\right)^{2/(n-1)}.
 \tag{2}
\end{equation}
(For $n=3$, resolving the indeterminate form gives
$\sigma=2\pi^2\alpha m/\hbar v$.)

The first condition for the applicability of this result is that $l\gg1$
when $\delta_l\sim1$; hence
\[
 \frac{m\alpha k^{n-2}}{\hbar^2}\gg1.
\]
A further condition follows from requiring the field $U(r)$ to have the form
under consideration already at distances
\[
 r\sim l/k\sim(m\alpha/\hbar^2k)^{1/(n-1)}
\]
(with $l$ obtained from $\delta_l\sim1$), which make the main contribution to
integral (1). If this form is attained only at $r\gg a$, where $a$ is the
characteristic size of the field, the resulting condition is
\[
 \frac{m\alpha}{\hbar^2ka^{n-1}}\gg1,
\]
which sets an upper limit on the permissible velocities. Recall that in the
opposite case, at sufficiently high velocities, when
$m\alpha/\hbar^2ka^{n-1}\ll1$, the dependence is $\sigma\sim k^{-2}$
(compare Problem 5 in \S~126).

\ProblemHeading 2. Find the angular distribution of scattering near an
extremum of the classical scattering angle $\theta(\rho)$ as a function of
the impact parameter $\rho=l/k$.

\SolutionHeading The presence of an extremum of $\theta(l)$ at $l=l_0$
means, according to (127.3), that the phase near this point has the form
\[
 \delta_l\approx\delta_{l_0}+\frac{\theta_0}{2}l'+\frac\alpha3l'^3,
\]
where $\theta_0=\theta(l_0)$ and $l'=l-l_0$ (we again choose the case of the
lower sign in (127.3) for definiteness); the constant $\alpha$ is negative or
positive for a maximum or a minimum, respectively, of $\theta(l)$. Instead of
(127.6), the scattering amplitude is
\[
 |f(\theta)|=\frac1k\left(\frac{l_0}{2\pi\sin\theta_0}\right)^{1/2}
 \left|\int_{-\infty}^{\infty}\exp\left[i\left(\frac{2\alpha}3l'^3-\theta'l'\right)\right]dl'\right|,
\]
where $\theta'=\theta-\theta_0$. Expressing the integral in terms of the Airy
function by (b.3), we finally obtain for the scattering
cross-section\footnote{This type of scattering occurs in the theory of the
rainbow and is therefore called \emph{rainbow scattering}.}
\[
 d\sigma=\frac{4\pi l_0}{\alpha^{2/3}k^2}
 \Phi^2\left(-\frac{\theta'}{\alpha^{1/3}}\right)d\theta.
\]
