\SourcePage{212}{215}
\section{The Bohr--Sommerfeld quantization rule}

States in the discrete energy spectrum are semiclassical when the quantum
number $n$, which gives the ordinal number of the state, is large. Indeed,
this number determines the number of nodes of the eigenfunction (see §~21).
The distance between adjacent nodes, however, is of the order of the de
Broglie wavelength. For large $n$ this distance is small, and hence the
wavelength is small compared with the dimensions of the region of motion.

Let us derive the condition that determines the quantum energy levels in the
semiclassical case. Consider for this purpose the finite one-dimensional
motion of a particle in a potential well; the classically accessible region
$b\leq x\leq a$ is bounded by two turning points\footnote{In classical
mechanics a particle in such a field would execute periodic motion with period
(the time required to travel from $b$ to $a$ and back)
\[
  T=2\int_b^a\frac{dx}{v}
\]
($v$ is the particle velocity).}.

\SourcePage{213}{216}
By rule (47.5), the boundary condition at $x=b$ gives, in the region to its
right, the wave function
\begin{equation}
  \psi=\frac C{\sqrt p}\cos\left(\frac1\hbar\int_b^x p\,dx-\frac\pi4\right).
  \tag{48.1}
\end{equation}
Applying the same rule to the region to the left of $x=a$, we obtain the same
wave function in the form
\[
  \psi=\frac{C'}{\sqrt p}\cos\left(\frac1\hbar\int_x^a p\,dx-\frac\pi4\right).
\]
For these two expressions to agree throughout the region, the sum of their
phases, which is constant, must be an integral multiple of $\pi$:
\[
  \frac1\hbar\int_b^a p\,dx-\frac\pi2=n\pi
\]
(with $C'=(-1)^nC$). Hence
\begin{equation}
  \frac1{2\pi\hbar}\oint p\,dx=n+\frac12,
  \tag{48.2}
\end{equation}
where the integral $\oint p\,dx=2\int_b^a p\,dx$ is taken over a complete
period of the particle's classical motion. This is the condition determining
the stationary states of a particle in the semiclassical case. It corresponds
to the Bohr--Sommerfeld quantization rule of the old quantum theory.

The quantity $I=(1/2\pi)\oint p\,dx$ is called the adiabatic invariant (see
I, §~49), so the quantization condition (48.2) may be written
\[
  I(E)=\hbar(n+1/2).
\]
It was already mentioned in §~41 that, when the parameters vary sufficiently
slowly, ``adiabatically,'' the system remains in the same quantum state---in
the present case, the state with a particular $n$. We see that in the
semiclassical limit this statement becomes the classical theorem that the
adiabatic invariant remains constant under a slow change of the parameters.

\SourcePage{214}{217}
It is readily seen that the integer $n$ equals the number of zeros of the wave
function and is therefore the ordinal number of the stationary state. Indeed,
the phase of the wave function (48.1) increases from $-\pi/4$ at $x=b$ to
$(n+1/4)\pi$ at $x=a$, so the cosine vanishes $n$ times in this interval
(outside the interval $b\ll x\ll a$, the wave function decreases monotonically
and has no zeros at finite distances)\footnote{Strictly, the zeros must be
counted using the exact form of the wave function near the turning points.
Such an analysis confirms the stated result.}.

As explained above, $n$ is large in the semiclassical case. It should be
stressed, nevertheless, that retaining the term $1/2$ beside $n$ in (48.2) is
legitimate: including the next correction terms in the phase of the wave
functions would introduce on the right-hand side of (48.2) only terms of order
$\lambda/L$, small compared with unity (see the remark at the end of
§~46)\footnote{In some cases the exact expression for the energy levels
$E(n)$, as a function of the quantum number $n$, obtained from the exact
Schrödinger equation has a form that persists as $n\to\infty$. Examples are
the energy levels in a Coulomb field and those of the harmonic oscillator. It
is natural that in these cases the quantization rule (48.2), applicable for
large $n$, yields an expression for $E(n)$ that agrees with the exact one.}.

To normalize the wave function it is enough to integrate $|\psi|^2$ only over
$b\leq x\leq a$, since $\psi(x)$ decays exponentially outside this interval.
Because the argument of the cosine in (48.1) varies rapidly, the square of the
cosine may, with sufficient accuracy, be replaced by its mean value $1/2$.
We then obtain
\[
  \int|\psi|^2dx\approx\frac{C^2}{2}\int_b^a\frac{dx}{p(x)}
  =\frac{\pi C^2}{2m\omega}=1,
\]
where $\omega=2\pi/T$ is the frequency of the classical periodic motion. Thus
the normalized semiclassical function is
\begin{equation}
  \psi=2\sqrt{\frac{m\omega}{\pi p}}
  \cos\left(\frac1\hbar\int_b^x p\,dx-\frac\pi4\right).
  \tag{48.3}
\end{equation}
It must be borne in mind that the frequency $\omega$ is a function of the
energy and is, in general, different for different levels.

Relation (48.2) also admits another interpretation. The integral
$\oint p\,dx$ is the area enclosed by the closed

\SourcePage{215}{218}
classical phase trajectory of the particle (that is, the curve in the $p,x$
plane---the particle's phase space). Dividing this area into cells of area
$2\pi\hbar$ gives $n$ cells in all. But $n$ is the number of quantum states
whose energies do not exceed the specified value (the value corresponding to
the phase trajectory under consideration). Thus, in the semiclassical case,
we may say that every quantum state corresponds to a cell of area $2\pi\hbar$
in phase space. In other words, the number of states assigned to a phase-space
volume element $\Delta p\Delta x$ is
\begin{equation}
  \frac{\Delta p\Delta x}{2\pi\hbar}.
  \tag{48.4}
\end{equation}
If the wave vector $k=p/\hbar$ is used instead of the momentum, this number is
$\Delta k\Delta x/2\pi$. As expected, it agrees with the familiar expression
for the number of eigenmodes of a wave field (see II, §~52).

The general character of the distribution of levels in the energy spectrum
can be found from the quantization rule (48.2). Let $\Delta E$ be the spacing
between two adjacent levels, that is, levels whose quantum numbers $n$ differ
by unity. Since $\Delta E$ is small (for large $n$) compared with the level
energy itself, (48.2) gives
\[
  \Delta E\oint\frac{\partial p}{\partial E}\,dx=2\pi\hbar.
\]
But $\partial E/\partial p=v$, and therefore
\[
  \oint\frac{\partial p}{\partial E}\,dx=\oint\frac{dx}{v}=T.
\]
Consequently,
\begin{equation}
  \Delta E=\frac{2\pi\hbar}{T}=\hbar\omega.
  \tag{48.5}
\end{equation}

Thus the separation of adjacent levels is $\hbar\omega$. For a sequence of
neighboring levels (whose quantum numbers differ by amounts small compared
with the numbers themselves), the corresponding frequencies $\omega$ may be
regarded as approximately equal. We conclude that within each small segment
of the semiclassical part of the spectrum the levels are equidistant, with
equal intervals $\hbar\omega$. This result could, moreover, have been
anticipated, since in the semiclassical case the frequencies corresponding to
transitions between different energy levels must be integral multiples of the
classical frequency $\omega$.

\SourcePage{216}{219}
It is instructive to examine what the matrix elements of a physical quantity
$f$ become in the classical limit. We start from the fact that in this limit
the mean value of $f$ in a quantum state must become simply the classical
value of that quantity, provided the state itself tends to motion of the
particle along a definite trajectory. Such a state is represented by a wave
packet (see §~6) formed by superposing a number of stationary states with
nearby energies. Its wave function has the form
\[
  \psi=\sum_n a_n\psi_n,
\]
where the coefficients $a_n$ differ appreciably from zero only in an interval
$\Delta n$ of values of the quantum number $n$, such that
$1\ll\Delta n\ll n$; the numbers $n$ are assumed large, in accordance with
the semiclassical character of the stationary states. By definition, the mean
value of $f$ is
\[
  \bar f=\int\psi^*f\psi\,dx
  =\sum_n\sum_m a_m^*a_nf_{mn}e^{i\omega_{mn}t},
\]
or, on replacing summation over $n,m$ by summation over $n$ and the difference
$s=m-n$,
\[
  \bar f=\sum_n\sum_s a_{n+s}^*a_nf_{n+s,n}e^{i\omega_st},
\]
where $\omega_{mn}=\omega s$ in accordance with (48.5).

Matrix elements $f_{mn}$ calculated with semiclassical wave functions decrease
rapidly in magnitude as the difference $m-n$ increases, while at the same time
they vary slowly with $n$ itself (for a fixed $m-n$). We may therefore write,
approximately,
\[
  \bar f=\sum_n\sum_s a_n^*a_nf_se^{i\omega st}
  =\sum_n|a_n|^2\sum_sf_se^{i\omega st},
\]
where $f_s=f_{n+s,n}$ and $n$ now denotes some mean value of the quantum
number in the interval $\Delta n$. Since $\sum_n|a_n|^2=1$, it follows that
\[
  \bar f=\sum_sf_se^{i\omega st}.
\]

\SourcePage{217}{220}
The resulting sum has the form of an ordinary Fourier series. Since $\bar f$
must tend to the classical quantity $f(t)$, we find that in the limit the
matrix elements $f_{mn}$ become the components $f_{m-n}$ in the Fourier-series
expansion of the classical function $f(t)$.

Similarly, the matrix elements for transitions between continuous-spectrum
states become the components of the Fourier-integral expansion of $f(t)$. In
this case the wave functions of the stationary states must be normalized to a
delta-function of the energy divided by $\hbar$.

All the results given above extend directly to systems with many degrees of
freedom executing finite motion for which the mechanical (classical) problem
admits complete separation of variables in the Hamilton--Jacobi method (the
so-called conditionally periodic motion; see I, §~52). Once the variables have
been separated, the problem for each degree of freedom reduces to a
one-dimensional one, and the corresponding quantization conditions are
\begin{equation}
  \oint p_i\,dq_i=2\pi\hbar(n_i+\gamma_i),
  \tag{48.6}
\end{equation}
where the integral is taken over a period of variation of the generalized
coordinate $q_i$, while $\gamma_i$ is a number of order unity depending on the
nature of the boundary conditions for that degree of freedom\footnote{For
motion in a centrally symmetric field, for example,
\[
  \oint p_r\,dr=2\pi\hbar(n_r+1/2),\qquad
  \oint p_\theta\,d\theta=2\pi\hbar(l-m+1/2),\qquad
  \oint p_\varphi\,d\varphi=2\pi\hbar m
\]
(where $n_r=n-l-1$ is the radial quantum number). The last equality follows
simply because $p_\varphi$ is the $z$-component of the angular momentum, equal
to $\hbar m$.}.

For arbitrary (not conditionally periodic) multidimensional motion, a
formulation of the semiclassical quantization conditions requires a deeper
analysis\footnote{See $J.,B.~Keller$ // Ann. Phys. 1958. V.~4. P.~180.}. The
concept of phase-space ``cells,'' on the other hand, always applies in the
same form (within the semiclassical approximation). This is clear from its
connection, noted above, with the number of eigenmodes of a wave field in a
specified volume of space. In the general case of a system with $s$ degrees
of freedom, a phase-space volume element contains

\SourcePage{218}{221}
\begin{equation}
  \Delta N=\frac{\Delta q_1\cdots\Delta q_s\,
  \Delta p_1\cdots\Delta p_s}{(2\pi\hbar)^s}
  \tag{48.7}
\end{equation}
quantum states\footnote{In particular, for one particle,
$d^3p/(2\pi\hbar)^3$ is the number of states assigned to the interval $d^3p$
of momentum values per unit volume of space. This explains why the two
interpretations of the normalization of the plane wave (15.8), noted in the
footnote on p.~68, agree.}.

\ProblemHeading

\textbf{1.} Find approximately the number of discrete energy levels of a
particle moving in a field $U(r)$ that satisfies the semiclassical condition.

\SolutionHeading The number of states ``assigned'' to the phase-space volume
corresponding to momenta in the interval $0\leq p\leq p_{\max}$ and particle
coordinates in the volume element $dV$ is
\[
  \frac{(4\pi/3)p_{\max}^3dV}{(2\pi\hbar)^3}.
\]
At a specified $r$, the particle may have (in its classical motion) a momentum
satisfying $E=(p^2/2m)+U(r)\leq0$. Substitution of
$p_{\max}=\sqrt{-2mU(r)}$ gives the total number of states in the discrete
spectrum:
\[
  \frac{\sqrt2m^{3/2}}{3\pi^2\hbar^3}\int(-U)^{3/2}\,dV,
\]
where the integration extends over the region of space in which $U<0$. This
integral diverges (the number of levels is infinite) if at infinity $U$
decreases as $r^{-s}$ with $s<2$, in agreement with the results of §~18.

\textbf{2.} Find the same quantity for a semiclassical centrally symmetric
field $U(r)$ (V.~L.~Pokrovskii).

\SolutionHeading In a centrally symmetric field, the number of states does
not equal the number of energy levels, because the latter are degenerate with
respect to the directions of the angular momentum. The required number may be
found by observing that the number of levels for a specified angular momentum
$M$ equals the number of (nondegenerate) levels of one-dimensional motion with
potential energy $U_{\text{eff}}=U(r)+M^2/(2mr^2)$. At a specified $r$ and for
energies $E\leq0$, the greatest possible value of the momentum $p_r$ is
$p_{r\max}=\sqrt{-2mU_{\text{eff}}}$. Hence the number of states (that is, the
number of levels) is
\[
  \frac1{2\pi\hbar}\oint dr\,dp_r
  =\frac{\sqrt{2m}}{\pi\hbar}\int
  \sqrt{-U-\frac{M^2}{2mr^2}}\,dr.
\]
The required total number of discrete levels is obtained by integrating this
expression over $dM/\hbar$ (which replaces summation over $l$ in the
semiclassical case), and is
\[
  \frac{\sqrt{2m}}{4\hbar^2}\int(-U)r\,dr.
\]
