\SourcePage{205}{208}
\chapter{The Semiclassical Case}
\section{The Wave Function in the Semiclassical Case}

When the de Broglie wavelengths of the particles are small compared with the
characteristic dimensions $L$ that determine the conditions of the particular
problem, the system behaves nearly classically. (This is analogous to the
passage from wave optics to geometrical optics as the wavelength tends to
zero.)

We shall now examine the properties of semiclassical systems in greater
detail. In the Schr\"odinger equation
\[
  \sum_a\frac{\hbar^2}{2m_a}\Delta_a\psi+(E-U)\psi=0
\]
make the formal substitution
\begin{equation}
  \psi=\exp\left(\frac i\hbar\sigma\right).
  \tag{46.1}
\end{equation}
The resulting equation for $\sigma$ is
\begin{equation}
  \sum_a\frac1{2m_a}(\nabla_a\sigma)^2
  -\sum_a\frac{i\hbar}{2m_a}\Delta_a\sigma=E-U.
  \tag{46.2}
\end{equation}
Since the system is assumed to have almost classical properties, we seek
$\sigma$ as a series
\begin{equation}
  \sigma=\sigma_0+\frac\hbar i\sigma_1+
  \left(\frac\hbar i\right)^2\sigma_2+\cdots,
  \tag{46.3}
\end{equation}
in powers of $\hbar$.

We begin with the simplest situation: one-dimensional motion of a single
particle. Equation (46.2) then reduces to
\begin{equation}
  \frac1{2m}\sigma'^2-\frac{i\hbar}{2m}\sigma''=E-U(x)
  \tag{46.4}
\end{equation}
(a prime denotes differentiation with respect to $x$).

In the first approximation, put $\sigma=\sigma_0$ and omit the term containing
$\hbar$ from the equation:
\[
  \frac1{2m}\sigma_0'^2=E-U(x).
\]

\SourcePage{206}{209}
It follows that
\[
  \sigma_0=\pm\int\sqrt{2m[E-U(x)]}\,dx.
\]
The integrand is precisely the classical momentum $p(x)$ of the particle,
expressed as a function of position. Defining $p(x)$ by taking the positive
sign of the square root, we have
\begin{equation}
  \sigma_0=\pm\int p\,dx,
  \qquad p=\sqrt{2m(E-U)},
  \tag{46.5}
\end{equation}
as is to be expected from the limiting expression (6.1) for the wave
function\footnote{As is well known, $\int p\,dx$ is the time-independent part
of the action. The complete mechanical action of the particle is
$S=-Et\pm\int p\,dx$. The term $Et$ is absent from $\sigma$ because we are
considering a time-independent wave function $\psi$.}.

The neglect made in (46.4) is justified only when the second term on the left
is small compared with the first. Thus
$\hbar|\sigma''/\sigma'^2|\ll1$, or
\[
  \left|\frac d{dx}\left(\frac\hbar{\sigma'}\right)\right|\ll1.
\]
To the first approximation, (46.5) gives $\sigma'=p$, and hence this condition
may be written
\begin{equation}
  \left|\frac{d\lambdabar}{dx}\right|\ll1,
  \tag{46.6}
\end{equation}
where $\lambdabar=\lambda/2\pi$, while
$\lambda(x)=2\pi\hbar/p(x)$ is the particle's de Broglie wavelength expressed
as a function of $x$ through the classical function $p(x)$. We have therefore
obtained a quantitative criterion for the semiclassical approximation: over a
distance of the order of the wavelength itself, the wavelength must vary only
slightly. The approximation ceases to apply in regions where this condition
is not fulfilled.

Condition (46.6) can also be put into another form. Since
\[
  \frac{dp}{dx}=\frac d{dx}\sqrt{2m(E-U)}=-\frac mp\frac{dU}{dx}=\frac{mF}{p},
\]
where $F=-dU/dx$ is the classical force exerted on the particle by the
external field, introducing this force gives
\begin{equation}
  \frac{m\hbar|F|}{p^3}\ll1.
  \tag{46.7}
\end{equation}

\SourcePage{207}{210}
This shows that the semiclassical approximation fails when the particle's
momentum becomes too small. In particular, it necessarily fails near turning
points, that is, near points at which classical mechanics would have the
particle stop and then reverse its direction of motion. Such points are
determined by $p(x)=0$, or $E=U(x)$. As $p\to0$, the de Broglie wavelength
tends to infinity and plainly cannot be regarded as small.

It should be stressed, however, that condition (46.6) or (46.7) may by itself
be insufficient to justify the semiclassical approximation. It was obtained
by estimating the different terms of the differential equation (46.4), whose
discarded term contains the highest derivative. What must actually be small
are successive terms in the expansion of the solution of this equation, and
smallness of the discarded term in the equation need not ensure this. For
example, suppose that the solution for $\sigma(x)$ contains a term that grows
with $x$ in an almost linear fashion. The smallness of its second derivative
does not prevent that term from accumulating a large value over a sufficiently
long distance. This situation generally arises when the field extends over
distances large compared with the characteristic length $L$ on which it
changes appreciably (see the remark below concerning (46.11)). The
semiclassical approximation then cannot be used to follow the wave function
over large distances.

We proceed to calculate the next term of expansion (46.3). Terms of first
order in $\hbar$ in (46.4) give
\[
  \sigma_0'\sigma_1'+\frac12\sigma_0''=0,
  \qquad
  \sigma_1'=-\frac{\sigma_0''}{2\sigma_0'}=-\frac{p'}{2p}.
\]
Integration yields
\begin{equation}
  \sigma_1=-\frac12\ln p
  \tag{46.8}
\end{equation}
(the integration constant is omitted).

Substitution into (46.1) and (46.3) gives the wave function
\begin{equation}
  \psi=\frac C{\sqrt p}\exp\left(\frac i\hbar\int p\,dx\right)
  +\frac{C'}{\sqrt p}\exp\left(-\frac i\hbar\int p\,dx\right).
  \tag{46.9}
\end{equation}
The factor $1/\sqrt p$ has a simple interpretation. The probability of finding
the particle between $x$ and $x+dx$ is determined by $|\psi|^2$ and is

\SourcePage{208}{211}
therefore, in the main, proportional to $1/p$. This is just what one expects
for a ``semiclassical particle,'' since in classical motion the time spent by
the particle in an interval $dx$ is inversely proportional to its velocity
(or momentum).

In classically forbidden regions, where $E<U(x)$, the function $p(x)$ is
purely imaginary and the exponential arguments are real. The general form of
the wave-equation solution in these regions is
\begin{equation}
  \psi=\frac C{\sqrt{|p|}}\exp\left(-\frac1\hbar\int|p|\,dx\right)
  +\frac{C'}{\sqrt{|p|}}\exp\left(\frac1\hbar\int|p|\,dx\right).
  \tag{46.10}
\end{equation}
One must bear in mind, however, that the accuracy of the semiclassical
approximation does not permit exponentially small terms in the wave function
to be retained against a background of exponentially large ones. In this
sense, keeping both terms of (46.10) at once is generally inadmissible.

Higher-order terms in the wave function are ordinarily unnecessary. We shall
nevertheless find one further term of expansion (46.3), in order to point out
some features concerning the accuracy of the semiclassical approximation.

The terms of order $\hbar^2$ in (46.4) give
\[
  \sigma_0'\sigma_2'+\frac12\sigma_1'^2+\frac12\sigma_1''=0,
\]
and therefore, on substituting (46.5) and (46.8) for $\sigma_0$ and
$\sigma_1$,
\[
  \sigma_2'=\frac{p''}{4p^2}-\frac{3p'^2}{8p^3}.
\]
Integrating (the first term being integrated by parts) and introducing the
force $F=pp'/m$, we obtain
\[
  \sigma_2=\frac{mF}{4p^3}+\frac{m^2}{8}\int\frac{F^2}{p^5}\,dx.
\]
In this approximation the wave function is
\[
  \psi=\exp[(i/\hbar)\sigma]
  =\exp[(i/\hbar)\sigma_0+\sigma_1](1-i\hbar\sigma_2),
\]
or
\begin{equation}
  \psi=\frac{\operatorname{const}}{\sqrt p}
  \left[1-\frac{im\hbar F}{4p^3}
  -\frac{im^2\hbar}{8}\int\frac{F^2}{p^5}\,dx\right]
  \exp\left(\frac i\hbar\int p\,dx\right).
  \tag{46.11}
\end{equation}
The appearance of imaginary correction terms in the prefactor is equivalent
to a real correction to the phase of the wave function (that is, an addition
to the integral

\SourcePage{209}{212}
$\hbar^{-1}\int p\,dx$ in its exponential). This correction is proportional
to $\hbar$, and thus has order $\lambda/L$.

The second and third terms in the square brackets in (46.11) must be small
compared with 1. For the former this requirement is the same as (46.7); for
the latter, estimating the integral leads to (46.7) only if $F^2$ tends to
zero sufficiently rapidly over distances $\sim L$.
