\SourcePage{212}{215}
\section{The Bohr--Sommerfeld quantization rule}

States belonging to the discrete energy spectrum become quasiclassical when
the quantum number $n$, which gives the ordinal number of the state, is large.
Indeed, this number specifies the number of nodes of the eigenfunction (see
§~21). The separation of two successive nodes, however, has the same order of
magnitude as the de Broglie wavelength. Hence, for large $n$ this separation
is small, and the wavelength is short in comparison with the dimensions of the
region of motion.

Let us derive the condition that fixes the quantum energy levels in the
quasiclassical case. Consider a particle executing bounded one-dimensional
motion in a potential well. The classically accessible interval
$b\leq x\leq a$ is bounded by two turning points\footnote{In classical
mechanics a particle in this field would move periodically, with period (the
time required to travel from $b$ to $a$ and back)
\[
  T=2\int_b^a\frac{dx}{v}
\]
($v$ is the particle velocity).}.

\SourcePage{213}{216}
By rule (47.5), the boundary condition at $x=b$ gives, in the region to its
right, the wave function
\begin{equation}
  \psi=\frac C{\sqrt p}\cos\left(\frac1\hbar\int_b^x p\,dx-\frac\pi4\right).
  \tag{48.1}
\end{equation}
Applying the same rule to the region on the left of $x=a$, we obtain another
form of the same wave function:
\[
  \psi=\frac{C'}{\sqrt p}\cos\left(\frac1\hbar\int_x^a p\,dx-\frac\pi4\right).
\]
For these expressions to agree throughout the interval, the sum of their
phases, which is constant, must be an integral multiple of $\pi$:
\[
  \frac1\hbar\int_b^a p\,dx-\frac\pi2=n\pi
\]
(with $C'=(-1)^nC$). It follows that
\begin{equation}
  \frac1{2\pi\hbar}\oint p\,dx=n+\frac12,
  \tag{48.2}
\end{equation}
where $\oint p\,dx=2\int_b^a p\,dx$ is evaluated over a complete period of
the particle's classical motion. This is the condition determining the
stationary states in the quasiclassical regime. It corresponds to the
Bohr--Sommerfeld quantization rule of the old quantum theory.

The quantity $I=(1/2\pi)\oint p\,dx$ is called the adiabatic invariant (see
I, §~49), and therefore (48.2) may be written
\[
  I(E)=\hbar(n+1/2).
\]
It was already noted in §~41 that, if the parameters change sufficiently
slowly, or ``adiabatically,'' a system remains in the same quantum state---in
the present case, the state having a particular $n$. We now see that in the
quasiclassical limit this statement becomes the classical theorem asserting
the constancy of the adiabatic invariant under a slow change of parameters.

\SourcePage{214}{217}
It is readily seen that the integer $n$ equals the number of zeros of the wave
function and is therefore the ordinal number of the stationary state. In
fact, the phase in (48.1) increases from $-\pi/4$ at $x=b$ to
$(n+1/4)\pi$ at $x=a$, so the cosine vanishes $n$ times within this interval
(outside the range $b\ll x\ll a$, the wave function decreases monotonically
and has no zeros at finite distances)\footnote{Strictly, the zeros must be
counted using the exact form of the wave function near the turning points. An
analysis of this kind confirms the stated result.}.

As explained above, $n$ is large in the quasiclassical case. Nevertheless,
retaining the term $1/2$ beside $n$ in (48.2) is legitimate. Inclusion of the
next correction terms in the phases of the wave functions would add to the
right-hand side of (48.2) only terms of order $\lambda/L$, which are small
compared with unity (see the remark at the end of §~46)\input{manuscript/chapters/ch07/fragments/sec48-footnote074-080-fresh-i-20260718.translation.en.tex}.

To normalize the wave function, it is sufficient to integrate $|\psi|^2$
only over $b\leq x\leq a$, since $\psi(x)$ decays exponentially outside this
interval. The argument of the cosine in (48.1) varies rapidly, so to the
required accuracy its square may be replaced by its mean value, $1/2$. Thus
\[
  \int|\psi|^2dx\approx\frac{C^2}{2}\int_b^a\frac{dx}{p(x)}
  =\frac{\pi C^2}{2m\omega}=1,
\]
where $\omega=2\pi/T$ is the frequency of the classical periodic motion. The
normalized quasiclassical function is consequently
\begingroup\small
\begin{equation}
  \psi=2\sqrt{\frac{m\omega}{\pi p}}
  \cos\left(\frac1\hbar\int_b^x p\,dx-\frac\pi4\right).
  \tag{48.3}
\end{equation}
\endgroup
It should be kept in mind that $\omega$ depends on the energy and, in
general, differs from one level to another.

\input{manuscript/chapters/ch07/fragments/sec48-src101-121-fresh-d-20260718.translation.en.tex}

The general pattern in which levels are distributed through the energy
spectrum can be deduced from the quantization rule (48.2). Let $\Delta E$
denote the separation of two adjacent levels, whose quantum numbers $n$ thus
differ by one. Since, for large $n$, $\Delta E$ is small in comparison with
the level energy itself, (48.2) gives
\[
  \Delta E\oint\frac{\partial p}{\partial E}\,dx=2\pi\hbar.
\]
But $\partial E/\partial p=v$, and hence
\[
  \oint\frac{\partial p}{\partial E}\,dx=\oint\frac{dx}{v}=T.
\]
We therefore find
\begin{equation}
  \Delta E=\frac{2\pi\hbar}{T}=\hbar\omega.
  \tag{48.5}
\end{equation}

Thus the spacing of adjacent levels is $\hbar\omega$. For a sequence of
nearby levels whose differences in $n$ are small compared with $n$ itself,
the corresponding frequencies $\omega$ may be regarded as approximately
equal. It follows that, within any small portion of the quasiclassical part of
the spectrum, the levels are equidistant, with the common interval
$\hbar\omega$.
\input{manuscript/chapters/ch07/fragments/sec48-src147-151-fresh-j-20260718.translation.en.tex}
It is useful to trace the limiting behavior of the matrix elements of an
arbitrary physical quantity $f$. We start from the requirement that the mean
value of $f$ in a quantum state must tend simply to its classical value,
provided that, in the limit, the state itself describes motion of the particle
along a definite trajectory. Such a state is represented by a wave packet
(see §~6), formed by superposing stationary states with neighboring energies.
Its wave function has the form
\[
  \psi=\sum_n a_n\psi_n,
\]
where the coefficients $a_n$ differ appreciably from zero only within an
interval $\Delta n$ of quantum-number values such that
$1\ll\Delta n\ll n$. The numbers $n$ are assumed large, as required for the
stationary states to be quasiclassical. By definition, the mean value of $f$
is
\[
  \bar f=\int\psi^*f\psi\,dx
  =\sum_n\sum_m a_m^*a_nf_{mn}e^{i\omega_{mn}t},
\]
or, replacing the summation over $n,m$ by summations over $n$ and the
difference $s=m-n$, we obtain
\[
  \bar f=\sum_n\sum_s a_{n+s}^*a_nf_{n+s,n}e^{i\omega_st},
\]
where, in accordance with (48.5), $\omega_{mn}=\omega s$.

When evaluated with quasiclassical wave functions, the matrix elements
$f_{mn}$ decrease rapidly in magnitude as the difference $m-n$ grows, while
varying slowly with $n$ itself when $m-n$ is fixed. We may consequently write,
to the same approximation,
\[
  \bar f=\sum_n\sum_s a_n^*a_nf_se^{i\omega st}
  =\sum_n|a_n|^2\sum_sf_se^{i\omega st},
\]
where $f_s=f_{n+s,n}$ and $n$ denotes some mean value of the quantum number in
the interval $\Delta n$. Since $\sum_n|a_n|^2=1$, it follows that
\[
  \bar f=\sum_sf_se^{i\omega st}.
\]

\SourcePage{217}{220}
The resulting sum has the form of an ordinary Fourier series. Since $\bar f$
must approach the classical quantity $f(t)$ in the limit, we conclude that
the matrix elements $f_{mn}$ tend to the coefficients $f_{m-n}$ in the
Fourier-series expansion of the classical function $f(t)$.

Similarly, matrix elements for transitions between states in the continuous
spectrum tend to the components in the Fourier-integral expansion of $f(t)$.
Here the wave functions of the stationary states must be normalized to a
delta function of the energy divided by $\hbar$.

\input{manuscript/chapters/ch07/fragments/sec48-src204-224-fresh-c-20260718.translation.en.tex}

For arbitrary multidimensional motion that is not conditionally periodic, a
formulation of the quasiclassical quantization conditions requires a more
subtle argument\footnote{See $J.,B.~Keller$ // Ann. Phys. 1958. V.~4.
P.~180.}. The notion of phase-space ``cells,'' however, always applies in the
same manner within the quasiclassical approximation. This is clear from the
connection, noted above, between this notion and the number of normal modes of
a wave field in a given spatial volume. In the general case of a system with
$s$ degrees of freedom, a phase-space volume element

\SourcePage{218}{221}
contains
\begin{equation}
  \Delta N=\frac{\Delta q_1\cdots\Delta q_s\,
  \Delta p_1\cdots\Delta p_s}{(2\pi\hbar)^s}
  \tag{48.7}
\end{equation}
quantum states\footnote{In particular, $d^3p/(2\pi\hbar)^3$ for a single
particle is the number of states associated with a momentum interval $d^3p$
in a unit volume of space. This accounts for the agreement between the two
interpretations of the plane-wave normalization (15.8) noted in the footnote
on p.~68.}.

\ProblemHeading

\input{manuscript/chapters/ch07/fragments/sec48-problem249-250-fresh-i-20260718.translation.en.tex}

\SolutionHeading The number of states ``belonging'' to the phase-space volume
for momenta in the range $0\leq p\leq p_{\max}$ and particle coordinates in
the volume element $dV$ is
\[
  \frac{(4\pi/3)p_{\max}^3dV}{(2\pi\hbar)^3}.
\]
At a given $r$, the particle may have, in its classical motion, any momentum
that obeys $E=(p^2/2m)+U(r)\leq0$. Substitution of
$p_{\max}=\sqrt{-2mU(r)}$ gives the total number of discrete-spectrum states:
\[
  \frac{\sqrt2m^{3/2}}{3\pi^2\hbar^3}\int(-U)^{3/2}\,dV,
\]
where the integration extends over the part of space in which $U<0$. This
integral diverges, and the number of levels is infinite, if $U$ decreases at
infinity as $r^{-s}$ with $s<2$, in agreement with the results of §~18.

\textbf{2.} Find the corresponding number for a quasiclassical central field
$U(r)$ (V.~L.~Pokrovskii).

\input{manuscript/chapters/ch07/fragments/sec48-src271-289-fresh-f-20260718.translation.en.tex}
