\SourcePage{219}{222}
\section{Semiclassical motion in a centrally symmetric field}

For motion in a centrally symmetric field, the particle's wave function
separates, as we know, into angular and radial parts. We first consider the
angular part.

The dependence of the angular wave function on the angle $\varphi$ (specified
by the quantum number $m$) is so simple that there is no question of seeking
approximate formulas for it. Its dependence on the polar angle $\theta$, on
the other hand, is semiclassical according to the general rule if the
corresponding quantum number $l$ is large (a more precise statement of this
condition is given below).

Here we shall derive a semiclassical expression for the angular function only
in the case of greatest practical importance, namely states with zero magnetic
quantum number ($m=0$)\footnote{The opposite case, $m=l$, must in the limit
correspond to motion in a classical orbit lying in the equatorial plane
$\theta=\pi/2$. Indeed,
\[
  P_l^l(\cos\theta)=\operatorname{const}\cdot\sin^l\theta;
\]
as $l\to\infty$, this function (and with it $|\psi|^2$) tends to zero for all
$\theta\ne\pi/2$.}. Apart from a constant factor, this function is the
Legendre polynomial $P_l(\cos\theta)$ (see (28.8)) and satisfies the
differential equation
\begin{equation}
  \frac{d^2P_l}{d\theta^2}+\cot\theta\frac{dP_l}{d\theta}
  +l(l+1)P_l=0.
  \tag{49.1}
\end{equation}
The substitution
\begin{equation}
  P_l=\frac{\chi}{\sqrt{\sin\theta}}
  \tag{49.2}
\end{equation}
reduces it to the equation
\begin{equation}
  \chi''+\left[\left(l+\frac12\right)^2+
  \frac{1}{4\sin^2\theta}\right]\chi=0,
  \tag{49.3}
\end{equation}
which contains no first derivative and has the form of a one-dimensional
Schrödinger equation.

In (49.3), the role of the ``de Broglie wavelength'' is played by
$\lambda\sim1/l$.

\SourcePage{220}{223}
The requirement that the derivative $d\lambda/dx$ be small (condition (46.6))
gives the inequalities
\begin{equation}
  \theta l\gg1,\qquad(\pi-\theta)l\gg1
  \tag{49.4}
\end{equation}
(the semiclassical conditions for the angular part of the wave function). For
large $l$ these conditions hold over almost the whole interval of values of
$\theta$, except only for angles very close to zero or to $\pi$.

When condition (49.4) holds, the second term in the square brackets in (49.3)
may be neglected in comparison with the first:
\[
  \chi''+\left(l+\frac12\right)^2\chi=0.
\]
The solution of this equation is
\begin{equation}
  \chi=\sqrt{\sin\theta}\,P_l(\cos\theta)
  =A\sin\left[\left(l+\frac12\right)\theta+\alpha\right]
  \tag{49.5}
\end{equation}
($A$ and $\alpha$ are constants).

For angles $\theta\ll1$, we may set $\cot\theta\approx1/\theta$ in (49.1).
Also replacing $l(l+1)$ approximately by $(l+1/2)^2$, we obtain
\[
  \frac{d^2P_l}{d\theta^2}+\frac1\theta\frac{dP_l}{d\theta}
  +\left(l+\frac12\right)^2P_l=0,
\]
whose solution is the Bessel function of order zero,
\begin{equation}
  P_l(\cos\theta)=J_0\left[\left(l+\frac12\right)\theta\right],
  \qquad\theta\ll1.
  \tag{49.6}
\end{equation}
The constant factor has been set equal to unity because $P_l$ must equal one
at $\theta=0$. The approximate expression (49.6) for $P_l$ is valid for all
angles $\theta\ll1$. In particular, it may be used in the region
$1/l\ll\theta\ll1$, where it must agree with (49.5), which is valid whenever
$\theta\gg1/l$. When $\theta l\gg1$, the Bessel function may be replaced by
its asymptotic expression for large argument, giving
\[
  P_l\approx\sqrt{\frac{2}{\pi l\theta}}
  \sin\left[\left(l+\frac12\right)\theta+\frac\pi4\right]
\]
(in the coefficient, $1/2$ may be neglected in comparison with $l$).
Comparison with (49.5) gives $A=\sqrt{2/\pi l}$ and $\alpha=\pi/4$. We thus
arrive at the following expression for $P_l(\cos\theta)$,

\SourcePage{221}{224}
applicable in the semiclassical case\footnote{Notice that it is precisely the
replacement of $l(l+1)$ by $(l+1/2)^2$ that has yielded an expression which
is multiplied by $(-1)^l$ when $\theta$ is replaced by $\pi-\theta$, as the
function $P_l(\cos\theta)$ must be.}:
\begin{equation}
  P_l(\cos\theta)=\sqrt{\frac{2}{\pi l\sin\theta}}
  \sin\left[\left(l+\frac12\right)\theta+\frac\pi4\right].
  \tag{49.7}
\end{equation}
The normalized spherical function $Y_{l0}$ therefore takes the form (compare
(28.8))
\begin{equation}
  Y_{l0}=\frac{1}{\pi\sqrt{\sin\theta}}
  \sin\left[\left(l+\frac12\right)\theta+\frac\pi4\right].
  \tag{49.8}
\end{equation}

We now turn to the radial part of the wave function. It was shown in §~32 that
the function $\chi(r)=rR(r)$ satisfies an equation identical with the
one-dimensional Schrödinger equation having potential energy
\[
  U_l(r)=U(r)+\frac{\hbar^2}{2mr^2}l(l+1).
\]
We may therefore apply the results of the preceding sections, with the
function $U_l(r)$ understood as the potential energy.

The case $l=0$ is the simplest. The centrifugal energy is absent, and if the
field $U(r)$ satisfies the required condition (46.6), the radial wave function
is semiclassical throughout space. At $r=0$ we must have $\chi=0$, and hence
the semiclassical function $\chi(r)$ is determined in accordance with formulas
(47.6).

If $l\ne0$, the centrifugal energy must also satisfy condition (46.6). In the
region of small $r$, where the centrifugal energy is of the order of the total
energy, the wavelength is $\lambda=\hbar/p\sim r/l$, and condition (46.6)
gives $l\gg1$. Thus, when $l$ is not large, the centrifugal energy violates
the semiclassical condition in the region of small $r$. It is readily verified
that the correct phase of the semiclassical wave function $\chi(r)$ is
obtained by using the formulas for one-dimensional motion and replacing the
coefficient $l(l+1)$ in the potential energy $U_l(r)$

\SourcePage{222}{225}
by $(l+1/2)^2$\footnote{Thus, in the simplest case of free motion ($U=0$), the
phase of the function calculated from (48.1), with $U_l$ from (49.9), agrees
at large $r$, as it should, with the phase of the function (33.12).}:
\begin{equation}
  U_l(r)=U(r)+\frac{\hbar^2}{2mr^2}\left(l+\frac12\right)^2.
  \tag{49.9}
\end{equation}

The applicability of the semiclassical approximation to a Coulomb field
$U=\pm\alpha/r$ requires separate consideration. Of the entire region of
motion, the most important part corresponds to distances $r$ for which
$|U|\sim|E|$, that is, $r\sim\alpha/|E|$. The semiclassical condition for
motion in this region reduces to requiring that the wavelength
$\lambdabar\sim\hbar/\sqrt{2m|E|}$ be small compared with the size
$\alpha/|E|$ of the region. This gives
\begin{equation}
  |E|\ll\frac{m\alpha^2}{\hbar^2},
  \tag{49.10}
\end{equation}
\input{manuscript/chapters/ch07/fragments/sec49-bohr.fresh.translation.en.tex}
Condition (49.10) may also be
written
\begin{equation}
  \frac{\hbar v}{\alpha}\gg1,
  \tag{49.11}
\end{equation}
where $v\sim\sqrt{|E|/m}$ is the particle velocity. Notice that this condition
is the reverse of condition (45.7) for the applicability of perturbation
theory to the Coulomb field.

As for the region of small distances ($|U(r)|\gg E$), in a repulsive Coulomb
field it is of no interest at all, since for $U>E$ the semiclassical wave
functions decay exponentially. In an attractive field with small $l$, however,
the particle may penetrate into a region where $|U|\gg|E|$, and the limits of
applicability of the semiclassical approximation there must be determined. We
use the general condition (46.7), putting
\[
  F=-\frac{dU}{dr}=\frac{\alpha}{r^2},\qquad
  p\approx\sqrt{2m|U|}\sim\sqrt{\frac{m\alpha}{r}}.
\]
We then find that the region of applicability of the semiclassical
approximation is restricted to distances
\begin{equation}
  r\gg\frac{\hbar^2}{m\alpha},
  \tag{49.12}
\end{equation}
that is, distances large compared with the ``radius'' of the first Bohr orbit.

\SourcePage{223}{226}
\ProblemHeading

Find the behavior of the wave function near the origin if, as $r\to0$, the
field tends to infinity as $\pm\alpha/r^s$, where $s>2$.

\SolutionHeading For sufficiently small $r$, the wavelength is
\[
  \lambdabar\sim\frac{\hbar}{\sqrt{m|U|}}
  \sim\frac{\hbar r^{s/2}}{\sqrt{m\alpha}},
\]
and therefore
\[
  \frac{d\lambdabar}{dr}\sim
  \frac{\hbar r^{s/2-1}}{\sqrt{m\alpha}}\ll1;
\]
thus the semiclassical condition is fulfilled. In an attractive field,
$U\to-\infty$ as $r\to0$. The region near the origin is then classically
accessible, and the radial wave function $\chi\sim1/\sqrt p$, whence
\[
  \psi\sim r^{s/4-1}.
\]
In a repulsive field the region of small $r$ is classically inaccessible. The
wave function then tends exponentially to zero as $r\to0$. Omitting the factor
multiplying the exponential function, we have
\[
  \psi\sim\exp\left[-\frac{2\sqrt{m\alpha}}{(s-2)\hbar}
  r^{-(s/2-1)}\right].
\]
