\SourcePage{223}{226}
\section{Passage through a potential barrier}

Consider a particle moving in a field of the type shown in Fig.~13. The field
contains a \emph{potential barrier}: an interval in which the potential energy
$U(x)$ is greater than the particle's total energy $E$. Classical mechanics
does not allow the particle to penetrate such a barrier. In quantum mechanics,
however, there is a nonzero probability that it will pass ``through the
barrier''; this phenomenon is also called the \emph{tunnel effect}\footnote{We
have already encountered examples of this kind in Problems 2 and 4 of
\S~25.}. When $U(x)$ satisfies the semiclassical conditions, a general
expression can be obtained for the barrier transmission coefficient. These
conditions imply, in particular, that the barrier is broad; consequently its
semiclassical transmission coefficient is small.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1.05cm,y=1.0cm]
  \draw[->] (-2.5,0)--(2.8,0) node[right] {$x$};
  \draw[->] (-1.7,-0.15)--(-1.7,2.25) node[right] {$U(x)$};
  \draw[dashed] (-2.3,.85)--(2.4,.85) node[above right] {$E$};
  \draw[thick] (-2.35,.18) .. controls (-1.55,.18) and (-1.2,1.65) .. (0,1.75)
    .. controls (1.0,1.65) and (1.45,.5) .. (2.35,.22);
  \draw[dashed] (-1.4,.85)--(-1.4,0) node[below=2pt] {$a$};
  \draw[dashed] (1.42,.85)--(1.42,0) node[below=2pt] {$b$};
  \node at (-2.05,.62) {$I$}; \node at (0,.62) {$II$}; \node at (1.9,.62) {$III$};
\end{tikzpicture}
\par\smallskip Fig. 13
\end{figure}

We first solve an auxiliary problem so that it will not interrupt the
calculation below. Suppose the semiclassical wave function to the right of the
turning point $x=b$ (where

\SourcePage{224}{227}
$U(x)<E$) is the travelling wave
\begin{equation}
 \psi=\frac{C}{\sqrt p}\exp\left(\frac{i}{\hbar}\int_b^x p\,dx+\frac{i\pi}{4}\right).
 \tag{50.1}
\end{equation}
We seek the wave function of the same state in the region $x<b$. We shall use
the same continuation in the complex $x$-plane as in \S~47.

Writing
\[
 E-U(x)\approx F_0(x-b),\qquad F_0>0,
\]
we put (50.1) in the form
\[
 \psi(x)=\frac{C}{(2mF_0)^{1/4}}\frac1{(x-b)^{1/4}}
 \exp\left\{\frac{i}{\hbar}(2mF_0)^{1/2}
 \int_b^x\sqrt{x-b}\,dx+\frac{i\pi}{4}\right\}.
\]
Continue this expression from right to left round a semicircle in the upper
half-plane:
\[
 x-b=\rho e^{i\varphi},\qquad
 i\int_b^x\sqrt{x-b}\,dx=\frac23\rho^{3/2}
 \left(-\sin\frac{3\varphi}{2}+i\cos\frac{3\varphi}{2}\right),
\]
with the phase $\varphi$ changing from 0 to $\pi$. During this continuation,
$|\psi(x)|$ first decreases and then increases. At its end the function is
\[
 \psi(x)=\frac{C}{(2mF_0)^{1/4}}\frac1{(b-x)^{1/4}e^{i\pi/4}}
 \exp\left\{\frac1\hbar\int_x^b(2mF_0)^{1/2}\sqrt{b-x}\,dx+\frac{i\pi}{4}\right\}.
\]
We have therefore obtained the connection rule\footnote{If the continuation
from right to left is instead made through the lower half-plane, $|\psi(x)|$
first grows and then falls. On the negative half-axis ($\varphi\to-\pi$) it
becomes exponentially small, and retaining it against the exponentially large
function (50.2) would be illegitimate. Over the part of the path on which
$\psi(x)$ is exponentially large, the inaccuracy of the semiclassical
approximation loses an exponentially small addition. As $\varphi\to-\pi$ this
addition could have become an exponentially large term, which is therefore
lost as well.}:
\begin{equation}
 \left.\frac{C}{\sqrt p}\exp\left\{\frac{i}{\hbar}\int_b^x p\,dx+
 \frac{i\pi}{4}\right\}\right|_{x>b}
 \longrightarrow
 \left.\frac{C}{\sqrt{|p|}}\exp\left\{\frac1\hbar
 \left|\int_b^x p\,dx\right|\right\}\right|_{x<b}.
 \tag{50.2}
\end{equation}

\SourcePage{225}{228}
It is important that this rule presupposes a particular wave in the
classically allowed region, namely one travelling to the right, and is to be
used specifically for continuation from that region into the classically
forbidden one.

We now return to the barrier transmission coefficient. Let a particle approach
the barrier from region $I$, moving from left to right. Beyond the barrier, in
region $III$, there is only the transmitted wave, which also travels to the
right. Write its wave function as
\begin{equation}
 \psi=\sqrt{\frac Dv}\exp\left(\frac{i}{\hbar}\int_b^x p\,dx+\frac{i\pi}{4}\right),
 \tag{50.3}
\end{equation}
where $v=p/m$ is the particle velocity and $D$ is the current density in the
wave. Rule (50.2) now gives the wave function within the barrier, in region
$II$:
\begin{equation}
 \psi=\sqrt{\frac D{|v|}}\exp\left(\frac1\hbar\left|\int_x^b p\,dx\right|\right)
 =\sqrt{\frac D{|v|}}\exp\left(\frac1\hbar\left|\int_a^b p\,dx\right|
 -\frac1\hbar\left|\int_a^x p\,dx\right|\right).
 \tag{50.4}
\end{equation}
Finally, rule (47.5) yields, in region $I$ before the barrier,
\[
 \psi=2\sqrt{\frac Dv}\exp\left(\frac1\hbar\int_a^b|p|\,dx\right)
 \cos\left(\frac1\hbar\int_x^a p\,dx-\frac\pi4\right).
\]
If we set
\begin{equation}
 D=\exp\left(-\frac2\hbar\int_a^b|p|\,dx\right),
 \tag{50.5}
\end{equation}
this function becomes
\[
 \begin{split}
 \psi&=\frac2{\sqrt v}\cos\left(\frac1\hbar\int_a^x p\,dx+\frac\pi4\right)\\
 &=\frac1{\sqrt v}\exp\left(\frac{i}{\hbar}\int_a^x p\,dx+\frac{i\pi}{4}\right)
 +\frac1{\sqrt v}\exp\left(-\frac{i}{\hbar}\int_a^x p\,dx-\frac{i\pi}{4}\right).
 \end{split}
\]

\SourcePage{226}{229}
The first term (which tends, as $x\to-\infty$, to the plane wave
$\psi\sim e^{ipx/\hbar}$) is the wave incident on the barrier, and the second
is the reflected wave. Our normalization assigns unit current density to the
incident wave. Thus $D$, the current density of the transmitted wave, is the
required barrier transmission coefficient. Formula (50.5) is applicable only
when its exponential argument is large, so that $D$ itself is small\footnote{The
exponential smallness of $D$ also explains why the incident and reflected
amplitudes found in region $I$ are equal: their exponentially small difference
is lost within the semiclassical approximation.}.

So far we have assumed that $U(x)$ obeys the semiclassical condition throughout
the barrier, apart from the immediate neighborhoods of the turning points. In
practice one often encounters barriers whose potential-energy curve rises or
falls so sharply on one side that the semiclassical approximation fails there.
The leading exponential factor in $D$ remains the one in (50.5), but the
pre-exponential factor, equal to unity in (50.5), changes. To calculate it one
must in principle find the exact wave function in the non-semiclassical region
and use its matching to fix the semiclassical wave function inside the barrier.

\begin{center}\ProblemHeading\end{center}

\textbf{1.} Find the transmission coefficient for the potential barrier shown
in Fig.~14: $U(x)=0$ for $x<0$, and $U(x)=U_0-Fx$ for $x>0$. Calculate only
the exponential factor.

\SolutionHeading A direct calculation gives
\[
 D\sim\exp\left[-\frac{4\sqrt{2m}}{3\hbar F}(U_0-E)^{3/2}\right].
\]

\textbf{2.} Find the probability for a particle of zero angular momentum to
escape from the spherically symmetric potential well $U(r)=-U_0$ for $r<r_0$,
$U(r)=\alpha/r$ for $r>r_0$ (Fig.~15)\footnote{This problem was first studied
by G.~F.~Gamow (1928), and by $R. W. Gurney$ and $E. U. Condon$ (1929), in
connection with the theory of radioactive $\alpha$ decay.}.

\SolutionHeading A spherically symmetric problem reduces to a one-dimensional
one, so the preceding formulas apply. Thus
\[
 w\sim\exp\left[-\frac2\hbar\int_{r_0}^{\alpha/E}
 \sqrt{2m\left(\frac\alpha r-E\right)}\,dr\right].
\]

\SourcePage{227}{230}
Evaluation of the integral gives finally
\[
 w\sim\exp\left\{-\frac{2\alpha}{\hbar}\sqrt{\frac{2m}{E}}
 \left[\arccos\sqrt{\frac{Er_0}{\alpha}}-
 \sqrt{\frac{Er_0}{\alpha}\left(1-\frac{Er_0}{\alpha}\right)}\right]\right\}.
\]

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm]
 \begin{scope}[xshift=-3.1cm]
  \draw[->] (-1.7,0)--(1.9,0) node[right] {$x$};
  \draw[->] (0,-1)--(0,2.2) node[right] {$U(x)$};
  \draw[thick] (-1.5,0)--(0,0)--(0,1.55)--(1.35,-.65);
  \draw[dashed] (-1,.85)--(1.15,.85) node[right] {$E$};
  \node[left] at (0,1.55) {$U_0$};
  \node at (0,-1.25) {Fig. 14};
 \end{scope}
 \begin{scope}[xshift=3.1cm]
  \draw[->] (-1.6,0)--(1.9,0) node[right] {$r$};
  \draw[->] (-1.2,-1)--(-1.2,2.2) node[left] {$U(r)$};
  \draw[thick] (-1.2,-.7)--(0,-.7)--(0,1.655);
  \draw[thick,domain=.05:1.75,samples=80] plot (\x,{.48/(\x+.24)});
  \node[left] at (-1.2,-.7) {$-U_0$}; \node[below right] at (0,0) {$r_0$};
  \node at (0,-1.25) {Fig. 15};
 \end{scope}
\end{tikzpicture}
\end{figure}

In the limiting case $r_0\to0$, this expression becomes
\[
 w\sim\exp\left(-\frac{\pi\alpha}{\hbar}\sqrt{\frac{2m}{E}}\right)
 =\exp\left(-\frac{2\pi\alpha}{\hbar v}\right).
\]
These formulas require a large exponent, that is, $\alpha/\hbar v\gg1$.
As expected, this is precisely condition (49.11) for semiclassical motion in a
Coulomb field.

\textbf{3.} The field $U(x)$ consists of two symmetric potential wells ($I$
and $II$ in Fig.~16) separated by a barrier. If the particle could not
penetrate the barrier, there would be levels corresponding to motion in either
well alone, with the same energies for the two wells. Passage through the
barrier splits every such level into a pair of nearby levels. Their states
describe a particle moving in both wells at once. Find the splitting, assuming
that $U(x)$ is semiclassical.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[x=1cm,y=1cm]
 \draw[->] (-2.8,0)--(2.9,0) node[right] {$x$};
 \draw[->] (0,-.15)--(0,2.9) node[right] {$U(x)$};
 \draw[thick] plot[smooth] coordinates {(-2.6,2.7)(-1.45,.55)(-1,.45)(0,1.35)(1,.45)(1.45,.55)(2.6,2.7)};
 \draw[dashed] (1,.87)--(1,0) node[below=2pt] {$a$};
 \draw[dashed] (.2,1.17)--(1.79,1.17) node[above right=-2pt] {$E_2$};
 \draw[dashed] (.3,1.02)--(1.71,1.02) node[right] {$E_0$};
 \draw[dashed] (.38,.87)--(1.63,.87) node[below right=-2pt] {$E_1$};
 \node at (-1.25,1.75) {$II$}; \node at (1.25,1.75) {$I$}; \node[above left] at (0,1.4) {$U_0$};
\end{tikzpicture}
\par\smallskip Fig. 16
\end{figure}

\SolutionHeading To neglect barrier penetration, construct an approximate
solution of the Schrödinger equation from the semiclassical function
$\psi_0(x)$ describing motion with some energy $E_0$ in one well, say $I$.
Thus it decays exponentially on both sides of that well. Normalize $\psi_0$
so that the integral of $\psi_0^2$ over well $I$ is unity. Allowing for the
small tunnelling probability splits $E_0$ into $E_1$ and $E_2$. The proper
zeroth-order functions belonging to these levels are the symmetric and
antisymmetric combinations of $\psi_0(x)$ and $\psi_0(-x)$:
\begin{equation}
 \psi_1(x)=\frac1{\sqrt2}[\psi_0(x)+\psi_0(-x)],\qquad
 \psi_2(x)=\frac1{\sqrt2}[\psi_0(x)-\psi_0(-x)]. \tag{1}
\end{equation}

\SourcePage{228}{231}
In well $I$, $\psi_0(-x)$ is negligibly small compared with $\psi_0(x)$; in
well $II$ the reverse holds. Hence $\psi_0(x)\psi_0(-x)$ is negligible
everywhere, and (1) is normalized so that the integrals of the squared
functions over wells $I$ and $II$ are unity.

Write the Schrödinger equations
\[
 \psi_0''+\frac{2m}{\hbar^2}(E_0-U)\psi_0=0,\qquad
 \psi_1''+\frac{2m}{\hbar^2}(E_1-U)\psi_1=0.
\]
Multiply the first by $\psi_1$ and the second by $\psi_0$, subtract them term
by term, and integrate over $dx$ from 0 to $\infty$. At $x=0$ we have
$\psi_1=\sqrt2\psi_0$ and $\psi_1'=0$, while
\[
 \int_0^\infty\psi_0\psi_1\,dx\approx\frac1{\sqrt2}
 \int_0^\infty\psi_0^2\,dx=\frac1{\sqrt2}.
\]
It follows that
\[
 E_1-E_0=-\frac{\hbar^2}{m}\psi_0(0)\psi_0'(0).
\]
The same calculation for $E_2-E_0$ gives the opposite sign. Therefore
\[
 E_2-E_1=\frac{2\hbar^2}{m}\psi_0(0)\psi_0'(0).
\]
Using (47.1), with the coefficient $C$ from (48.3), gives
\[
 \psi_0(0)=\sqrt{\frac{\omega}{2\pi v_0}}
 \exp\left(-\frac1\hbar\int_0^a|p|\,dx\right),\qquad
 \psi_0'(0)=\frac{mv_0}{\hbar}\psi_0(0),
\]
where $v_0=\sqrt{2(U_0-E_0)/m}$. We thus obtain
\[
 E_2-E_1=\frac{\omega\hbar}{\pi}
 \exp\left(-\frac1\hbar\int_{-a}^a|p|\,dx\right),
\]
where $a$ is the turning point associated with the energy $E_0$ (see Fig.~16).

\textbf{4.} Find the exact transmission coefficient $D$, without assuming it
small, for the parabolic potential barrier $U(x)=-kx^2/2$ ($E. C. Kemble$,
1935)\footnote{The result also applies to transmission sufficiently near the
top of any barrier $U(x)$ whose dependence on $x$ is quadratic near its
maximum.}.

\SolutionHeading For arbitrary $k$ and $E$, the motion is semiclassical at
sufficiently large $|x|$, where
\[
 p=\sqrt{2m\left(E+\frac12kx^2\right)}\approx x\sqrt{mk}
 +E\sqrt{\frac mk}\frac1x,
\]
and the asymptotic form of the Schrödinger-equation solution is
\[
 \psi=\operatorname{const}\cdot\xi^{\pm i\varepsilon-1/2}
 \exp(\pm i\xi^2/2),
\]

\SourcePage{229}{232}
where
\[
 \xi=x\left(\frac{mk}{\hbar^2}\right)^{1/4},\qquad
 \varepsilon=\frac E\hbar\sqrt{\frac mk}.
\]
We need the solution which, as $x\to+\infty$, contains only the wave
transmitted through the barrier, travelling from left to right. Set
\begin{align}
 \psi&=B\xi^{i\varepsilon-1/2}\exp(i\xi^2/2)
 &&\text{as }x\to\infty, \tag{1}\\
 \psi&=(-\xi)^{-i\varepsilon-1/2}\exp(-i\xi^2/2)
 +A(-\xi)^{i\varepsilon-1/2}\exp(i\xi^2/2)
 &&\text{as }x\to-\infty. \tag{2}
\end{align}
The first term in (2) is the incident wave and the second the reflected wave;
a wave propagates in the direction in which its phase increases. The relation
between $A$ and $B$ follows from the validity of the asymptotic form of $\psi$
throughout the sufficiently distant part of the complex $\xi$-plane. Follow
the change in (1) along a semicircle of large radius $\rho$ in the upper
half-plane:
\[
 \xi=\rho e^{i\varphi},\qquad
 i\xi^2=\rho^2(-\sin2\varphi+i\cos2\varphi),
\]
where $\varphi$ runs from 0 to $\pi$. After this continuation, (1) becomes the
second term of (2), with coefficient
\begin{equation}
 A=B(e^{i\pi})^{i\varepsilon-1/2}=-iBe^{-\pi\varepsilon}. \tag{3}
\end{equation}
On the portion $\pi/2<\varphi<\pi$, where
$|\exp(i\xi^2/2)|$ is exponentially large, an exponentially small quantity
that should have produced the first term in (2) is lost\footnote{Continuation
through the lower half-plane would not determine $A$. On the portion
$-\pi<\varphi<-\pi/2$ adjoining its left end, where $\psi$ is given by (2),
the term containing $\exp(i\xi^2/2)$ is exponentially small in comparison with
the one containing $\exp(-i\xi^2/2)$.}.

With the normalization of the incident wave chosen in (2), conservation of
particle number gives
\begin{equation}
 |A|^2+|B|^2=1. \tag{4}
\end{equation}
Equations (3) and (4) yield the required transmission coefficient
\[
 D=|B|^2=1/(1+e^{-2\pi\varepsilon}).
\]
This result holds for every $E$. If the energy is negative and large in
absolute value, $D\approx e^{-2\pi|\varepsilon|}$, in agreement with (50.5).
For $E>0$ the quantity
\[
 R=1-D=1/(1+e^{2\pi\varepsilon})
\]
is the above-barrier reflection coefficient.
