\chapter{Fundamental Concepts of Quantum Mechanics}

\section{The Uncertainty Principle}

\SourcePage{10}{13}
When classical mechanics and electrodynamics are applied in an attempt to
explain atomic phenomena, they give results sharply at variance with
experiment. This is already seen most clearly in the contradiction produced
by applying ordinary electrodynamics to a model of the atom in which electrons
move around the nucleus in classical orbits. Since any accelerated charge radiates, electrons
in such orbits would be obliged to emit electromagnetic waves without
interruption. Through radiation they would lose energy
and eventually fall into the nucleus. Classical electrodynamics would therefore render the atom
unstable---a conclusion entirely at odds with reality.

So profound a conflict between theory and experiment shows that a theory
applicable to atomic phenomena---phenomena involving particles of very small
mass in very small regions of space---requires a fundamental alteration of
the basic classical ideas and laws.

A convenient point of departure for identifying these changes is the
experimentally observed phenomenon known as electron diffraction\footnote{In
fact, electron diffraction was discovered after quantum mechanics had been
created. Our presentation, however, does not follow the historical sequence
in which the theory developed; it is arranged instead to show as clearly as
possible how the fundamental principles of quantum mechanics are connected
with observed phenomena.}. When a uniform beam of electrons passes through a
crystal, the transmitted beam displays a succession of intensity maxima and
minima entirely analogous to the diffraction pattern observed for
electromagnetic waves. Under certain conditions, therefore, the behavior of
material particles---electrons---shows features characteristic of wave
processes.

\SourcePage{11}{14}
The depth of this phenomenon's conflict with ordinary ideas of motion is best
shown by the following thought experiment, an idealization of electron
diffraction by a crystal. Imagine a barrier impermeable to electrons, with two
slits cut in it. If we observe an electron beam\footnote{The beam is assumed
to be so dilute that interactions among its particles play no part.} passing
through one slit while the other is closed, we obtain some intensity
distribution on a solid screen behind the slit. Opening the second slit and
closing the first gives another distribution. With both slits open at once,
ordinary reasoning would lead us to expect simply the superposition of those
two distributions: each electron, moving on its own trajectory, passes
through one slit without affecting electrons passing through the other.
Electron diffraction shows instead that the actual result is a diffraction
pattern which, because of interference, is by no means the sum of the patterns
from the separate slits. Plainly, this result cannot be reconciled in any way
with the idea that electrons move along trajectories.

The mechanics governing atomic phenomena---the so-called \emph{quantum} or
\emph{wave mechanics}---must therefore rest on concepts of motion fundamentally
different from those of classical mechanics. Quantum mechanics has no concept
of a particle trajectory. This fact constitutes the content of the so-called
\emph{uncertainty principle}, a fundamental principle of quantum
mechanics, due to \emph{Heisenberg} (\emph{W.~Heisenberg},
1927)\footnote{It is noteworthy that the complete mathematical apparatus of
quantum mechanics had already been constructed by \emph{W.~Heisenberg} and
\emph{E.~Schrödinger} in 1925--1926---prior to formulation of the uncertainty
principle, which lays bare the physical content of that apparatus.}. Since it
rejects the customary ideas of classical mechanics, the uncertainty principle
may be said to have negative content. By itself it is naturally quite
insufficient as the basis for constructing a new mechanics of particles. Such
a theory must of course rest on certain positive propositions, which will be
considered below (\S~2). Before those propositions can be formulated, however,
the character of the problems posed to quantum mechanics must first be made
clear.

\SourcePage{12}{15}
For this purpose, let us first consider the special relation between quantum
and classical mechanics.

Ordinarily, a more general theory can be formulated as a logically closed
system independently of the less general theory that arises as its limiting
case. Relativistic mechanics, for example, can be constructed from its own
fundamental principles without any reference to Newtonian mechanics. The
fundamental propositions of quantum mechanics, on the other hand, cannot in
principle be formulated without invoking classical mechanics.

The absence of a definite trajectory for an electron\footnote{In this and the
next section, for brevity, we speak of an electron while meaning any quantum
object in general: a particle or system of particles which obeys quantum but
not classical mechanics.} also deprives the electron, considered by itself,
of any other dynamical characteristics\footnote{Here we mean quantities
characterizing the electron's motion, not those characterizing the electron
as a particle (charge and mass), which are parameters.}. It is therefore clear
that no logically closed mechanics could be constructed for a system
consisting solely of quantum objects. A quantitative description of electron
motion requires the presence of physical objects which obey classical
mechanics to sufficient accuracy. When an electron interacts with a
``classical object,'' the state of that object is, in general, altered. Both the character
and the magnitude of this alteration are governed by the electron's state and can therefore
serve as its quantitative characteristic.

In this connection, the ``classical object'' is usually called an
``apparatus,'' and its interaction with the electron is spoken of as a
``measurement.'' It must be emphasized, however, that this does not mean a
process of ``measurement'' involving a physicist as observer. In quantum
mechanics, measurement means any interaction between a classical and a
quantum object, occurring apart from and independently of any observer. The
recognition of the profound role of the measurement concept in quantum
mechanics is due to \emph{Bohr} (\emph{N.~Bohr}).

We have defined an apparatus as a physical object which obeys classical
mechanics to sufficient accuracy; a body of sufficiently great mass is one
example. It should not be supposed, however, that an apparatus must be
macroscopic. Under suitable conditions an object known to be microscopic can
also play the part of an apparatus, since%
\input{manuscript/chapters/ch01/fragments/sec01-p013-p016.translation.en}
\input{manuscript/chapters/ch01/fragments/sec02-p016-p019.translation.en}
\input{manuscript/chapters/ch01/fragments/sec03-p019-p025.translation.en}
\input{manuscript/chapters/ch01/fragments/sec04-p025-p029.translation.en}
\input{manuscript/chapters/ch01/fragments/sec05-p029-p034.translation.en}
\input{manuscript/chapters/ch01/fragments/sec06-p034-p036.translation.en}
\input{manuscript/chapters/ch01/fragments/sec07-p036-p041.translation.en}
