\SourcePage{430}{433}
\section{Symmetry transformations}

As in the diatomic case, the classification of the terms of a polyatomic molecule is closely connected with its symmetry. We therefore begin by studying the kinds of symmetry a molecule may possess.

The symmetry of a body is specified by the collection of motions that bring the body into coincidence with itself; these motions are called \emph{symmetry transformations}. Every possible symmetry transformation can be represented as a combination of one or more of three basic types. These three essentially distinct types are: a \emph{rotation} of the body through a definite angle about an axis, a \emph{reflection} in a plane, and a \emph{translation} of the body through some distance. Of these, the last type can clearly occur only for an unbounded medium (a crystal lattice). A body of finite dimensions (in particular, a molecule) can be symmetric only under rotations and reflections.

If a body coincides with itself after a rotation through $2\pi/n$ about an axis, that axis is called an \emph{$n$-fold symmetry axis}. The number $n$ may have any integral value $n=2,3,\ldots$; $n=1$ corresponds to a rotation through $2\pi$, or equivalently through zero, and hence to the identity transformation. We denote the operation of rotation through $2\pi/n$ about the given axis by $C_n$. Repeating this operation two, three, \ldots times gives rotations through $2(2\pi/n),3(2\pi/n),\ldots$, which likewise bring the body into coincidence with itself; these rotations may be denoted by $C_n^2,C_n^3,\ldots$ Clearly, if $n$ is divisible by $p$, then
\begin{equation}
C_n^p=C_{n/p}.
\tag{91.1}
\end{equation}
In particular, after performing the rotation $n$ times we return to the initial position, i.e. carry out the \emph{identity transformation}; the latter is conventionally denoted by $E$, so that
\begin{equation}
C_n^n=E.
\tag{91.2}
\end{equation}

\SourcePage{431}{434}
If a body coincides with itself after reflection in a plane, that plane is called a \emph{plane of symmetry}. We denote the operation of reflection in a plane by $\sigma$. Evidently, two successive reflections in the same plane give the identity transformation,
\begin{equation}
\sigma^2=E.
\tag{91.3}
\end{equation}

The combined application of a rotation and a reflection leads to the so-called \emph{rotation--reflection axes}. A body has an $n$-fold rotation--reflection axis if it coincides with itself after rotation through $2\pi/n$ about this axis followed by reflection in a plane perpendicular to the axis (Fig.~32). It is readily seen that this is a new kind of symmetry only when $n$ is even. Indeed, if $n$ is odd, repeating the rotation--reflection transformation $n$ times is equivalent simply to reflection in the plane perpendicular to the axis (the total angle of rotation is $2\pi$, while an odd number of reflections in the same plane is a single reflection). Repeating the transformation a further $n$ times shows that the rotation--reflection axis reduces to the simultaneous presence of an independent $n$-fold symmetry axis and a plane of symmetry perpendicular to it. If $n$ is even, however, $n$ repetitions of the rotation--reflection transformation return the body to its initial position.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[line width=.45pt,scale=.85]
 \draw (-2.0,0)--(1.4,0)--(2.3,.75)--(-1.1,.75)--cycle;
 \draw (0,-2.1)--(0,2.0);
 \draw[dashed] (0,-2.1)--(0,-.05) (0,.75)--(0,2.0);
 \fill (0,.38) circle (1.1pt) node[left] {$O$};
 \filldraw[fill=white] (0,1.55) circle (1.4pt) node[right] {$P$};
 \filldraw[fill=white] (0,-1.7) circle (1.4pt) node[right] {$P'$};
 \draw[dashed] (0,1.55) arc[start angle=90,end angle=15,radius=.7];
 \node at (.72,1.25) {$2\pi/n$};
 \draw (.05,-1.25)--(.4,-1.45);
\end{tikzpicture}
\caption{}
\label{fig:91-32}
\end{figure}

We denote a rotation--reflection transformation by $S_n$. Denoting reflection in the plane perpendicular to the given axis by $\sigma_h$, we may write, by definition,
\begin{equation}
S_n=C_n\sigma_h=\sigma_hC_n
\tag{91.4}
\end{equation}
(the order in which $C_n$ and $\sigma_h$ are performed clearly does not affect the result).

An important special case is a twofold rotation--reflection axis. A rotation through $\pi$ followed by reflection in a plane perpendicular to the rotation axis is readily seen to constitute the transformation of \emph{inversion}: a point $P$ of the body is carried into a point $P'$ on the continuation of the line joining $P$ to the point $O$ where the axis meets the plane, with $OP=OP'$.

\SourcePage{432}{435}
A body symmetric under this transformation is said to possess a center of symmetry (we denote inversion by $I$):
\begin{equation}
I\equiv S_2=C_2\sigma_h.
\tag{91.5}
\end{equation}
It is also evident that $I\sigma_h=C_2$ and $IC_2=\sigma_h$; in other words, a twofold axis, a plane of symmetry perpendicular to it, and a center of symmetry at their intersection are mutually dependent: the presence of any two automatically entails the third.

We now state several purely geometrical properties of rotations and reflections that are useful in studying the symmetry of bodies.

The product of two rotations about axes intersecting at a point is a rotation about a third axis through the same point. The product of two reflections in intersecting planes is equivalent to a rotation; its axis clearly coincides with the line in which the planes intersect, and a simple geometrical construction shows that its angle is twice the angle between them. Denote rotation through an angle $\varphi$ about the axis by $C(\varphi)$ and reflections in two planes containing the axis by $\sigma_v$ and $\sigma'_v$\footnote{The subscript $v$ conventionally distinguishes reflection in a plane containing the given axis (a ``vertical'' plane), while $h$ denotes reflection in a plane perpendicular to the axis (a ``horizontal'' plane).}. The statement can then be written
\begin{equation}
\sigma_v\sigma'_v=C(2\varphi),
\tag{91.6}
\end{equation}
where $\varphi$ is the angle between the two planes. The order of the two reflections is important: $\sigma_v\sigma'_v$ gives a rotation directed from the plane $\sigma'_v$ toward $\sigma_v$, whereas interchanging the factors gives rotation in the opposite direction. Multiplying (91.6) from the left by $\sigma_v$, we obtain
\begin{equation}
\sigma'_v=\sigma_vC(2\varphi);
\tag{91.7}
\end{equation}
thus, the product of a rotation and reflection in a plane containing its axis is equivalent to reflection in another plane meeting the first at an angle equal to half the angle of rotation. It follows in particular that a twofold symmetry axis and two mutually perpendicular planes of symmetry containing it are mutually dependent: the presence of two requires the third.

\SourcePage{433}{436}
We shall show that the product of rotations through $\pi$ about two axes meeting at an angle $\varphi$ ($Oa$ and $Ob$ in Fig.~33) is a rotation through $2\varphi$ about an axis perpendicular to both ($PP'$ in Fig.~33). It is clear in advance that the resulting transformation is also a rotation. After the first rotation (about $Oa$), the point $P$ goes to $P'$, and after the second (about $Ob$) it returns to its initial position. The line $PP'$ therefore remains fixed and is consequently the rotation axis. To determine the angle of rotation, it suffices to note that the axis $Oa$ remains fixed during the first rotation and is carried by the second into the position $Oa'$, which makes an angle $2\varphi$ with $Oa$. The same argument shows that reversing the order of the two transformations gives a rotation in the opposite direction.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[line width=.45pt,scale=.9]
 \draw (-2,0)--(1.7,0)--(2.5,.8)--(-1.2,.8)--cycle;
 \draw (0,-1.6)--(0,2.0);
 \filldraw[fill=white] (0,1.75) circle (1.3pt) node[right] {$P$};
 \filldraw[fill=white] (0,-1.4) circle (1.3pt) node[right] {$P'$};
 \fill (0,.4) circle (1pt) node[left] {$O$};
 \draw (0,.4)--(1.2,.65) node[right] {$b$};
 \draw (0,.4)--(.95,.12) node[right] {$a$};
 \draw (0,.4)--(.65,.9) node[above] {$a'$};
 \draw (.42,.5) arc[start angle=14,end angle=40,radius=.42];
 \draw (.43,.49) arc[start angle=40,end angle=72,radius=.43];
 \node at (.63,.43) {$\varphi$}; \node at (.42,.72) {$\varphi$};
\end{tikzpicture}
\caption{}
\label{fig:91-33}
\end{figure}

Although the result of two successive transformations generally depends on their order, in a number of cases the order is immaterial: the transformations commute. This occurs for the following transformations:

1) two rotations about the same axis;

2) two reflections in mutually perpendicular planes (they are equivalent to a rotation through $\pi$ about the line in which the planes intersect);

3) two rotations through $\pi$ about mutually perpendicular axes (they are equivalent to a rotation through the same angle about the third perpendicular axis);

4) a rotation and reflection in a plane perpendicular to the rotation axis;

5) any rotation (or reflection) and inversion in a point lying on the rotation axis (or in the reflection plane); this follows from 1) and 4).
